### 2022-06-29 src/tex/ms.tex
[showyourwork] first commit

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}

% Begin!
\begin{document}

% Title
\title{An open source scientific article}

% Author list
\author{@matiscke}

% Abstract with filler text
\begin{abstract}
    Lorem ipsum dolor sit amet, consectetuer adipiscing elit. 
    Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis. 
    Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna. 
    Donec vehicula augue eu neque, morbi tristique senectus et netus et. 
    Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum. 
    Phasellus eu tellus sit amet tortor gravida placerat. 
    Integer sapien est, iaculis in, pretium quis, viverra ac, nunc. 
    Praesent eget sem vel leo ultrices bibendum. 
    Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac. 
    Curabitur auctor semper nulla donec varius orci eget risus. 
    Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam. 
    Duis eget orci sit amet orci dignissim rutrum.
\end{abstract}

% Main body with filler text
\section{Introduction}
\label{sec:intro}

Lorem ipsum dolor sit amet, consectetuer adipiscing elit. 
Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis. 
Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna. 
Donec vehicula augue eu neque, morbi tristique senectus et netus et. 
Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum. 
Phasellus eu tellus sit amet tortor gravida placerat. 
Integer sapien est, iaculis in, pretium quis, viverra ac, nunc. 
Praesent eget sem vel leo ultrices bibendum. 
Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac. 
Curabitur auctor semper nulla donec varius orci eget risus. 
Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam. 
Duis eget orci sit amet orci dignissim rutrum.

Nam dui ligula, fringilla a, euismod sodales, sollici- tudin vel, wisi. 
Morbi auctor lorem non justo, nam lacus libero, pretium at, lobortis vitae. 
Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna. 
Morbi ac orci et nisl hendrerit mollis, suspendisse ut massa, cras nec ante. 
Pellentesque a nulla cum sociis natoque penatibus et magnis dis parturient. 
Aliquam tincidunt urna, nulla ullamcorper vestibulum turpis. 
Pellentesque cursus luctus mauris \citep{Luger2021}.

\bibliography{bib}

\end{document}

### 2022-06-29 src/tex/ms.tex
bring over preamble and tex boilerplate from other paper

@@
 \documentclass[twocolumn]{aastex631}
-\usepackage{showyourwork}
+
+\input{preamble.tex}
 
@@
 % Title
-\title{An open source scientific article}
+\title{An open source scientific article about extrasolar magma oceans}
 
 % Author list
-\author{@matiscke}
+\author[0000-0001-8355-2107]{Martin Schlecker}
+\affiliation{Department of Astronomy/Steward Observatory, The University of Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
+\author{al.}
 
-% Abstract with filler text
+% Abstract
 \begin{abstract}
-    Lorem ipsum dolor sit amet, consectetuer adipiscing elit. 
-    Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis. 
-    Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna. 
-    Donec vehicula augue eu neque, morbi tristique senectus et netus et. 
-    Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum. 
-    Phasellus eu tellus sit amet tortor gravida placerat. 
-    Integer sapien est, iaculis in, pretium quis, viverra ac, nunc. 
-    Praesent eget sem vel leo ultrices bibendum. 
-    Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac. 
-    Curabitur auctor semper nulla donec varius orci eget risus. 
-    Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam. 
-    Duis eget orci sit amet orci dignissim rutrum.
+    $\ldots$ magma oceans $\ldots$
+
+    Here, we assess the ability of space and ground-based telescopes to test this hypothesis using Bioverse, a simulation framework that leverages contextual information from the overall planet population.
+    We argue that in the near future, ESA's PLATO mission and NASA's Roman Space Telescope will be the most promising endeavors to constrain this demographic feature.
+    For each of these missions, we identify the key mission design drivers that enable a statistically sound detection.
+    We also show the unique synergy of these missions in answering this question, and what survey strategy optimizes the statistical power of the combined dataset.
+
+    $\ldots$ its measurement will also provide insights into which stars harbor the nearest habitable worlds.
 \end{abstract}
 
-% Main body with filler text
+% Main body
 \section{Introduction}
-\label{sec:intro}
+%\Blindtext[4]
 
-Lorem ipsum dolor sit amet, consectetuer adipiscing elit. 
-Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis. 
-Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna. 
-Donec vehicula augue eu neque, morbi tristique senectus et netus et. 
-Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum. 
-Phasellus eu tellus sit amet tortor gravida placerat. 
-Integer sapien est, iaculis in, pretium quis, viverra ac, nunc. 
-Praesent eget sem vel leo ultrices bibendum. 
-Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac. 
-Curabitur auctor semper nulla donec varius orci eget risus. 
-Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam. 
-Duis eget orci sit amet orci dignissim rutrum.
+\begin{figure*}
+    \begin{centering}
+        \includegraphics[width=\hsize]{../static/flowchart}
+        \caption{Workflow of our hypothesis testing with Bioverse.
+In the first block, a stellar sample is generated based on XXX.
+The stars are then populated with planetary systems from XXX, to which a population-level trend can be applied.
+The planets' respective transit probabilities are computed.
+The second block simulates the exoplanet survey whereby selection effects and detection biases are introduced.
+Finally, the third block deals with testing a hypothesis based on the data from the simulated survey.
+By iterating through these steps, we compute the statistical power of testing the hypothesis given the assumed survey design.}
+        \label{fig:flowchart}
+    \end{centering}
+\end{figure*}
 
-Nam dui ligula, fringilla a, euismod sodales, sollici- tudin vel, wisi. 
-Morbi auctor lorem non justo, nam lacus libero, pretium at, lobortis vitae. 
-Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna. 
-Morbi ac orci et nisl hendrerit mollis, suspendisse ut massa, cras nec ante. 
-Pellentesque a nulla cum sociis natoque penatibus et magnis dis parturient. 
-Aliquam tincidunt urna, nulla ullamcorper vestibulum turpis. 
-Pellentesque cursus luctus mauris \citep{Luger2021}.
+%\begin{deluxetable*}{cc}
+%\tablecaption{Confirmed and suggested trends in the exoplanet population}
+%%\tablewidth{\pagewidth}
+%\tablehead{\colhead{Trend} & \colhead{Reference}}
+%\startdata
+%missing sub-Saturn mass valley in M-dwarf planetary systems &  Schlecker et al., subm. \\
+%$\cdots$ & $\cdots$
+%\enddata
+%\tablecomments{We show a list of population-level trends that have been reported in demographic studies or predicted from planet formation theories.}
+%\end{deluxetable*}
+%
+%\begin{deluxetable*}{cccccc}
+%\tablecaption{Confirmed and suggested trends in the exoplanet population}
+%%\tablewidth{\pagewidth}
+%\tablehead{\colhead{No.} & \colhead{Trend} & \colhead{Reference} & \colhead{Terrestrials}&\colhead{Roman}&\colhead{Plato}}
+%\startdata
+%%\hline
+%%Theoretically predicted&&&&&
+%%\\
+%%\hline
+%%1&High occurrence of low-mass planets at a few au&\citep{Penny2019}&FALSE&TRUE&FALSE \\
+%%2&(super-)Earth occurrence drops again for very low-mass stars&Mulders2021, (Schlecker2021a)&TRUE&TRUE&TRUE
+%%\\
+%%3&Hot Jupiter - cold Jupiter relation&Schlaufman \& Winn (2016)&FALSE&TRUE&FALSE
+%%\\
+%%4&super-Earth composition-architecture link&Schlecker+2021a&TRUE&FALSE&FALSE
+%%\\
+%
+%\enddata
+%\tablecomments{We show a list of population-level trends that have been reported in demographic studies or predicted from planet formation theories.}
+%\end{deluxetable*}
 
+Several patterns in the planetary parameter space have been reported in demographic studies or predicted from planet formation theories.
+
+\section{Synthetic star and planet sample}
+
+\section{The XXX Hypothesis}
+
+\section{Survey simulation}
+
+\subsection{Testing the XXX hypothesis with the Nancy Grace Roman Telescope}
+
+\section{Discussion}
+
+\subsection{Mission design trades}\label{sec:mission-design-trades}
+\citet{Penny2019} mention that a significant increase in planet yield could be achieved if the telescope's slew speed was increased.
+
+\section{Conclusions}
+
+\bibliographystyle{aasjournal}
 \bibliography{bib}
-
 \end{document}

### 2022-07-29 src/tex/ms.tex
make showyourwork find the static figure
add blindtext just to please Latex
add section headings

@@
 \section{Introduction}
-%\Blindtext[4]
+\Blindtext[4]
 
@@
     \begin{centering}
-        \includegraphics[width=\hsize]{../static/flowchart}
+        \includegraphics[width=\hsize]{figures/flowchart.pdf}
         \caption{Workflow of our hypothesis testing with Bioverse.
@@
 
-\section{The XXX Hypothesis}
+\section{The magma ocean hypothesis}
+
+\subsection{Global magma oceans}
+\todo[inline]{introduce magma oceans and their influence on planetary radii~\citep{Dorn2021}.\\ Might also be already fully covered in the Introduction section}
+
+\subsection{Demographic imprint of magma oceans}
+\todo{describe the expected imprint on the exoplanet population. \\Provide parametrizations from \citet{Dorn2021}'s models.}
+
+\subsubsection{Parametrization}
+\todo{specific parametrization, dependency on bulk planet/orbit params (doesn't have to be its own subsubsection)}
+
+\begin{note}
+The power from the host star per unit area at the position of a planet or stellar insolation $S$ in units of Earth's insolation is given by
+    \begin{equation}
+        \frac{S}{S_\oplus} = \left(\frac{L_\star}{L_\odot}\right) \left(\frac{au}{a}\right)^2 .
+    \end{equation}
+
+We further define the solar-equivalent semi-major axis $a_{eff} = a (L_\star/L_\odot)^{-1/2}$, at which a planet experiences the same insolation as a Solar System planet at an orbital distance $a$.
+\end{note}
+
+\subsection{Synthetic planet populations with and without magma oceans}
+\todo{describe the way we inject magma oceans into the synthetic planet population with Bioverse}
 

### 2022-08-02 src/tex/ms.tex
add additional literature file PhD.bib
During the evolution of the manuscript, both PhD.bib and bib.bib are automatically being updated by Zotero.
add a paragraph on false positives for a magma ocean pattern

@@
     \begin{centering}
-        \includegraphics[width=\hsize]{figures/flowchart.pdf}
+        \includegraphics[width=\hsize]{figures/flowchart}
         \caption{Workflow of our hypothesis testing with Bioverse.
@@
 
-\subsection{Testing the XXX hypothesis with the Nancy Grace Roman Telescope}
+\subsection{Testing the magma ocean hypothesis with current and planned exoplanet missions}
+\todo{PLATO, Ariel, LIFE?, Nautilus\\ What are threshold missions that are able to detect this signal?
+=>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.
+ }
 
 \section{Discussion}
+
+\subsection{False positive scenarios}
+\todo{what other processes could be confused with a magma ocean signal?}
+\begin{note}
+Global magma oceans are not the only physical mechanism that may cause a decrease in transit radii for a subset of planets.
+Other causes of shrinking planet sizes have been put forward, the most widely discussed ones being atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
+    Both processes are being traded as potentially sculpting the observed radius bimodality of small, close-in exoplanets from the \textit{Kepler} mission~\citep{Fulton2017,VanEylen2018}.
+    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
+    This region is distinct from the one affected by global magma oceans, since... \todo{...elaborate why they cannot be confused}
+
+    Another potential false positive contribution comes from the "Neptune desert``, a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
+    The shape of this region is such that smaller planets become less frequent the closer to the star they are, which to some degree resembles the pattern introduced by the insolation dependency of the magma ocean probability.
+    \todo{explore if this can cause confusion}
+\end{note}
 
@@
 \bibliographystyle{aasjournal}
-\bibliography{bib}
+\bibliography{bib,PhD}
 \end{document}

### 2022-08-02 src/tex/ms.tex
update flowchart
add figure showing which planets have magma oceans

@@
 In the first block, a stellar sample is generated based on XXX.
-The stars are then populated with planetary systems from XXX, to which a population-level trend can be applied.
+The stars are then populated with planets from XXX, which may be assigned a magma ocean based on the model described in Sect.~\ref{sec:mo_model}.
 The planets' respective transit probabilities are computed.
@@
 \subsubsection{Parametrization}
+\label{sec:mo_model}
 \todo{specific parametrization, dependency on bulk planet/orbit params (doesn't have to be its own subsubsection)}
@@
 \todo{describe the way we inject magma oceans into the synthetic planet population with Bioverse}
+\begin{figure*}
+    \begin{centering}
+        \includegraphics[width=\hsize]{figures/has_magmaocean}
+        \caption{Synthetic planets with and without magmaoceans as a function of Solar-equivalent semi-major axis}
+        \label{fig:has_magmaocean}
+    \end{centering}
+\end{figure*}
 

### 2022-08-02 src/tex/ms.tex
include file endings for figures
define likelihood function

@@
     \begin{centering}
-        \includegraphics[width=\hsize]{figures/flowchart}
+        \includegraphics[width=\hsize]{figures/flowchart.pdf}
         \caption{Workflow of our hypothesis testing with Bioverse.
@@
 
+\subsubsection{Definition of the magma ocean hypothesis and null hypothesis}
+\todo{Define the null and alternative hypotheses}
+...
+
+For the likelihood function, we assumed that the planet radii $R_{\mathrm{P}, i}$ are measured with a normally distributed uncertainty $\sigma_{R_\mathrm{P}, i}$ and adopted a normal distribution
+\begin{equation}
+\mathcal{L}(R_\mathrm{P} \mid \boldsymbol{\theta})=\prod_{i}^{N} \frac{1}{\sqrt{2 \pi \sigma_{R_\mathrm{P}, i}^{2}}} \exp \left(-\frac{\left(R_{\mathrm{P}, i}-h\left(\boldsymbol{\theta}, a_{\mathrm{eff}, i}\right)\right)^{2}}{2 \sigma_{R_\mathrm{P}, i}^{2}}\right).
+\end{equation}
+Here, $h\left(\boldsymbol{\theta}, a_{\mathrm{eff}, i}\right)$ corresponds to the functional form of the magma ocean hypothesis\todo{link to eqn defining the hypothesis} .
+We tested this hypothesis against the null hypothesis $h_\mathrm{null} (\theta, R_\mathrm{P}) = \theta$, which states that there is no relationship between the measured planet radii and (scaled) semi-major axes.
+
 \subsection{Synthetic planet populations with and without magma oceans}
@@
     \begin{centering}
-        \includegraphics[width=\hsize]{figures/has_magmaocean}
+        \includegraphics[width=\hsize]{figures/has_magmaocean.pdf}
         \caption{Synthetic planets with and without magmaoceans as a function of Solar-equivalent semi-major axis}

### 2022-08-06 src/tex/ms.tex
restructure, include magmaocean model schematic

@@
 
-%\begin{deluxetable*}{cc}
-%\tablecaption{Confirmed and suggested trends in the exoplanet population}
-%%\tablewidth{\pagewidth}
-%\tablehead{\colhead{Trend} & \colhead{Reference}}
-%\startdata
-%missing sub-Saturn mass valley in M-dwarf planetary systems &  Schlecker et al., subm. \\
-%$\cdots$ & $\cdots$
-%\enddata
-%\tablecomments{We show a list of population-level trends that have been reported in demographic studies or predicted from planet formation theories.}
-%\end{deluxetable*}
-%
-%\begin{deluxetable*}{cccccc}
-%\tablecaption{Confirmed and suggested trends in the exoplanet population}
-%%\tablewidth{\pagewidth}
-%\tablehead{\colhead{No.} & \colhead{Trend} & \colhead{Reference} & \colhead{Terrestrials}&\colhead{Roman}&\colhead{Plato}}
-%\startdata
-%%\hline
-%%Theoretically predicted&&&&&
-%%\\
-%%\hline
-%%1&High occurrence of low-mass planets at a few au&\citep{Penny2019}&FALSE&TRUE&FALSE \\
-%%2&(super-)Earth occurrence drops again for very low-mass stars&Mulders2021, (Schlecker2021a)&TRUE&TRUE&TRUE
-%%\\
-%%3&Hot Jupiter - cold Jupiter relation&Schlaufman \& Winn (2016)&FALSE&TRUE&FALSE
-%%\\
-%%4&super-Earth composition-architecture link&Schlecker+2021a&TRUE&FALSE&FALSE
-%%\\
-%
-%\enddata
-%\tablecomments{We show a list of population-level trends that have been reported in demographic studies or predicted from planet formation theories.}
-%\end{deluxetable*}
+\section{Global magma oceans and their imprint on exoplanet demographics}
+\todo[inline]{"Methods Section" for the geophysical models.}
+\subsection{Global magma oceans}
+\todo[inline]{introduce geophysical model.\\ Introduce its parametrization either here or further down when modeling demographic imprint.\\ Figure~\ref{fig:magmaocean_model} goes here.}
 
-Several patterns in the planetary parameter space have been reported in demographic studies or predicted from planet formation theories.
-
-\section{Synthetic star and planet sample}
-
-\section{The magma ocean hypothesis}
-
-\subsection{Global magma oceans}
-\todo[inline]{introduce magma oceans and their influence on planetary radii~\citep{Dorn2021}.\\ Might also be already fully covered in the Introduction section}
+%\begin{figure}
+\begin{figure}[!ht]
+    \begin{centering}
+        \includegraphics[width=\hsize]{figures/magmaocean_model.jpg}
+        \caption{Schematics of the geophysical model framework.}
+        \label{fig:magmaocean_model}
+    \end{centering}
+\end{figure}
 
@@
 \todo{specific parametrization, dependency on bulk planet/orbit params (doesn't have to be its own subsubsection)}
+
+\Blindtext[1]
 
@@
 \end{note}
+
+For the likelihood function, we assumed that the planet radii $R_{\mathrm{P}, i}$ are measured with a normally distributed uncertainty $\sigma_{R_\mathrm{P}, i}$ and adopted a normal distribution
+\begin{equation}
+\mathcal{L}(R_\mathrm{P} \mid \boldsymbol{\theta})=\prod_{i}^{N} \frac{1}{\sqrt{2 \pi \sigma_{R_\mathrm{P}, i}^{2}}} \exp \left(-\frac{\left(R_{\mathrm{P}, i}-h\left(\boldsymbol{\theta}, a_{\mathrm{eff}, i}\right)\right)^{2}}{2 \sigma_{R_\mathrm{P}, i}^{2}}\right).
+\end{equation}
+Here, $h\left(\boldsymbol{\theta}, a_{\mathrm{eff}, i}\right)$ corresponds to the functional form of the magma ocean hypothesis\todo{link to eqn defining the hypothesis} .
+We tested this hypothesis against the null hypothesis $h_\mathrm{null} (\theta, R_\mathrm{P}) = \theta$, which states that there is no relationship between the measured planet radii and (scaled) semi-major axes.
+
+\subsubsection{Synthetic planet populations with and without magma oceans}
+\todo{describe the way we inject magma oceans into the synthetic planet population with Bioverse}
+\begin{figure*}
+    \begin{centering}
+        \includegraphics[width=\hsize]{figures/has_magmaocean.pdf}
+        \caption{Synthetic planets with and without magmaoceans as a function of Solar-equivalent semi-major axis}
+        \label{fig:has_magmaocean}
+    \end{centering}
+\end{figure*}
+
+\subsection{Testing the magma ocean hypothesis with current and planned exoplanet missions}
+\todo{Introduce survey simulation(s).
+PLATO, Ariel, LIFE?, Nautilus\\ What are threshold missions that are able to detect this signal?
+=>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
 
@@
 We tested this hypothesis against the null hypothesis $h_\mathrm{null} (\theta, R_\mathrm{P}) = \theta$, which states that there is no relationship between the measured planet radii and (scaled) semi-major axes.
-
-\subsection{Synthetic planet populations with and without magma oceans}
-\todo{describe the way we inject magma oceans into the synthetic planet population with Bioverse}
-\begin{figure*}
-    \begin{centering}
-        \includegraphics[width=\hsize]{figures/has_magmaocean.pdf}
-        \caption{Synthetic planets with and without magmaoceans as a function of Solar-equivalent semi-major axis}
-        \label{fig:has_magmaocean}
-    \end{centering}
-\end{figure*}
-
-\section{Survey simulation}
-
-\subsection{Testing the magma ocean hypothesis with current and planned exoplanet missions}
-\todo{PLATO, Ariel, LIFE?, Nautilus\\ What are threshold missions that are able to detect this signal?
-=>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.
- }
 
@@
 
+\begin{note}
+    \textit{Here or in the introduction:} For observations of rocky exoplents, the currently best-probed regime is hot, close-in planets.
+    These bodies experience climatic conditions that are similar to the environment of the inner Solar System bodies at early stages of their formation.
+    Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.
+\end{note}
+
 \bibliographystyle{aasjournal}

### 2022-08-06 src/tex/ms.tex
change title
add SAG13 planetary sample and some notes

@@
 % Define document class
-\documentclass[twocolumn]{aastex631}
+%\documentclass[twocolumn]{aastex631}
+\documentclass[modern]{aastex631}
 
@@
 % Title
-\title{An open source scientific article about extrasolar magma oceans}
+\title{The imprint of global magma oceans on exoplanet demographics}
 
@@
 \section{Introduction}
-\Blindtext[4]
+Several patterns in the planetary parameter space have been reported in demographic studies or predicted from planet formation theories.
+\begin{note}
+    ...long-term goal/overarching objective: derive the geophysical history of rocky extrasolar planets.
+   ...
+\todo[inline]{introduce magma oceans and their influence on planetary radii~\citep{Dorn2021}.}
+    ...current/future observations of planets that are currently in the runaway greenhouse phase may constrain properties of their planetary mantles...
+make connection between interior and atmospheres.
+\end{note}
 
@@
 
-For the likelihood function, we assumed that the planet radii $R_{\mathrm{P}, i}$ are measured with a normally distributed uncertainty $\sigma_{R_\mathrm{P}, i}$ and adopted a normal distribution
+\section{Testing the magma ocean hypothesis}
+\todo[inline]{"Methods Section" of the hypothesis testing part.
+Introduce the idea of constraining magma ocean frequencies and/or physics with Bayesian stats on exoplanet population; introduce Bioverse; report methods of the hypothesis tests.}
+
+\subsection{Synthetic star and planet sample}
+\subsubsection{Stellar sample from Gaia DR3}
+\todo{Which star sample was actually used?}
+
+\subsubsection{Planetary sample from SAG~13}
+\todo[inline]{What is the source of our occurrence rate density from which we draw?
+We could also use a generative model that directly approximates an observed or theoretical density \citep[e.g.,][]{Schlecker2021b}.}
+\begin{note}
+    The NASA Exoplanet Program Analysis Group chartered Science Analysis Group 13 (SAG 13)...
+    The inferred occurrence rate density can be expressed as a power law in planet radius and period
     \begin{equation}
-\mathcal{L}(R_\mathrm{P} \mid \boldsymbol{\theta})=\prod_{i}^{N} \frac{1}{\sqrt{2 \pi \sigma_{R_\mathrm{P}, i}^{2}}} \exp \left(-\frac{\left(R_{\mathrm{P}, i}-h\left(\boldsymbol{\theta}, a_{\mathrm{eff}, i}\right)\right)^{2}}{2 \sigma_{R_\mathrm{P}, i}^{2}}\right).
+        \frac{\partial^2n}{\partial \log R \, \partial \log P} = \Gamma R^{\alpha} P^{\beta}
     \end{equation}
-Here, $h\left(\boldsymbol{\theta}, a_{\mathrm{eff}, i}\right)$ corresponds to the functional form of the magma ocean hypothesis\todo{link to eqn defining the hypothesis} .
-We tested this hypothesis against the null hypothesis $h_\mathrm{null} (\theta, R_\mathrm{P}) = \theta$, which states that there is no relationship between the measured planet radii and (scaled) semi-major axes.
+that is broken in planet radius, resulting in the free parameters $\Gamma$, $\alpha$, $\beta$, and a breakpoint radius $R_\mathrm{break}$.
+\end{note}
 

### 2022-08-22 src/tex/ms.tex
prepare acknowledgements
add example corner plot, caption

@@
 
+\subsection{Signature and testability of the magma ocean hypothesis with a $<nominal telescope size>$ class space telescope survey}
+
+\begin{figure}[ht!]
+%    \script{figurescript.py}
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/example_cornerplot.pdf}
+        \caption{
+        Retrieved posterior distribution of magma ocean model.
+The density maps in each panel show relationships between and marginalized distributions of the model parameters as they would be retrieved with a targeted $<nominal telescope size>$ survey of $<survey duration>$ duration.
+True values of the injected magma ocean parameters are shown in orange.
+...
+        }
+        \label{fig:cornerplot}
+    \end{centering}
+\end{figure}
+
 \section{Discussion}
+
+\subsection{Atmospheric signatures}
+\todo[inline]{discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (Tim's talk)}
 
@@
 \begin{note}
-    \textit{Here or in the introduction:} For observations of rocky exoplents, the currently best-probed regime is hot, close-in planets.
+    \textit{Here or in the introduction:} For observations of rocky exoplanets, the currently best-probed regime that of hot, close-in planets.
     These bodies experience climatic conditions that are similar to the environment of the inner Solar System bodies at early stages of their formation.
@@
 
+%\begin{acknowledgments}
+%The authors thank Terry-Ann Suer for insightful discussions.
+%\end{acknowledgements}
+
 \bibliographystyle{aasjournal}

### 2022-08-22 src/tex/ms.tex
add example statpower plot, caption

@@
 
+\begin{note}
+   The fidelity of a future detection or falsification of a magma ocean signature will depend on the significance with which the null hypothesis can be excluded.
+   As a consequence, instrumentation and survey strategy aiming at testing the hypothesis should aim at maximizing the probability of a true positive detection given the existence of the effect.
+   This is called the statistical power of the test.
+   Before turning to mission design trades that influence the statistical power in Sect.~\ref{sec:mission-design-trades}, we explore its behavior as a function of various free parameters of the magma ocean model.
+    ...
+\begin{figure*}[ht!]
+%    \script{figurescript.py}
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/example_statistical_power.pdf}
+        \caption{
+        Statistical power of a magma ocean hypothesis test as a function of model parameters.
+        }
+        \label{fig:statistical_power}
+    \end{centering}
+\end{figure*}
+
+\end{note}
+
 \section{Discussion}

### 2022-08-22 src/tex/ms.tex
add example plot of N_mo_detections, caption

@@
 
+\subsection{Statistical power of different mission designs}\label{sec:statpower_missions}
+\todo[inline]{show detectability of the magma ocean signature for different mission designs}
+
+\begin{figure}[ht!]
+%    \script{figurescript.py}
+    \begin{centering}
+
+        \includegraphics[width=\linewidth]{figures/example_Ndetected_surveys.pdf}
+        \caption{
+        Detections of magma ocean-bearing planets as a function of magma ocean occurrence for different survey designs.
+        }
+        \label{fig:example_Ndetected_surveys}
+    \end{centering}
+\end{figure}
+
 \section{Discussion}
@@
 \subsection{Mission design trades}\label{sec:mission-design-trades}
+\todo[inline]{Maps to Sect. \ref{sec:statpower_missions}.}
 \citet{Penny2019} mention that a significant increase in planet yield could be achieved if the telescope's slew speed was increased.

### 2022-08-26 src/tex/ms.tex
add example plot of a key posterior for different surveys
add software cites

@@
 \todo[inline]{Maps to Sect. \ref{sec:statpower_missions}.}
+
+\begin{note}
     \citet{Penny2019} mention that a significant increase in planet yield could be achieved if the telescope's slew speed was increased.
+\end{note}
+
+\begin{figure}[ht!]
+%    \script{figurescript.py}
+    \begin{centering}
+
+        \includegraphics[width=\linewidth]{figures/example_posterior_surveys.pdf}
+        \caption{
+        Retrieved posterior distribution of the $<key parameter>$ for different survey realizations.
+        The orange line corresponds to the true value of the injected signal.
+        }
+        \label{fig:posterior_surveys}
+    \end{centering}
+\end{figure}
 
@@
 
+\begin{large}\textit{Author contributions:}\end{large}
+
+\software{
+Bioverse~\citep{Bixel2021},
+Astropy~\citep[][]{AstropyCollaboration2018},
+NumPy~\citep[][]{Harris2020},
+SciPy~\citep[][]{Virtanen2020},
+corner.py~\citep{Foreman-Mackey2016b}.
+}
+
 \bibliographystyle{aasjournal}

### 2022-08-29 src/tex/ms.tex
introduce star and planet samples

@@
 
-    We further define the solar-equivalent semi-major axis $a_{eff} = a (L_\star/L_\odot)^{-1/2}$, at which a planet experiences the same insolation as a Solar System planet at an orbital distance $a$.
+    We further define the solar-equivalent semi-major axis $a_{eff} = a (L_\star/L_\odot)^{-1/2}$, at which a planet experiences the same insolation as a Solar System planet at an orbital distance~$a$.
 \end{note}
@@
 \subsection{Synthetic star and planet sample}
+\begin{note}
+    The first step in our analysis is to generate a synthetic sample of stars and planets.
+    First, we create a stellar population based on the observed population in the solar neighborhood, to which we then assign planetary systems based on occurrence rates derived from the \kepler mission.
+    In our fiducial model setup, we consider stars with a maximum distance to the solar system of \dmax.
+\end{note}
+
 \subsubsection{Stellar sample from Gaia DR3}
@@
 
-\subsubsection{Planetary sample from SAG~13}
+\subsubsection{Planetary ocurrence rates in orbital period and radius}
 \todo[inline]{What is the source of our occurrence rate density from which we draw?
@@
 
+\subsubsection{Transit probability}
+\todo[inline]{describe how we model inclinations and transits of synthetic planets}
+\begin{note}
+    Not all planets are transiting from our point of view.
+    We model the occurrence of transits by assuming random, isotropic orientations of planetary orbits and calculating the transit impact parameter $b = \cos(i)/R_\star$ for each planet.
+    Only planets with $|b| < 1$ are transiting and further considered.
+    For these cases we calculate the transit depth
+    \begin{equation}\label{eq:transitdepth}
+        \delta = \left( \frac{R_\mathrm{P}}{R_\star} \right)^2
+    \end{equation}\todo[inline]{check if transit depth or duration are actually used further down}
+    and transit duration
+    \begin{equation}\label{eq:transitduration}
+        t_{\mathrm{T}} = \frac{R_\star P}{\pi a} \sqrt{1 - b^2},
+    \end{equation}
+    which are relevant for the detection probability of the respective planet~(see Sect.~\ref{sec:sensitivity}).
+    After excluding all non-transiting planets, the size of our nominal sample of potentially observable planets drops to $N = \Nplanets$.
+\end{note}
+
 \subsubsection{Synthetic planet populations with and without magma oceans}
@@
 \end{figure*}
+
+\begin{note}
+\todo[inline]{Transition to tests of the trend:}
+Figure~\ref{fig:has_magmaocean} shows that the subpopulation of planets with magma oceans is apparent in the synthetic dataset.
+    In the following, we test if and under what conditions this effect is large enough to be detected with high significance.
+\end{note}
 

### 2022-08-30 src/tex/ms.tex
Introduce definition of surveys

@@
 \todo{describe the way we inject magma oceans into the synthetic planet population with Bioverse}
+\begin{note}
+
+\end{note}
+
 \begin{figure*}
@@
 =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
+\begin{note}
+    In previous steps, we generated a population of synthetic planets that orbit synthetic stars, and we have injected the statistical trend expected from the presence of magma oceans on a fraction of the planets.
+    Only a subset of these planets would be detectable by a transit survey, and their properties can be probed only with a finite precision.
+    To test the detectability of a magma ocean signature, we simulated several transit surveys with different designs and strategies and explored their ability to recover the magma ocean trend and constrain its parameters.
+
+    We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ...
+
+\end{note}
 

### 2022-09-02 src/tex/ms.tex
new section about habitability

@@
 
+\subsection{Implications for planetary habitability}\label{sec:habitability}
+\todo[inline]{Magma oceans influence the amount of water available at the surface and atmosphere, which is 1.
+commonly used as environmental marker to assess habitability, and 2.
+influences the planet's climate (greenhouse effect).}
+
 \subsection{Atmospheric signatures}
@@
 SciPy~\citep[][]{Virtanen2020},
-corner.py~\citep{Foreman-Mackey2016b}.
+corner.py~\citep{Foreman-Mackey2016b},
+dynesty~\citep{Speagle2020}.
 }

### 2022-09-08 src/tex/ms.tex
Minor edits, mostly describing methods

@@
     Here, we assess the ability of space and ground-based telescopes to test this hypothesis using Bioverse, a simulation framework that leverages contextual information from the overall planet population.
-    We argue that in the near future, ESA's PLATO mission and NASA's Roman Space Telescope will be the most promising endeavors to constrain this demographic feature.
-    For each of these missions, we identify the key mission design drivers that enable a statistically sound detection.
-    We also show the unique synergy of these missions in answering this question, and what survey strategy optimizes the statistical power of the combined dataset.
+    $\ldots$
 
-    $\ldots$ its measurement will also provide insights into which stars harbor the nearest habitable worlds.
+%    We argue that in the near future, ESA's PLATO mission and NASA's Roman Space Telescope will be the most promising endeavors to constrain this demographic feature.
+%    For each of these missions, we identify the key mission design drivers that enable a statistically sound detection.
+%    We also show the unique synergy of these missions in answering this question, and what survey strategy optimizes the statistical power of the combined dataset.
+%    $\ldots$ its measurement will also provide insights into which stars harbor the nearest habitable worlds.
 \end{abstract}
@@
 \todo{specific parametrization, dependency on bulk planet/orbit params (doesn't have to be its own subsubsection)}
-
-\Blindtext[1]
 
@@
     Only a subset of these planets would be detectable by a transit survey, and their properties can be probed only with a finite precision.
-    To test the detectability of a magma ocean signature, we simulated several transit surveys with different designs and strategies and explored their ability to recover the magma ocean trend and constrain its parameters.
+    To test the detectability of a magma ocean signature, we simulated several transit surveys with different designs and strategies and explored their capability to recover the magma ocean trend and constrain its parameters.
+    We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
+    ...
 
     We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ...
+\todo[]{double-check which survey params are needed}
 

### 2022-09-15 src/scripts/plot_luminosity_tracks.py
Add luminosity tracks figure and specify its package dependencies
first actually compiled figure added to the repo. Seems to work!

@@ -1,0 +1,69 @@
+import matplotlib as plt
+import pandas as pd
+import matplotlib.pyplot as plt
+import glob
+from natsort import natsorted
+from scipy import interpolate
+import numpy as np
+import seaborn as sns
+
+import paths
+
+# sns.set(style="ticks")
+
+
+# Change color cycle: https://matplotlib.org/3.1.1/users/dflt_style_changes.html#colors-in-default-property-cycle
+# https://seaborn.pydata.org/tutorial/color_palettes.html
+from cycler import cycler
+plt.rcParams['axes.prop_cycle'] = cycler(color=reversed(sns.cubehelix_palette(8)))
+
+lum_tracks = natsorted(glob.glob(str(paths.data) + "/luminosity_tracks/Lum_m*.txt"))
+print(lum_tracks)
+
+for lum_track in reversed(lum_tracks):
+
+    luminosity_df = pd.read_csv(lum_track)
+
+    star_mass = str(lum_track[-7:-4])
+
+    ages            = luminosity_df["age"]*1e+3
+    luminosities    = luminosity_df["lum"]
+
+    p = plt.plot(ages, luminosities, label=star_mass+" $M_{\odot}$")
+
+    # Plot interpolated luminosities
+    interpolate_luminosity  = interpolate.interp1d(ages, luminosities)
+    age_grid                = ages
+    
+    # Plot interpolated values
+    # lum_interpolated    = interpolate_luminosity(age_grid)
+    # plt.plot(age_grid, lum_interpolated, '--')
+
+    # Plot Sun as reference point
+    if star_mass == "1.0":
+        # Retrieve last color
+        c = p[-1].get_color()
+        plt.plot(4603, interpolate_luminosity([4603]), 'o', color=c)
+        plt.annotate('Current Sun',
+            xy=(4603, interpolate_luminosity([4603])), xycoords='data',
+            xytext=(0, -10), textcoords='offset pixels',
+            horizontalalignment='center',
+            verticalalignment='top', color=c)
+        
+        # Moon-forming impact
+        # plt.annotate('', xy=(30, 0.55), xycoords='data', xytext=(100, 0.55), textcoords='data', arrowprops=dict(arrowstyle="|-|, widthA=0.2, widthB=0.2", fc=c, ec=c, linewidth=1.2), horizontalalignment='center', verticalalignment='center')
[…]

### 2022-09-15 src/tex/ms.tex
Add figure showing synthetic planets (and those transiting)
Add luminosity tracks figure and specify its package dependencies
first actually compiled figure added to the repo. Seems to work!
add main assumptions and other notes

@@
 \section{Introduction}
+\todo[inline]{papers: \citet{Hamano2013,Barth2021,Downey2022}. \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, }
+
 Several patterns in the planetary parameter space have been reported in demographic studies or predicted from planet formation theories.
@@
     ...long-term goal/overarching objective: derive the geophysical history of rocky extrasolar planets.
+More concrete: Constrain the limits of runaway greenhouse transitions, and thus the inner edges of the habitable zone.
+Also: How close is Earth from this runaway greenhouse limit?
    ...
@@
 \section{Global magma oceans and their imprint on exoplanet demographics}
-\todo[inline]{"Methods Section" for the geophysical models.}
+\todo[inline]{"Methods Section" for the geophysical models.\\
+~\\
+\textbf{Main assumptions:}\\
+- Assume that every planet has \textit{some} water. Enough for runaway GH under the right conditions. Baseline water content: 20\% \\
+- Limitation to small planets without significant atmospheric envelopes.\\
+- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
+- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
+}
 \subsection{Global magma oceans}
@@
 
+\subsubsection{Stellar luminosity tracks}
+\todo[inline]{if implemented, describe the luminosity evolution of stars}
+
+\begin{figure}[ht!]
+    \script{plot_luminosity_tracks.py}
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/luminosity_tracks.pdf}
+        \caption{
+            Bolometric luminosity tracks of stars with different masses.
+        }
+        \label{fig:luminosity_tracks}
+    \end{centering}
+\end{figure}
+
 \subsubsection{Transit probability}
@@
 
+\begin{figure}
+    \begin{centering}
+        \includegraphics[width=\hsize]{figures/synthetic_population.pdf}
+        \caption{Total synthetic planet population and transiting planets.}
+        \label{fig:synthetic_population}
+    \end{centering}
+\end{figure}
+
 \begin{figure*}
@@
 \todo{Introduce survey simulation(s).
-PLATO, Ariel, LIFE?, Nautilus\\ What are threshold missions that are able to detect this signal?
+PLATO, Cheops, Ariel, LIFE?, Nautilus\\ What are threshold missions that are able to detect this signal?
 =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
@@
     \todo{explore if this can cause confusion}
+    \todo{is there overlap/confusion with the radius/density trend in \citet{Luque2022b}?}
 \end{note}
@@
 %\begin{acknowledgments}
-%The authors thank Terry-Ann Suer for insightful discussions.
+%The authors thank Paul Molliere, Arnaud Salavdor, and Terry-Ann Suer for insightful discussions.
+%We are grateful to Martin Turbet for providing the mass-radius relationship data.
 %\end{acknowledgements}

### 2022-09-15 src/scripts/plot_luminosity_tracks.py
show lum_tracks for debugging

@@ -54,6 +54,17 @@ for lum_track in reversed(lum_tracks):
         # plt.annotate('', xy=(30, 0.55), xycoords='data', xytext=(100, 0.55), textcoords='data', arrowprops=dict(arrowstyle="|-|, widthA=0.2, widthB=0.2", fc=c, ec=c, linewidth=1.2), horizontalalignment='center', verticalalignment='center')
         # plt.annotate('Moon-forming giant impact', xy=(65, 0.54), xycoords='data', xytext=(0, -5), textcoords='offset pixels', horizontalalignment='center', verticalalignment='top', color=c, fontsize=8)
 
+
+
+# DEBUG
+
+plt.text(0.1, 0.5, lum_tracks, fontsize='xx-small')
+
+
+
+
+
+
 plt.yscale("log")
 plt.xscale("log")
 

### 2022-09-15 src/scripts/plot_luminosity_tracks.py
need.to.find.this.BUG!

@@ -58,6 +58,7 @@ for lum_track in reversed(lum_tracks):
 
 # DEBUG
 
+plt.text(0.1, 0.7, 'DEBUGGING', fontsize='small')
 plt.text(0.1, 0.5, lum_tracks, fontsize='xx-small')
 
 

### 2022-09-15 src/scripts/plot_luminosity_tracks.py
another debugging attempt

@@ -11,7 +11,6 @@ import paths
 
 # sns.set(style="ticks")
 
-
 # Change color cycle: https://matplotlib.org/3.1.1/users/dflt_style_changes.html#colors-in-default-property-cycle
 # https://seaborn.pydata.org/tutorial/color_palettes.html
 from cycler import cycler
@@ -19,6 +18,16 @@ plt.rcParams['axes.prop_cycle'] = cycler(color=reversed(sns.cubehelix_palette(8)
 
 lum_tracks = natsorted(glob.glob(str(paths.data) + "/luminosity_tracks/Lum_m*.txt"))
 print(lum_tracks)
+
+
+
+# DEBUG
+plt.text(0.1, 0.7, 'DEBUGGING', fontsize='small')
+plt.text(0.1, 0.5, lum_tracks, fontsize='xx-small')
+
+
+
+
 
 for lum_track in reversed(lum_tracks):
 
@@ -53,16 +62,6 @@ for lum_track in reversed(lum_tracks):
         # Moon-forming impact
         # plt.annotate('', xy=(30, 0.55), xycoords='data', xytext=(100, 0.55), textcoords='data', arrowprops=dict(arrowstyle="|-|, widthA=0.2, widthB=0.2", fc=c, ec=c, linewidth=1.2), horizontalalignment='center', verticalalignment='center')
         # plt.annotate('Moon-forming giant impact', xy=(65, 0.54), xycoords='data', xytext=(0, -5), textcoords='offset pixels', horizontalalignment='center', verticalalignment='top', color=c, fontsize=8)
-
-
-
-# DEBUG
-
-plt.text(0.1, 0.7, 'DEBUGGING', fontsize='small')
-plt.text(0.1, 0.5, lum_tracks, fontsize='xx-small')
-
-
-
 
 
 

### 2022-09-15 src/scripts/plot_luminosity_tracks.py
remove subfolder in ...data/

@@ -16,7 +16,7 @@ import paths
 from cycler import cycler
 plt.rcParams['axes.prop_cycle'] = cycler(color=reversed(sns.cubehelix_palette(8)))
 
-lum_tracks = natsorted(glob.glob(str(paths.data) + "/luminosity_tracks/Lum_m*.txt"))
+lum_tracks = natsorted(glob.glob(str(paths.data) + "/Lum_m*.txt"))
 print(lum_tracks)
 
 

### 2022-09-24 src/tex/ms.tex
add description of stellar luminosity tracks

@@
 \todo[inline]{if implemented, describe the luminosity evolution of stars}
+\begin{note}
+    Planetary systems are hosted by stars of a wide range of ages, and stellar luminosities evolve with time.
+    Since the emergence of a runaway greenhouse phase on a planet is highly dependent on the level of radiation it receives, and thus on the luminosity of the host star, we assigned age-dependent luminosities to our synthetic stars.
+    Stellar ages are notoriously poor-constrained (CITE); we thus drew random ages from a uniform distribution from \SI{0}{\giga\year} to \SI{10}{\giga\year}.
+    We then assigned each star a luminosity from the mass-dependent evolutionary models in \citet{Baraffe1998}.
+    Figure~\ref{fig:luminosity_tracks} shows the corresponding luminosity evolution as a function of stellar mass and age.
+
+\end{note}
 
@@
 
-\section{Discussion}
+\section{Discussion}\label{sec:discussion}
+\todo[inline]{\textbf{Potential discussion points: } what is the fraction of Kepler planets w/ runaway GH.}
 
@@
 
+\section*{Data Availability}
+All data sets and software required to reproduce our results and this article are available through GitHub and Zenodo.
+The code to reproduce the figures can be accessed via the icon links associated with the respective figure caption.
+
 \software{

### 2022-09-24 src/scripts/plot_luminosity_tracks.py

@@ -11,14 +11,17 @@ import paths
 
 # sns.set(style="ticks")
 
-
 # Change color cycle: https://matplotlib.org/3.1.1/users/dflt_style_changes.html#colors-in-default-property-cycle
 # https://seaborn.pydata.org/tutorial/color_palettes.html
 from cycler import cycler
 plt.rcParams['axes.prop_cycle'] = cycler(color=reversed(sns.cubehelix_palette(8)))
 
-lum_tracks = natsorted(glob.glob(str(paths.data) + "/luminosity_tracks/Lum_m*.txt"))
+lum_tracks = natsorted(glob.glob(str(paths.data) + "/Lum_m*.txt"))
 print(lum_tracks)
+
+# DEBUG
+# plt.text(0.1, 0.7, 'DEBUGGING', fontsize='small')
+# plt.text(0.1, 0.5, lum_tracks, fontsize='xx-small')
 
 for lum_track in reversed(lum_tracks):
 
@@ -54,16 +57,12 @@ for lum_track in reversed(lum_tracks):
         # plt.annotate('', xy=(30, 0.55), xycoords='data', xytext=(100, 0.55), textcoords='data', arrowprops=dict(arrowstyle="|-|, widthA=0.2, widthB=0.2", fc=c, ec=c, linewidth=1.2), horizontalalignment='center', verticalalignment='center')
         # plt.annotate('Moon-forming giant impact', xy=(65, 0.54), xycoords='data', xytext=(0, -5), textcoords='offset pixels', horizontalalignment='center', verticalalignment='top', color=c, fontsize=8)
 
-plt.yscale("log")
-plt.xscale("log")
-
+plt.loglog()
 plt.ylabel("Bolometric luminosity ($L/L_{\odot}$)")
 plt.xlabel("Time (Myr)")
 
 sns.despine()
-
 plt.legend(ncol=1)
-
 plt.tight_layout(pad=0.2, w_pad=0.5, h_pad=0.5) # https://matplotlib.org/users/tight_layout_guide.html
 
 plt.savefig(paths.figures / "luminosity_tracks.pdf")

### 2022-09-28 src/tex/ms.tex
change instellation to new value

@@
 \begin{note}
-    The power from the host star per unit area at the position of a planet or stellar insolation $S$ in units of Earth's insolation is given by
+    The power from the host star per unit area at the position of a planet or instellation $S$ in units of Earth's insolation is given by
     \begin{equation}
@@
     \end{equation}
-
-    We further define the solar-equivalent semi-major axis $a_{eff} = a (L_\star/L_\odot)^{-1/2}$, at which a planet experiences the same insolation as a Solar System planet at an orbital distance~$a$.
+    Here, we assume global redistribution of incoming flux and a fixed albedo of $0.3$, comparable to Earth's [CITE].
+    The absolute value of $S_\oplus$, by which all synthetic planets are scaled, is then \SI{238}{\watt\per\square\meter}.
+    We further define the solar-equivalent semi-major axis $a_{eff} = a (L_\star/L_\odot)^{-1/2}$, at which a planet experiences the same instellation as a Solar System planet at an orbital distance~$a$.
 \end{note}
@@
     Another potential false positive contribution comes from the "Neptune desert``, a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
-    The shape of this region is such that smaller planets become less frequent the closer to the star they are, which to some degree resembles the pattern introduced by the insolation dependency of the magma ocean probability.
+    The shape of this region is such that smaller planets become less frequent the closer to the star they are, which to some degree resembles the pattern introduced by the instellation dependency of the magma ocean probability.
     \todo{explore if this can cause confusion}

### 2022-09-30 src/tex/ms.tex
describe mass-radius relationship

@@
 \subsubsection{Stellar sample from Gaia DR3}
-\todo{Which star sample was actually used?}
-
-\subsubsection{Planetary ocurrence rates in orbital period and radius}
-\todo[inline]{What is the source of our occurrence rate density from which we draw?
-We could also use a generative model that directly approximates an observed or theoretical density \citep[e.g.,][]{Schlecker2021b}.}
-\begin{note}
-    The NASA Exoplanet Program Analysis Group chartered Science Analysis Group 13 (SAG 13)...
-    The inferred occurrence rate density can be expressed as a power law in planet radius and period
-    \begin{equation}
-        \frac{\partial^2n}{\partial \log R \, \partial \log P} = \Gamma R^{\alpha} P^{\beta}
-    \end{equation}
-that is broken in planet radius, resulting in the free parameters $\Gamma$, $\alpha$, $\beta$, and a breakpoint radius $R_\mathrm{break}$.
-\end{note}
+\todo[inline]{KEVIN: describe stellar population}
 
 \subsubsection{Stellar luminosity tracks}
-\todo[inline]{if implemented, describe the luminosity evolution of stars}
 \begin{note}
@@
 \end{figure}
+
+\subsubsection{Orbital and bulk properties of the planet sample}
+
+\todo[inline]{GALEN: Describe underlying planet occurrence rates of synthetic population, its Mstar dependency, any caveats or assumptions. Probably replace SAG13-based content below.}
+\begin{note}
+    The NASA Exoplanet Program Analysis Group chartered Science Analysis Group 13 (SAG 13)...
+    The inferred occurrence rate density can be expressed as a power law in planet radius and period
+    \begin{equation}
+        \frac{\partial^2n}{\partial \log R \, \partial \log P} = \Gamma R^{\alpha} P^{\beta}
+    \end{equation}
+that is broken in planet radius, resulting in the free parameters $\Gamma$, $\alpha$, $\beta$, and a breakpoint radius $R_\mathrm{break}$.
+\end{note}
+
+\begin{note}
+    ...
+    In this study, we consider only rocky planets with a maximum mass of \SI{2}{\Mearth}. % (Turbet+2020 goes only up to 2 Mearth)
+
+    ...
+    To assign masses to our planets, we use the theoretical mass-radius relationship derived in \citet{Zeng2016}.
+    The default relation our planets follow is one assuming a pure $\mathrm{MgSiO_3}$ composition. 
+    \todo[inline]{JUSTIFY this choice}
+    This represents the baseline bulk density before any magma ocean-related effects are applied.
+\end{note}
 

### 2022-10-13 src/tex/ms.tex
drop a train of thoughts in the introduction

@@
 \section{Introduction}
-\todo[inline]{papers: \citet{Hamano2013,Barth2021,Downey2022}. \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, }
+\todo[inline]{papers: \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.
+\citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,}}
 
-Several patterns in the planetary parameter space have been reported in demographic studies or predicted from planet formation theories.
 \begin{note}
+(Several patterns in the planetary parameter space have been reported in demographic studies or predicted from planet formation theories.)
+    Venus and Earth, while having accreted from the same mass reservoir and despite their similar bulk properties, evolved into planets with very different environmental conditions on their surfaces.
+    At the formation of the Solar System, both planets underwent a giant impact phase [CITE!] that melted their mantles, leading to a magma ocean stage.
+    Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
+    \citet{Hamano2013} showed that such a late desiccation process is not needed to explain the stark differences: for a planet rich in volatiles and receiving high enough radiation levels from its host star, a steam atmosphere limits the outgoing radiation from the (molten) planetary surface.
+    This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}, enough to remove the entire water reservoir from a rocky planet by hydrodynamic escape.
+    Under the right conditions, different orbital distances alone can thus decide about a planet's water content and ultimately its habitability.
+    If received flux makes the difference between our own habitable planet and dry, dead Venus, there is good reason to believe that such patterns exist in exoplanet systems as well.
+
     ...long-term goal/overarching objective: derive the geophysical history of rocky extrasolar planets.
@@
 \todo{Introduce survey simulation(s).
-PLATO, Cheops, Ariel, LIFE?, Nautilus\\ What are threshold missions that are able to detect this signal?
+PLATO, Cheops, Ariel, LIFE (reach out to team; Hendrik Linz produced a LIFE Technology Overview document)?, Nautilus\\ What are threshold missions that are able to detect this signal?
 =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
@@
 %\begin{acknowledgments}
-%The authors thank Paul Molliere, Arnaud Salavdor, and Terry-Ann Suer for insightful discussions.
+%The authors thank Yann Fabel, Paul Molliere, and Terry-Ann Suer for insightful discussions.
 %We are grateful to Martin Turbet for providing the mass-radius relationship data.

### 2022-10-26 src/tex/ms.tex
add description of null and alternative hypothesis and corresponding figure

@@
 \todo{Define the null and alternative hypotheses}
+\begin{note}
+    As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
+    \begin{equation}
+        h_{\mathrm{null}}(\theta, S) = \theta,
+    \end{equation}
+    where $\theta$ is the set of parameters defining the radius distribution.
+    We further define an alternative hypothesis that describes radius changes due to steam atmospheres and water in the melt.
+    As motivated above, this hypothesis takes the form of a step function in instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
+    Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
+    The magma ocean hypothesis is then defined as
+\begin{equation}
+    h_{\mathrm{MO}}(\theta, S) =
+%        \begin{cases}
+%        1 & \text{if } x \in \mathbb{Q}\\
+%        0 & \text{if } x \in \mathbb{R}\setminus\mathbb{Q}
+%    \end{cases}
+        \begin{cases}
+        h_{\mathrm{null}} & \text{for } S \leq S_\mathrm{thresh}\\
+        <R_\mathrm{P}> (f_\mathrm{steam}, f_\mathrm{water}) & \text{for } S > S_\mathrm{thresh}.
+    \end{cases}
+\end{equation}
+    Here, $f_\mathrm{steam}$ and $f_\mathrm{water}$ are predictions from the steam atmosphere and water incorporation models, respectively.
+    They are assumed to act additively on the planet radii and thus on their average $<R_\mathrm{P}>$.
+
+%        \begin{cases}
+%        \end{cases}
+
+\begin{figure}
+    \begin{centering}
+        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
+        \caption{Average radius predictions of the null hypothesis and magma ocean hypothesis.
+The sharp boundary of the runaway greenhouse regime causes a discontinuity at the threshold instellation $S_\mathrm{thresh}$, which we set to \SI{280}{\watt\per\meter\squared} in the fiducial case.}
+        \label{fig:HnullHmo}
+    \end{centering}
+\end{figure}
+    Figure~\ref{fig:HnullHmo} shows the null and alternative hypotheses as a function of instellation.
+    The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
+
+\end{note}
 ...

### 2022-11-23 src/tex/ms.tex
add a (preliminary) bullet-point list of conclusions

@@
 \begin{note}
-    \textit{Here or in the introduction:} For observations of rocky exoplanets, the currently best-probed regime that of hot, close-in planets.
+    \textit{Here or in the introduction:} For observations of rocky exoplanets, the currently best-probed regime is that of warm, close-in planets.
     These bodies experience climatic conditions that are similar to the environment of the inner Solar System bodies at early stages of their formation.
@@
 \end{note}
+
+The key findings of our work are as follows:
+\begin{enumerate}
+    \item Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.
+    \item Since XXX\SI{}{\percent} of confirmed exoplanets in this size range are in the runaway greenhouse regime, we suggest that this imprint is detectable with transit measurements of a large enough sample and sufficient precision.
+    \item Using Bioverse, we find that the planned PLATO transit survey will provide a sufficient sample to test the magma ocean hypothesis.
+    \item The diagnostic power of transit missions in testing this hypothesis can be maximized by ...
+    \item A detection of the feature will constrain the water inventory of rocky exoplanets, providing important context in the assessment of their habitability.
+\end{enumerate}
 

### 2022-11-23 src/tex/ms.tex
map conclusions back to Discussion section

@@
 \section{Testing the magma ocean hypothesis}
-\todo[inline]{"Methods Section" of the hypothesis testing part.
+\todo[inline]{Methods and results section of the hypothesis testing part.
 Introduce the idea of constraining magma ocean frequencies and/or physics with Bayesian stats on exoplanet population; introduce Bioverse; report methods of the hypothesis tests.}
@@
 
-\subsection{Signature and testability of the magma ocean hypothesis with a $<nominal telescope size>$ class space telescope survey}
+\subsection{Signature and testability of the magma ocean hypothesis with the PLATO transit survey}
 
@@
 \todo[inline]{\textbf{Potential discussion points: } what is the fraction of Kepler planets w/ runaway GH.}
+\subsection{Expected imprint of runaway greenhouse phases on exoplanet demographics}
+
+\subsection{Detectability of the magma ocean demographic feature}
+
+\subsection{Diagnostic power of upcoming exoplanet missions}\todo[inline]{PLATO: needed sample size, precision}
+
+\subsection{Mission design trades}\label{sec:mission-design-trades}
 
@@
 
-\subsection{Atmospheric signatures}
+\subsection{OPTIONAL...: Atmospheric signatures}
 \todo[inline]{discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (Tim's talk)}
@@
 
-\subsection{Mission design trades}\label{sec:mission-design-trades}
 \todo[inline]{Maps to Sect. \ref{sec:statpower_missions}.}

### 2022-11-23 src/tex/ms.tex
add a paragraph on PLATO

@@
 \subsection{Signature and testability of the magma ocean hypothesis with the PLATO transit survey}
+PLATO (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
+The PLATO team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).
+In the FGK sample of the mission, a minimum of 500 Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
+We base our assessment of PLATO's diagnostic power to test the magma ocean hypothesis on these estimates.
 

### 2022-11-23 src/tex/ms.tex
Introduce Bioverse, general workflow

@@
 
-\begin{figure*}
-    \begin{centering}
-        \includegraphics[width=\hsize]{figures/flowchart.pdf}
-        \caption{Workflow of our hypothesis testing with Bioverse.
-In the first block, a stellar sample is generated based on XXX.
-The stars are then populated with planets from XXX, which may be assigned a magma ocean based on the model described in Sect.~\ref{sec:mo_model}.
-The planets' respective transit probabilities are computed.
-The second block simulates the exoplanet survey whereby selection effects and detection biases are introduced.
-Finally, the third block deals with testing a hypothesis based on the data from the simulated survey.
-By iterating through these steps, we compute the statistical power of testing the hypothesis given the assumed survey design.}
-        \label{fig:flowchart}
-    \end{centering}
-\end{figure*}
-
 \section{Global magma oceans and their imprint on exoplanet demographics}
@@
 Introduce the idea of constraining magma ocean frequencies and/or physics with Bayesian stats on exoplanet population; introduce Bioverse; report methods of the hypothesis tests.}
+The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
+This section describes our approach.
+
+The main task was to determine -- for different survey configurations -- the confidence level with which a magma ocean signal can be detected in simulated exoplanet observations.
+For this purpose we used and expanded the Bioverse framework~\citep{Bixel2020,Bixel2021}\footnote{Bioverse is actively maintained and documented open source software written in Python.
+Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic stellar and planetary samples into which we injected magma ocean signals, simulated observations of this sample, and computed Bayesian evidences in favor of the magma ocean hypothesis (see Fig.~\ref{fig:flowchart}).
+\begin{figure*}
+    \begin{centering}
+        \includegraphics[width=\hsize]{figures/flowchart.pdf}
+        \caption{Workflow of our hypothesis testing with Bioverse.
+In the first block, a stellar sample is generated based on XXX.
+The stars are then populated with planets from XXX, which may be assigned a magma ocean based on the model described in Sect.~\ref{sec:mo_model}.
+The planets' respective transit probabilities are computed.
+The second block simulates the exoplanet survey whereby selection effects and detection biases are introduced.
+Finally, the third block deals with testing a hypothesis based on the data from the simulated survey.
+By iterating through these steps, we compute the statistical power of testing the hypothesis given the assumed survey design.}
+        \label{fig:flowchart}
+    \end{centering}
+\end{figure*}
 
 \subsection{Synthetic star and planet sample}
-\begin{note}
 The first step in our analysis is to generate a synthetic sample of stars and planets.
-    First, we create a stellar population based on the observed population in the solar neighborhood, to which we then assign planetary systems based on occurrence rates derived from the \kepler mission.
-    In our fiducial model setup, we consider stars with a maximum distance to the solar system of \dmax.
-\end{note}
+First, we created a stellar population based on the observed population in the solar neighborhood, to which we then assign planetary systems based on occurrence rates derived from the \kepler mission.
 
 \subsubsection{Stellar sample from Gaia DR3}
+Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
 \todo[inline]{KEVIN: describe stellar population}
@@
         \caption{
-            Bolometric luminosity tracks of stars with different masses.
+            Bolometric luminosity tracks of stars with different masses, computed from stellar evolution models in \citet{Baraffe1998}.
+            Low-mass stars, which make up the majority of stars in the solar neighborhood, undergo an extended early phase of several magnitudes higher luminosity before entering a lifetime of relative faintness.
         }

### 2022-11-23 src/tex/ms.tex
prepare description of planetary sample

@@
 \subsubsection{Stellar luminosity tracks}
-\begin{note}
     Planetary systems are hosted by stars of a wide range of ages, and stellar luminosities evolve with time.
@@
     Figure~\ref{fig:luminosity_tracks} shows the corresponding luminosity evolution as a function of stellar mass and age.
-
-\end{note}
-
 \begin{figure}[ht!]
@@
 \subsubsection{Orbital and bulk properties of the planet sample}
-
-\todo[inline]{GALEN: Describe underlying planet occurrence rates of synthetic population, its Mstar dependency, any caveats or assumptions. Probably replace SAG13-based content below.}
-\begin{note}
-    The NASA Exoplanet Program Analysis Group chartered Science Analysis Group 13 (SAG 13)...
-    The inferred occurrence rate density can be expressed as a power law in planet radius and period
+Next, we assigned planetary systems to the synthetic stellar sample.
+The orbital parameters and bulk properties of these simulated planets are informed by statistics from the \kepler mission; we adopted the planetary occurrence rates recently derived in \citep{Bergsten2022}.
+Following \citep{Youdin2011a}, their inferred occurrence rate density can be expressed in the form
 \begin{equation}
-        \frac{\partial^2n}{\partial \log R \, \partial \log P} = \Gamma R^{\alpha} P^{\beta}
+    \frac{\mathrm{d}^2n}{\mathrm{d}R \, \partial P} = F_0 C_n g(R, P),
 \end{equation}
-that is broken in planet radius, resulting in the free parameters $\Gamma$, $\alpha$, $\beta$, and a breakpoint radius $R_\mathrm{break}$.
-\end{note}
+where $F_0$ ...
+\todo[inline]{GALEN: Describe underlying planet occurrence rates of synthetic population, its Mstar dependency, any caveats or assumptions.}
 
@@
 \subsubsection{Transit probability}
-\todo[inline]{describe how we model inclinations and transits of synthetic planets}
 \begin{note}
@@
     We model the occurrence of transits by assuming random, isotropic orientations of planetary orbits and calculating the transit impact parameter $b = \cos(i)/R_\star$ for each planet.
-    Only planets with $|b| < 1$ are transiting and further considered.
+    Only planets with $|b| < 1$ are potentially observable in transit and further considered.
     For these cases we calculate the transit depth
@@
     which are relevant for the detection probability of the respective planet~(see Sect.~\ref{sec:sensitivity}).
-    After excluding all non-transiting planets, the size of our nominal sample of potentially observable planets drops to $N = \Nplanets$.
+    Excluding all non-transiting planets diminishes the sample to \SI{\percentageTransiting}{\percent} of its original size.
 \end{note}
@@
 Other causes of shrinking planet sizes have been put forward, the most widely discussed ones being atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
-    Both processes are being traded as potentially sculpting the observed radius bimodality of small, close-in exoplanets from the \textit{Kepler} mission~\citep{Fulton2017,VanEylen2018}.
+    Both processes are being traded as potentially sculpting the observed radius bimodality of small, close-in exoplanets from the  mission~\citep{Fulton2017,VanEylen2018}.
     Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.

### 2022-11-28 src/tex/ms.tex
describe planetary sample selection cuts
Separate subsections on star and planet sample

@@
 
-\subsection{Synthetic star and planet sample}
+\subsection{Synthetic star sample}
 The first step in our analysis is to generate a synthetic sample of stars and planets.
-First, we created a stellar population based on the observed population in the solar neighborhood, to which we then assign planetary systems based on occurrence rates derived from the \kepler mission.
+First, we created a stellar population based on the observed population in the solar neighborhood, to which we then assign planetary systems based on occurrence rates derived from the \kepler\ mission.
 
@@
 
-\subsubsection{Orbital and bulk properties of the planet sample}
+\subsection{Synthetic planet sample}
 Next, we assigned planetary systems to the synthetic stellar sample.
-The orbital parameters and bulk properties of these simulated planets are informed by statistics from the \kepler mission; we adopted the planetary occurrence rates recently derived in \citep{Bergsten2022}.
+The orbital parameters and bulk properties of these simulated planets are informed by statistics from the \kepler\ mission.
+
+\subsubsection{Planet occurrence rates in orbital period and radius}
+We adopted the planetary occurrence rates recently derived in \citep{Bergsten2022}.
 Following \citep{Youdin2011a}, their inferred occurrence rate density can be expressed in the form
@@
 
-\begin{note}
-    ...
-    In this study, we consider only rocky planets with a maximum mass of \SI{2}{\Mearth}. % (Turbet+2020 goes only up to 2 Mearth)
+\subsubsection{Eccentricities, orbital inclinations, and planet masses}
 
 ...
+
 To assign masses to our planets, we use the theoretical mass-radius relationship derived in \citet{Zeng2016}.
@@
 This represents the baseline bulk density before any magma ocean-related effects are applied.
+...
+
+\subsubsection{Sample selection}
+\begin{note}
+    ...
+    In this study, we consider only rocky planets with masses \SIrange{0.1}{2}{\Mearth}. % (Turbet+2020 covers only 0.1 to 2 Mearth)
+    As we are interested in the effects of runaway greenhouse phases of planets with volatiles and atmosphere, we further excluded lava worlds with extremely high instellation that does not allow long-term preservation of an atmosphere.
+    We thus removed all planets with an instellation $S > \SI{2000}{\watt\per\square\meter}$.
 \end{note}

### 2022-11-28 src/tex/ms.tex
describe choice of eccentricities, inclinations

@@
 \subsubsection{Eccentricities, orbital inclinations, and planet masses}
+Eccentric orbits alter the probability of a planet to transit \citep[e.g.,][]{Barnes2007a}.
+The distribution of eccentricities $e$ of exoplanets has been found to resemble a Beta function \citep{Kipping2013b}, which we chose to draw synthetic eccentricities from.
+Following~\citet{Kipping2013b}, we used a Beta distribution with parameters $a=0.867$ and $b=3.03$, and truncated the distribution at $e = 0.8$.
+
+Another parameter with an even higher impact on transit probabilities is the orbital inclination toward the observer $i$.
+Assuming isotropic alignments of orbits, we assigned each planet an inclination drawn from a distribution uniform in $\cos(i)$.
 

### 2022-11-30 src/tex/ms.tex
define sample selections

@@
 More concrete: Constrain the limits of runaway greenhouse transitions, and thus the inner edges of the habitable zone.
-Also: How close is Earth from this runaway greenhouse limit?
+Also: How close is Earth from this runaway greenhouse irradiation limit?
    ...
@@
 
-...
-
-To assign masses to our planets, we use the theoretical mass-radius relationship derived in \citet{Zeng2016}.
+To assign masses to our planets, we use the semi-empirical mass-radius relationship derived in \citet{Zeng2016}.
 The default relation our planets follow is one assuming a pure $\mathrm{MgSiO_3}$ composition.
-\todo[inline]{JUSTIFY this choice}
+\todo[inline]{JUSTIFY this choice (TIM?)}
 This represents the baseline bulk density before any magma ocean-related effects are applied.
-...
 
@@
 \begin{note}
-    ...
-    In this study, we consider only rocky planets with masses \SIrange{0.1}{2}{\Mearth}. % (Turbet+2020 covers only 0.1 to 2 Mearth)
-    As we are interested in the effects of runaway greenhouse phases of planets with volatiles and atmosphere, we further excluded lava worlds with extremely high instellation that does not allow long-term preservation of an atmosphere.
-    We thus removed all planets with an instellation $S > \SI{2000}{\watt\per\square\meter}$.
+    \todo[inline]{use criterion of Zahnle \& Caitling 2017 instead? $0.8 S^{0.25} < R$}
+    The runaway greenhouse effect becomes obsolete where no atmosphere can be maintained due to proximity to the host star and resulting atmospheric erosion.
+    We thus excluded planets with extremely high instellation that does not allow long-term preservation of an atmosphere and clear our sample from all planets with an instellation $S > \SI{2000}{\watt\per\square\meter}$.
+    We further consider only rocky planets with masses below \SI{2}{\Mearth}. % (Turbet+2020 covers only 0.1 to 2 Mearth)
 \end{note}

### 2022-12-01 src/tex/ms.tex
describe MO injection with Bioverse; introduce the three structure cases

@@
 \todo[inline]{Methods and results section of the hypothesis testing part.
-Introduce the idea of constraining magma ocean frequencies and/or physics with Bayesian stats on exoplanet population; introduce Bioverse; report methods of the hypothesis tests.}
+Introduce the idea of constraining magma ocean frequencies and/or physics with Bayesian stats on exoplanet population; introduce Bioverse; report methods of the hypothesis tests.\\
+Caveats: see \citet{Turbet2019} Sect. 3, \citet{Bower2019}}
 The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
@@
 
-\subsection{Synthetic planet sample}
+\subsection{Synthetic planet sample}\label{sec:syn_planets}
 Next, we assigned planetary systems to the synthetic stellar sample.
@@
 \begin{note}
+    Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
+    To enable a quantitative assessment of this detectability, we injected the signal into the simulated planet sample before observing it with simulated surveys.
+    While we search for the magma ocean pattern in demographic quantities such as average planet radii, the injected changes happen on the planetary level: We changed each planet's radius based on its individual set of properties and the associated predictions from steam atmosphere and water incorporation models.
+    Relevant properties are a planet's mass $M$, its instellation $S$, and its bulk water inventory expressed as a water mass fraction $x_{H_2O}$.
+    We consider three different cases:
+    \begin{itemize}
+   \item[A.] No radius change.
+    This case serves as our null hypothesis, where planetary radii are not affected by the amount of irradiation the planets receive.
+    It also applies outside the runaway greenhouse limit, where we assume a solid planetary surface and no steam atmosphere.
+        A planet that lies \textit{within} the runaway greenhouse limit but that is already completely desiccated will have the same radius.
+\todo{Mention if this case is considered in the model or not (as of now, no).}
+   \item[B.] Radius inflation due to a steam atmosphere.
+    We assigned planets within the runaway greenhouse limit steam atmospheres, which increase atmospheric scale heights significantly.
+   \item[C.] Radius decrease of inflated planets due to water in the melt.
+    Due to greenhouse forcing, planetary surfaces below steam atmospheres are thought to be molten.
+    Part of the water will then partition into the melt, removing it from the steam atmosphere, which in turn decreases the radius inflation.
+    This effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes typically reach measurable values.
+    \end{itemize}
+\end{note}
 
-\end{note}
+In case A, a planet simply retains the radius it was assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
+Case B assumes an inflated radius due to a steam atmosphere for runaway greenhouse planets.
+To quantify this radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
+Their calculations start from the same underlying mass-radius relations that we use for case-A-planets~\citep{Zeng2016}.
+...
 
@@
 %The authors thank Yann Fabel, Paul Molliere, and Terry-Ann Suer for insightful discussions.
-%We are grateful to Martin Turbet for providing the mass-radius relationship data.
+%We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
 %\end{acknowledgements}

### 2022-12-03 src/tex/ms.tex
describe case B) steam atmospheres, dry melt

@@
     This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}, enough to remove the entire water reservoir from a rocky planet by hydrodynamic escape.
+    \todo[inline]{important for motivating the existence of the pattern: ``the solidification timescale can become comparable to the main-sequence lifetime of the star (Hamano et al.
+2013, 2015)'' \citep{Lichtenberg2022} }
     Under the right conditions, different orbital distances alone can thus decide about a planet's water content and ultimately its habitability.
@@
 To quantify this radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
-Their calculations start from the same underlying mass-radius relations that we use for case-A-planets~\citep{Zeng2016}.
+\begin{note}
+For the rocky interiors, their calculations rely on the mass-radius relations from ~\citep{Zeng2016} that we use for case-A-planets.
+For each planet receiving higher instellation than the runaway greenhouse threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
+We chose a threshold dayside-averaged instellation of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$, which was found as a limit for the emitted flux in a runaway greenhouse situation~\citep{Goldblatt2013,Leconte2013}.
+We note that an additional radius change due to potentially molten interiors~\citep{Bower2019} is not taken into account here.
+\end{note}
 ...

### 2022-12-03 src/tex/ms.tex
add plot comparing M-R relations of different cases

@@
 ...
+
+\begin{figure}
+    \begin{centering}
+        \includegraphics[width=\hsize]{figures/radiuscomparison.pdf}
+        \caption{Mass-radius relationships for a water mass fraction of \wrr\ and different planet states: A) fully solid planets without magma oceans.
+B) planets with steam atmospheres.
+C) As B), but including the effect of water incorporation in the melt.}
+        \label{fig:radiuscomparison}
+    \end{centering}
+\end{figure}
+
+Figure~\ref{fig:radiuscomparison} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of \wrr.
+In general, steam atmospheres (case B) cause a significant radius increase, which is slightly reduced when water incorporation in the melt is allowed (case C).
 

### 2022-12-06 src/tex/ms.tex
edit case A and B injections

@@
 \section{Introduction}
-\todo[inline]{papers: \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.
-\citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,}}
+\todo[inline]{papers: \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
 
@@
 In case A, a planet simply retains the radius it was assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
-Case B assumes an inflated radius due to a steam atmosphere for runaway greenhouse planets.
-To quantify this radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
-\begin{note}
+There is no additional dependency on orbital distance or instellation.
+
+Case B assumes an inflated radius due to a steam atmosphere for all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$.
+This value was found to be a limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Leconte2013}.
+To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
 For the rocky interiors, their calculations rely on the mass-radius relations from ~\citep{Zeng2016} that we use for case-A-planets.
-For each planet receiving higher instellation than the runaway greenhouse threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
-We chose a threshold dayside-averaged instellation of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$, which was found as a limit for the emitted flux in a runaway greenhouse situation~\citep{Goldblatt2013,Leconte2013}.
+For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
 We note that an additional radius change due to potentially molten interiors~\citep{Bower2019} is not taken into account here.
-\end{note}
 ...

### 2022-12-06 src/scripts/plotstyle.py
add matplotlib styles

@@ -1,0 +1,99 @@
+""" Config file with my personal preference of default matplotlibrc settings.
+"""
+from matplotlib import rc, rcParams
+from cycler import cycler
+
+
+def get_color_palette():
+    """ return a custom made color palette."""
+    return ['#029e73', '#fc4f30', '#008fd5',
+            '#e5ae38', '#810f7c', '#00035b', '#005249',
+            '#8b8b8b', '#fe828c']
+
+
+def styleplots():
+    # Set plot style
+    spinewidth = 1.8
+
+    font = {'family': 'sans',
+            'weight': 'normal',
+            'size': 13}  # 9 pt corresponds to caption size of 1-col A&A
+    legend = {'handlelength': 0.75,
+              'handletextpad': 0.5}
+    figure = {'dpi': 100,
+              'subplot.left': 0.16,  # the left side of the subplots of the figure
+              'subplot.bottom': 0.21,  # the bottom of the subplots of the figure
+              'subplot.right': 0.98,  # the right side of the subplots of the figure
+              'subplot.top': 0.97,  # the top of the subplots of the figure
+              'subplot.hspace': 0.0}  # height reserved for space between subplots
+
+    rc('font', **font)
+    rc('legend', **legend)
+    rc('figure', **figure)
+    rc('lines', linewidth=2.5)
+
+    rc('axes', linewidth=spinewidth)
+    rc('xtick.major', width=spinewidth)
+    rc('ytick.major', width=spinewidth)
+    rc('xtick.minor', width=.5 * spinewidth)
+    rc('ytick.minor', width=.5 * spinewidth)
+    rc('image', cmap='inferno')
+    rc('hist', bins=20)
+    rc('patch', edgecolor='black')
+    rc('savefig', bbox='tight', dpi=200)
+
+    rcParams['axes.prop_cycle'] = cycler(color=get_color_palette())
+
+
+def set_size(width='aa', subplot=[1,1], scale=1):
+    """ Set aesthetic figure dimensions to avoid scaling in latex.
+
+    Parameters
+    ----------
+    width: float or string
+        either width in pts or one of the following strings:
+        'aa' : A&A column width
+        'aaDouble' : A&A total text width
+    subplot : list
+        subplot dimensions in [rows, columns]
+    scale: float
+        fraction of the width which you wish the figure to occupy
+
+    Returns
+    -------
+    fig_dim: tuple
[…]

### 2022-12-07 src/tex/ms.tex
describe case C, update figure

@@
 \subsubsection{Synthetic planet populations with and without magma oceans}
-\todo{describe the way we inject magma oceans into the synthetic planet population with Bioverse}
 \begin{note}
@@
     We assigned planets within the runaway greenhouse limit steam atmospheres, which increase atmospheric scale heights significantly.
-   \item[C.] Radius decrease of inflated planets due to water in the melt.
+   \item[C.] As B, but including radius decrease of inflated planets due to water in the melt.
     Due to greenhouse forcing, planetary surfaces below steam atmospheres are thought to be molten.
@@
 
-In case A, a planet simply retains the radius it was assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
+In case~A, a planet simply retains the radius it was assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
 There is no additional dependency on orbital distance or instellation.
 
-Case B assumes an inflated radius due to a steam atmosphere for all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$.
+Case~B assumes an inflated radius due to a steam atmosphere for all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$.
 This value was found to be a limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Leconte2013}.
 To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
-For the rocky interiors, their calculations rely on the mass-radius relations from ~\citep{Zeng2016} that we use for case-A-planets.
+For the rocky interiors, their calculations rely on the same mass-radius relations from ~\citep{Zeng2016} that we applied for case~A-planets.
 For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
 We note that an additional radius change due to potentially molten interiors~\citep{Bower2019} is not taken into account here.
-...
+
+Case~C contains case~B, but also includes a radius decrease due to water partitioning in the melt.
+The magnitude of this effect is again dependent on the planet mass and on the water mass fraction.
+We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
+We then add the (in almost all cases negative) radius deviations to the planet radii computed for case~B.
 
@@
         \includegraphics[width=\hsize]{figures/radiuscomparison.pdf}
-        \caption{Mass-radius relationships for a water mass fraction of \wrr\ and different planet states: A) fully solid planets without magma oceans.
+        \caption{Mass-radius relationships for a water mass fraction $x_{H_2O}= \wrr$ and different planet states: A) fully solid planets without magma oceans.
 B) planets with steam atmospheres.
@@
 
-Figure~\ref{fig:radiuscomparison} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of \wrr.
+Figure~\ref{fig:radiuscomparison} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of $x_{H_2O}= \wrr$.
 In general, steam atmospheres (case B) cause a significant radius increase, which is slightly reduced when water incorporation in the melt is allowed (case C).

### 2022-12-07 src/scripts/plot_luminosity_tracks.py
edit lum tracks plot

@@ -6,22 +6,21 @@ from natsort import natsorted
 from scipy import interpolate
 import numpy as np
 import seaborn as sns
+import cmocean
 
 import paths
+import plotstyle
 
 # sns.set(style="ticks")
 
 # Change color cycle: https://matplotlib.org/3.1.1/users/dflt_style_changes.html#colors-in-default-property-cycle
 # https://seaborn.pydata.org/tutorial/color_palettes.html
 from cycler import cycler
-plt.rcParams['axes.prop_cycle'] = cycler(color=reversed(sns.cubehelix_palette(8)))
+# plt.rcParams['axes.prop_cycle'] = cycler(color=reversed(sns.cubehelix_palette(8)))
+plt.rcParams['axes.prop_cycle'] = cycler(color=reversed(sns.color_palette("cmo.thermal", n_colors=8, desat=None)))
 
 lum_tracks = natsorted(glob.glob(str(paths.data) + "/Lum_m*.txt"))
-print(lum_tracks)
-
-# DEBUG
-# plt.text(0.1, 0.7, 'DEBUGGING', fontsize='small')
-# plt.text(0.1, 0.5, lum_tracks, fontsize='xx-small')
+# print(lum_tracks)
 
 for lum_track in reversed(lum_tracks):
 
@@ -46,11 +45,11 @@ for lum_track in reversed(lum_tracks):
     if star_mass == "1.0":
         # Retrieve last color
         c = p[-1].get_color()
-        plt.plot(4603, interpolate_luminosity([4603]), 'o', color=c)
-        plt.annotate('Current Sun',
+        plt.plot(4603, interpolate_luminosity([4603]), 'o', color=c, ms=8)
+        plt.annotate('current Sun',
             xy=(4603, interpolate_luminosity([4603])), xycoords='data',
-            xytext=(0, -10), textcoords='offset pixels',
-            horizontalalignment='center',
+            xytext=(-9, +12), textcoords='offset pixels',
+            horizontalalignment='right',
             verticalalignment='top', color=c)
         
         # Moon-forming impact
@@ -62,7 +61,10 @@ plt.ylabel("Bolometric luminosity ($L/L_{\odot}$)")
 plt.xlabel("Time (Myr)")
 
 sns.despine()
-plt.legend(ncol=1)
+# plt.legend(loc='lower left', ncol=4, bbox_to_anchor=(0.0, 1.),
[…]

### 2022-12-08 src/tex/ms.tex
remove synthetic population scatter plot
will be combined with hypothesis plot
Unify MO/no MO and hypotheses plots, describe in text

@@
 Figure~\ref{fig:radiuscomparison} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of $x_{H_2O}= \wrr$.
-In general, steam atmospheres (case B) cause a significant radius increase, which is slightly reduced when water incorporation in the melt is allowed (case C).
-
-\begin{figure}
-    \begin{centering}
-        \includegraphics[width=\hsize]{figures/synthetic_population.pdf}
-        \caption{Total synthetic planet population and transiting planets.}
-        \label{fig:synthetic_population}
-    \end{centering}
-\end{figure}
-
-\begin{figure*}
-    \begin{centering}
-        \includegraphics[width=\hsize]{figures/has_magmaocean.pdf}
-        \caption{Synthetic planets with and without magmaoceans as a function of Solar-equivalent semi-major axis}
-        \label{fig:has_magmaocean}
-    \end{centering}
-\end{figure*}
+In general, steam atmospheres (case~B) cause a significant radius increase, which is slightly reduced when water incorporation in the melt is allowed (case~C).
+Our nominal simulation setup is such that all planets receiving an instellation $S < S_\mathrm{thresh}$ follow relation~A, and all planets with $S > S_\mathrm{thresh}$ follow relation~C.
 
@@
         \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
-        \caption{Average radius predictions of the null hypothesis and magma ocean hypothesis.
-The sharp boundary of the runaway greenhouse regime causes a discontinuity at the threshold instellation $S_\mathrm{thresh}$, which we set to \SI{280}{\watt\per\meter\squared} in the fiducial case.}
+        \caption{Planets above and below the runaway greenhouse threshold.
+        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
+        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of instellation.
+Only the planets marked as transiting are observable.
+Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, radii are inflated by the amount indicated with gray lines.
+The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue).
+This magma ocean hypothesis can be tested agains the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
         \label{fig:HnullHmo}

### 2022-12-09 src/tex/ms.tex
transition to testing the runaway GH transition

@@
 Our nominal simulation setup is such that all planets receiving an instellation $S < S_\mathrm{thresh}$ follow relation~A, and all planets with $S > S_\mathrm{thresh}$ follow relation~C.
-
-\begin{note}
-\todo[inline]{Transition to tests of the trend:}
-Figure~\ref{fig:has_magmaocean} shows that the subpopulation of planets with magma oceans is apparent in the synthetic dataset.
-    In the following, we test if and under what conditions this effect is large enough to be detected with high significance.
-\end{note}
+This introduces a sharp discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
+In the following, we test if and under what conditions this trend is large enough to be detected with high significance.
 

### 2022-12-09 src/tex/ms.tex
list main drivers of diagnostic power

@@
 
-The main task was to determine -- for different survey configurations -- the confidence level with which a magma ocean signal can be detected in simulated exoplanet observations.
+The main task is to determine -- for different survey configurations -- the confidence level with which a magma ocean signal can be detected in simulated exoplanet observations.
 For this purpose we used and expanded the Bioverse framework~\citep{Bixel2020,Bixel2021}\footnote{Bioverse is actively maintained and documented open source software written in Python.
@@
 =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
-\begin{note}
 In previous steps, we generated a population of synthetic planets that orbit synthetic stars, and we have injected the statistical trend expected from the presence of magma oceans on a fraction of the planets.
@@
 We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
-...
 
-    We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ...
+%We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ...
 \todo[]{double-check which survey params are needed}
-
-\end{note}
+We identify the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a given transit survey:
+\begin{itemize}
+    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction.
+As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern.
+    \item Sample size: Since we are looking for a population-level trend, the statistical significance will increase with a larger sample size.
+    \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be.
+Good \textit{accuracy} is less important, as long as it does not have a systematic error scaling with stellar irradiance.
+    \item Availability and precision of mass measurements: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space.
+If transiting planets can be followed up to obtain mass measurements, the statistical significance increases.
+\end{itemize}
+Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.
+    This can be different for young host stars when their ages can not be well constrained; in particular the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).
 
@@
 \end{note}
+
 ...
@@
 
-    Another potential false positive contribution comes from the "Neptune desert``, a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
+    Another false positive contribution may come from potential unknown variations in the occurrence rate gradients in radius-instellation space.
+    They can impact the statistical inference of the transitions, especially if these variations are similar to the expected pattern.
+    Although an abrupt pattern at the expected runaway greenhouse transition seems unlikely, examples of steep occurrence rate density changes exist.
+    An example is the "Neptune desert``, a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
     The shape of this region is such that smaller planets become less frequent the closer to the star they are, which to some degree resembles the pattern introduced by the instellation dependency of the magma ocean probability.
-    \todo{explore if this can cause confusion}
+    However, the Neptune desert occurs at smaller orbital periods and its location depends on a planet's size~\citep{Szabo2011}, which is not expected for the runaway greenhouse transition.
     \todo{is there overlap/confusion with the radius/density trend in \citet{Luque2022b}?}

### 2022-12-10 src/tex/ms.tex
introduce survey simulation module

@@
 
-\subsubsection{Sample selection}
-\begin{note}
-    \todo[inline]{use criterion of Zahnle \& Caitling 2017 instead? $0.8 S^{0.25} < R$}
-    The runaway greenhouse effect becomes obsolete where no atmosphere can be maintained due to proximity to the host star and resulting atmospheric erosion.
-    We thus excluded planets with extremely high instellation that does not allow long-term preservation of an atmosphere and clear our sample from all planets with an instellation $S > \SI{2000}{\watt\per\square\meter}$.
-    We further consider only rocky planets with masses below \SI{2}{\Mearth}. % (Turbet+2020 covers only 0.1 to 2 Mearth)
-\end{note}
-
 \subsubsection{Transit probability}
@@
 Only a subset of these planets would be detectable by a transit survey, and their properties can be probed only with a finite precision.
-To test the detectability of a magma ocean signature, we simulated several transit surveys with different designs and strategies and explored their capability to recover the magma ocean trend and constrain its parameters.
+To test the detectability of a runaway greenhouse transition, we simulated several transit surveys with different designs and strategies and explored their capability to recover the magma ocean trend and constrain its parameters.
 We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
@@
     This can be different for young host stars when their ages can not be well constrained; in particular the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).
+
+\subsubsection{Survey simulations}
+\begin{note}
+The survey module of Bioverse converts the synthetic planet sample into a set of uncertainty-laden measurements of some parameters of a subset of that sample.
+   This task includes selection of the targets, application of any detection biases, and conducting simulated measurements.
+   All of the above are specific to the particular survey.
+
+Current and near-future missions focusing on Earth-sized planets target several host star spectral types, and their exact sample selection is based on specific mission objectives.
+    We thus chose a broad selection function centered on general detectability of transits with state-of-the-art technology.
+
+\end{note}
+
+\subsubsection{Sample selection}
+\begin{note}
+    \todo[inline]{use criterion of Zahnle \& Caitling 2017 instead? $0.8 S^{0.25} < R$}
+    The runaway greenhouse effect becomes obsolete where no atmosphere can be maintained due to proximity to the host star and resulting atmospheric erosion.
+    We thus excluded planets with extremely high instellation that does not allow long-term preservation of an atmosphere and clear our sample from all planets with an instellation $S > \SI{2000}{\watt\per\square\meter}$.
+    We further consider only rocky planets with masses below \SI{2}{\Mearth}. % (Turbet+2020 covers only 0.1 to 2 Mearth)
+\end{note}
 

### 2022-12-13 src/tex/ms.tex
Include detection biases for PLATO, CHEOPS

@@
 
-\section{Testing the magma ocean hypothesis}
+\section{Observational test of the runaway greenhouse transition}
 \todo[inline]{Methods and results section of the hypothesis testing part.
@@
 
-\subsection{Testing the magma ocean hypothesis with current and planned exoplanet missions}
+\subsection{Detecting the runaway greenhouse transition with current and planned exoplanet missions}
 \todo{Introduce survey simulation(s).
@@
 
-\subsubsection{Sample selection}
+\subsubsection{Detection bias and sample selection}\label{sec:sensitivity}
+Not all transiting planets are detectable with the same likelihood and detection biases have an impact on the demographic measurements we are interested in.
+A detailed characterization of the detection biases of individual missions would not be justifiable given the uncertainties of the theoretical predictions.
+Instead, we derived generic observing limits that reflect the limitations of state-of-the-art transit surveys.
+
+The runaway greenhouse transition is expected to occur at orbits sufficiently close to let mission duration not be a limiting factor: $S_\mathrm{thresh}$ occurs close to Earth's orbital period for \mbox{G-type} stars and on significantly shorter periods for the more prevalent M~dwarf hosts~\citep{Goldblatt2012}.
+However, for a transit signal to be detected in a transit survey, it must reach a sufficient signal-to-noise ratio (S/N).
+This S/N is sensitive on the photometric precision that can be achieved~\citep[e.g.,][]{Burke2015,Hardegree-Ullman2019}.
+In its in-orbit commissioning phase, the \cheops\ mission demonstrated a photometric precision of \SI{75}{\ppm} for faint ($V\approx 12$) stars~\citep{Benz2021}.
+In comparison, \plato's expected precision for similar stars was estimated to be \SI{\sim 50}{\ppm}~\citep[][Matuszewski et al., in prep.]{plato2017}.
+To represent these photometric limits, we chose a minimum transit depth of \SI{75}{\ppm} as a detection limit and consider only measurements of planets exceeding this threshold.
+
 \begin{note}
     \todo[inline]{use criterion of Zahnle \& Caitling 2017 instead? $0.8 S^{0.25} < R$}
+    Another selection concerns the climate regime of interest.
     The runaway greenhouse effect becomes obsolete where no atmosphere can be maintained due to proximity to the host star and resulting atmospheric erosion.

### 2022-12-15 src/tex/ms.tex
introduce hypothesis tests, form of hypothesis

@@
 
-\subsubsection{Definition of the magma ocean hypothesis and null hypothesis}
-\todo{Define the null and alternative hypotheses}
+\subsubsection{Definition of the runaway greenhouse hypothesis}
+We now turn to assessing the simulated surveys in terms of their ability to detect the runaway greenhouse transition and their constraining power on parameters of this trend.
+To do this, we rely on a Bayesian hypothesis testing approach where we quantify the evidence of a hypothesis over another based on the (simulated) data.
+For our specific problem, this implies comparing evidences for a demographic imprint of the runaway greenhouse effect to its absence.
+
+\begin{note}
+%    Bayesian model comparison requires that an alternative mode be specified against which the comparison is made.
+\end{note}
+
 \begin{note}
@@
     The magma ocean hypothesis is then defined as
-\begin{equation}
+\begin{equation}\label{eq:mo_hypo}
     h_{\mathrm{MO}}(\theta, S) =
@@
 \begin{equation}
-    \mathcal{L}(R_\mathrm{P} \mid \boldsymbol{\theta})=\prod_{i}^{N} \frac{1}{\sqrt{2 \pi \sigma_{R_\mathrm{P}, i}^{2}}} \exp \left(-\frac{\left(R_{\mathrm{P}, i}-h\left(\boldsymbol{\theta}, a_{\mathrm{eff}, i}\right)\right)^{2}}{2 \sigma_{R_\mathrm{P}, i}^{2}}\right).
+    \mathcal{L}(R_\mathrm{P} \mid \boldsymbol{\theta})=\prod_{i}^{N} \frac{1}{\sqrt{2 \pi \sigma_{R_\mathrm{P}, i}^{2}}} \exp \left(-\frac{\left(R_{\mathrm{P}, i}-h\left(\boldsymbol{\theta}, S_i\right)\right)^{2}}{2 \sigma_{R_\mathrm{P}, i}^{2}}\right).
 \end{equation}
-Here, $h\left(\boldsymbol{\theta}, a_{\mathrm{eff}, i}\right)$ corresponds to the functional form of the magma ocean hypothesis\todo{link to eqn defining the hypothesis} .
-We tested this hypothesis against the null hypothesis $h_\mathrm{null} (\theta, R_\mathrm{P}) = \theta$, which states that there is no relationship between the measured planet radii and (scaled) semi-major axes.
+Here, $h\left(\boldsymbol{\theta}, S_i\right)$ corresponds to the functional form of the magma ocean hypothesis~(Eqn.~\ref{eq:mo_hypo}).
+We tested this hypothesis against the null hypothesis $h_\mathrm{null} (\theta, R_\mathrm{P}) = \theta$, which states that there is no relationship between the measured planet radii and received instellation.
 
@@
 %\begin{acknowledgments}
-%The authors thank Yann Fabel, Paul Molliere, and Terry-Ann Suer for insightful discussions.
+%The authors thank Yann Fabel, Paul Molliere, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
 %We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.

### 2022-12-20 src/tex/ms.tex
introduce moving average, likelihood function

@@
 \section{Introduction}
-\todo[inline]{papers: \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
+\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
 
@@
     They are assumed to act additively on the planet radii and thus on their average $<R_\mathrm{P}>$.
-
-%        \begin{cases}
-%        \end{cases}
-
 \begin{figure}
@@
     The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
-
 \end{note}
 
-...
+\begin{note}
+The measured radii $R_\mathrm{P, i}$ cannot be directly used for the hypothesis tests as they include intrinsic scatter that is not caused by measurement errors.
+$h_{\mathrm{MO}}$ and $h_{\mathrm{null}}$ should thus be tested against a statistical estimator that represents the population mean.
+To avoid binning and the artificial patterns it may introduce, we chose to test our hypotheses against a simple moving average $SMA$ along the instellation axis with a window of size \windowsize\ centered around each measurement.
+We further computed the uncertainty of this moving average by propagating the individual measurement errors and applying a rolling standard error of the mean. \todo[inline]{specify error propagation}
 
-For the likelihood function, we assumed that the planet radii $R_{\mathrm{P}, i}$ are measured with a normally distributed uncertainty $\sigma_{R_\mathrm{P}, i}$ and adopted a normal distribution
+As our procedure involves random sampling of the model parameters $\theta$, we need to define the probability of obtaining a dataset given the model parameters, i.e., a likelihood function.
+We assumed here that the individual moving averages $SMA_i$ are measured with a normally distributed uncertainty $\sigma_{SMA_i}$ and adopted a normal distribution
 \begin{equation}
-    \mathcal{L}(R_\mathrm{P} \mid \boldsymbol{\theta})=\prod_{i}^{N} \frac{1}{\sqrt{2 \pi \sigma_{R_\mathrm{P}, i}^{2}}} \exp \left(-\frac{\left(R_{\mathrm{P}, i}-h\left(\boldsymbol{\theta}, S_i\right)\right)^{2}}{2 \sigma_{R_\mathrm{P}, i}^{2}}\right).
+    \mathcal{L}(SMA \mid \boldsymbol{\theta})=\prod_{i}^{N} \frac{1}{\sqrt{2 \pi \sigma_{SMA_i}^{2}}} \exp \left(-\frac{\left(SMA_i - h\left(\boldsymbol{\theta}, S_i\right)\right)^{2}}{2 \sigma_{SMA_i}^{2}}\right).
 \end{equation}

### 2022-12-20 src/tex/ms.tex
state model parameters for H_MO, H_null

@@
 \begin{note}
+    The only free parameter of the null hypothesis, which assumes the average radius to be independent of instellation, is the predicted mean radius $<R_\mathrm{P}>$.
+    The functional form of the magma ocean hypothesis is more complex: Besides the mean radius of planets outside the runaway greenhouse threshold $<R_\mathrm{P}>_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$.
+    Further, the parameter encoding the amount of radius change is the planetary water mass fraction~$x_{H_2O}$.
+\end{note}
+
+\begin{note}
 The measured radii $R_\mathrm{P, i}$ cannot be directly used for the hypothesis tests as they include intrinsic scatter that is not caused by measurement errors.

### 2022-12-20 src/tex/ms.tex
explain priors, nested sampling

@@
 Here, $h\left(\boldsymbol{\theta}, S_i\right)$ corresponds to the functional form of the magma ocean hypothesis~(Eqn.~\ref{eq:mo_hypo}).
-We tested this hypothesis against the null hypothesis $h_\mathrm{null} (\theta, R_\mathrm{P}) = \theta$, which states that there is no relationship between the measured planet radii and received instellation.
+We tested this hypothesis against the null hypothesis $h_\mathrm{null} (\theta, S) = \theta$, which states that there is no relationship between the measured planet radii and received instellation.
+\end{note}
 
-\subsection{Signature and testability of the magma ocean hypothesis with the PLATO transit survey}
+\subsubsection{Hypothesis tests}
+We are now ready to assess the relative plausibility of $h_{\mathrm{MO}}$ and $h_{\mathrm{null}}$ given the synthetic data we have generated, assigning equal a~priori~ probabilities to these models.
+This is done by comparing the Bayesian evidence $\mathcal{Z}$ of the models, which we estimate with the nested sampling~\citep{Skilling2004} algorithm \emph{dynesty}~\citep{Speagle2020}.
+Our criterion to reject the null hypothesis is
+\begin{equation}
+\Delta \ln \mathcal{Z}  = \ln \mathcal{Z} - \ln \mathcal{Z}_\mathrm{null}  > 3.
+\end{equation}
+The chosen threshold roughly corresponds to the $p = 0.05$ threshold for statistical significance popular in frequentist approaches.\todo{CITATION}
+
+A sensible choice of priors is central for evidence estimation via nested sampling.
+As the parameters we are interested in recovering, $S_\mathrm{thresh}$ and $x_{H_2O}$, are poorly constrained by previous data, we used relatively uninformative priors to sample the entire physically plausible parameter space.
+For $S_\mathrm{thresh}$, we chose a log-uniform prior in $[\prSmin,\prSmax]\, \SI{}{\watt\per\meter\squared}$.
+Similarly, we sample $x_{H_2O}$ from a log-uniform distribution.
+Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Sect.~\ref{sec:mo_model}).
+Finally, we adopt a broad, uniform prior for $<R_\mathrm{P}>$ bound by $[\prRmin, \prRmax] \, \mathrm{R_\oplus}$.
+We then ran the nested sampler to compute the log-evidence and sample the posterior distributions.
+
+\subsection{Signature of the magma ocean hypothesis}
+
+\subsection{Testability of the magma ocean hypothesis with the PLATO transit survey}
 PLATO (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
@@
    As a consequence, instrumentation and survey strategy aiming at testing the hypothesis should aim at maximizing the probability of a true positive detection given the existence of the effect.
-   This is called the statistical power of the test.
+   This is called the \textit{statistical power}: the probability of correctly rejecting a false null hypothesis.
    Before turning to mission design trades that influence the statistical power in Sect.~\ref{sec:mission-design-trades}, we explore its behavior as a function of various free parameters of the magma ocean model.

### 2022-12-20 src/tex/ms.tex
make mathematical notation consistent

@@
 \subsubsection{Synthetic planet populations with and without magma oceans}
-\begin{note}
     Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
@@
     \end{itemize}
-\end{note}
 
@@
 
-\begin{note}
     As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
@@
         h_{\mathrm{null}} & \text{for } S \leq S_\mathrm{thresh}\\
-        <R_\mathrm{P}> (f_\mathrm{steam}, f_\mathrm{water}) & \text{for } S > S_\mathrm{thresh}.
+        \langle R_\mathrm{P}\rangle (f_\mathrm{steam}, f_\mathrm{water}) & \text{for } S > S_\mathrm{thresh}.
     \end{cases}
@@
     Here, $f_\mathrm{steam}$ and $f_\mathrm{water}$ are predictions from the steam atmosphere and water incorporation models, respectively.
-    They are assumed to act additively on the planet radii and thus on their average $<R_\mathrm{P}>$.
+    They are assumed to act additively on the planet radii and thus on their average $\langle R_\mathrm{P}\rangle $.
 \begin{figure}
@@
     The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
-\end{note}
 
 \begin{note}
-    The only free parameter of the null hypothesis, which assumes the average radius to be independent of instellation, is the predicted mean radius $<R_\mathrm{P}>$.
-    The functional form of the magma ocean hypothesis is more complex: Besides the mean radius of planets outside the runaway greenhouse threshold $<R_\mathrm{P}>_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$.
+    The only free parameter of the null hypothesis, which assumes the average radius to be independent of instellation, is the predicted mean radius $\langle R_\mathrm{P}\rangle $.
+    The functional form of the magma ocean hypothesis is more complex: Besides the mean radius of planets outside the runaway greenhouse threshold $\langle R_\mathrm{P}\rangle_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$.
     Further, the parameter encoding the amount of radius change is the planetary water mass fraction~$x_{H_2O}$.
@@
 Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Sect.~\ref{sec:mo_model}).
-Finally, we adopt a broad, uniform prior for $<R_\mathrm{P}>$ bound by $[\prRmin, \prRmax] \, \mathrm{R_\oplus}$.
-We then ran the nested sampler to compute the log-evidence and sample the posterior distributions.
+Finally, we adopt a broad, uniform prior for $\langle R_\mathrm{P}\rangle_\mathrm{out}$ bound by $[\prRmin, \prRmax] \, \mathrm{R_\oplus}$.
+We then ran the nested sampler to compute the evidence and sample the posterior distributions.
 

### 2022-12-22 src/tex/ms.tex
separate methods and results

@@
 \section{Observational test of the runaway greenhouse transition}
-\todo[inline]{Methods and results section of the hypothesis testing part.
+\todo[inline]{Methods section of the hypothesis testing part.
 Introduce the idea of constraining magma ocean frequencies and/or physics with Bayesian stats on exoplanet population; introduce Bioverse; report methods of the hypothesis tests.\\
@@
 
-\subsection{Detecting the runaway greenhouse transition with current and planned exoplanet missions}
-\todo{Introduce survey simulation(s).
-PLATO, Cheops, Ariel, LIFE (reach out to team; Hendrik Linz produced a LIFE Technology Overview document)?, Nautilus\\ What are threshold missions that are able to detect this signal?
-=>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
-In previous steps, we generated a population of synthetic planets that orbit synthetic stars, and we have injected the statistical trend expected from the presence of magma oceans on a fraction of the planets.
-Only a subset of these planets would be detectable by a transit survey, and their properties can be probed only with a finite precision.
-To test the detectability of a runaway greenhouse transition, we simulated several transit surveys with different designs and strategies and explored their capability to recover the magma ocean trend and constrain its parameters.
-We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
-
-%We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ...
-\todo[]{double-check which survey params are needed}
-We identify the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a given transit survey:
-\begin{itemize}
-    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction.
-As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern.
-    \item Sample size: Since we are looking for a population-level trend, the statistical significance will increase with a larger sample size.
-    \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be.
-Good \textit{accuracy} is less important, as long as it does not have a systematic error scaling with stellar irradiance.
-    \item Availability and precision of mass measurements: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space.
-If transiting planets can be followed up to obtain mass measurements, the statistical significance increases.
-\end{itemize}
-Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.
-    This can be different for young host stars when their ages can not be well constrained; in particular the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).
-
-\subsubsection{Survey simulations}
+\subsection{Survey simulations}
 \begin{note}
@@
 
-\subsubsection{Detection bias and sample selection}\label{sec:sensitivity}
+\subsection{Detection bias and sample selection}\label{sec:sensitivity}
 Not all transiting planets are detectable with the same likelihood and detection biases have an impact on the demographic measurements we are interested in.
@@
 
-\subsubsection{Definition of the runaway greenhouse hypothesis}
+\subsection{Definition of the runaway greenhouse hypothesis}
 We now turn to assessing the simulated surveys in terms of their ability to detect the runaway greenhouse transition and their constraining power on parameters of this trend.
@@
 The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue).
-This magma ocean hypothesis can be tested agains the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
+This magma ocean hypothesis can be tested against the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
         \label{fig:HnullHmo}
@@
 
-\subsubsection{Hypothesis tests}
+\subsection{Hypothesis tests}
 We are now ready to assess the relative plausibility of $h_{\mathrm{MO}}$ and $h_{\mathrm{null}}$ given the synthetic data we have generated, assigning equal a~priori~ probabilities to these models.
@@
 
+\section{Results}
 \subsection{Signature of the magma ocean hypothesis}
 
-\subsection{Testability of the magma ocean hypothesis with the PLATO transit survey}
+\subsection{Detecting the runaway greenhouse transition with current and planned exoplanet missions}
+\todo{Introduce survey simulation(s).
+PLATO, Cheops, Ariel, LIFE (reach out to team; Hendrik Linz produced a LIFE Technology Overview document)?, Nautilus\\ What are threshold missions that are able to detect this signal?
+=>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
+In previous steps, we generated a population of synthetic planets that orbit synthetic stars, and we have injected the statistical trend expected from the presence of magma oceans on a fraction of the planets.
+Only a subset of these planets would be detectable by a transit survey, and their properties can be probed only with a finite precision.
+To test the detectability of a runaway greenhouse transition, we simulated several transit surveys with different designs and strategies and explored their capability to recover the magma ocean trend and constrain its parameters.
+We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
+
+%We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ...
+\todo[]{double-check which survey params are needed}
+We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a given transit survey:
+\begin{itemize}
+    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction.
+As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern.
+    \item Sample size: Since we are looking for a population-level trend, the statistical significance will increase with a larger sample size.
+    \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be.
+Good \textit{accuracy} is less important, as long as it does not have a systematic error scaling with stellar irradiance.
+    \item Availability and precision of mass measurements: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space.
+If transiting planets can be followed up to obtain mass measurements, the statistical significance increases.
+\end{itemize}
+Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.
+This can be different for young host stars when their ages can not be well constrained; in particular the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).
+
 PLATO (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
@@
 \todo[inline]{\textbf{Potential discussion points: } what is the fraction of Kepler planets w/ runaway GH.}
-\subsection{Expected imprint of runaway greenhouse phases on exoplanet demographics}
+\subsection{Detectability of the runaway greenhouse threshold}
 
-\subsection{Detectability of the magma ocean demographic feature}
+\subsubsection{False positive scenarios}
+\todo{what other processes could be confused with a magma ocean signal?}
+\begin{note}
+    Global magma oceans are not the only physical mechanism that may cause a decrease in transit radii for a subset of planets.
+    Other causes of shrinking planet sizes have been put forward, the most widely discussed ones being atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
+    Both processes are being traded as potentially sculpting the observed radius bimodality of small, close-in exoplanets from the  mission~\citep{Fulton2017,VanEylen2018}.
+    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
+    This region is distinct from the one affected by global magma oceans, since... \todo{...elaborate why they cannot be confused}
+
+    Another false positive contribution may come from potential unknown variations in the occurrence rate gradients in radius-instellation space.
+    They can impact the statistical inference of the transitions, especially if these variations are similar to the expected pattern.
+    Although an abrupt pattern at the expected runaway greenhouse transition seems unlikely, examples of steep occurrence rate density changes exist.
+    An example is the "Neptune desert``, a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
+    The shape of this region is such that smaller planets become less frequent the closer to the star they are, which to some degree resembles the pattern introduced by the instellation dependency of the magma ocean probability.
+    However, the Neptune desert occurs at smaller orbital periods and its location depends on a planet's size~\citep{Szabo2011}, which is not expected for the runaway greenhouse transition.
+    \todo{is there overlap/confusion with the radius/density trend in \citet{Luque2022b}?}
+\end{note}
+
+\todo[inline]{Maps to Sect. \ref{sec:statpower_missions}.}
+
+\begin{note}
+    \citet{Penny2019} mention that a significant increase in planet yield could be achieved if the telescope's slew speed was increased.
+\end{note}
+
+\begin{figure}[ht!]
+%    \script{figurescript.py}
+    \begin{centering}
+
+        \includegraphics[width=\linewidth]{figures/example_posterior_surveys.pdf}
+        \caption{
+            Retrieved posterior distribution of the $<key parameter>$ for different survey realizations.
+            The orange line corresponds to the true value of the injected signal.
+        }
+        \label{fig:posterior_surveys}
+    \end{centering}
+\end{figure}
+
+\subsubsection{OPTIONAL...: Atmospheric signatures}
+\todo[inline]{discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (Tim's talk)}
 

### 2022-12-22 src/tex/ms.tex
update prospective conclusions, restructure discussion accordingly

@@
 
-\subsection{OPTIONAL...: Atmospheric signatures}
-\todo[inline]{discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (Tim's talk)}
-
-\subsection{False positive scenarios}
-\todo{what other processes could be confused with a magma ocean signal?}
-\begin{note}
-Global magma oceans are not the only physical mechanism that may cause a decrease in transit radii for a subset of planets.
-Other causes of shrinking planet sizes have been put forward, the most widely discussed ones being atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
-    Both processes are being traded as potentially sculpting the observed radius bimodality of small, close-in exoplanets from the  mission~\citep{Fulton2017,VanEylen2018}.
-    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
-    This region is distinct from the one affected by global magma oceans, since... \todo{...elaborate why they cannot be confused}
-
-    Another false positive contribution may come from potential unknown variations in the occurrence rate gradients in radius-instellation space.
-    They can impact the statistical inference of the transitions, especially if these variations are similar to the expected pattern.
-    Although an abrupt pattern at the expected runaway greenhouse transition seems unlikely, examples of steep occurrence rate density changes exist.
-    An example is the "Neptune desert``, a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
-    The shape of this region is such that smaller planets become less frequent the closer to the star they are, which to some degree resembles the pattern introduced by the instellation dependency of the magma ocean probability.
-    However, the Neptune desert occurs at smaller orbital periods and its location depends on a planet's size~\citep{Szabo2011}, which is not expected for the runaway greenhouse transition.
-    \todo{is there overlap/confusion with the radius/density trend in \citet{Luque2022b}?}
-\end{note}
-
-\todo[inline]{Maps to Sect. \ref{sec:statpower_missions}.}
-
-\begin{note}
-    \citet{Penny2019} mention that a significant increase in planet yield could be achieved if the telescope's slew speed was increased.
-\end{note}
-
-\begin{figure}[ht!]
-%    \script{figurescript.py}
-    \begin{centering}
-
-        \includegraphics[width=\linewidth]{figures/example_posterior_surveys.pdf}
-        \caption{
-        Retrieved posterior distribution of the $<key parameter>$ for different survey realizations.
-        The orange line corresponds to the true value of the injected signal.
-        }
-        \label{fig:posterior_surveys}
-    \end{centering}
-\end{figure}
-
 \section{Conclusions}
@@
     Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.
+    Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.
+    In this study, we assessed the feasibility of a detection of this pattern with current and near-future exoplanet missions.
 \end{note}
@@
 \begin{enumerate}
-    \item Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.
-    \item Since XXX\SI{}{\percent} of confirmed exoplanets in this size range are in the runaway greenhouse regime, we suggest that this imprint is detectable with transit measurements of a large enough sample and sufficient precision.
-    \item Using Bioverse, we find that the planned PLATO transit survey will provide a sufficient sample to test the magma ocean hypothesis.
+    \item The demographic imprint of the runaway greenhouse threshold should be detectable with transit measurements of a large enough sample and sufficient precision.
+    \item We find that the planned PLATO transit survey will provide a sufficient sample to test the magma ocean hypothesis.
     \item The diagnostic power of transit missions in testing this hypothesis can be maximized by ...

### 2022-12-22 src/tex/ms.tex
add figure showing optimistic scenario

@@
 The main task is to determine -- for different survey configurations -- the confidence level with which a magma ocean signal can be detected in simulated exoplanet observations.
-For this purpose we used and expanded the Bioverse framework~\citep{Bixel2020,Bixel2021}\footnote{Bioverse is actively maintained and documented open source software written in Python.
+For this purpose we expanded the Bioverse framework~\citep{Bixel2020,Bixel2021}\footnote{Bioverse is actively maintained and documented open source software written in Python.
 Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic stellar and planetary samples into which we injected magma ocean signals, simulated observations of this sample, and computed Bayesian evidences in favor of the magma ocean hypothesis (see Fig.~\ref{fig:flowchart}).
@@
         \caption{Workflow of our hypothesis testing with Bioverse.
-In the first block, a stellar sample is generated based on XXX.
-The stars are then populated with planets from XXX, which may be assigned a magma ocean based on the model described in Sect.~\ref{sec:mo_model}.
+In the first block, a stellar sample is generated based on the Gaia nearby stars catalog~\citep{Smart2021}\todo[]{KEVIN pls verify}.
+The stars are then populated with planets based on Kepler demographics, and they may be assigned a magma ocean based on the model described in Sect.~\ref{sec:mo_model}.
 The planets' respective transit probabilities are computed.
@@
 \subsection{Signature of the magma ocean hypothesis}
+\todo[inline]{show general detectability}
+
+\begin{figure}[ht!]
+%    \script{figurescript.py}
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
+        \caption{
+        Detection of the runaway greenhouse threshold in the optimistic case.
+        From simulated radius and instellation measurements of a large ($N = \Nplanets$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:mo_hypo}, random draws from the posterior in red).
+            The pattern is detected with high significance.
+        }
+        \label{fig:optimistic_R-S}
+    \end{centering}
+\end{figure}
+Figure~\ref{fig:optimistic_R-S} shows a prototypical detection of the runaway greenhouse threshold.
+For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \Nplanets$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
+We assumed a water mass fractions of \wrr\ for each planet.
+In this case, the pattern was detected with high significance ($\ln \mathcal{Z} \approx 100$).
 

### 2022-12-22 src/tex/ms.tex
remove figure with schematics of the geophysical model
Merge methods sections into one

@@
 
-\section{Global magma oceans and their imprint on exoplanet demographics}
-\todo[inline]{"Methods Section" for the geophysical models.\\
-~\\
-\textbf{Main assumptions:}\\
-- Assume that every planet has \textit{some} water. Enough for runaway GH under the right conditions. Baseline water content: 20\% \\
-- Limitation to small planets without significant atmospheric envelopes.\\
-- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
-- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
-}
-\subsection{Global magma oceans}
-\todo[inline]{introduce geophysical model.\\ Introduce its parametrization either here or further down when modeling demographic imprint.\\ Figure~\ref{fig:magmaocean_model} goes here.}
-
-%\begin{figure}
-\begin{figure}[!ht]
-    \begin{centering}
-        \includegraphics[width=\hsize]{figures/magmaocean_model.jpg}
-        \caption{Schematics of the geophysical model framework.}
-        \label{fig:magmaocean_model}
-    \end{centering}
-\end{figure}
-
-\subsection{Demographic imprint of magma oceans}
-\todo{describe the expected imprint on the exoplanet population. \\Provide parametrizations from \citet{Dorn2021}'s models.}
-
-\subsubsection{Parametrization}
-\label{sec:mo_model}
-\todo{specific parametrization, dependency on bulk planet/orbit params (doesn't have to be its own subsubsection)}
-
-\begin{note}
-    The power from the host star per unit area at the position of a planet or instellation $S$ in units of Earth's insolation is given by
-    \begin{equation}
-        \frac{S}{S_\oplus} = \left(\frac{L_\star}{L_\odot}\right) \left(\frac{au}{a}\right)^2 .
-    \end{equation}
-    Here, we assume global redistribution of incoming flux and a fixed albedo of $0.3$, comparable to Earth's [CITE].
-    The absolute value of $S_\oplus$, by which all synthetic planets are scaled, is then \SI{238}{\watt\per\square\meter}.
-    We further define the solar-equivalent semi-major axis $a_{eff} = a (L_\star/L_\odot)^{-1/2}$, at which a planet experiences the same instellation as a Solar System planet at an orbital distance~$a$.
-\end{note}
-
-\section{Observational test of the runaway greenhouse transition}
-\todo[inline]{Methods section of the hypothesis testing part.
-Introduce the idea of constraining magma ocean frequencies and/or physics with Bayesian stats on exoplanet population; introduce Bioverse; report methods of the hypothesis tests.\\
+\section{Methods}\label{sec:methods}
+\todo[inline]{Introduce the idea of constraining magma ocean frequencies and/or physics with Bayesian stats on exoplanet population; introduce Bioverse; report methods of the hypothesis tests.\\
 Caveats: see \citet{Turbet2019} Sect. 3, \citet{Bower2019}}
@@
 
-\subsubsection{Synthetic planet populations with and without magma oceans}
+\subsection{Planet radius changes due to runaway greenhouse climates}\label{sec:mo_model}
+\todo[inline]{"Methods Section" for the geophysical models. Describe specific parametrization, dependency on bulk planet/orbit params\\
+~\\
+\textbf{Main assumptions:}\\
+- Assume that every planet has \textit{some} water. Enough for runaway GH under the right conditions. Baseline water content: 20\% \\
+- Limitation to small planets without significant atmospheric envelopes.\\
+- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
+- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
+}
+...
+
+\begin{note}
+    The power from the host star per unit area at the position of a planet or instellation $S$ in units of Earth's insolation is given by
+    \begin{equation}
+        \frac{S}{S_\oplus} = \left(\frac{L_\star}{L_\odot}\right) \left(\frac{au}{a}\right)^2 .
+    \end{equation}
+    Here, we assume global redistribution of incoming flux and a fixed albedo of $0.3$, comparable to Earth's [CITE].
+    The absolute value of $S_\oplus$, by which all synthetic planets are scaled, is then \SI{238}{\watt\per\square\meter}.
+    We further define the solar-equivalent semi-major axis $a_{eff} = a (L_\star/L_\odot)^{-1/2}$, at which a planet experiences the same instellation as a Solar System planet at an orbital distance~$a$.
+\end{note}
+
     Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.

### 2022-12-22 src/tex/ms.tex
Restructure methods section
...into three subsections reflecting the three pillars of Bioverse

@@
 
-\subsection{Synthetic star sample}
-The first step in our analysis is to generate a synthetic sample of stars and planets.
-First, we created a stellar population based on the observed population in the solar neighborhood, to which we then assign planetary systems based on occurrence rates derived from the \kepler\ mission.
+\subsection{Synthetic star and planet sample}
+The first step in our analysis is to generate a sample of stars that host synthetic planets.
 
@@
 
-\subsection{Synthetic planet sample}\label{sec:syn_planets}
-Next, we assigned planetary systems to the synthetic stellar sample.
-The orbital parameters and bulk properties of these simulated planets are informed by statistics from the \kepler\ mission.
-
-\subsubsection{Planet occurrence rates in orbital period and radius}
-We adopted the planetary occurrence rates recently derived in \citep{Bergsten2022}.
+\subsubsection{Synthetic planet sample}\label{sec:syn_planets}
+Next, we assigned to the stellar sample planetary systems with frequencies, orbital parameters, and bulk properties derived from the \kepler\ mission.
+We adopted the planetary occurrence rates in orbital period and radius recently derived in~\citep{Bergsten2022}.
 Following \citep{Youdin2011a}, their inferred occurrence rate density can be expressed in the form
@@
 \end{equation}
-where $F_0$ ...
-\todo[inline]{GALEN: Describe underlying planet occurrence rates of synthetic population, its Mstar dependency, any caveats or assumptions.}
+where $F_0$
+...
+\todo[inline]{GALEN: Briefly describe underlying planet occurrence rates of synthetic population, its Mstar dependency, any caveats or assumptions.}
 
@@
 
-\subsection{Planet radius changes due to runaway greenhouse climates}\label{sec:mo_model}
+\subsubsection{Planet radius changes due to runaway greenhouse climates}\label{sec:mo_model}
 \todo[inline]{"Methods Section" for the geophysical models. Describe specific parametrization, dependency on bulk planet/orbit params\\
@@
 
-\subsection{Detection bias and sample selection}\label{sec:sensitivity}
+\subsubsection{Detection bias and sample selection}\label{sec:sensitivity}
 Not all transiting planets are detectable with the same likelihood and detection biases have an impact on the demographic measurements we are interested in.
@@
 
-\subsection{Definition of the runaway greenhouse hypothesis}
+\subsection{Hypothesis tests}
 We now turn to assessing the simulated surveys in terms of their ability to detect the runaway greenhouse transition and their constraining power on parameters of this trend.
@@
 For our specific problem, this implies comparing evidences for a demographic imprint of the runaway greenhouse effect to its absence.
-
+\subsubsection{Definition of the runaway greenhouse hypothesis}
 \begin{note}
@@
 \end{note}
-
     As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
@@
 
-\subsection{Hypothesis tests}
+\subsubsection{Bayesian model comparison}
 We are now ready to assess the relative plausibility of $h_{\mathrm{MO}}$ and $h_{\mathrm{null}}$ given the synthetic data we have generated, assigning equal a~priori~ probabilities to these models.

### 2022-12-22 src/tex/ms.tex
streamline Methods section

@@
 \section{Methods}\label{sec:methods}
-\todo[inline]{Introduce the idea of constraining magma ocean frequencies and/or physics with Bayesian stats on exoplanet population; introduce Bioverse; report methods of the hypothesis tests.\\
-Caveats: see \citet{Turbet2019} Sect. 3, \citet{Bower2019}}
-The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
-This section describes our approach.
+\todo[inline]{Caveats: see \citet{Turbet2019} Sect. 3, \citet{Bower2019}}
+%The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
 
@@
 \todo[inline]{KEVIN: describe stellar population}
+...
 
@@
 \todo[inline]{JUSTIFY this choice (TIM?)}
-This represents the baseline bulk density before any magma ocean-related effects are applied.
+This represents the baseline bulk density before any climate-related effects are applied.
 
@@
 \begin{note}
-    Not all planets are transiting from our point of view.
-    We model the occurrence of transits by assuming random, isotropic orientations of planetary orbits and calculating the transit impact parameter $b = \cos(i)/R_\star$ for each planet.
-    Only planets with $|b| < 1$ are potentially observable in transit and further considered.
+    We model the occurrence of transits by assuming isotropic orientations of planetary orbits, calculating the impact parameters $b = \cos(i)/R_\star$ and considering only observable planets with $|b| < 1$.
     For these cases we calculate the transit depth
     \begin{equation}\label{eq:transitdepth}
-        \delta = \left( \frac{R_\mathrm{P}}{R_\star} \right)^2
-    \end{equation}\todo[inline]{check if transit depth or duration are actually used further down}
-    and transit duration
-    \begin{equation}\label{eq:transitduration}
-        t_{\mathrm{T}} = \frac{R_\star P}{\pi a} \sqrt{1 - b^2},
+        \delta = \left( \frac{R_\mathrm{P}}{R_\star} \right)^2,
     \end{equation}
-    which are relevant for the detection probability of the respective planet~(see Sect.~\ref{sec:sensitivity}).
+    which is relevant for the detection probability of the respective planet~(see Sect.~\ref{sec:sensitivity}).
     Excluding all non-transiting planets diminishes the sample to \SI{\percentageTransiting}{\percent} of its original size.
@@
     The absolute value of $S_\oplus$, by which all synthetic planets are scaled, is then \SI{238}{\watt\per\square\meter}.
-    We further define the solar-equivalent semi-major axis $a_{eff} = a (L_\star/L_\odot)^{-1/2}$, at which a planet experiences the same instellation as a Solar System planet at an orbital distance~$a$.
+%    We further define the solar-equivalent semi-major axis $a_{eff} = a (L_\star/L_\odot)^{-1/2}$, at which a planet experiences the same instellation as a Solar System planet at an orbital distance~$a$.
 \end{note}
@@
 The survey module of Bioverse converts the synthetic planet sample into a set of uncertainty-laden measurements of some parameters of a subset of that sample.
-   This task includes selection of the targets, application of any detection biases, and conducting simulated measurements.
+   This task includes selection of the targets, application of detection biases, and conducting simulated measurements.
    All of the above are specific to the particular survey.
-
-Current and near-future missions focusing on Earth-sized planets target several host star spectral types, and their exact sample selection is based on specific mission objectives.
-    We thus chose a broad selection function centered on general detectability of transits with state-of-the-art technology.
 
@@
 
-\subsubsection{Detection bias and sample selection}\label{sec:sensitivity}
+\subsubsection{Detection bias and sample selection}\label{sec:sensitivity} % Turbet2019 lists precision for CHEOPS, PLATO?, TESS?
 Not all transiting planets are detectable with the same likelihood and detection biases have an impact on the demographic measurements we are interested in.
@@
 To represent these photometric limits, we chose a minimum transit depth of \SI{75}{\ppm} as a detection limit and consider only measurements of planets exceeding this threshold.
+We further exclude target stars with \gaia\ magnitudes $M_\mathrm{G} > \minMagnitude$.
 
@@
     \todo[inline]{use criterion of Zahnle \& Caitling 2017 instead? $0.8 S^{0.25} < R$}
-    Another selection concerns the climate regime of interest.
+%    Current and near-future missions focusing on Earth-sized planets target several host star spectral types, and their exact sample selection is based on specific mission objectives.
+%    We thus chose a broad selection function centered on general detectability of transits with state-of-the-art technology.
+
     The runaway greenhouse effect becomes obsolete where no atmosphere can be maintained due to proximity to the host star and resulting atmospheric erosion.
@@
 For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \Nplanets$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
-We assumed a water mass fractions of \wrr\ for each planet.
+We assumed a water mass fraction of \wrr\ for each planet.
 In this case, the pattern was detected with high significance ($\ln \mathcal{Z} \approx 100$).
@@
 Only a subset of these planets would be detectable by a transit survey, and their properties can be probed only with a finite precision.
-To test the detectability of a runaway greenhouse transition, we simulated several transit surveys with different designs and strategies and explored their capability to recover the magma ocean trend and constrain its parameters.
+To test the detectability of a runaway greenhouse transition, we simulated several transit surveys with different designs and strategies and explored their capability to recover the trend and constrain its parameters.
 We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
@@
         \caption{
-        Retrieved posterior distribution of magma ocean model.
+        Retrieved posterior distribution of the magma ocean model.
 The density maps in each panel show relationships between and marginalized distributions of the model parameters as they would be retrieved with a targeted $<nominal telescope size>$ survey of $<survey duration>$ duration.

### 2022-12-22 src/tex/ms.tex
"magma ocean hypothesis" -> "runaway greenhouse hypothesis"
...and relatives

@@
 
-The main task is to determine -- for different survey configurations -- the confidence level with which a magma ocean signal can be detected in simulated exoplanet observations.
+We determined -- for different survey configurations -- the confidence level with which the runaway greenhouse threshold can be detected in simulated exoplanet observations.
 For this purpose we expanded the Bioverse framework~\citep{Bixel2020,Bixel2021}\footnote{Bioverse is actively maintained and documented open source software written in Python.
-Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic stellar and planetary samples into which we injected magma ocean signals, simulated observations of this sample, and computed Bayesian evidences in favor of the magma ocean hypothesis (see Fig.~\ref{fig:flowchart}).
+Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic stellar and planetary samples into which we injected the runaway greenhouse signal, simulated observations of this sample, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).
 \begin{figure*}
@@
     To enable a quantitative assessment of this detectability, we injected the signal into the simulated planet sample before observing it with simulated surveys.
-    While we search for the magma ocean pattern in demographic quantities such as average planet radii, the injected changes happen on the planetary level: We changed each planet's radius based on its individual set of properties and the associated predictions from steam atmosphere and water incorporation models.
+    While we search for the runaway greenhouse pattern in demographic quantities such as average planet radii, the injected changes happen on the planetary level: We changed each planet's radius based on its individual set of properties and the associated predictions from steam atmosphere and water incorporation models.
     Relevant properties are a planet's mass $M$, its instellation $S$, and its bulk water inventory expressed as a water mass fraction $x_{H_2O}$.
@@
     Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
-    The magma ocean hypothesis is then defined as
-\begin{equation}\label{eq:mo_hypo}
-    h_{\mathrm{MO}}(\theta, S) =
-%        \begin{cases}
-%        1 & \text{if } x \in \mathbb{Q}\\
-%        0 & \text{if } x \in \mathbb{R}\setminus\mathbb{Q}
-%    \end{cases}
+    The runaway greenhouse hypothesis is then defined as
+\begin{equation}\label{eq:rgh_hypo}
+    h_{\mathrm{rgh}}(\theta, S) =
         \begin{cases}
@@
 The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue).
-This magma ocean hypothesis can be tested against the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
+This runaway greenhouse hypothesis can be tested against the null hypothesis $h_\mathrm{null}$ (green), where average radii are independent of instellation.}
         \label{fig:HnullHmo}
@@
     The only free parameter of the null hypothesis, which assumes the average radius to be independent of instellation, is the predicted mean radius $\langle R_\mathrm{P}\rangle $.
-    The functional form of the magma ocean hypothesis is more complex: Besides the mean radius of planets outside the runaway greenhouse threshold $\langle R_\mathrm{P}\rangle_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$.
+    The functional form of the runaway greenhouse hypothesis is more complex: Besides the mean radius of planets outside the threshold $\langle R_\mathrm{P}\rangle_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$.
     Further, the parameter encoding the amount of radius change is the planetary water mass fraction~$x_{H_2O}$.
@@
 The measured radii $R_\mathrm{P, i}$ cannot be directly used for the hypothesis tests as they include intrinsic scatter that is not caused by measurement errors.
-$h_{\mathrm{MO}}$ and $h_{\mathrm{null}}$ should thus be tested against a statistical estimator that represents the population mean.
+$h_{\mathrm{rgh}}$ and $h_{\mathrm{null}}$ should thus be tested against a statistical estimator that represents the population mean.
 To avoid binning and the artificial patterns it may introduce, we chose to test our hypotheses against a simple moving average $SMA$ along the instellation axis with a window of size \windowsize\ centered around each measurement.
@@
 \end{equation}
-Here, $h\left(\boldsymbol{\theta}, S_i\right)$ corresponds to the functional form of the magma ocean hypothesis~(Eqn.~\ref{eq:mo_hypo}).
+Here, $h\left(\boldsymbol{\theta}, S_i\right)$ corresponds to the functional form of the runaway greenhouse hypothesis~(Eqn.~\ref{eq:rgh_hypo}).
 We tested this hypothesis against the null hypothesis $h_\mathrm{null} (\theta, S) = \theta$, which states that there is no relationship between the measured planet radii and received instellation.
@@
 \subsubsection{Bayesian model comparison}
-We are now ready to assess the relative plausibility of $h_{\mathrm{MO}}$ and $h_{\mathrm{null}}$ given the synthetic data we have generated, assigning equal a~priori~ probabilities to these models.
+We are now ready to assess the relative plausibility of $h_{\mathrm{rgh}}$ and $h_{\mathrm{null}}$ given the synthetic data we have generated, assigning equal a~priori~ probabilities to these models.
 This is done by comparing the Bayesian evidence $\mathcal{Z}$ of the models, which we estimate with the nested sampling~\citep{Skilling2004} algorithm \emph{dynesty}~\citep{Speagle2020}.
@@
 \section{Results}
-\subsection{Signature of the magma ocean hypothesis}
+\subsection{Signature of the runaway greenhouse hypothesis}
 \todo[inline]{show general detectability}
@@
         Detection of the runaway greenhouse threshold in the optimistic case.
-        From simulated radius and instellation measurements of a large ($N = \Nplanets$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:mo_hypo}, random draws from the posterior in red).
+        From simulated radius and instellation measurements of a large ($N = \Nplanets$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
             The pattern is detected with high significance.
@@
 In the FGK sample of the mission, a minimum of 500 Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
-We base our assessment of PLATO's diagnostic power to test the magma ocean hypothesis on these estimates.
+We base our assessment of PLATO's diagnostic power to test the runaway greenhouse hypothesis on these estimates.
 
@@
         \caption{
-        Statistical power of a magma ocean hypothesis test as a function of model parameters.
+        Statistical power of the runaway greenhouse hypothesis test as a function of model parameters.
         }
@@
 \subsection{Statistical power of different mission designs}\label{sec:statpower_missions}
-\todo[inline]{show detectability of the magma ocean signature for different mission designs}
+\todo[inline]{show detectability of the runaway greenhouse signature for different mission designs}
 
@@
     \item The demographic imprint of the runaway greenhouse threshold should be detectable with transit measurements of a large enough sample and sufficient precision.
-    \item We find that the planned PLATO transit survey will provide a sufficient sample to test the magma ocean hypothesis.
+    \item We find that the planned PLATO transit survey will provide a sufficient sample to test the runaway greenhouse hypothesis.
     \item The diagnostic power of transit missions in testing this hypothesis can be maximized by ...

### 2022-12-23 src/tex/ms.tex
add high-level conclusions and corresponding result/discussion structure

@@
 \section{Introduction}
+...just dropping material and references for now...
 \todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
@@
 \begin{note}
-(Several patterns in the planetary parameter space have been reported in demographic studies or predicted from planet formation theories.)
+For observations of rocky exoplanets, the currently best-probed regime is that of warm, close-in planets.
+These bodies experience climatic conditions that are similar to the environment of the inner Solar System bodies at early stages of their formation.
+Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.
+
+...
     Venus and Earth, while having accreted from the same mass reservoir and despite their similar bulk properties, evolved into planets with very different environmental conditions on their surfaces.
@@
 make connection between interior and atmospheres.
+    ...
+    Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.
 \end{note}
@@
 \section{Results}
-\subsection{Signature of the runaway greenhouse hypothesis}
-\todo[inline]{show general detectability}
+\subsection{Statistical signature of the runaway greenhouse threshold}
+
+\subsection{Testability of the runaway greenhouse hypothesis}
+\todo[inline]{show general detectability. Fiducial (optimistic) scenario.}
 
@@
 
+% Additional \subsection for habitability? To reflect structure in Discussion, Conclusions?
+
 \section{Discussion}\label{sec:discussion}
 \todo[inline]{\textbf{Potential discussion points: } what is the fraction of Kepler planets w/ runaway GH.}
+
+\subsection{Statistical imprint of exoplanet climates}
+\todo[inline]{the predicted climate tipping point causes a demographic imprint.}
+
 \subsection{Detectability of the runaway greenhouse threshold}
+\todo[inline]{the demographic imprint is detectable. Under what conditions?}
 
@@
 \section{Conclusions}
-
 \begin{note}
-    \textit{Here or in the introduction:} For observations of rocky exoplanets, the currently best-probed regime is that of warm, close-in planets.
-    These bodies experience climatic conditions that are similar to the environment of the inner Solar System bodies at early stages of their formation.
-    Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.
-    Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.
-    In this study, we assessed the feasibility of a detection of this pattern with current and near-future exoplanet missions.
+%    Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.
+    We have explored the potential of contemporary exoplanet missions to statistically detect a radius discontinuity that is predicted to result from runaway greenhouse phases.
+    ...Bioverse...
 \end{note}
@@
 \begin{enumerate}
-    \item The demographic imprint of the runaway greenhouse threshold should be detectable with transit measurements of a large enough sample and sufficient precision.
+    \item The predicted tipping point in planetary climates causes a discontinuity in the radius and density distribution of small (DEFINITION) exoplanets.
+    \item Its demographic imprint of the runaway greenhouse threshold should be detectable with transit measurements of a large enough sample and sufficient precision.
     \item We find that the planned PLATO transit survey will provide a sufficient sample to test the runaway greenhouse hypothesis.

### 2023-01-03 src/tex/ms.tex
use consistent symbols for hypotheses

@@
     \begin{equation}
-        h_{\mathrm{null}}(\theta, S) = \theta,
+        H_0(\theta, S) = \theta,
     \end{equation}
@@
 \begin{equation}\label{eq:rgh_hypo}
-    h_{\mathrm{rgh}}(\theta, S) =
+    H_{\mathrm{rgh}}(\theta, S) =
         \begin{cases}
-            h_{\mathrm{null}} & \text{for } S \leq S_\mathrm{thresh}\\
+            H_0 & \text{for } S \leq S_\mathrm{thresh}\\
             \langle R_\mathrm{P}\rangle (f_\mathrm{steam}, f_\mathrm{water}) & \text{for } S > S_\mathrm{thresh}.
@@
 The measured radii $R_\mathrm{P, i}$ cannot be directly used for the hypothesis tests as they include intrinsic scatter that is not caused by measurement errors.
-$h_{\mathrm{rgh}}$ and $h_{\mathrm{null}}$ should thus be tested against a statistical estimator that represents the population mean.
+$H_{\mathrm{rgh}}$ and $H_0$ should thus be tested against a statistical estimator that represents the population mean.
 To avoid binning and the artificial patterns it may introduce, we chose to test our hypotheses against a simple moving average $SMA$ along the instellation axis with a window of size \windowsize\ centered around each measurement.
@@
 \subsubsection{Bayesian model comparison}
-We are now ready to assess the relative plausibility of $h_{\mathrm{rgh}}$ and $h_{\mathrm{null}}$ given the synthetic data we have generated, assigning equal a~priori~ probabilities to these models.
+We are now ready to assess the relative plausibility of $H_{\mathrm{rgh}}$ and $H_0$ given the synthetic data we have generated, assigning equal a~priori~ probabilities to these models.
 This is done by comparing the Bayesian evidence $\mathcal{Z}$ of the models, which we estimate with the nested sampling~\citep{Skilling2004} algorithm \emph{dynesty}~\citep{Speagle2020}.

### 2023-01-03 src/tex/ms.tex
move HnullHmo figure to results, describe expected imprint

@@
     They are assumed to act additively on the planet radii and thus on their average $\langle R_\mathrm{P}\rangle $.
-\begin{figure}
-    \begin{centering}
-        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
-        \caption{Planets above and below the runaway greenhouse threshold.
-        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
-        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of instellation.
-Only the planets marked as transiting are observable.
-Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, radii are inflated by the amount indicated with gray lines.
-The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue).
-This runaway greenhouse hypothesis can be tested against the null hypothesis $h_\mathrm{null}$ (green), where average radii are independent of instellation.}
-        \label{fig:HnullHmo}
-    \end{centering}
-\end{figure}
-    Figure~\ref{fig:HnullHmo} shows the null and alternative hypotheses as a function of instellation.
-    The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
 
@@
 \subsection{Statistical signature of the runaway greenhouse threshold}
+To characterize the population-level imprint of individual radius changes, we generated a generic planet population with an injected runaway greenhouse effect assuming a water fraction $x_\mathrm{H_2O} = \wrr$.
+\begin{figure}
+    \begin{centering}
+        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
+        \caption{Planets above and below the runaway greenhouse threshold.
+        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
+        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of instellation.
+Only the planets marked as transiting are observable.
+Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, radii are inflated by the amount indicated with gray lines.
+The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue).
+This runaway greenhouse hypothesis can be tested against the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
+        \label{fig:HnullHmo}
+    \end{centering}
+\end{figure}
+Figure~\ref{fig:HnullHmo} shows the resulting planetary radii.
+When ordered in orbital period space, the different planet types overlap, diluting the demographic imprint.
+With net instellation as an independent variable, planets above and below the runaway greenhouse threshold separate.
+We also show the predictions of observable average planet radii from the statistical hypotheses defined above.
+%The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
+Within the runaway greenhouse limit, an average radius change of \avgRadiusChange\ occurs.
 

### 2023-01-04 src/tex/ms.tex
update flowchart

@@
         \includegraphics[width=\hsize]{figures/flowchart.pdf}
-        \caption{Workflow of our hypothesis testing with Bioverse.
-In the first block, a stellar sample is generated based on the Gaia nearby stars catalog~\citep{Smart2021}\todo[]{KEVIN pls verify}.
-The stars are then populated with planets based on Kepler demographics, and they may be assigned a magma ocean based on the model described in Sect.~\ref{sec:mo_model}.
-The planets' respective transit probabilities are computed.
-The second block simulates the exoplanet survey whereby selection effects and detection biases are introduced.
-Finally, the third block deals with testing a hypothesis based on the data from the simulated survey.
+        \caption{Workflow of our hypothesis testing with Bioverse. In the first block, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
+            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:mo_model}.
+        The second block simulates an exoplanet survey, whereby selection effects and detection biases are introduced.
+Finally, the third block deals with testing the runaway greenhouse hypothesis based on data from the survey simulation.
         By iterating through these steps, we compute the statistical power of testing the hypothesis given the assumed survey design.}

### 2023-01-04 src/tex/ms.tex
minor edits

@@
 
+\todo[inline]{ARNAUD: revise definition of $S$ following our discussion on 2022-12-22. Introduce net instellation and clarify what is meant by it.}
 \begin{note}
@@
 \begin{note}
-The survey module of Bioverse converts the synthetic planet sample into a set of uncertainty-laden measurements of some parameters of a subset of that sample.
+The survey module of Bioverse converts the synthetic planet sample into a set of uncertainty-laden measurements on a subset of that sample.
    This task includes selection of the targets, application of detection biases, and conducting simulated measurements.
@@
 Here, $h\left(\boldsymbol{\theta}, S_i\right)$ corresponds to the functional form of the runaway greenhouse hypothesis~(Eqn.~\ref{eq:rgh_hypo}).
-We tested this hypothesis against the null hypothesis $h_\mathrm{null} (\theta, S) = \theta$, which states that there is no relationship between the measured planet radii and received instellation.
+We tested this hypothesis against the null hypothesis $H_0 (\theta, S) = \theta$, which states that there is no relationship between the measured planet radii and received instellation.
 \end{note}
@@
 \begin{equation}
-\Delta \ln \mathcal{Z}  = \ln \mathcal{Z} - \ln \mathcal{Z}_\mathrm{null}  > 3.
+\Delta \ln \mathcal{Z}  = \ln \mathcal{Z}_\mathrm{rgh} - \ln \mathcal{Z}_0  > 3.
 \end{equation}
@@
 \section{Discussion}\label{sec:discussion}
-\todo[inline]{\textbf{Potential discussion points: } what is the fraction of Kepler planets w/ runaway GH.}
+\todo[inline]{\textbf{Potential discussion points: } what is the fraction of \kepler\ planets w/ runaway GH.}
 
@@
 \subsection{Detectability of the runaway greenhouse threshold}
-\todo[inline]{the demographic imprint is detectable. Under what conditions?}
+\todo[inline]{the demographic imprint is detectable. Under what conditions?\\
+Mention multiplanet systems.}
 
@@
     \item The predicted tipping point in planetary climates causes a discontinuity in the radius and density distribution of small (DEFINITION) exoplanets.
-    \item Its demographic imprint of the runaway greenhouse threshold should be detectable with transit measurements of a large enough sample and sufficient precision.
+    \item The demographic imprint of the runaway greenhouse threshold should be detectable with transit measurements of a large enough sample and sufficient precision.
     \item We find that the planned PLATO transit survey will provide a sufficient sample to test the runaway greenhouse hypothesis.

### 2023-01-04 src/tex/ms.tex
introduce 'background' population without runaway GH
new free parameter f_rgh = fraction of planets with S > S_thresh that are in a runaway climate.

@@
 In general, steam atmospheres (case~B) cause a significant radius increase, which is slightly reduced when water incorporation in the melt is allowed (case~C).
-Our nominal simulation setup is such that all planets receiving an instellation $S < S_\mathrm{thresh}$ follow relation~A, and all planets with $S > S_\mathrm{thresh}$ follow relation~C.
-This introduces a sharp discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
+
+Our simulation setup is such that all planets receiving an instellation $S < S_\mathrm{thresh}$ follow relation~A, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow relation~C.
+This introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
 In the following, we test if and under what conditions this trend is large enough to be detected with high significance.
@@
         \begin{cases}
-            H_0 & \text{for } S \leq S_\mathrm{thresh}\\
-            \langle R_\mathrm{P}\rangle (f_\mathrm{steam}, f_\mathrm{water}) & \text{for } S > S_\mathrm{thresh}.
+            H_0, &  S \leq S_\mathrm{thresh}\\
+            \langle R_\mathrm{P}\rangle (f_\mathrm{rgh},\Delta R_\mathrm{stm}, \Delta R_\mathrm{wtr}), &  S > S_\mathrm{thresh}.
         \end{cases}
 \end{equation}
-    Here, $f_\mathrm{steam}$ and $f_\mathrm{water}$ are predictions from the steam atmosphere and water incorporation models, respectively.
+    Here, $f_\mathrm{rgh}$ is the fraction of planets experiencing a runaway greenhouse effect.
+    $\Delta R_\mathrm{stm}$ and $\Delta R_\mathrm{wtr}$ are predicted radius changes from the steam atmosphere and water incorporation models, respectively.
     They are assumed to act additively on the planet radii and thus on their average $\langle R_\mathrm{P}\rangle $.
@@
     The only free parameter of the null hypothesis, which assumes the average radius to be independent of instellation, is the predicted mean radius $\langle R_\mathrm{P}\rangle $.
-    The functional form of the runaway greenhouse hypothesis is more complex: Besides the mean radius of planets outside the threshold $\langle R_\mathrm{P}\rangle_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$.
-    Further, the parameter encoding the amount of radius change is the planetary water mass fraction~$x_{H_2O}$.
+    The functional form of the runaway greenhouse hypothesis is more complex: Besides the mean radius of planets outside the threshold $\langle R_\mathrm{P}\rangle_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$, the planetary water mass fraction~$x_{H_2O}$ encoding the amount of radius change, and the fraction of planets above the threshold instellation that are in a runaway greenhouse climate $f_\mathrm{rgh}$.
 \end{note}
@@
 Only the planets marked as transiting are observable.
-Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, radii are inflated by the amount indicated with gray lines.
+Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, the radii of a fraction of planets are inflated by the amount indicated with gray lines.
 The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue).
@@
 For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \Nplanets$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
-We assumed a water mass fraction of \wrr\ for each planet.
-In this case, the pattern was detected with high significance ($\ln \mathcal{Z} \approx 100$).
+The fraction of planets with runaway greenhouse climates was \frgh, and we assumed a water mass fraction of \wrr\ for each planet.
+In this case, the pattern was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).
 
@@
 \subsection{Statistical imprint of exoplanet climates}
-\todo[inline]{the predicted climate tipping point causes a demographic imprint.}
+\todo[inline]{the predicted climate tipping point causes a demographic imprint.\\
+Discuss our ignorance on $f_\mathrm{rgh}$ and its impact on the imprint.}
 

### 2023-01-05 src/tex/ms.tex
move main drivers to discussion section
add retrieval for optimistic scenario

@@
 For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \Nplanets$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
-The fraction of planets with runaway greenhouse climates was \frgh, and we assumed a water mass fraction of \wrr\ for each planet.
+We assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \wrr\ for each planet.
 In this case, the pattern was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).
+
+\begin{figure}[ht!]
+%    \script{figurescript.py}
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/corner.pdf}
+        \caption{
+            Retrieved posterior distribution of key parameters in the optimistic scenario.
+The density maps in each panel show relationships between and marginalized distributions of the model parameters as they could be retrieved with a high-precision transit survey and a sample of \Nplanets\ planets.
+True values of the parameters for the injected effect are shown in orange.
+The threshold instellation can be well constrained; the predominant water fraction and the fraction of planets with runaway greenhouse climates are degenerate.
+        }
+        \label{fig:cornerplot}
+    \end{centering}
+\end{figure}
+With such a strong signal, we can attempt an inference of the parameters defining the injected effect.
+Figure~\ref{fig:cornerplot} shows the posterior distributions of $S_\mathrm{thresh}$, $x_{H_2O}$, and $f_\mathrm{rgh}$ as determined by the nested sampler.
+The threshold instellation can be accurately constrained, $S_\mathrm{thresh}$ = \retrSthresh.
+Both a higher water mass fraction and a higher fraction of planets in a runaway greenhouse state lead to larger average radii, thus these parameters are strongly correlated and cannot be well constrained without independent measurements.
 
@@
 
-%We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ...
-\todo[]{double-check which survey params are needed}
-We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a given transit survey:
-\begin{itemize}
-    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction.
-As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern.
-    \item Sample size: Since we are looking for a population-level trend, the statistical significance will increase with a larger sample size.
-    \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be.
-Good \textit{accuracy} is less important, as long as it does not have a systematic error scaling with stellar irradiance.
-    \item Availability and precision of mass measurements: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space.
-If transiting planets can be followed up to obtain mass measurements, the statistical significance increases.
-\end{itemize}
-Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.
-This can be different for young host stars when their ages can not be well constrained; in particular the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).
-
 PLATO (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
@@
 We base our assessment of PLATO's diagnostic power to test the runaway greenhouse hypothesis on these estimates.
-
-\begin{figure}[ht!]
-%    \script{figurescript.py}
-    \begin{centering}
-        \includegraphics[width=\linewidth]{figures/example_cornerplot.pdf}
-        \caption{
-        Retrieved posterior distribution of the magma ocean model.
-The density maps in each panel show relationships between and marginalized distributions of the model parameters as they would be retrieved with a targeted $<nominal telescope size>$ survey of $<survey duration>$ duration.
-True values of the injected magma ocean parameters are shown in orange.
-...
-        }
-        \label{fig:cornerplot}
-    \end{centering}
-\end{figure}
 
@@
 Mention multiplanet systems.}
+
+%We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ...
+\todo[]{double-check which survey params are needed}
+We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a given transit survey:
+\begin{itemize}
+    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction.
+As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern.
+    \item Sample size: Since we are looking for a population-level trend, the statistical significance will increase with a larger sample size.
+    \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be.
+Good \textit{accuracy} is less important, as long as it does not have a systematic error scaling with stellar irradiance.
+    \item Availability and precision of mass measurements: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space.
+If transiting planets can be followed up to obtain mass measurements, the statistical significance increases.
+\end{itemize}
+Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.
+This can be different for young host stars when their ages can not be well constrained; in particular the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).
 

### 2023-01-06 src/tex/ms.tex
separate mission-specific and optimistic case

@@
 
-\begin{note}
-   The fidelity of a future detection or falsification of a magma ocean signature will depend on the significance with which the null hypothesis can be excluded.
-   As a consequence, instrumentation and survey strategy aiming at testing the hypothesis should aim at maximizing the probability of a true positive detection given the existence of the effect.
-   This is called the \textit{statistical power}: the probability of correctly rejecting a false null hypothesis.
-   Before turning to mission design trades that influence the statistical power in Sect.~\ref{sec:mission-design-trades}, we explore its behavior as a function of various free parameters of the magma ocean model.
-    ...
-\begin{figure*}[ht!]
-%    \script{figurescript.py}
-    \begin{centering}
-        \includegraphics[width=\linewidth]{figures/example_statistical_power.pdf}
-        \caption{
-        Statistical power of the runaway greenhouse hypothesis test as a function of model parameters.
-        }
-        \label{fig:statistical_power}
-    \end{centering}
-\end{figure*}
-
-\end{note}
-
 \subsection{Statistical power of different mission designs}\label{sec:statpower_missions}
@@
 \subsection{Diagnostic power of upcoming exoplanet missions}\todo[inline]{PLATO: needed sample size, precision}
+The fidelity of a future detection or falsification of a magma ocean signature will depend on the significance with which the null hypothesis can be excluded.
+As a consequence, instrumentation and survey strategy aiming at testing the hypothesis should aim at maximizing the probability of a true positive detection given the existence of the effect.
+...
 
 \subsection{Mission design trades}\label{sec:mission-design-trades}
+We now turn to mission design trades that influence the statistical power for testing the runaway greenhouse hypothesis.
+...
 

### 2023-01-06 src/tex/ms.tex
new figure: stat. power grid in x_H2O, f_rgh

@@
         \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
-        \caption{Planets above and below the runaway greenhouse threshold.
+        \caption{Synthetic planets above and below the runaway greenhouse threshold.
         \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
@@
 Both a higher water mass fraction and a higher fraction of planets in a runaway greenhouse state lead to larger average radii, thus these parameters are strongly correlated and cannot be well constrained without independent measurements.
+
+\begin{note}
+    Given an optimistic survey design, what influences the probability of correctly rejecting a false null hypothesis?
+    \begin{figure}[ht!]
+%    \script{figurescript.py}
+        \begin{centering}
+            \includegraphics[width=\linewidth]{figures/optimistic_statpwr_H2O-f.pdf}
+            \caption{
+                Statistical power of the runaway greenhouse hypothesis test as a function of model parameters.
+                For a sample size $N = \Nplanets$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and fraction of greenhouse climate planets.
+                Higher values in either parameter result in a more reliable detection.
+            }
+            \label{fig:statpwr_H2O-f}
+        \end{centering}
+    \end{figure}
+    Figure~\ref{fig:statpwr_H2O-f} explores the statistical power achieved in the above scenario for different combinations of the poorly constrained parameters $x_{H_2O}$ and $f_\mathrm{rgh}$.
+    The higher the water content and thus the radius change, and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
+    As long as this fraction is larger than $\sim 0.1$, a sample size of \Nplanets\ is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $x_{H_2O} = 0.001$.
+\end{note}
 

### 2023-01-07 src/tex/ms.tex
minor edits and fixes

@@
 \todo[inline]{JUSTIFY this choice (TIM?)}
+\todo[inline]{run a test: do we expect that the choice of M-R relation impacts the results? Perhaps have an appendix section.}
 This represents the baseline bulk density before any climate-related effects are applied.
@@
 Our simulation setup is such that all planets receiving an instellation $S < S_\mathrm{thresh}$ follow relation~A, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow relation~C.
-This introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
+The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
+This parametrization introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
 In the following, we test if and under what conditions this trend is large enough to be detected with high significance.
@@
 
-\begin{note}
-    \todo[inline]{use criterion of Zahnle \& Caitling 2017 instead? $0.8 S^{0.25} < R$}
 %    Current and near-future missions focusing on Earth-sized planets target several host star spectral types, and their exact sample selection is based on specific mission objectives.
 %    We thus chose a broad selection function centered on general detectability of transits with state-of-the-art technology.
-
 The runaway greenhouse effect becomes obsolete where no atmosphere can be maintained due to proximity to the host star and resulting atmospheric erosion.
 We thus excluded planets with extremely high instellation that does not allow long-term preservation of an atmosphere and clear our sample from all planets with an instellation $S > \SI{2000}{\watt\per\square\meter}$.
+\todo[inline]{use criterion of Zahnle \& Caitling 2017 instead? $0.8 S^{0.25} < R$}
 We further consider only rocky planets with masses below \SI{2}{\Mearth}. % (Turbet+2020 covers only 0.1 to 2 Mearth)
-\end{note}
 
@@
 
-\begin{note}
 The only free parameter of the null hypothesis, which assumes the average radius to be independent of instellation, is the predicted mean radius $\langle R_\mathrm{P}\rangle $.
 The functional form of the runaway greenhouse hypothesis is more complex: Besides the mean radius of planets outside the threshold $\langle R_\mathrm{P}\rangle_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$, the planetary water mass fraction~$x_{H_2O}$ encoding the amount of radius change, and the fraction of planets above the threshold instellation that are in a runaway greenhouse climate $f_\mathrm{rgh}$.
-\end{note}
 
-\begin{note}
 The measured radii $R_\mathrm{P, i}$ cannot be directly used for the hypothesis tests as they include intrinsic scatter that is not caused by measurement errors.
@@
 \begin{equation}
-    \mathcal{L}(SMA \mid \boldsymbol{\theta})=\prod_{i}^{N} \frac{1}{\sqrt{2 \pi \sigma_{SMA_i}^{2}}} \exp \left(-\frac{\left(SMA_i - h\left(\boldsymbol{\theta}, S_i\right)\right)^{2}}{2 \sigma_{SMA_i}^{2}}\right).
+    \mathcal{L}(SMA \mid \boldsymbol{\theta})=\prod_{i}^{N} \frac{1}{\sqrt{2 \pi \sigma_{SMA_i}^{2}}} \exp \left(-\frac{\left(SMA_i - H\left(\boldsymbol{\theta}, S_i\right)\right)^{2}}{2 \sigma_{SMA_i}^{2}}\right).
 \end{equation}
-Here, $h\left(\boldsymbol{\theta}, S_i\right)$ corresponds to the functional form of the runaway greenhouse hypothesis~(Eqn.~\ref{eq:rgh_hypo}).
-We tested this hypothesis against the null hypothesis $H_0 (\theta, S) = \theta$, which states that there is no relationship between the measured planet radii and received instellation.
-\end{note}
+Here, $H\left(\boldsymbol{\theta}, S_i\right)$ corresponds to the functional form of the runaway greenhouse or null hypothesis.
 
@@
 For $S_\mathrm{thresh}$, we chose a log-uniform prior in $[\prSmin,\prSmax]\, \SI{}{\watt\per\meter\squared}$.
-Similarly, we sample $x_{H_2O}$ from a log-uniform distribution.
+Similarly, we sample $x_{H_2O}$ from a log-uniform distribution to imply scale-invariant ignorance.
 Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Sect.~\ref{sec:mo_model}).
@@
 \subsection{Testability of the runaway greenhouse hypothesis}
-\todo[inline]{show general detectability. Fiducial (optimistic) scenario.}
-
 \begin{figure}[ht!]
@@
 =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
-In previous steps, we generated a population of synthetic planets that orbit synthetic stars, and we have injected the statistical trend expected from the presence of magma oceans on a fraction of the planets.
-Only a subset of these planets would be detectable by a transit survey, and their properties can be probed only with a finite precision.
+
+The planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
 To test the detectability of a runaway greenhouse transition, we simulated several transit surveys with different designs and strategies and explored their capability to recover the trend and constrain its parameters.
 We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
-
-PLATO (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
+\plato\~(PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
 The PLATO team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).

### 2023-01-10 src/tex/ms.tex
Define PLATO mission

@@
 
-%    We argue that in the near future, ESA's PLATO mission and NASA's Roman Space Telescope will be the most promising endeavors to constrain this demographic feature.
+%    We argue that in the near future, ESA's \plato\ mission and NASA's Roman Space Telescope will be the most promising endeavors to constrain this demographic feature.
 %    For each of these missions, we identify the key mission design drivers that enable a statistically sound detection.
@@
 \subsection{Survey simulations}
-\begin{note}
 The survey module of Bioverse converts the synthetic planet sample into a set of uncertainty-laden measurements on a subset of that sample.
@@
    All of the above are specific to the particular survey.
-
-\end{note}
 
@@
 \todo{Introduce survey simulation(s).
-PLATO, Cheops, Ariel, LIFE (reach out to team; Hendrik Linz produced a LIFE Technology Overview document)?, Nautilus\\ What are threshold missions that are able to detect this signal?
+, Cheops, Ariel, LIFE (reach out to team; Hendrik Linz produced a LIFE Technology Overview document)?, Nautilus\\ What are threshold missions that are able to detect this signal?
 =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
@@
 We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
-\plato\~(PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
-The PLATO team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).
+
+\subsubsection{PLATO}
+\plato\ (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
+The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).
 In the FGK sample of the mission, a minimum of 500 Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
-We base our assessment of PLATO's diagnostic power to test the runaway greenhouse hypothesis on these estimates.
+\plato's definition study report~\citep{plato2017} further states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).
+We base our assessment of \plato's diagnostic power on these estimates.
 
@@
 
-\subsection{Diagnostic power of upcoming exoplanet missions}\todo[inline]{PLATO: needed sample size, precision}
+\subsection{Diagnostic power of upcoming exoplanet missions}\todo[inline]{\plato: needed sample size, precision}
 The fidelity of a future detection or falsification of a magma ocean signature will depend on the significance with which the null hypothesis can be excluded.
@@
     \item The demographic imprint of the runaway greenhouse threshold should be detectable with transit measurements of a large enough sample and sufficient precision.
-    \item We find that the planned PLATO transit survey will provide a sufficient sample to test the runaway greenhouse hypothesis.
+    \item We find that the planned \plato\ transit survey will provide a sufficient sample to test the runaway greenhouse hypothesis.
     \item The diagnostic power of transit missions in testing this hypothesis can be maximized by ...

### 2023-01-14 src/scripts/utils.py
let python write variables to LaTeX

@@ -1,0 +1,26 @@
+# Code that we need that doesn't produce figures
+
+import paths
+
+def save_var_latex(key, value, datafile=paths.tex/'variables.dat'):
+    # https://stackoverflow.com/questions/59408823/pdf-latex-with-python-script-custom-python-variables-into-latex-output on 2023-01-13
+    import csv
+    import os
+
+    dict_var = {}
+
+    file_path = os.path.join(os.getcwd(), datafile)
+
+    try:
+        with open(file_path, newline="") as file:
+            reader = csv.reader(file)
+            for row in reader:
+                dict_var[row[0]] = row[1]
+    except FileNotFoundError:
+        pass
+
+    dict_var[key] = value
+
+    with open(file_path, "w") as f:
+        for key in dict_var.keys():
+            f.write(f"{key},{dict_var[key]}\n")

### 2023-01-14 src/tex/ms.tex
let python write variables to LaTeX
add figure with different PLATO surveys
Figure is going to need some further polishing.

@@
 Figure~\ref{fig:cornerplot} shows the posterior distributions of $S_\mathrm{thresh}$, $x_{H_2O}$, and $f_\mathrm{rgh}$ as determined by the nested sampler.
-The threshold instellation can be accurately constrained, $S_\mathrm{thresh}$ = \retrSthresh.
+The threshold instellation can be accurately constrained, $S_\mathrm{thresh} = \var{retrSthresh}$.
 Both a higher water mass fraction and a higher fraction of planets in a runaway greenhouse state lead to larger average radii, thus these parameters are strongly correlated and cannot be well constrained without independent measurements.
@@
 We base our assessment of \plato's diagnostic power on these estimates.
+
+    \begin{figure}[ht!]
+%    \script{figurescript.py}
+        \begin{centering}
+            \includegraphics[width=\linewidth]{figures/plato_fgrid.pdf}
+            \caption{
+            Expected evidences as a function of the fraction of planets harboring runaway greenhouse climates for different versions of the \plato\ survey.
+            $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
+            For a large planet yield of $N = \var{Nplato}$, even small planet fractions $\gtrsim \var{platoMinfrgh}$ allow a detection.
+                A much smaller sample of 100 planets is sufficient if their masses are constrained to within \SI{10}{\percent} (dashed green line).
+                Without such follow-up measurements, sufficient diagnostic power can only be achieved with this sample if $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred}$.
+               Even smaller samples are unlikely to yield a significant detection.
+            }
+            \label{fig:plato_fgrid}
+        \end{centering}
+    \end{figure}
+
+For a volume-limited sample of \var{Nplato} planets, considering all host star spectral types, we find that \plato\ is able to obtain a significant detection if the fraction of runaway greenhouse planets is larger than $\sim \var{platoMinfrgh}$ (see Fig.~\ref{fig:plato_fgrid}).
+The minimum needed fraction rises to \var{platoMinfrghHundred} for $N = 100$.
+For much smaller samples, even an optimistically strong signal can no longer be detected.
 

### 2023-01-19 src/tex/ms.tex
create combined figure exploring sample size and FGK vs. M.

@@
 
-    \begin{figure}[ht!]
+    \begin{figure*}[ht!]
 %    \script{figurescript.py}
@@
             $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
-            For a large planet yield of $N = \var{Nplato}$, even small planet fractions $\gtrsim \var{platoMinfrgh}$ allow a detection.
+            \textit{Left:} For a large planet yield of $N = \var{Nplato}$, even small planet fractions $\gtrsim \var{platoMinfrgh}$ allow a detection.
                 A much smaller sample of 100 planets is sufficient if their masses are constrained to within \SI{10}{\percent} (dashed green line).
@@
                Even smaller samples are unlikely to yield a significant detection.
+            \textit{Right:} Evidences when only FGK or only M~dwarfs are considered.
+                Only M~dwarfs host enough planets on both sides of the threshold instellation to allow a detection of the runaway greenhouse signal.
             }
@@
         \end{centering}
-    \end{figure}
+    \end{figure*}
 
@@
 ...
+\begin{note}
+...FOLLOW-UP...
+
+The advantage of mass measurements for the planet sample can be enhanced with a wise selection of follow-up targets, i.e. favoring those planets residing close to $S_\mathrm{thresh}$.
+\end{note}
 

### 2023-01-19 src/tex/ms.tex
remove example plots
add results section on pure radius vs bulk density measurements

@@
 \subsection{Statistical power of different mission designs}\label{sec:statpower_missions}
-\todo[inline]{show detectability of the runaway greenhouse signature for different mission designs}
+\subsubsection{Additional planet mass measurements}
+The situation differs when additional information beyond planet radii is available for the characterized planet population.
+Since the planet bulk density $\rho \propto R^{-3}$, the signal is stronger when measured in bulk density instead of radius:
+With an optimistic choice of geophysical parameters (see above), the average measured radius change is \SI{+15}{\percent} whereas the average density change is \SI{-33}{\percent}.
+Constraining bulk densities requires planetary mass measurements.
+We now examine the value of a follow-up campaign to constrain planet masses with respect to the detectability of the runaway greenhouse threshold.
 
-\begin{figure}[ht!]
-%    \script{figurescript.py}
-    \begin{centering}
+For uncertainties in density measurements, we propagate normally distributed errors of mass measurements assuming $\sigma_\mathrm{M_\mathrm{P}} = \SI{10}{\percent}$, as expected for \plato.
+As can be seen in Fig.~\ref{fig:plato_fgrid}, the detectability is enhanced when radius constraints can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
+Consequently, we achieve a statistically significant detection with smaller samples or lower fractions $f_\mathrm{rgh}$.
+A density-based survey of 100 targets is roughly equivalent to a radius-based survey of 500 targets.
+At $N=100$, pure radius measurements require $f_\mathrm{rgh} \gtrsim 0.5$ whereas bulk density measurements enable a detection from $f_\mathrm{rgh} \gtrsim 0.2$.
 
-        \includegraphics[width=\linewidth]{figures/example_Ndetected_surveys.pdf}
-        \caption{
-        Detections of magma ocean-bearing planets as a function of magma ocean occurrence for different survey designs.
-        }
-        \label{fig:example_Ndetected_surveys}
-    \end{centering}
-\end{figure}
+\subsubsection{Spectral type of target stars}
 

### 2023-01-19 src/tex/ms.tex
add results section on FGK vs. M

@@
 
-\subsubsection{Spectral type of target stars}
+\subsubsection{Spectral type of target stars}\label{sec:results_FGK_M}
+Since receiving radiation is sensitive on the spectral type of the host star, the share of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
+We tested the detectability of the runaway greenhouse signal when only FGK or only M~dwarfs are considered (see Fig.~\ref{fig:plato_fgrid}).
+The samples are volume and magnitude-limited ($d_\mathrm{max} = \SI{55}{\parsec}, M_\mathrm{G} < 11$), which makes the M~dwarf sample significantly larger with \var{N_M} planets compared to \var{N_FGK} planets in the FGK sample.
+No significant detection is possible in the pure FGK sample, independent of the assumed geophysical parameters.
+In the M~dwarf sample, the evidence threshold is reached around $f_\mathrm{rgh} \sim 0.5$, similar to the case above where all spectral types are considered.
 
@@
 
+\begin{note}
+    ...FGK vs. M...
+
+To date, the majority of planets with radius measurements orbit FGK~dwarfs and lie in the runaway greenhouse regime.
+    ...
+
+    Our tests with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
+The reason is, besides their large number in a volume/magnitude-limited sample, that many detected M~dwarf planets are located near the threshold instellation.
+As in the cases above, an M~dwarf survey profits from follow-up measurements of planetary masses with an order of magnitude increase in evidence.
+\end{note}
+
 \subsection{Implications for planetary habitability}\label{sec:habitability}

### 2023-01-20 src/tex/ms.tex
discuss CHEOPS and Ariel

@@
 \todo{Introduce survey simulation(s).
-, Cheops, Ariel, LIFE (reach out to team; Hendrik Linz produced a LIFE Technology Overview document)?, Nautilus\\ What are threshold missions that are able to detect this signal?
+PLATO, (Ariel, LIFE(reach out to team; Hendrik Linz produced a LIFE Technology Overview document)) ?, Nautilus\\ What are threshold missions that are able to detect this signal?
 =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
@@
 
-\subsubsection{PLATO}
+%\subsubsection{PLATO}
 \plato\ (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
@@
 
+What other planned missions are suited to test the runaway greenhouse hypothesis?
+The ongoing \cheops\ mission was designed as a follow-up mission to search for transits of planets discovered with other techniques, in particular with radial velocity measurements~\citep{Benz2021}.
+As such, it will provide precise radius constraints on a sample of small planets; however, only a small number of planets with periods \SI{> 50}{\day} are being observed.
+This largely limits \cheops' coverage to planets within the runaway greenhouse regime, preventing a detection of the transition.
+
+As \cheops, \ariel~\citep{Puig2016} will be a follow-up mission that is not designed to provide a large number of new radius measurements.
+\ariel's primary targets are larger planets in the range of sub-Neptune to Jupiter-like planets.
+We thus do not expect a significant contribution to exploring the inner edge of the habitable zone for Earth-sized planets.
+
 \subsection{Statistical power of different mission designs}\label{sec:statpower_missions}
 \subsubsection{Additional planet mass measurements}
-The situation differs when additional information beyond planet radii is available for the characterized planet population.
+The dependencies of a detection change when additional information beyond planet radii is available for the characterized planet population.
 Since the planet bulk density $\rho \propto R^{-3}$, the signal is stronger when measured in bulk density instead of radius:
@@
 \begin{enumerate}
-    \item The predicted tipping point in planetary climates causes a discontinuity in the radius and density distribution of small (DEFINITION) exoplanets.
+    \item The predicted tipping point in planetary climates causes a discontinuity in the radius and density distribution of small exoplanets.
     \item The demographic imprint of the runaway greenhouse threshold should be detectable with transit measurements of a large enough sample and sufficient precision.

### 2023-01-21 src/tex/ms.tex
discuss sensitivity of detectability on x_H2O and f_rgh

@@
     The higher the water content and thus the radius change, and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
-    As long as this fraction is larger than $\sim 0.1$, a sample size of \Nplanets\ is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $x_{H_2O} = 0.001$.
+     $f_\mathrm{rgh}$ dominates this trend: It enters linearly into the average planet radius, whereas the contribution of $x_{H_2O}$ - as predicted by the geophysical models - is sublinear with a power-law exponent of $\sim 0.3$.
+    As long as $f_\mathrm{rgh}$ is larger than $\sim 0.1$, a sample size of \Nplanets\ is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $x_{H_2O} = 0.001$.
 \end{note}

### 2023-01-21 src/tex/ms.tex
move text around

@@
 The planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
-To test the detectability of a runaway greenhouse transition, we simulated several transit surveys with different designs and strategies and explored their capability to recover the trend and constrain its parameters.
-We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
 
@@
 \subsection{Statistical power of different mission designs}\label{sec:statpower_missions}
+To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated several transit surveys with different designs and strategies and measured their capability to recover the trend and constrain its parameters.
+We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
+
 \subsubsection{Additional planet mass measurements}
@@
 In the M~dwarf sample, the evidence threshold is reached around $f_\mathrm{rgh} \sim 0.5$, similar to the case above where all spectral types are considered.
+
+\subsubsection{Constraining the threshold instellation}
+\todo[inline]{for different survey modes, compare the constraints on  $S_{thresh}$.
+Plot posteriors from random draws of simulations.}
 

### 2023-01-30 src/tex/ms.tex
update PLATO figure with finer grid

@@
             \caption{
-            Expected evidences as a function of the fraction of planets harboring runaway greenhouse climates for different versions of the \plato\ survey.
-            $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
-            \textit{Left:} For a large planet yield of $N = \var{Nplato}$, even small planet fractions $\gtrsim \var{platoMinfrgh}$ allow a detection.
-                A much smaller sample of 100 planets is sufficient if their masses are constrained to within \SI{10}{\percent} (dashed green line).
+            Expected delta-evidences as a function of the fraction of planets harboring runaway greenhouse climates for different versions of the \plato\ survey.
+            Median and $[16\mathrm{th}, 84\mathrm{th}]$ percentiles of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
+            \textit{Left:} For a large planet yield of $N = \var{Nplato}$, even small planet fractions $\sim \var{platoMinfrgh}$ allow a detection.
+                A sample of 100 planets is sufficient if their masses are constrained to within \SI{10}{\percent} (dotted green line).
                 Without such follow-up measurements, sufficient diagnostic power can only be achieved with this sample if $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred}$.
@@
             \textit{Right:} Evidences when only FGK or only M~dwarfs are considered.
-                Only M~dwarfs host enough planets on both sides of the threshold instellation to allow a detection of the runaway greenhouse signal.
+                Only M~dwarfs host enough planets on both sides of the threshold instellation to allow a reliable detection of the runaway greenhouse signal.
             }
@@
 A density-based survey of 100 targets is roughly equivalent to a radius-based survey of 500 targets.
-At $N=100$, pure radius measurements require $f_\mathrm{rgh} \gtrsim 0.5$ whereas bulk density measurements enable a detection from $f_\mathrm{rgh} \gtrsim 0.2$.
+At $N=100$, pure radius measurements require $f_\mathrm{rgh} \gtrsim 0.4$ whereas bulk density measurements enable a detection from $f_\mathrm{rgh} \gtrsim 0.1$.
 
@@
 No significant detection is possible in the pure FGK sample, independent of the assumed geophysical parameters.
-In the M~dwarf sample, the evidence threshold is reached around $f_\mathrm{rgh} \sim 0.5$, similar to the case above where all spectral types are considered.
+In the M~dwarf sample, the evidence threshold is reached around $f_\mathrm{rgh} \sim 0.2$, similar to the case above where all spectral types are considered.
 

### 2023-01-30 src/tex/ms.tex
comment on M dwarf rgh phases

@@
 As in the cases above, an M~dwarf survey profits from follow-up measurements of planetary masses with an order of magnitude increase in evidence.
+
+    An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of \SI{100}{\mega\year}~CITE.
+    This increases the probability of observing any given planet in the sample during the runaway greenhouse phase, essentially driving $f_\mathrm{rgh}$ to higher values.
 \end{note}

### 2023-01-30 src/tex/ms.tex
update and expand conclusions
based on new results

@@
 =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
-
-The planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
-
+Under real-world conditions, the planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
 %\subsubsection{PLATO}
@@
 \begin{note}
-    ...FGK vs. M...
+    ...FGK vs. M...:
 
@@
 \section{Conclusions}
-\begin{note}
 % Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.
-    We have explored the potential of contemporary exoplanet missions to statistically detect a radius discontinuity that is predicted to result from runaway greenhouse phases.
-    ...Bioverse...
-\end{note}
-
-The key findings of our work are as follows:
+Runaway greenhouse climates on Earth-like planets are thought to cause significant inflation of their radii.
+Using Bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius discontinuity resulting from this inflation in the exoplanet population.
+Our key findings are as follows:
 \begin{enumerate}
-    \item The predicted tipping point in planetary climates causes a discontinuity in the radius and density distribution of small exoplanets.
-    \item The demographic imprint of the runaway greenhouse threshold should be detectable with transit measurements of a large enough sample and sufficient precision.
-    \item We find that the planned \plato\ transit survey will provide a sufficient sample to test the runaway greenhouse hypothesis.
-    \item The diagnostic power of transit missions in testing this hypothesis can be maximized by ...
+    \item The predicted tipping point in planetary climates causes a discontinuity in the radius and density distribution of small exoplanets with respect to the incident host star irradiation.
+    \item The demographic imprint of the runaway greenhouse threshold should be detectable with high-precision transit measurements.
+For a large sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of planets and if typical water mass fractions are above $\sim 10^{-3}$.
+    \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to detect the predicted trend.
+Assuming the projected photometric precision, \plato\ will be able to test the runaway greenhouse hypothesis for planet yields $\gtrsim 100$.
+    \item The diagnostic power of transit missions in testing this hypothesis can be increased through a follow-up campaign providing planet mass measurements.
+This can reduce the required planet yield by up to a factor of five.
+An optimized survey will further focus on M~dwarfs to ensure sufficient targets on both sides of the expected threshold instellation.
     \item A detection of the feature will constrain the water inventory of rocky exoplanets, providing important context in the assessment of their habitability.

### 2023-02-07 src/tex/ms.tex
select FGK/M samples to reflect PLATO's expected LOP fields

@@
 We tested the detectability of the runaway greenhouse signal when only FGK or only M~dwarfs are considered (see Fig.~\ref{fig:plato_fgrid}).
-The samples are volume and magnitude-limited ($d_\mathrm{max} = \SI{55}{\parsec}, M_\mathrm{G} < 11$), which makes the M~dwarf sample significantly larger with \var{N_M} planets compared to \var{N_FGK} planets in the FGK sample.
+%($d_\mathrm{max} = \SI{75}{\parsec}, M_\mathrm{G} < 11$)
+The samples are volume and magnitude-limited to reflect the target counts of \plato's provisional Long-duration Observation Phase fields~\citep[$15996$ FGK stars in the P1 and P2 samples, $33948$ M~stars in the P4 sample, ][]{Nascimbeni2022}.
+The resulting M~dwarf planet sample is significantly larger with $\var{N_M}\pm \var{N_M_err}$ planets compared to $\var{N_FGK}\pm \var{N_FGK_err}$ planets in the FGK sample.
 No significant detection is possible in the pure FGK sample, independent of the assumed geophysical parameters.

### 2023-02-07 src/tex/ms.tex
minor edits

@@
 % Define document class
-%\documentclass[twocolumn]{aastex631}
-\documentclass[modern]{aastex631}
+\documentclass[twocolumn]{aastex631}
+\usepackage{amssymb}
+%\documentclass[modern]{aastex631}
 
@@
 \todo{Introduce survey simulation(s).
-PLATO, (Ariel, LIFE(reach out to team; Hendrik Linz produced a LIFE Technology Overview document)) ?, Nautilus\\ What are threshold missions that are able to detect this signal?
+PLATO, Cheops, (Ariel, LIFE/LIFE precursor, Nautilus)\\ What are threshold missions that are able to detect this signal?
 =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
@@
 The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).
-In the FGK sample of the mission, a minimum of 500 Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
+In the FGK sample of the mission, several hundred Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
 \plato's definition study report~\citep{plato2017} further states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).
@@
             Median and $[16\mathrm{th}, 84\mathrm{th}]$ percentiles of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
-            \textit{Left:} For a large planet yield of $N = \var{Nplato}$, even small planet fractions $\sim \var{platoMinfrgh}$ allow a detection.
+            \textit{Left:} For a large planet yield of $N = \var{N_plato}$, even small planet fractions $\sim \var{platoMinfrgh}$ allow a detection.
                 A sample of 100 planets is sufficient if their masses are constrained to within \SI{10}{\percent} (dotted green line).
@@
 
-For a volume-limited sample of \var{Nplato} planets, considering all host star spectral types, we find that \plato\ is able to obtain a significant detection if the fraction of runaway greenhouse planets is larger than $\sim \var{platoMinfrgh}$ (see Fig.~\ref{fig:plato_fgrid}).
+For a volume-limited sample, considering all host star spectral types, we find that a yield of \var{N_plato} planets enables a significant detection if the fraction of runaway greenhouse planets is larger than $\sim \var{platoMinfrgh}$ (see Fig.~\ref{fig:plato_fgrid}).
 The minimum needed fraction rises to \var{platoMinfrghHundred} for $N = 100$.
@@
 Plot posteriors from random draws of simulations.}
+A key constraint resulting from a successful detection is a measurement of $S_{thresh}$.
+Here, we assess the ability of different mission concepts explored above to constrain this parameter.
+...
 
@@
 
-\todo[inline]{Maps to Sect. \ref{sec:statpower_missions}.}
-
-\begin{note}
-    \citet{Penny2019} mention that a significant increase in planet yield could be achieved if the telescope's slew speed was increased.
-\end{note}
+%\todo[inline]{Maps to Sect. \ref{sec:statpower_missions}.:}
+%\begin{note}
+%    \citet{Penny2019} mention that a significant increase in planet yield could be achieved if the telescope's slew speed was increased.
+%\end{note}
 
@@
 ...
+\subsubsection{The value of follow-up campaigns}
 \begin{note}
-...FOLLOW-UP...
+...FOLLOW-UP...radial velocity follow-up measurements...
 
@@
 
+\subsubsection{Target star spectral types}
 \begin{note}

### 2023-02-10 src/tex/ms.tex
increase sample size of optimistic case

@@
         Detection of the runaway greenhouse threshold in the optimistic case.
-        From simulated radius and instellation measurements of a large ($N = \Nplanets$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
+        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
             The pattern is detected with high significance.
@@
 Figure~\ref{fig:optimistic_R-S} shows a prototypical detection of the runaway greenhouse threshold.
-For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \Nplanets$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
+For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
 We assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \wrr\ for each planet.
@@
             Retrieved posterior distribution of key parameters in the optimistic scenario.
-The density maps in each panel show relationships between and marginalized distributions of the model parameters as they could be retrieved with a high-precision transit survey and a sample of \Nplanets\ planets.
+The density maps in each panel show relationships between and marginalized distributions of the model parameters as they could be retrieved with a high-precision transit survey and a sample of \var{N_optimistic} planets.
 True values of the parameters for the injected effect are shown in orange.
@@
 Figure~\ref{fig:cornerplot} shows the posterior distributions of $S_\mathrm{thresh}$, $x_{H_2O}$, and $f_\mathrm{rgh}$ as determined by the nested sampler.
-The threshold instellation can be accurately constrained, $S_\mathrm{thresh} = \var{retrSthresh}$.
+The threshold instellation can be accurately constrained with a median $\hat{S}_\mathrm{thresh} = \var{retrSthresh}$.
 Both a higher water mass fraction and a higher fraction of planets in a runaway greenhouse state lead to larger average radii, thus these parameters are strongly correlated and cannot be well constrained without independent measurements.
@@
                 Statistical power of the runaway greenhouse hypothesis test as a function of model parameters.
-                For a sample size $N = \Nplanets$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and fraction of greenhouse climate planets.
+                For a sample size $N = \var{N_optimistic}$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and fraction of greenhouse climate planets.
                 Higher values in either parameter result in a more reliable detection.
@@
      $f_\mathrm{rgh}$ dominates this trend: It enters linearly into the average planet radius, whereas the contribution of $x_{H_2O}$ - as predicted by the geophysical models - is sublinear with a power-law exponent of $\sim 0.3$.
-    As long as $f_\mathrm{rgh}$ is larger than $\sim 0.1$, a sample size of \Nplanets\ is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $x_{H_2O} = 0.001$.
+    As long as $f_\mathrm{rgh}$ is larger than $\sim 0.1$, a sample size of \var{N_optimistic} is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $x_{H_2O} = 0.001$.
 \end{note}

### 2023-02-10 src/tex/ms.tex
add new PLATO simulations
uniform prior in S_thresh, not log-uniform
constrain threshold instellation with different survey realizations

@@
 
-\subsection{Detecting the runaway greenhouse transition with current and planned exoplanet missions}
+\subsection{Detecting the runaway greenhouse transition with current and planned exoplanet missions}\label{sec:res_samplesize}
 \todo{Introduce survey simulation(s).
@@
 Consequently, we achieve a statistically significant detection with smaller samples or lower fractions $f_\mathrm{rgh}$.
-A density-based survey of 100 targets is roughly equivalent to a radius-based survey of 500 targets.
+A density-based survey of 100 targets is roughly equivalent to a radius-based survey of \var{N_plato} targets.
 At $N=100$, pure radius measurements require $f_\mathrm{rgh} \gtrsim 0.4$ whereas bulk density measurements enable a detection from $f_\mathrm{rgh} \gtrsim 0.1$.
@@
 \subsubsection{Constraining the threshold instellation}
-\todo[inline]{for different survey modes, compare the constraints on  $S_{thresh}$.
-Plot posteriors from random draws of simulations.}
+\begin{figure}[ht!]
+%    \script{figurescript.py}
+    \begin{centering}
+
+        \includegraphics[width=\linewidth]{figures/S_thresh_posteriors.pdf}
+        \caption{
+            Retrieved posterior distributions of the threshold instellation for different survey realizations.
+            All cases assume $x_{H_2O}$ and a planet sample size $N=100\pm10$; The fraction of planets with runaway greenhouse climates varies across rows.
+            Orange lines show the true value of the injected signal.
+            Accuracy and precision of the constraint on $S_\mathrm{thresh}$ generally improve with higher $f_\mathrm{rgh}$.
+            Bulk density-based inferences improve the constraints.
+        }
+        \label{fig:posterior_surveys}
+    \end{centering}
+\end{figure}
 A key constraint resulting from a successful detection is a measurement of $S_{thresh}$.
 Here, we assess the ability of different mission concepts explored above to constrain this parameter.
-...
+Figure~\ref{fig:posterior_surveys} shows posterior distributions from a grid of inferences together with the true values of the injected signal for different fractions $f_\mathrm{rgh}$.
+We consider three cases: only radius measurements, radius and mass measurements, and radius and mass measurements of only planets orbiting M~dwarfs.
+The simulations otherwise represent the simulated PLATO example described above.
+We chose a planetary sample size of $N=100\pm10$, as this was found to be a threshold case in Sect.~\ref{sec:res_samplesize}.
+
+Accurate retrievals from radius measurements alone require high fractions of greenhouse climate-bearing planets to achieve an accurate constraint on $S_\mathrm{thresh}$.
+If, on the other hand, a follow-up campaign adds planetary mass estimates and we fit the signal in bulk density space, useful constraints are possible under more pessimistic conditions.
+A sample containing only planets around M~dwarf yields a comparable performance as the overall sample.
 
@@
 %\end{note}
-
-\begin{figure}[ht!]
-%    \script{figurescript.py}
-    \begin{centering}
-
-        \includegraphics[width=\linewidth]{figures/example_posterior_surveys.pdf}
-        \caption{
-            Retrieved posterior distribution of the $<key parameter>$ for different survey realizations.
-            The orange line corresponds to the true value of the injected signal.
-        }
-        \label{fig:posterior_surveys}
-    \end{centering}
-\end{figure}
 

### 2023-02-10 src/tex/ms.tex
update statistical power grid
based on new optimistic sample size

@@
 
-\begin{note}
 Given an optimistic survey design, what influences the probability of correctly rejecting a false null hypothesis?
@@
             Higher values in either parameter result in a more reliable detection.
+            For water mass fractions $\gtrsim 10^{-4}$, the statistical power largely depends on the fraction of greenhouse climate planets in the sample.
         }
@@
 The higher the water content and thus the radius change, and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
-     $f_\mathrm{rgh}$ dominates this trend: It enters linearly into the average planet radius, whereas the contribution of $x_{H_2O}$ - as predicted by the geophysical models - is sublinear with a power-law exponent of $\sim 0.3$.
-    As long as $f_\mathrm{rgh}$ is larger than $\sim 0.1$, a sample size of \var{N_optimistic} is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $x_{H_2O} = 0.001$.
-\end{note}
+ For all but very low water fractions, $f_\mathrm{rgh}$ dominates this trend: It enters linearly into the average planet radius, whereas the contribution of $x_{H_2O}$ - as predicted by the geophysical models - is sublinear with a power-law exponent of $\sim 0.3$.
+As long as $f_\mathrm{rgh}$ is larger than $\sim 0.2$, a sample size of \var{N_optimistic} is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $10^{-3}$.
 

### 2023-02-15 src/tex/ms.tex
add figure w/ schematic radius evolution

@@
         \includegraphics[width=\hsize]{figures/radiuscomparison.pdf}
-        \caption{Mass-radius relationships for a water mass fraction $x_{H_2O}= \wrr$ and different planet states: A) fully solid planets without magma oceans.
+    \end{centering}
+    \begin{centering}
+        \includegraphics[width=\hsize]{figures/radiusevolution.pdf}
+        \caption{
+            \textit{Top}: Mass-radius relationships for a water mass fraction $x_{H_2O}= \wrr$ and different planet states: A) planets with a solid mantle and no steam atmosphere.
 B) planets with steam atmospheres.
-C) As B), but including the effect of water incorporation in the melt.}
-        \label{fig:radiuscomparison}
+C) as B), but including the effect of water incorporation in the melt.
+            \textit{Bottom}: Radius evolution of different planet types, illustrating degeneracies among planet classes.
+Shown is a schematic time evolution of the transit radius normalized to the atmosphere-free radius $R_\mathrm{0}$ for different scenarios.
+Planets can move between planet classes through processes such as atmospheric loss and desiccation, which ultimately ends a runaway greenhouse phase on a timescale dependent on a planet's water content.}
+        \label{fig:radiusevolution}
     \end{centering}
 \end{figure}
+Figure~\ref{fig:radiusevolution} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of $x_{H_2O}= \wrr$.
+Steam atmospheres (case~B) cause a significant radius increase, which is slightly reduced when water incorporation in the melt is allowed (case~C).
+In the following, we only distinguish between cases~A and~C.
 
-Figure~\ref{fig:radiuscomparison} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of $x_{H_2O}= \wrr$.
-In general, steam atmospheres (case~B) cause a significant radius increase, which is slightly reduced when water incorporation in the melt is allowed (case~C).
-
+As illustrated in Fig.~\ref{fig:radiusevolution}~(bottom panel), different processes can lead to the classification of a planet into either of these categories.
+For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a case~C planet, diluting the expected demographic signature.
+Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to category~A~\citep[e.g.,][]{Hamano2013}.
+To account for these effects, we introduce a free parameter $f_\mathrm{rgh}$ representing the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
 Our simulation setup is such that all planets receiving an instellation $S < S_\mathrm{thresh}$ follow relation~A, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow relation~C.

### 2023-02-15 src/tex/ms.tex
streamline description of planet classes

@@
     This case serves as our null hypothesis, where planetary radii are not affected by the amount of irradiation the planets receive.
-    It also applies outside the runaway greenhouse limit, where we assume a solid planetary surface and no steam atmosphere.
-        A planet that lies \textit{within} the runaway greenhouse limit but that is already completely desiccated will have the same radius.
-\todo{Mention if this case is considered in the model or not (as of now, no).}
+    A solid planetary surface is assumed and no steam atmosphere is present.
    \item[B.] Radius inflation due to a steam atmosphere.
-    We assigned planets within the runaway greenhouse limit steam atmospheres, which increase atmospheric scale heights significantly.
-   \item[C.] As B, but including radius decrease of inflated planets due to water in the melt.
-    Due to greenhouse forcing, planetary surfaces below steam atmospheres are thought to be molten.
-    Part of the water will then partition into the melt, removing it from the steam atmosphere, which in turn decreases the radius inflation.
-    This effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes typically reach measurable values.
+    Atmospheric scale heights are significantly increased.
+   \item[C.] As B, but including radius decrease of inflated planets due to water partitioning in the molten mantle.
     \end{itemize}
@@
 Case~C contains case~B, but also includes a radius decrease due to water partitioning in the melt.
+Due to greenhouse forcing, planetary surfaces below steam atmospheres are expected to be molten.
+Part of the water will then partition into the melt, removing it from the steam atmosphere, which in turn decreases the radius inflation.
+This effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes typically reach measurable values~\citep{Dorn2021}.
 The magnitude of this effect is again dependent on the planet mass and on the water mass fraction.
@@
 Our simulation setup is such that all planets receiving an instellation $S < S_\mathrm{thresh}$ follow relation~A, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow relation~C.
-The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
+%The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
 This parametrization introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.

### 2023-02-15 src/tex/ms.tex
fix python-generated latex variables

@@
 \documentclass[twocolumn]{aastex631}
-\usepackage{amssymb}
 %\documentclass[modern]{aastex631}
@@
     which is relevant for the detection probability of the respective planet~(see Sect.~\ref{sec:sensitivity}).
-    Excluding all non-transiting planets diminishes the sample to \SI{\percentageTransiting}{\percent} of its original size.
+    Excluding all non-transiting planets diminishes the sample to \var{percentage_transiting} of its original size.
 \end{note}
@@
 \subsubsection{Planet radius changes due to runaway greenhouse climates}\label{sec:mo_model}
-\todo[inline]{"Methods Section" for the geophysical models. Describe specific parametrization, dependency on bulk planet/orbit params\\
-~\\
+\todo[inline]{
 \textbf{Main assumptions:}\\
-- Assume that every planet has \textit{some} water. Enough for runaway GH under the right conditions. Baseline water content: 20\% \\
+- Assume that every planet has \textit{some} water. Enough for runaway GH under the right conditions.  \\
 - Limitation to small planets without significant atmospheric envelopes.\\
@@
     However, the Neptune desert occurs at smaller orbital periods and its location depends on a planet's size~\citep{Szabo2011}, which is not expected for the runaway greenhouse transition.
-    \todo{is there overlap/confusion with the radius/density trend in \citet{Luque2022b}?}
+    \todo{is there overlap/confusion with the radius/density trend in \citet{Luque2022}?}
 \end{note}

### 2023-02-15 src/tex/ms.tex
uniform prior for S_thresh

@@
 As the parameters we are interested in recovering, $S_\mathrm{thresh}$ and $x_{H_2O}$, are poorly constrained by previous data, we used relatively uninformative priors to sample the entire physically plausible parameter space.
-For $S_\mathrm{thresh}$, we chose a log-uniform prior in $[\prSmin,\prSmax]\, \SI{}{\watt\per\meter\squared}$.
-Similarly, we sample $x_{H_2O}$ from a log-uniform distribution to imply scale-invariant ignorance.
+For $S_\mathrm{thresh}$, we chose a uniform prior in $[\prSmin,\prSmax]\, \SI{}{\watt\per\meter\squared}$.
+We sample $x_{H_2O}$ from a log-uniform distribution to imply scale-invariant ignorance.
 Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Sect.~\ref{sec:mo_model}).

### 2023-02-15 src/tex/ms.tex
justify choice of MR relation
call S "net instellation"

@@
 The default relation our planets follow is one assuming a pure $\mathrm{MgSiO_3}$ composition.
-\todo[inline]{JUSTIFY this choice (TIM?)}
+While interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure and detectability.
 \todo[inline]{run a test: do we expect that the choice of M-R relation impacts the results? Perhaps have an appendix section.}
@@
 \begin{note}
-    The power from the host star per unit area at the position of a planet or instellation $S$ in units of Earth's insolation is given by
+    The power from the host star per unit area at the position of a planet or net instellation $S$ in units of Earth's insolation is given by
     \begin{equation}
@@
     While we search for the runaway greenhouse pattern in demographic quantities such as average planet radii, the injected changes happen on the planetary level: We changed each planet's radius based on its individual set of properties and the associated predictions from steam atmosphere and water incorporation models.
-    Relevant properties are a planet's mass $M$, its instellation $S$, and its bulk water inventory expressed as a water mass fraction $x_{H_2O}$.
+    Relevant properties are a planet's mass $M$, its net instellation $S$, and its bulk water inventory expressed as a water mass fraction $x_{H_2O}$.
     We consider three different cases:
@@
 To account for these effects, we introduce a free parameter $f_\mathrm{rgh}$ representing the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
-Our simulation setup is such that all planets receiving an instellation $S < S_\mathrm{thresh}$ follow relation~A, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow relation~C.
+Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow relation~A, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow relation~C.
 %The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
@@
 The runaway greenhouse effect becomes obsolete where no atmosphere can be maintained due to proximity to the host star and resulting atmospheric erosion.
-We thus excluded planets with extremely high instellation that does not allow long-term preservation of an atmosphere and clear our sample from all planets with an instellation $S > \SI{2000}{\watt\per\square\meter}$.
+We thus excluded planets with extremely high instellation that does not allow long-term preservation of an atmosphere and clear our sample from all planets with a net instellation $S > \SI{2000}{\watt\per\square\meter}$.
 \todo[inline]{use criterion of Zahnle \& Caitling 2017 instead? $0.8 S^{0.25} < R$}
@@
     We further define an alternative hypothesis that describes radius changes due to steam atmospheres and water in the melt.
-    As motivated above, this hypothesis takes the form of a step function in instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
+    As motivated above, this hypothesis takes the form of a step function in net instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
     Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
@@
         \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
-        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of instellation.
+        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation.
 Only the planets marked as transiting are observable.
@@
     An example is the "Neptune desert``, a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
-    The shape of this region is such that smaller planets become less frequent the closer to the star they are, which to some degree resembles the pattern introduced by the instellation dependency of the magma ocean probability.
+    The shape of this region is such that smaller planets become less frequent the closer to the star they are, which to some degree resembles the pattern introduced by the instellation dependency of the runaway greenhouse transition.
     However, the Neptune desert occurs at smaller orbital periods and its location depends on a planet's size~\citep{Szabo2011}, which is not expected for the runaway greenhouse transition.

### 2023-02-16 src/tex/ms.tex
include some small updates

@@
 \textbf{Main assumptions:}\\
-- Assume that every planet has \textit{some} water. Enough for runaway GH under the right conditions.  \\
+- every planet (modulo $f_\mathrm{rgh}$) has \textit{some} water. Enough for runaway GH under the right conditions.  \\
 - Limitation to small planets without significant atmospheric envelopes.\\
@@
 %    We further define the solar-equivalent semi-major axis $a_{eff} = a (L_\star/L_\odot)^{-1/2}$, at which a planet experiences the same instellation as a Solar System planet at an orbital distance~$a$.
+   We do not take into account additional heating sources such as tidal effects~\citep[e.g.][]{Barnes2013}.
 \end{note}
@@
 The runaway greenhouse effect becomes obsolete where no atmosphere can be maintained due to proximity to the host star and resulting atmospheric erosion.
-We thus excluded planets with extremely high instellation that does not allow long-term preservation of an atmosphere and clear our sample from all planets with a net instellation $S > \SI{2000}{\watt\per\square\meter}$.
-\todo[inline]{use criterion of Zahnle \& Caitling 2017 instead? $0.8 S^{0.25} < R$}
+Ensuring to stay well clear of such regions, we clear our sample from all planets with a net instellation $S > \SI{2000}{\watt\per\square\meter}$.
+%\todo[inline]{use criterion of Zahnle \& Caitling 2017 instead? $0.8 S^{0.25} < R$}
 We further consider only rocky planets with masses below \SI{2}{\Mearth}. % (Turbet+2020 covers only 0.1 to 2 Mearth)
@@
 \end{equation}
-The chosen threshold roughly corresponds to the $p = 0.05$ threshold for statistical significance popular in frequentist approaches.\todo{CITATION}
+%The chosen threshold roughly corresponds to the $p = 0.05$ threshold for statistical significance popular in frequentist approaches.\todo{CITATION}
 
@@
             Expected delta-evidences as a function of the fraction of planets harboring runaway greenhouse climates for different versions of the \plato\ survey.
-            Median and $[16\mathrm{th}, 84\mathrm{th}]$ percentiles of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
+            The median values of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
             \textit{Left:} For a large planet yield of $N = \var{N_plato}$, even small planet fractions $\sim \var{platoMinfrgh}$ allow a detection.
@@
 \todo[inline]{the predicted climate tipping point causes a demographic imprint.\\
+...detection would provide an empirical confirmation of the inner edge of the HZ.\\
 Discuss our ignorance on $f_\mathrm{rgh}$ and its impact on the imprint.}
@@
 An optimized survey will further focus on M~dwarfs to ensure sufficient targets on both sides of the expected threshold instellation.
-    \item A detection of the feature will constrain the water inventory of rocky exoplanets, providing important context in the assessment of their habitability.
+    \item Testing the runaway greenhouse hypothesis will constrain the water inventory of rocky exoplanets, providing important context in the assessment of their habitability.
+         A detection will provide an empirical confirmation of the habitable zone concept and localize its inner edge.
+
 \end{enumerate}

### 2023-02-17 src/tex/ms.tex
draft discussion for statistical imprint

@@
 Case~B assumes an inflated radius due to a steam atmosphere for all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$.
-This value was found to be a limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Leconte2013}.
+While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Leconte2013}.
 To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
@@
 For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a case~C planet, diluting the expected demographic signature.
-Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to category~A~\citep[e.g.,][]{Hamano2013}.
+Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to category~A on a timescale of $10^8\SI{}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
 To account for these effects, we introduce a free parameter $f_\mathrm{rgh}$ representing the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
@@
 \section{Discussion}\label{sec:discussion}
-\todo[inline]{\textbf{Potential discussion points: } what is the fraction of \kepler\ planets w/ runaway GH.}
+\todo[inline]{\textbf{Potential discussion points: }\\
+How close is Earth to runaway? Could anthropogenic climate change push us over the edge? (-> \citet{Goldblatt2012}; what is the max. expected change in Earth's albedo due to deglaciation?}
 
 \subsection{Statistical imprint of exoplanet climates}
-\todo[inline]{the predicted climate tipping point causes a demographic imprint.\\
-...detection would provide an empirical confirmation of the inner edge of the HZ.\\
-Discuss our ignorance on $f_\mathrm{rgh}$ and its impact on the imprint.}
+\todo[inline]{what is the fraction of Kepler planets w/ run-
+away GH?}
+\begin{note}
+The runaway greenhouse transition is a robust prediction from climate models, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
+Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
+On the planetary level, however, such undertaking is hindered by degeneracies: From internal composition to atmospheric loss to volatile content, many factors influence a planet's size.
+It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019}.
+Using Bioverse to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we showed that the population of greenhouse-inflated planets create a signature in the radius and density distribution that is well in the detectable range.
+
+However, the idealized step-function central to our hypothesis tests is blurred by several factors.
+Its amplitude is affected by ``sub-Neptunes'' with significant Hydrogen/Helium envelopes, which can disguise as inflated rocky planets.
+They might not contribute to the demographic signal we are looking for, although a similar transition was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
+A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, turning them from planets appearing inflated into much smaller \mbox{(super-)Earths}.
+Further dilution comes from the population of planets that lack an atmosphere or sufficient volatiles that can create a steam atmosphere.
+There is also a time dependence, both because stellar luminosities change with age and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.
+Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, or surface gravity~\citep{Pierrehumbert2022}.
+
+All of the above mechanisms dilute the demographic imprint of the runaway greenhouse transition with contributions that are challenging to quantify.
+The ``dilution factor" we introduced embodies our ignorance of these contributions.
+Importantly, while this may offset our estimates of the signal's amplitude, its shape remains the same.
+
+Our finding that the expected trend is strong except for extremely low water mass fractions not only gives reason for optimism regarding its detection, it also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
+Should a detection be possible, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
+
+\end{note}
 
@@
 
+    \todo[inline]{transit depth difference is larger in M dwarfs \citep{Turbet2019}}
+
     An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of \SI{100}{\mega\year}~CITE.

### 2023-02-17 src/tex/ms.tex
'dilution factor' instead of cumbersome formula

@@
 Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to category~A on a timescale of $10^8\SI{}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
-To account for these effects, we introduce a free parameter $f_\mathrm{rgh}$ representing the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
+To account for these effects we introduce a dilution parameter $f_\mathrm{rgh}$, which represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
 Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow relation~A, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow relation~C.
@@
 The only free parameter of the null hypothesis, which assumes the average radius to be independent of instellation, is the predicted mean radius $\langle R_\mathrm{P}\rangle $.
-The functional form of the runaway greenhouse hypothesis is more complex: Besides the mean radius of planets outside the threshold $\langle R_\mathrm{P}\rangle_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$, the planetary water mass fraction~$x_{H_2O}$ encoding the amount of radius change, and the fraction of planets above the threshold instellation that are in a runaway greenhouse climate $f_\mathrm{rgh}$.
+The functional form of the runaway greenhouse hypothesis is more complex: Besides the mean radius of planets outside the threshold $\langle R_\mathrm{P}\rangle_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$, the planetary water mass fraction~$x_{H_2O}$, and the dilution factor $f_\mathrm{rgh}$.
 
@@
 The threshold instellation can be accurately constrained with a median $\hat{S}_\mathrm{thresh} = \var{retrSthresh}$.
-Both a higher water mass fraction and a higher fraction of planets in a runaway greenhouse state lead to larger average radii, thus these parameters are strongly correlated and cannot be well constrained without independent measurements.
+Both a higher water mass fraction and a higher dilution factor lead to larger average radii, thus these parameters are strongly correlated and cannot be well constrained without independent measurements.
 
@@
             Statistical power of the runaway greenhouse hypothesis test as a function of model parameters.
-            For a sample size $N = \var{N_optimistic}$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and fraction of greenhouse climate planets.
+            For a sample size $N = \var{N_optimistic}$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and dilution factor.
             Higher values in either parameter result in a more reliable detection.
@@
             The median values of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
-            \textit{Left:} For a large planet yield of $N = \var{N_plato}$, even small planet fractions $\sim \var{platoMinfrgh}$ allow a detection.
+            \textit{Left:} For a large planet yield of $N = \var{N_plato}$, even small dilution factors $\sim \var{platoMinfrgh}$ allow a detection.
                 A sample of 100 planets is sufficient if their masses are constrained to within \SI{10}{\percent} (dotted green line).
@@
 As can be seen in Fig.~\ref{fig:plato_fgrid}, the detectability is enhanced when radius constraints can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
-Consequently, we achieve a statistically significant detection with smaller samples or lower fractions $f_\mathrm{rgh}$.
+Consequently, we achieve a statistically significant detection with smaller samples or lower dilution factors $f_\mathrm{rgh}$.
 A density-based survey of 100 targets is roughly equivalent to a radius-based survey of \var{N_plato} targets.
@@
     \begin{centering}
-
         \includegraphics[width=\linewidth]{figures/S_thresh_posteriors.pdf}

### 2023-02-17 src/tex/ms.tex
adapt figures to dilution factor, apply some eye candy

@@
             \caption{
-            Expected delta-evidences as a function of the fraction of planets harboring runaway greenhouse climates for different versions of the \plato\ survey.
+            Expected delta-evidences as a function of the fraction of planets having runaway greenhouse climates for different versions of the \plato\ survey.
             The median values of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
@@
             Retrieved posterior distributions of the threshold instellation for different survey realizations.
-            All cases assume $x_{H_2O}$ and a planet sample size $N=100\pm10$; The fraction of planets with runaway greenhouse climates varies across rows.
+            All cases assume $x_{H_2O} = \wrr$ and a planet sample size $N=100\pm10$; The fraction of planets with runaway greenhouse climates varies across rows.
             Orange lines show the true value of the injected signal.
             Accuracy and precision of the constraint on $S_\mathrm{thresh}$ generally improve with higher $f_\mathrm{rgh}$.
-            Bulk density-based inferences improve the constraints.
+            Bulk density-based inferences improve the constraints, and M~dwarf samples yield the highest accuracy.
         }

### 2023-02-22 src/tex/ms.tex
add a zoom-out sentence

@@
 \todo[inline]{\textbf{Potential discussion points: }\\
-How close is Earth to runaway? Could anthropogenic climate change push us over the edge? (-> \citet{Goldblatt2012}; what is the max. expected change in Earth's albedo due to deglaciation?}
+(How close is Earth to runaway? Could anthropogenic climate change push us over the edge? (-> \citet{Goldblatt2012,Popp2016}; what is the max. expected change in Earth's albedo due to deglaciation?)}
 
@@
 
+The habitable zone is widely employed in target prioritization for exoplanet missions, and it will provide context for interpreting potential signatures of life.
+As we have demonstrated, it appears realistic that an empirical confirmation of the habitable zone concept is imminent.
+This will be a key contribution to the search for extraterrestrial life in the Universe.
+
 %\begin{acknowledgments}
-%The authors thank Yann Fabel, Paul Molliere, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
+%The authors thank Yann Fabel, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
 %We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
-%\end{acknowledgements}
+%\end{acknowledgments}
 

### 2023-02-22 src/tex/ms.tex
edit conclusions

@@
 \section{Conclusions}
-% Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.
-Runaway greenhouse climates on Earth-like planets are thought to cause significant inflation of their radii.
+
+Runaway greenhouse climates on rocky planets cause significant inflation of their radii.
 Using Bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius discontinuity resulting from this inflation in the exoplanet population.
@@
     \item The demographic imprint of the runaway greenhouse threshold should be detectable with high-precision transit measurements.
-For a large sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of planets and if typical water mass fractions are above $\sim 10^{-3}$.
+For a large sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.
     \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to detect the predicted trend.
@@
     \item The diagnostic power of transit missions in testing this hypothesis can be increased through a follow-up campaign providing planet mass measurements.
-This can reduce the required planet yield by up to a factor of five.
+This can reduce the required planet yield by about a factor of three.
 An optimized survey will further focus on M~dwarfs to ensure sufficient targets on both sides of the expected threshold instellation.

### 2023-02-22 src/tex/ms.tex
draft discussion on the detectability of the rgh transition

@@
 \section{Methods}\label{sec:methods}
-\todo[inline]{Caveats: see \citet{Turbet2019} Sect. 3, \citet{Bower2019}}
+\todo[inline]{Caveats: see \citet{Turbet2019} Sect. 3, \citet{Bower2019}. Mention we only consider instellation-induced rgh.}
 %The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
@@
 
-\subsection{Testability of the runaway greenhouse hypothesis}
+\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
 \begin{figure}[ht!]
@@
 Our finding that the expected trend is strong except for extremely low water mass fractions not only gives reason for optimism regarding its detection, it also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
-Should a detection be possible, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
+Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
 
@@
 
-\subsection{Detectability of the runaway greenhouse threshold}
+\subsection{Detectability of the runaway greenhouse transition}
 \todo[inline]{the demographic imprint is detectable. Under what conditions?\\
 Mention multiplanet systems.}
+We showed in Sect.~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
+However, even assuming capable instrumentation and an optimal survey design, geophysical and demographic unknowns may jeopardize this undertaking.
+The detectability of the runaway greenhouse threshold is thus a function of both survey design and astrophysical circumstances.
 
@@
 \todo[]{double-check which survey params are needed}
-We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a given transit survey:
+We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey:
 \begin{itemize}
+    \item Occurrence rate of planets forming steam atmospheres: The runaway greenhouse climate relies on the production of large amounts of water vapor that then acts as a greenhouse gas.
+This process works only on planets with an atmosphere and of sufficiently volatile-rich composition\todo[inline]{how much?}.
+The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
+    \item Planetary evolution and duration of the steam atmosphere phase: An inflated steam atmosphere can only be sustained until the planet has lost its water.
+Depending on host star spectral type and planetary composition, planets spend from a few Myr to hundreds of Myr in runaway greenhouse climates~\citep[e.g.,][]{Luger2015}, and only planets observed during this phase will contribute to the signal.
     \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction.
 As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern.
-    \item Sample size: Since we are looking for a population-level trend, the statistical significance will increase with a larger sample size.
+    \item Sample size: Since we are looking for a population-level trend, the statistical significance increases with a larger sample size.
     \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be.
@@
 \subsubsection{False positive scenarios}
-\todo{what other processes could be confused with a magma ocean signal?}
+\todo{what other processes could be confused with a magma ocean signal?\\
+If there's no other subsubsection here, make this a (the final?) subsection in ``Discussion''.}
 \begin{note}
@@
 As a consequence, instrumentation and survey strategy aiming at testing the hypothesis should aim at maximizing the probability of a true positive detection given the existence of the effect.
+...
+
+The required sample size and precision are well in reach of near-future exoplanet missions such as PLATO.
 ...
@@
 \begin{large}\textit{Author contributions:}\end{large}
+%M.S., D.A., and T.L. conceived the project.
+%M.S. wrote the manuscript and carried out the hypothesis tests and statistical analyses.
+%...
+%K.H.-U. provided the star generator based on the Gaia catalog.
+%G.B. implemented the planet generator in the Bioverse framework.
+%All authors provided comments and suggestions on the manuscript.
 

### 2023-02-23 src/tex/ms.tex
defend stellar age distribution

@@
 
-\subsubsection{Stellar luminosity tracks}
+\subsubsection{Stellar luminosity evolution}
 Planetary systems are hosted by stars of a wide range of ages, and stellar luminosities evolve with time.
 Since the emergence of a runaway greenhouse phase on a planet is highly dependent on the level of radiation it receives, and thus on the luminosity of the host star, we assigned age-dependent luminosities to our synthetic stars.
-    Stellar ages are notoriously poor-constrained (CITE); we thus drew random ages from a uniform distribution from \SI{0}{\giga\year} to \SI{10}{\giga\year}.
+
+While stellar ages are notoriously poor-constrained~\cite[e.g.,][]{Adams2005}, the age distribution of planet host stars in the Solar neighborhood was shown to be broadly consistent with uniform~\citep{Reid2007,Gaidos2023}.
+For our synthetic stars, we thus drew random ages from a uniform distribution from \SI{0}{\giga\year} to \SI{10}{\giga\year}.
 We then assigned each star a luminosity from the mass-dependent evolutionary models in \citet{Baraffe1998}.

### 2023-02-23 src/tex/ms.tex
Update M-R and radius evo figures for new cases

@@
         \caption{
-            \textit{Top}: Mass-radius relationships for a water mass fraction $x_{H_2O}= \wrr$ and different planet states: A) planets with a solid mantle and no steam atmosphere.
-B) planets with steam atmospheres.
-C) as B), but including the effect of water incorporation in the melt.
+            \textit{Top}: Mass-radius relationships for a water mass fraction $x_{H_2O}= \wrr$ and different planet states.
+Green: planets with a solid mantle and no steam atmosphere.
+Dashed: planets with steam atmospheres.
+Blue: planets with steam atmospheres and including the effect of water incorporation in the melt.
             \textit{Bottom}: Radius evolution of different planet types, illustrating degeneracies among planet classes.

### 2023-02-23 src/tex/ms.tex
streamline planet categories

@@
 Relevant properties are a planet's mass $M$, its net instellation $S$, and its bulk water inventory expressed as a water mass fraction $x_{H_2O}$.
-    We consider three different cases:
-    \begin{itemize}
-   \item[A.] No radius change.
-    This case serves as our null hypothesis, where planetary radii are not affected by the amount of irradiation the planets receive.
-    A solid planetary surface is assumed and no steam atmosphere is present.
-   \item[B.] Radius inflation due to a steam atmosphere.
-    Atmospheric scale heights are significantly increased.
-   \item[C.] As B, but including radius decrease of inflated planets due to water partitioning in the molten mantle.
-    \end{itemize}
+We consider the following cases (see Fig.~\ref{fig:radiusevolution}):
+%\begin{itemize}
+%\item[A.] No radius change.
+%This case serves as our null hypothesis, where planetary radii are not affected by the amount of irradiation the planets receive.
+%A solid planetary surface is assumed and no steam atmosphere is present.
+%\item[B.] Radius inflation due to a steam atmosphere.
+%Atmospheric scale heights are significantly increased.
+%\item[C.] As B, but including radius decrease of inflated planets due to water partitioning in the molten mantle.
+%\end{itemize}
 
-In case~A, a planet simply retains the radius it was assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
-There is no additional dependency on orbital distance or instellation.
+\textit{Non-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
+This case serves as our null hypothesis.
 
-Case~B assumes an inflated radius due to a steam atmosphere for all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$.
+For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$, we assume an inflated radius due to a steam atmosphere.
 While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Leconte2013}.
 To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
-For the rocky interiors, their calculations rely on the same mass-radius relations from ~\citep{Zeng2016} that we applied for case~A-planets.
+Their calculations rely on the same mass-radius relations from ~\citep{Zeng2016} for rocky interiors that we apply for our non-runaway planets.
 For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
-We note that an additional radius change due to potentially molten interiors~\citep{Bower2019} is not taken into account here.
+This scenario does not take into account the expected additional radius change due to interiors that are molten due to greenhouse forcing~\citep{Bower2019}.
+It further assumes a \textit{dry melt} without dissolved volatiles.
 
-Case~C contains case~B, but also includes a radius decrease due to water partitioning in the melt.
-Due to greenhouse forcing, planetary surfaces below steam atmospheres are expected to be molten.
-Part of the water will then partition into the melt, removing it from the steam atmosphere, which in turn decreases the radius inflation.
-This effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes typically reach measurable values~\citep{Dorn2021}.
+\textit{Wet melt} planets follow the rules above, but we take into account a radius decrease from water partitioning in the melt:
+In the situation of a magma ocean below a steam atmosphere, part of the water will partition into the melt.
+This removes it from the steam atmosphere, which in turn decreases the radius inflation.
+The effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes typically reach measurable values~\citep{Dorn2021}.
 The magnitude of this effect is again dependent on the planet mass and on the water mass fraction.
 We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
-We then add the (in almost all cases negative) radius deviations to the planet radii computed for case~B.
+We then add the (in almost all cases negative) radius deviations to the planet radii computed for the dry melt case.
 
@@
 Blue: planets with steam atmospheres and including the effect of water incorporation in the melt.
-            \textit{Bottom}: Radius evolution of different planet types, illustrating degeneracies among planet classes.
+            \textit{Bottom}: Radius evolution of different planet types, illustrating degeneracies and potential for confusion among planet classes.
 Shown is a schematic time evolution of the transit radius normalized to the atmosphere-free radius $R_\mathrm{0}$ for different scenarios.
@@
 Figure~\ref{fig:radiusevolution} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of $x_{H_2O}= \wrr$.
-Steam atmospheres (case~B) cause a significant radius increase, which is slightly reduced when water incorporation in the melt is allowed (case~C).
-In the following, we only distinguish between cases~A and~C.
+Steam atmospheres cause a significant radius increase, which is slightly reduced when water incorporation in the melt is considered.
+In the following, we only distinguish between the non-runaway greenhouse and wet melt scenarios.
 
 As illustrated in Fig.~\ref{fig:radiusevolution}~(bottom panel), different processes can lead to the classification of a planet into either of these categories.
-For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a case~C planet, diluting the expected demographic signature.
-Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to category~A on a timescale of $10^8\SI{}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
-To account for these effects we introduce a dilution parameter $f_\mathrm{rgh}$, which represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
-Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow relation~A, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow relation~C.
+For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature.
+Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the non-runaway greenhouse category on a timescale of $10^8\SI{}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
+To account for these effects, we introduce a dilution parameter $f_\mathrm{rgh}$, which represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
+Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the non-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
 %The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.

### 2023-02-24 src/tex/ms.tex
add some citations and numbers

@@
 2013, 2015)'' \citep{Lichtenberg2022} }
-    Under the right conditions, different orbital distances alone can thus decide about a planet's water content and ultimately its habitability.
+    Consequently, it is aptly considered to be the inner boundary of the habitable zone~\citep{Kasting1993,Kopparapu2013}.
+
+    (Under the right conditions, different orbital distances alone can thus decide about a planet's water content and ultimately its habitability.)
     If received flux makes the difference between our own habitable planet and dry, dead Venus, there is good reason to believe that such patterns exist in exoplanet systems as well.
@@
 Next, we assigned to the stellar sample planetary systems with frequencies, orbital parameters, and bulk properties derived from the \kepler\ mission.
-We adopted the planetary occurrence rates in orbital period and radius recently derived in~\citep{Bergsten2022}.
-Following \citep{Youdin2011a}, their inferred occurrence rate density can be expressed in the form
+We adopted the planetary occurrence rates in orbital period and radius recently derived in \citet{Bergsten2022}.
+Following \citet{Youdin2011a}, their inferred occurrence rate density can be expressed in the form
 \begin{equation}
@@
 }
+Extreme climates have a direct influence on planetary sizes as measured by transit photometry~\citep{Turbet2019}.
+Here, we use predictions on atmospheric thickness from geophysical models to derive the transit radii of planets with temperate climates and those with instellation-induced runaway greenhouse climates, respectively.
 ...
@@
 For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$, we assume an inflated radius due to a steam atmosphere.
-While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Leconte2013}.
+While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013}.
 To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
-Their calculations rely on the same mass-radius relations from ~\citep{Zeng2016} for rocky interiors that we apply for our non-runaway planets.
+Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our non-runaway planets~\citep{Zeng2016}.
 For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
@@
     \item Occurrence rate of planets forming steam atmospheres: The runaway greenhouse climate relies on the production of large amounts of water vapor that then acts as a greenhouse gas.
-This process works only on planets with an atmosphere and of sufficiently volatile-rich composition\todo[inline]{how much?}.
+This process works only on planets with an atmosphere and of sufficiently volatile-rich composition\todo[inline]{TIM: how much?}.
 The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
     \item Planetary evolution and duration of the steam atmosphere phase: An inflated steam atmosphere can only be sustained until the planet has lost its water.
-Depending on host star spectral type and planetary composition, planets spend from a few Myr to hundreds of Myr in runaway greenhouse climates~\citep[e.g.,][]{Luger2015}, and only planets observed during this phase will contribute to the signal.
+Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the signal.
     \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction.

### 2023-02-27 src/tex/ms.tex
draft discussion of missions diagnostic power

@@
     \end{equation}
-    Here, we assume global redistribution of incoming flux and a fixed albedo of $0.3$, comparable to Earth's [CITE].
+    Here, we assume global redistribution of incoming flux and a fixed albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
     The absolute value of $S_\oplus$, by which all synthetic planets are scaled, is then \SI{238}{\watt\per\square\meter}.
@@
 \end{figure}
-Figure~\ref{fig:optimistic_R-S} shows a prototypical detection of the runaway greenhouse threshold.
+Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the runaway greenhouse transition.
 For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
@@
 
-%We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ...
-\todo[]{double-check which survey params are needed}
+\subsubsection{Key factors influencing tests of the runaway greenhouse hypothesis}
 We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey:
@@
 
-\subsection{Diagnostic power of upcoming exoplanet missions}\todo[inline]{\plato: needed sample size, precision}
-The fidelity of a future detection or falsification of a magma ocean signature will depend on the significance with which the null hypothesis can be excluded.
-As a consequence, instrumentation and survey strategy aiming at testing the hypothesis should aim at maximizing the probability of a true positive detection given the existence of the effect.
-...
+\subsection{Diagnostic power of near-future exoplanet missions}\label{sec:dis_samplesize}
+The fidelity of a future detection or falsification of a runaway greenhouse signature will depend on the significance with which the null hypothesis can be excluded.
+As a consequence, instrumentation and survey strategy aiming at testing the hypothesis should aim at maximizing the probability of a true positive detection.
+\begin{note}
+In Sect.~\ref{sec:res_samplesize} we explore the diagnostic power of space missions that are currently flying or in the planning stage.
+We focus here on transit survey missions, as they provide the most direct radius measurements of a practical planet sample.
 
-The required sample size and precision are well in reach of near-future exoplanet missions such as PLATO.
-...
+As discussed above, the key drivers from a mission design perspective are sample size and photometric precision.
+We found that along these axes, \plato\ will be the most favorable among the upcoming transit missions.
+%The required sample size and precision are well in reach of near-future exoplanet missions such as PLATO.
+With an expected transit radius precision of \SI{3}{\percent}~\citep{plato2017} and a yield of $> 100$ planets characterized with such high precision~\citep{Rauer2021}, \plato\ is comparable to the optimistic survey we introduced in Sect.~\ref{sec:res_testability}.
+If successful, it should readily detect the predicted statistical imprint or, in case of a non-detection, provide strong upper limits on the occurrence rate of runaway greenhouse planets.
+
+The latter depends on the lifetimes of runaway greenhouse phases, which are a function of the initial water inventory of the planets~\citep{Hamano2015}.
+It seems thus feasible to derive the typical water content of rocky planets from these occurrence estimates.
+\end{note}
 
@@
 %\begin{acknowledgments}
-%The authors thank Yann Fabel, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
+%The authors thank Gabriele Cugno, Yann Fabel, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
 %We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.

### 2023-03-01 src/tex/ms.tex
draft discussion on follow-up campaigns

@@
 In the FGK sample of the mission, several hundred Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
-\plato's definition study report~\citep{plato2017} further states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).
+\plato's definition study report~\citep{plato2017} further states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity~(RV) follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).
 We base our assessment of \plato's diagnostic power on these estimates.
@@
 
-\subsubsection{Additional planet mass measurements}
+\subsubsection{Additional planet mass measurements}\label{sec:res_followup}
 The dependencies of a detection change when additional information beyond planet radii is available for the characterized planet population.
@@
 \subsection{Mission design trades}\label{sec:mission-design-trades}
+\todo[inline]{not sure if subsubsections needed}
 We now turn to mission design trades that influence the statistical power for testing the runaway greenhouse hypothesis.
-...
+
 \subsubsection{The value of follow-up campaigns}
 \begin{note}
-...FOLLOW-UP...radial velocity follow-up measurements...
-
-The advantage of mass measurements for the planet sample can be enhanced with a wise selection of follow-up targets, i.e. favoring those planets residing close to $S_\mathrm{thresh}$.
+As runaway greenhouse phases leave a stronger imprint in bulk density than in planet radius (see Sect.~\ref{sec:res_followup}), it would be beneficial to obtain constraints on planetary masses and test the runaway greenhouse hypothesis in density space instead of radius space.
+Precise ground-based radial velocity measurements are expensive, and therefore such follow-up efforts may only be available for a subsample of a mission's targets.
+For a mission design similar to PLATO, a density-based test on about a third of the overall sample is equivalent to a pure radius-based analysis.
+The diagnostic power of such a test may be improved by simultaneously fitting for the trend in radius space in the subsample without RV follow-up.
+An optimized mission in search for the inner edge of the habitable zone will further enhance its information content via a wise selection of follow-up targets, i.e., favoring those planets residing close to the expected instellation threshold $S_\mathrm{thresh}$.
 \end{note}

### 2023-03-04 src/tex/ms.tex
draft discussion on FGK vs. M

@@
 \todo[inline]{\textbf{Potential discussion points: }\\
-(How close is Earth to runaway? Could anthropogenic climate change push us over the edge? (-> \citet{Goldblatt2012,Popp2016}; what is the max. expected change in Earth's albedo due to deglaciation?)}
+(How close is Earth to runaway? (-> \citet{Goldblatt2012,Popp2016}; what is the max. expected change in Earth's albedo due to deglaciation?)}
 
 \subsection{Statistical imprint of exoplanet climates}
-\todo[inline]{what is the fraction of Kepler planets w/ run-
-away GH?}
 \begin{note}
@@
 \subsubsection{Target star spectral types}
-\begin{note}
-    ...FGK vs. M...:
+To date, the majority of planets with radius measurements orbit FGK~dwarfs, and, based on the instellation they receive, most of them lie in the runaway greenhouse regime~\citep{Thompson2018}.
+Obviously, a step-function-like signal in the exoplanet demographics like the one illustrated above can not be constrained well if only one side of the discontinuity is being sampled.
+This, however, is the situation for planetary systems around Sun-like stars -- their habitable zones are so distant that transiting planets within them are very rare due to pure geometrical reasons.
+It was, among other reasons, the sharp drop in transit probability with orbital distance that has prompted a number of recent transit surveys to specifically target M~dwarfs~\citep[e.g.,][]{Irwin2009,Obermeier2016,Delrez2018,Sebastian2021,Dietrich2023}, but the sample of terrestrial planets orbiting them is still small~\citep[e.g.,][]{Berger2020,Hardegree-Ullman2020a}.
 
-To date, the majority of planets with radius measurements orbit FGK~dwarfs and lie in the runaway greenhouse regime.
-    ...
-
-    Our tests with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
-The reason is, besides their large number in a volume/magnitude-limited sample, that many detected M~dwarf planets are located near the threshold instellation.
-As in the cases above, an M~dwarf survey profits from follow-up measurements of planetary masses with an order of magnitude increase in evidence.
-
-    \todo[inline]{transit depth difference is larger in M dwarfs \citep{Turbet2019}}
-
-    An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of \SI{100}{\mega\year}~CITE.
+M~dwarf systems are also key for detecting the runaway greenhouse transition:
+Our calculations with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
+Besides their large number in a volume/magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular beyond it, i.e., in the optimistic habitable zone.
+The transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars][]{Turbet2019}, enhancing the demographic signal it leaves.
+An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of \SI{100}{\mega\year}~\citep{Luger2015}.
 This increases the probability of observing any given planet in the sample during the runaway greenhouse phase, essentially driving $f_\mathrm{rgh}$ to higher values.
-\end{note}
+Therefore, in addition to the high scientific value of boosted detections of potentially habitable planets, M dwarfs are also indispensable for the discovery and characterization of the runaway greenhouse transition.
+As with a pure volume-limited sample, a targeted M~dwarf survey, too, profits from follow-up measurements of planetary masses with an order of magnitude increase in evidence.
 

### 2023-03-04 src/tex/ms.tex
mention value added by Roman's transiters

@@
 
-%    We argue that in the near future, ESA's \plato\ mission and NASA's Roman Space Telescope will be the most promising endeavors to constrain this demographic feature.
-%    For each of these missions, we identify the key mission design drivers that enable a statistically sound detection.
-%    We also show the unique synergy of these missions in answering this question, and what survey strategy optimizes the statistical power of the combined dataset.
+%    We argue that in the near future, ESA's \plato\ mission will be the most promising endeavors to constrain this demographic feature.
+%    We identify the key mission design drivers that enable a statistically sound detection.
+%    We also show what survey strategy optimizes the statistical power of the combined dataset.
 %    $\ldots$ its measurement will also provide insights into which stars harbor the nearest habitable worlds.
@@
 For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$, we assume an inflated radius due to a steam atmosphere.
-While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013}.
+While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013,Boukrouche2021}.
 To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
@@
 
+While not primarily designed to detect transiting planets, the \rst~\citep{Spergel2015} is expected to yield $\sim 10^5$ transiting planets on short orbits in the course of its mission~\citep{Montet2017}.
+In a fraction of the cases, planets as small as \SI{2}{\Rearth} will be detected.
+About 200 planets smaller than Neptune are expected to be found around M~dwarfs and thus may reach into the habitable zone.
+The \rst\ could thus provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.
+
 \subsection{Statistical power of different mission designs}\label{sec:statpower_missions}

### 2023-03-06 src/tex/ms.tex
turn comments into actionable ToDos

@@
 \todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
+\todo[inline]{``Points for the intro:
+Bioverse\\
+Transit survey to measure mass/radius of planets\\
+Does the HZ have an inner edge?\\
+Run-away GH\\
+Magma oceans\\
+Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman.
+Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
+~\\
+In contrast, PLATO and Nautilus are \\
+Possibility of follow-up to measureme masses
+''}
 
@@
 \section{Methods}\label{sec:methods}
-\todo[inline]{Caveats: see \citet{Turbet2019} Sect. 3, \citet{Bower2019}. Mention we only consider instellation-induced rgh.}
 %The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
-
+\todo[inline]{Start methods description with brief statement of goal and overview of the steps. The first sentence is similar, but would be good to separate goal from method here.
+}
 We determined -- for different survey configurations -- the confidence level with which the runaway greenhouse threshold can be detected in simulated exoplanet observations.
@@
 Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic stellar and planetary samples into which we injected the runaway greenhouse signal, simulated observations of this sample, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).
+\todo[inline]{Bioverse will need a short introduction before the modifications are mentioned. Probably should get a paragraph in the intro}
 \begin{figure*}
@@
 \subsubsection{Planet radius changes due to runaway greenhouse climates}\label{sec:mo_model}
-\todo[inline]{
-\textbf{Main assumptions:}\\
-- every planet (modulo $f_\mathrm{rgh}$) has \textit{some} water. Enough for runaway GH under the right conditions.  \\
-- Limitation to small planets without significant atmospheric envelopes.\\
-- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
-- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
-}
-Extreme climates have a direct influence on planetary sizes as measured by transit photometry~\citep{Turbet2019}.
-Here, we use predictions on atmospheric thickness from geophysical models to derive the transit radii of planets with temperate climates and those with instellation-induced runaway greenhouse climates, respectively.
-...
+%\todo[inline]{
+%\textbf{Main assumptions:}\\
+%- every planet (modulo $f_\mathrm{rgh}$) has \textit{some} water. Enough for runaway GH under the right conditions.  \\
+%- Limitation to small planets without significant atmospheric envelopes.\\
+%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
+%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
+%}
+%Extreme climates have a direct influence on planetary sizes as measured by transit photometry~\citep{Turbet2019}.
+%Here, we use predictions on atmospheric thickness from geophysical models to derive the transit radii of planets with temperate climates and those with instellation-induced runaway greenhouse climates, respectively.
+%...
 
@@
 \subsection{Detecting the runaway greenhouse transition with current and planned exoplanet missions}\label{sec:res_samplesize}
-\todo{Introduce survey simulation(s).
+%\todo{Introduce survey simulation(s).
 PLATO, Cheops, (Ariel, LIFE/LIFE precursor, Nautilus)\\ What are threshold missions that are able to detect this signal?
@@
 (How close is Earth to runaway? (-> \citet{Goldblatt2012,Popp2016}; what is the max. expected change in Earth's albedo due to deglaciation?)}
+\todo[inline]{Caveats: see \citet{Turbet2019} Sect. 3, \citet{Bower2019}. Mention we only consider instellation-induced rgh.\\
+- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
+- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
+- assumption of a single albedo. ``if there is a range of albedos, will that wash away the signal?''
+- Impact of assumptions on findings, e.g.
+M-R relation.
+``We should have a discussion section dedicated to the impact of assumptions on the findings - including mass-radius.
+A good way to go about this is to present bracketing cases; or to discuss why the discountinuity in the results would not be affected by changes that led to continuous offsets in the results.
+Or both.
+But it should not be in the methods section.''}
 
@@
 We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey:
+\todo[inline]{avoid bullet points within sub-subsections}
 \begin{itemize}
@@
 \todo[inline]{discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (Tim's talk)}
+\todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
 

### 2023-03-06 src/tex/ms.tex
move M-R to new discussion section w/ impact of assumptions

@@
 % Define document class
-\documentclass[twocolumn]{aastex631}
-%\documentclass[modern]{aastex631}
+%\documentclass[twocolumn]{aastex631}
+\documentclass[modern]{aastex631}
 
@@
 
-To assign masses to our planets, we use the semi-empirical mass-radius relationship derived in \citet{Zeng2016}.
-The default relation our planets follow is one assuming a pure $\mathrm{MgSiO_3}$ composition.
-While interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure and detectability.
-\todo[inline]{run a test: do we expect that the choice of M-R relation impacts the results? Perhaps have an appendix section.}
+To assign masses to our planets, we use the semi-empirical mass-radius relationship assuming a pure $\mathrm{MgSiO_3}$ composition from \citet{Zeng2016}.
 This represents the baseline bulk density before any climate-related effects are applied.
@@
 
+\subsection{Impact of assumptions on our findings}
+\subsubsection{Baseline mass-radius relationship}
+Our baseline mass-radius relationship assuming pure $\mathrm{MgSiO_3}$ interiors~\citep{Zeng2016} might not be representative of the rocky planet population.
+However, while interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure.
+The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure.\todo[inline]{TIM: do you have a reference for this statement?}
+We thus consider our mass-radius relationship a conservative case and do not expect a negative impact on the detectability of the runaway greenhouse transition due to deviations from it.
+\todo[inline]{if feasible: run a test with a different M-R relation. do we expect that the choice of M-R relation impacts the results? Perhaps have an appendix section.}
+
 \section{Conclusions}

### 2023-03-06 src/tex/ms.tex
discuss assumptions regarding water mass fraction

@@
         \caption{
-            \textit{Top}: Mass-radius relationships for a water mass fraction $x_{H_2O}= \wrr$ and different planet states.
+            \textit{Top}: Mass-radius relationships for a water mass fraction $x_{H_2O}= \var{wrr}$ and different planet states.
 Green: planets with a solid mantle and no steam atmosphere.
@@
 \end{figure}
-Figure~\ref{fig:radiusevolution} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of $x_{H_2O}= \wrr$.
+Figure~\ref{fig:radiusevolution} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of $x_{H_2O}= \var{wrr}$.
 Steam atmospheres cause a significant radius increase, which is slightly reduced when water incorporation in the melt is considered.
@@
 \subsection{Statistical signature of the runaway greenhouse threshold}
-To characterize the population-level imprint of individual radius changes, we generated a generic planet population with an injected runaway greenhouse effect assuming a water fraction $x_\mathrm{H_2O} = \wrr$.
+To characterize the population-level imprint of individual radius changes, we generated a generic planet population with an injected runaway greenhouse effect assuming a water fraction $x_\mathrm{H_2O} = \var{wrr}$.
 \begin{figure}
@@
 For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
-We assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \wrr\ for each planet.
+We assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
 In this case, the pattern was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).
@@
             Retrieved posterior distributions of the threshold instellation for different survey realizations.
-            All cases assume $x_{H_2O} = \wrr$ and a planet sample size $N=100\pm10$; The fraction of planets with runaway greenhouse climates varies across rows.
+            All cases assume $x_{H_2O} = \var{wrr}$ and a planet sample size $N=100\pm10$; The fraction of planets with runaway greenhouse climates varies across rows.
             Orange lines show the true value of the injected signal.
@@
 
+\subsubsection{Water mass fraction}
+The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
+Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent weight fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021}.
+Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
+This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the statistical radius inflation trend, also since the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
+Cases of pure rocky composition and very low volatile contants can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
+Assuming a distribution of water mass fractions instead of a fixed value would thus not significantly change our results.
+
 \section{Conclusions}

### 2023-03-06 src/tex/ms.tex
edit false positives

@@
 \begin{note}
-    Global magma oceans are not the only physical mechanism that may cause a decrease in transit radii for a subset of planets.
-    Other causes of shrinking planet sizes have been put forward, the most widely discussed ones being atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
-    Both processes are being traded as potentially sculpting the observed radius bimodality of small, close-in exoplanets from the  mission~\citep{Fulton2017,VanEylen2018}.
-    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
-    This region is distinct from the one affected by global magma oceans, since... \todo{...elaborate why they cannot be confused}
+    Runaway greenhouse climates are not the only physical mechanism that may cause a change in transit radius for a subset of planets.
+    Some of the most prominent alternative scenarios involve atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
+    Both processes are being traded as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
+%    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
+    Either of them leads to a decrease of planet radius for planets close to their host star, which is distinct from the radius inflation introduced by runaway greenhouse climates.
 

### 2023-03-06 src/tex/ms.tex
move wrr/f_rgh degeneracy to discussion

@@
 %The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
-Within the runaway greenhouse limit, an average radius change of \avgRadiusChange\ occurs.
+Within the runaway greenhouse regime, an average radius change of \avgRadiusChange\ occurs.
 
@@
 Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the runaway greenhouse transition.
-For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
+For the sake of clarity, we chose an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
 We assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
@@
 True values of the parameters for the injected effect are shown in orange.
-The threshold instellation can be well constrained; the predominant water fraction and the fraction of planets with runaway greenhouse climates are degenerate.
+The threshold instellation can be reasonably constrained; the predominant water fraction and the fraction of planets with runaway greenhouse climates are degenerate.
         }
@@
 The threshold instellation can be accurately constrained with a median $\hat{S}_\mathrm{thresh} = \var{retrSthresh}$.
-Both a higher water mass fraction and a higher dilution factor lead to larger average radii, thus these parameters are strongly correlated and cannot be well constrained without independent measurements.
+Both a higher water mass fraction and a higher dilution factor lead to larger average radii, thus these parameters are strongly correlated.
 
-Given an optimistic survey design, what influences the probability of correctly rejecting a false null hypothesis?
 \begin{figure}[ht!]
@@
 Figure~\ref{fig:statpwr_H2O-f} explores the statistical power achieved in the above scenario for different combinations of the poorly constrained parameters $x_{H_2O}$ and $f_\mathrm{rgh}$.
-The higher the water content and thus the radius change, and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
+It is highest for large water inventories and large dilution factors.
  For all but very low water fractions, $f_\mathrm{rgh}$ dominates this trend: It enters linearly into the average planet radius, whereas the contribution of $x_{H_2O}$ - as predicted by the geophysical models - is sublinear with a power-law exponent of $\sim 0.3$.
@@
 
-\subsubsection{Constraining the threshold instellation}
+\subsubsection{Constraining the threshold instellation}\label{sec:res_constrain-S_thresh}
 \begin{figure}[ht!]
@@
 \subsubsection{Key factors influencing tests of the runaway greenhouse hypothesis}
+What influences the probability of correctly rejecting a false null hypothesis?
 We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey:
@@
 Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the signal.
-    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction.
-As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern.
+    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is sensitive on the water mass fraction.
+As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern:
+        The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
     \item Sample size: Since we are looking for a population-level trend, the statistical significance increases with a larger sample size.
@@
 \begin{note}
-In Sect.~\ref{sec:res_samplesize} we explore the diagnostic power of space missions that are currently flying or in the planning stage.
+In Sect.~\ref{sec:res_samplesize} we explore the diagnostic power of space missions that are currently operating or are in the planning stage.
 We focus here on transit survey missions, as they provide the most direct radius measurements of a practical planet sample.
@@
 
-\subsection{Implications for planetary habitability}\label{sec:habitability}
+\subsection{Constraining planetary habitability}\label{sec:habitability}
 \todo[inline]{Magma oceans influence the amount of water available at the surface and atmosphere, which is 1.
@@
 influences the planet's climate (greenhouse effect).}
+A potential for liquid water on the surface of a planet is commonly used as an environmental marker to assess its habitability. REF
+The runaway greenhouse transition represents an upper bound on received irradiation for this condition.
+Its detection would thus not only empirically confirm the habitable zone concept but also help to locate it in the observationally available planetary parameter space.
+In Sect.~\ref{sec:res_constrain-S_thresh}, we show that the threshold instellation at which the runaway greenhouse transition occurs can be reasonably constrained without imposing overly optimistic conditions on the underlying planet population, instrumentation, or survey strategy.
+\todo[inline]{elaborate based on Sect.~\ref{sec:res_constrain-S_thresh}}
+
+The situation is different for the planetary water inventory and the fraction of planets with runaway greenhouse climates:
+Since these parameters are degenerate, they cannot be well constrained without independent measurements.
+\todo[inline]{discuss additional measurements that could lift the degeneracy. e.g., adding atmospheric compositional information i.e., PLATO + ARIEL or similar}
 

### 2023-03-07 src/tex/ms.tex
integrate comments on methods section

@@
 \subsection{Synthetic star and planet sample}
-The first step in our analysis is to generate a sample of stars that host synthetic planets.
+In the following, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, as well as the orbital parameters of the planets.
 
 \subsubsection{Stellar sample from Gaia DR3}
-Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
+%Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
 \todo[inline]{KEVIN: describe stellar population}
@@
 
-While stellar ages are notoriously poor-constrained~\cite[e.g.,][]{Adams2005}, the age distribution of planet host stars in the Solar neighborhood was shown to be broadly consistent with uniform~\citep{Reid2007,Gaidos2023}.
+While stellar ages are notoriously poorly constrained~\cite[e.g.,][]{Adams2005}, the age distribution of planet host stars in the Solar neighborhood was shown to be broadly consistent with uniform~\citep{Reid2007,Gaidos2023}.
 For our synthetic stars, we thus drew random ages from a uniform distribution from \SI{0}{\giga\year} to \SI{10}{\giga\year}.
@@
 
-\subsubsection{Eccentricities, orbital inclinations, and planet masses}
+\subsubsection{Orbit parameters and planet masses}
 Eccentric orbits alter the probability of a planet to transit \citep[e.g.,][]{Barnes2007a}.
@@
 Following~\citet{Kipping2013b}, we used a Beta distribution with parameters $a=0.867$ and $b=3.03$, and truncated the distribution at $e = 0.8$.
-
-Another parameter with an even higher impact on transit probabilities is the orbital inclination toward the observer $i$.
+%Another parameter with an even higher impact on transit probabilities is the orbital inclination toward the observer $i$.
 Assuming isotropic alignments of orbits, we assigned each planet an inclination drawn from a distribution uniform in $\cos(i)$.
 
-To assign masses to our planets, we use the semi-empirical mass-radius relationship assuming a pure $\mathrm{MgSiO_3}$ composition from \citet{Zeng2016}.
+To assign masses to our planets, we use the semi-empirical mass-radius relationship assuming a pure $\mathrm{MgSiO_3}$ composition from \citet{Zeng2016} (see green line in Fig.~\ref{fig:radiusevolution}).
 This represents the baseline bulk density before any climate-related effects are applied.
@@
 \subsubsection{Transit probability}
-\begin{note}
 We model the occurrence of transits by assuming isotropic orientations of planetary orbits, calculating the impact parameters $b = \cos(i)/R_\star$ and considering only observable planets with $|b| < 1$.
@@
 Excluding all non-transiting planets diminishes the sample to \var{percentage_transiting} of its original size.
-\end{note}
 
@@
 %}
-%Extreme climates have a direct influence on planetary sizes as measured by transit photometry~\citep{Turbet2019}.
-%Here, we use predictions on atmospheric thickness from geophysical models to derive the transit radii of planets with temperate climates and those with instellation-induced runaway greenhouse climates, respectively.
-%...
+The climate state of a planet has a direct influence on its size as measured by transit photometry~\citep{Turbet2019}.
+Here, we use predictions on transit atmospheric thickness from geophysical models to derive the transit radii of planets with temperate climates and those with instellation-induced runaway greenhouse climates, respectively.
+...
 
@@
 
-While not primarily designed to detect transiting planets, the \rst~\citep{Spergel2015} is expected to yield $\sim 10^5$ transiting planets on short orbits in the course of its mission~\citep{Montet2017}.
+While not primarily designed to detect transiting planets, the \rst~\citep{Spergel2015} is expected to yield $\sim 10^5$ transiting planets on short orbits and constrain their radii in the course of its mission~\citep{Montet2017}.
 In a fraction of the cases, planets as small as \SI{2}{\Rearth} will be detected.

### 2023-03-07 src/tex/ms.tex
test additions
Add two paragraphs on stellar characterization.

@@
 
-    Here, we assess the ability of space and ground-based telescopes to test this hypothesis using Bioverse, a simulation framework that leverages contextual information from the overall planet population.
+    Here, we assess the ability of space and ground-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages contextual information from the overall planet population.
     $\ldots$
@@
 \todo[inline]{``Points for the intro:
-Bioverse\\
+\bioverse\\
 Transit survey to measure mass/radius of planets\\
@@
 We determined -- for different survey configurations -- the confidence level with which the runaway greenhouse threshold can be detected in simulated exoplanet observations.
-For this purpose we expanded the Bioverse framework~\citep{Bixel2020,Bixel2021}\footnote{Bioverse is actively maintained and documented open source software written in Python.
+For this purpose we expanded the \bioverse\ framework~\citep{Bixel2020,Bixel2021}\footnote{\bioverse\ is actively maintained and documented open source software written in Python.
 Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic stellar and planetary samples into which we injected the runaway greenhouse signal, simulated observations of this sample, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).
-\todo[inline]{Bioverse will need a short introduction before the modifications are mentioned. Probably should get a paragraph in the intro}
+\todo[inline]{\bioverse\ will need a short introduction before the modifications are mentioned. Probably should get a paragraph in the intro}
 \begin{figure*}
@@
         \includegraphics[width=\hsize]{figures/flowchart.pdf}
-        \caption{Workflow of our hypothesis testing with Bioverse. In the first block, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
+        \caption{Workflow of our hypothesis testing with \bioverse. In the first block, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
             A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:mo_model}.
@@
 %Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
-\todo[inline]{KEVIN: describe stellar population}
-...
+%\todo[inline]{KEVIN: describe stellar population}
+\todo[inline]{KEVIN: update reference to my paper once accepted}
+The original \texttt{Bioverse} stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function.
+Exquisite parallax and photometric data from the \textit{Gaia} mission made it possible for Hardegree-Ullman et al.
+(submitted) to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \texttt{Bioverse}.
+Here, we briefly describe how stellar parameters $T_{\mathrm{eff}}$, $L_{\star}$, $R_{\star}$, and $M_{\star}$ were derived for \texttt{Bioverse}, but for a full description, please refer to Hardegree-Ullman et al.
+(submitted).
+
+Hardegree-Ullman et al.
+(submitted) used the Gaia Catalogue of Nearby Stars \citep[hereafter GCNS,][]{GCNS2021} as the basis for deriving stellar parameters for the \texttt{Bioverse} catalog.
+The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands \citep{GaiaCollaboration2021} and $K_S$ band photometry from 2MASS \citep{Cutri2003}, all of which were used in stellar classification.
+From this information, colors and absolute magnitudes were computed, and initial color-magnitude cuts were made to remove non-main sequence stars.
+Effective temperatures were computed primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
+Next, luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
+Stellar radii were then computed with the effective temperatures and lumonisities using the Stefan-Boltzmann law, or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M dwarfs.
+Finally, masses were derived from either the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim0.7\,M_{\odot}$, and that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
+The derived stellar parameters from Hardegree-Ullman et al.
+(submitted) were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
 
@@
 \subsection{Survey simulations}
-The survey module of Bioverse converts the synthetic planet sample into a set of uncertainty-laden measurements on a subset of that sample.
-   This task includes selection of the targets, application of detection biases, and conducting simulated measurements.
-   All of the above are specific to the particular survey.
+The survey module of \bioverse\ converts the synthetic planet sample into a set of uncertainty-laden measurements on a subset of that sample.
+This task includes selection of the targets, application of detection biases, and conducting simulated measurements, all of which are specific to the particular survey.
+For each planet-level measurement such as radius or instellation, we draw the measured value from a normal distribution centered on the true value with a standard deviation set by the survey's precision.
 
@@
             Retrieved posterior distribution of key parameters in the optimistic scenario.
-The density maps in each panel show relationships between and marginalized distributions of the model parameters as they could be retrieved with a high-precision transit survey and a sample of \var{N_optimistic} planets.
+The density maps in each panel show relationships between and marginalized distributions of the threshold instellation $S_\mathrm{thresh}$, the water mass fraction $x_{H_2O}$, and the dilution factor $f_\mathrm{rgh}$ as they could be retrieved with a high-precision transit survey and a sample of \var{N_optimistic} planets.
 True values of the parameters for the injected effect are shown in orange.
@@
 \end{figure}
-Figure~\ref{fig:statpwr_H2O-f} explores the statistical power achieved in the above scenario for different combinations of the poorly constrained parameters $x_{H_2O}$ and $f_\mathrm{rgh}$.
+Figure~\ref{fig:statpwr_H2O-f} explores the statistical power of the hypothesis test achieved in the above scenario for different combinations of the poorly constrained parameters $x_{H_2O}$ and $f_\mathrm{rgh}$.
 It is highest for large water inventories and large dilution factors.
@@
 
-\subsection{Detecting the runaway greenhouse transition with current and planned exoplanet missions}\label{sec:res_samplesize}
+\subsection{Detecting the runaway greenhouse transition with \plato}\label{sec:res_samplesize}
 %\todo{Introduce survey simulation(s).
@@
             \caption{
-            Expected delta-evidences as a function of the fraction of planets having runaway greenhouse climates for different versions of the \plato\ survey.
+            Expected delta-evidences as a function of the fraction of planets with runaway greenhouse climates for different versions of the \plato\ survey.
             The median values of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
@@
 \subsection{Statistical power of different mission designs}\label{sec:statpower_missions}
-To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated several transit surveys with different designs and strategies and measured their capability to recover the trend and constrain its parameters.
+To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
 We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
@@
 \subsubsection{Spectral type of target stars}\label{sec:results_FGK_M}
-Since receiving radiation is sensitive on the spectral type of the host star, the share of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
+Since the receiving radiation is sensitive on the spectral type of the host star, the share of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
 We tested the detectability of the runaway greenhouse signal when only FGK or only M~dwarfs are considered (see Fig.~\ref{fig:plato_fgrid}).
@@
 If, on the other hand, a follow-up campaign adds planetary mass estimates and we fit the signal in bulk density space, useful constraints are possible under more pessimistic conditions.
-A sample containing only planets around M~dwarf yields a comparable performance as the overall sample.
+A sample containing only planets around M~dwarf yields a performance comparable to that of the overall sample.
 
@@
 It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019}.
-Using Bioverse to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we showed that the population of greenhouse-inflated planets create a signature in the radius and density distribution that is well in the detectable range.
+Using \bioverse\ to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we showed that the population of greenhouse-inflated planets create a signature in the radius and density distribution that is well in the detectable range.
 
@@
 Our calculations with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
-Besides their large number in a volume/magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular beyond it, i.e., in the optimistic habitable zone.
-The transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars][]{Turbet2019}, enhancing the demographic signal it leaves.
+Besides their large number in a volume/magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular on orbits further out, i.e., in the optimistic habitable zone.
+In fact, we showed that the FGK part of the planet sample barely contributes to the statistical power.
+Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars][]{Turbet2019}, enhancing the demographic signal it leaves.
 An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of \SI{100}{\mega\year}~\citep{Luger2015}.
@@
 This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the statistical radius inflation trend, also since the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
-Cases of pure rocky composition and very low volatile contants can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
+Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
 Assuming a distribution of water mass fractions instead of a fixed value would thus not significantly change our results.
@@
 Runaway greenhouse climates on rocky planets cause significant inflation of their radii.
-Using Bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius discontinuity resulting from this inflation in the exoplanet population.
+Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius discontinuity resulting from this inflation in the exoplanet population.
 Our key findings are as follows:
@@
 %K.H.-U. provided the star generator based on the Gaia catalog.
-%G.B. implemented the planet generator in the Bioverse framework.
+%G.B. implemented the planet generator in the \bioverse\ framework.
 %All authors provided comments and suggestions on the manuscript.
@@
 \software{
-Bioverse~\citep{Bixel2021},
+\bioverse ~\citep{Bixel2021},
 Astropy~\citep[][]{AstropyCollaboration2018},

### 2023-03-07 src/tex/ms.tex
edit and add to Results section

--- 2023-03-07 src/tex/ms.tex
@@
 %Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
-%\todo[inline]{KEVIN: describe stellar population}
-\todo[inline]{KEVIN: update reference to my paper once accepted}
-The original \texttt{Bioverse} stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function.
-Exquisite parallax and photometric data from the \textit{Gaia} mission made it possible for Hardegree-Ullman et al.
-(submitted) to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \texttt{Bioverse}.
-Here, we briefly describe how stellar parameters $T_{\mathrm{eff}}$, $L_{\star}$, $R_{\star}$, and $M_{\star}$ were derived for \texttt{Bioverse}, but for a full description, please refer to Hardegree-Ullman et al.
-(submitted).
-
-Hardegree-Ullman et al.
-(submitted) used the Gaia Catalogue of Nearby Stars \citep[hereafter GCNS,][]{GCNS2021} as the basis for deriving stellar parameters for the \texttt{Bioverse} catalog.
-The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands \citep{GaiaCollaboration2021} and $K_S$ band photometry from 2MASS \citep{Cutri2003}, all of which were used in stellar classification.
-From this information, colors and absolute magnitudes were computed, and initial color-magnitude cuts were made to remove non-main sequence stars.
-Effective temperatures were computed primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
-Next, luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
-Stellar radii were then computed with the effective temperatures and lumonisities using the Stefan-Boltzmann law, or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M dwarfs.
-Finally, masses were derived from either the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim0.7\,M_{\odot}$, and that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
-The derived stellar parameters from Hardegree-Ullman et al.
-(submitted) were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
+\todo[inline]{KEVIN: describe stellar population}
+...
 

### 2023-03-07 src/tex/ms.tex
tidy the library

@@
 %Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
-\todo[inline]{KEVIN: describe stellar population}
-...
+%\todo[inline]{KEVIN: describe stellar population}
+\todo[inline]{KEVIN: update reference to my paper once accepted}
+The original \texttt{Bioverse} stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function.
+Exquisite parallax and photometric data from the \textit{Gaia} mission made it possible for Hardegree-Ullman et al.
+(submitted) to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \texttt{Bioverse}.
+Here, we briefly describe how stellar parameters $T_{\mathrm{eff}}$, $L_{\star}$, $R_{\star}$, and $M_{\star}$ were derived for \texttt{Bioverse}, but for a full description, please refer to Hardegree-Ullman et al.
+(submitted).
+
+Hardegree-Ullman et al.
+(submitted) used the Gaia Catalogue of Nearby Stars \citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \texttt{Bioverse} catalog.
+The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands \citep{GaiaCollaboration2021} and $K_S$ band photometry from 2MASS \citep{Cutri2003}, all of which were used in stellar classification.
+From this information, colors and absolute magnitudes were computed, and initial color-magnitude cuts were made to remove non-main sequence stars.
+Effective temperatures were computed primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
+Next, luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
+Stellar radii were then computed with the effective temperatures and lumonisities using the Stefan-Boltzmann law, or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M dwarfs.
+Finally, masses were derived from either the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim0.7\,M_{\odot}$, and that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
+The derived stellar parameters from Hardegree-Ullman et al.
+(submitted) were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
 
@@
 \bibliographystyle{aasjournal}
-\bibliography{bib,PhD}
+\bibliography{bib,PhD,coauthors}
 \end{document}

### 2023-03-07 src/tex/ms.tex
minor edits

@@
 The original \texttt{Bioverse} stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function.
-Exquisite parallax and photometric data from the \textit{Gaia} mission made it possible for Hardegree-Ullman et al.
-(submitted) to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \texttt{Bioverse}.
-Here, we briefly describe how stellar parameters $T_{\mathrm{eff}}$, $L_{\star}$, $R_{\star}$, and $M_{\star}$ were derived for \texttt{Bioverse}, but for a full description, please refer to Hardegree-Ullman et al.
-(submitted).
+Exquisite parallax and photometric data from the \textit{Gaia} mission made it possible to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \texttt{Bioverse}~(Hardegree-Ullman et al., submitted).
+Here, we briefly describe how stellar parameters $T_{\mathrm{eff}}$, $L_{\star}$, $R_{\star}$, and $M_{\star}$ were derived.
 
@@
 Stellar radii were then computed with the effective temperatures and lumonisities using the Stefan-Boltzmann law, or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M dwarfs.
-Finally, masses were derived from either the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim0.7\,M_{\odot}$, and that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
+Finally, masses were derived from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim0.7\M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
 The derived stellar parameters from Hardegree-Ullman et al.

### 2023-03-07 src/tex/ms.tex
wooops

--- 2023-03-07 src/tex/ms.tex
@@
 %Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
-\todo[inline]{KEVIN: describe stellar population}
-...
+%\todo[inline]{KEVIN: describe stellar population}
+\todo[inline]{KEVIN: update reference to my paper once accepted}
+The original \bioverse\ stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function.
+Exquisite parallax and photometric data from the \textit{Gaia} mission made it possible to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \bioverse ~(Hardegree-Ullman et al., submitted).
+Here, we briefly describe how stellar parameters $T_{\mathrm{eff}}$, $L_{\star}$, $R_{\star}$, and $M_{\star}$ were derived.
+
+Hardegree-Ullman et al. (submitted) used the Gaia Catalogue of Nearby Stars \citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
+The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2021} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
+From this information, colors and absolute magnitudes were computed, and initial color-magnitude cuts were made to remove non-main sequence stars.
+Effective temperatures were computed primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
+Next, luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
+Stellar radii were then computed with the effective temperatures and luminosities using the Stefan-Boltzmann law, or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M dwarfs.
+Finally, masses were derived from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
+The derived stellar parameters from Hardegree-Ullman et al.
+(submitted) were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
 

### 2023-03-08 src/tex/ms.tex
add what we do with that
explore alternative missions only in the Discussion

@@
 (submitted) were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
+From this catalog, \bioverse\ samples stars within an isotropic distance from the Solar System as required by the sample size.
 
@@
 
-What other planned missions are suited to test the runaway greenhouse hypothesis?
-The ongoing \cheops\ mission was designed as a follow-up mission to search for transits of planets discovered with other techniques, in particular with radial velocity measurements~\citep{Benz2021}.
-As such, it will provide precise radius constraints on a sample of small planets; however, only a small number of planets with periods \SI{> 50}{\day} are being observed.
-This largely limits \cheops' coverage to planets within the runaway greenhouse regime, preventing a detection of the transition.
-
-As \cheops, \ariel~\citep{Puig2016} will be a follow-up mission that is not designed to provide a large number of new radius measurements.
-\ariel's primary targets are larger planets in the range of sub-Neptune to Jupiter-like planets.
-We thus do not expect a significant contribution to exploring the inner edge of the habitable zone for Earth-sized planets.
-
-While not primarily designed to detect transiting planets, the \rst~\citep{Spergel2015} is expected to yield $\sim 10^5$ transiting planets on short orbits and constrain their radii in the course of its mission~\citep{Montet2017}.
-In a fraction of the cases, planets as small as \SI{2}{\Rearth} will be detected.
-About 200 planets smaller than Neptune are expected to be found around M~dwarfs and thus may reach into the habitable zone.
-The \rst\ could thus provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.
-
 \subsection{Statistical power of different mission designs}\label{sec:statpower_missions}
@@
         The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
-    \item Sample size: Since we are looking for a population-level trend, the statistical significance increases with a larger sample size.
+    \item Size and composition of the planetary sample: Evidently, the significance of a statistical trend increases with a larger sample size.
+In addition, the sample must include planets on both sides of the instellation threshold, which is only sufficiently given for low-mass host stars.
     \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be.
@@
 The fidelity of a future detection or falsification of a runaway greenhouse signature will depend on the significance with which the null hypothesis can be excluded.
-As a consequence, instrumentation and survey strategy aiming at testing the hypothesis should aim at maximizing the probability of a true positive detection.
-\begin{note}
-In Sect.~\ref{sec:res_samplesize} we explore the diagnostic power of space missions that are currently operating or are in the planning stage.
-We focus here on transit survey missions, as they provide the most direct radius measurements of a practical planet sample.
+As a consequence, instrumentation and survey strategy aiming at probing the runaway greenhouse threshold should aim at maximizing the probability of a true positive detection.
 
-As discussed above, the key drivers from a mission design perspective are sample size and photometric precision.
+As discussed above, key drivers from a mission design perspective are sample size, photometric precision, and the availability of planets around low-mass host stars.
 We found that along these axes, \plato\ will be the most favorable among the upcoming transit missions.
-%The required sample size and precision are well in reach of near-future exoplanet missions such as PLATO.
 With an expected transit radius precision of \SI{3}{\percent}~\citep{plato2017} and a yield of $> 100$ planets characterized with such high precision~\citep{Rauer2021}, \plato\ is comparable to the optimistic survey we introduced in Sect.~\ref{sec:res_testability}.
 If successful, it should readily detect the predicted statistical imprint or, in case of a non-detection, provide strong upper limits on the occurrence rate of runaway greenhouse planets.
+The latter depends on the lifetimes of runaway greenhouse phases, which are a function of the initial water inventory of the planets~\citep{Hamano2015}.
+Overall, it seems feasible to derive the typical water content of rocky planets from these occurrence estimates.
 
-The latter depends on the lifetimes of runaway greenhouse phases, which are a function of the initial water inventory of the planets~\citep{Hamano2015}.
-It seems thus feasible to derive the typical water content of rocky planets from these occurrence estimates.
-\end{note}
+What other planned missions are suited to test the runaway greenhouse hypothesis?
+The ongoing \cheops\ mission was designed as a follow-up mission to search for transits of planets discovered with other techniques, in particular with radial velocity measurements~\citep{Benz2021}.
+As such, it will provide precise radius constraints on a sample of small planets; however, only a small number of planets with periods \SI{> 50}{\day} are being observed.
+This largely limits \cheops' coverage to planets within the runaway greenhouse regime, preventing a detection of the transition.
+
+As \cheops, \ariel~\citep{Puig2016} will be a follow-up mission that is not designed to provide a large number of new radius measurements.
+\ariel's primary targets are larger planets in the range of sub-Neptune to Jupiter-like planets.
+We thus do not expect a significant contribution to exploring the inner edge of the habitable zone for Earth-sized planets.
+
+While not primarily designed to detect transiting planets, the \rst~\citep{Spergel2015} is expected to yield $\sim 10^5$ transiting planets on short orbits and constrain their radii in the course of its mission~\citep{Montet2017}.
+In a fraction of the cases, planets as small as \SI{2}{\Rearth} will be detected.
+About 200 planets smaller than Neptune are expected to be found around M~dwarfs and thus may reach into the habitable zone.
+The \rst\ could thus provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.
+
+Taken together, \plato\ appears to be the most promising exoplanet mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.
 

### 2023-03-08 src/tex/ms.tex
free "Results" from trace methodologies and discussions

@@
 To represent these photometric limits, we chose a minimum transit depth of \SI{75}{\ppm} as a detection limit and consider only measurements of planets exceeding this threshold.
+For uncertainties in bulk density measurements, we propagate normally distributed errors of mass measurements assuming $\sigma_\mathrm{M_\mathrm{P}} = \SI{10}{\percent}$, as expected for \plato.
+\todo[inline]{20230307 after restructuring, check where the above sentence should go}
 We further exclude target stars with \gaia\ magnitudes $M_\mathrm{G} > \minMagnitude$.
@@
 
-For uncertainties in density measurements, we propagate normally distributed errors of mass measurements assuming $\sigma_\mathrm{M_\mathrm{P}} = \SI{10}{\percent}$, as expected for \plato.
 As can be seen in Fig.~\ref{fig:plato_fgrid}, the detectability is enhanced when radius constraints can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
@@
           This can reduce the required planet yield by about a factor of three.
-An optimized survey will further focus on M~dwarfs to ensure sufficient targets on both sides of the expected threshold instellation.
+          Only an adequate sample of planets orbiting M~dwarfs will ensure sufficient targets on both sides of the expected threshold instellation.
     \item Testing the runaway greenhouse hypothesis will constrain the water inventory of rocky exoplanets, providing important context in the assessment of their habitability.

### 2023-03-08 src/tex/ms.tex
Split up Methods section

@@
 
-\section{Methods}\label{sec:methods}
+\section{Baseline Model}\label{sec:met_baseline}
 %The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
@@
         \caption{Workflow of our hypothesis testing with \bioverse. In the first block, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
-            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:mo_model}.
+            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:met_rghmodel}.
         The second block simulates an exoplanet survey, whereby selection effects and detection biases are introduced.
@@
 \end{figure*}
-
-\subsection{Synthetic star and planet sample}
 In the following, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, as well as the orbital parameters of the planets.
 
-\subsubsection{Stellar sample from Gaia DR3}
+\subsection{Stellar sample from Gaia DR3}
 %Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
@@
 
-\subsubsection{Stellar luminosity evolution}
+\subsection{Stellar luminosity evolution}
 Planetary systems are hosted by stars of a wide range of ages, and stellar luminosities evolve with time.
@@
 
-\subsubsection{Synthetic planet sample}\label{sec:syn_planets}
+\subsection{Synthetic planet sample}\label{sec:syn_planets}
 Next, we assigned to the stellar sample planetary systems with frequencies, orbital parameters, and bulk properties derived from the \kepler\ mission.
@@
 
-\subsubsection{Orbit parameters and planet masses}
+\subsection{Orbit parameters and planet masses}
 Eccentric orbits alter the probability of a planet to transit \citep[e.g.,][]{Barnes2007a}.
@@
 
-\subsubsection{Transit probability}
+\subsection{Transit probability}
 We model the occurrence of transits by assuming isotropic orientations of planetary orbits, calculating the impact parameters $b = \cos(i)/R_\star$ and considering only observable planets with $|b| < 1$.
@@
 
-\subsubsection{Planet radius changes due to runaway greenhouse climates}\label{sec:mo_model}
+\section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
 %\todo[inline]{
@@
 
-\subsection{Survey simulations}
+\section{Exoplanet Survey Simulations and Hypothesis Testing}
 The survey module of \bioverse\ converts the synthetic planet sample into a set of uncertainty-laden measurements on a subset of that sample.
@@
 
-\subsubsection{Detection bias and sample selection}\label{sec:sensitivity} % Turbet2019 lists precision for CHEOPS, PLATO?, TESS?
+\subsection{Detection bias, sample selection, and sensitivity}\label{sec:sensitivity} % Turbet2019 lists precision for CHEOPS, PLATO?, TESS?
 Not all transiting planets are detectable with the same likelihood and detection biases have an impact on the demographic measurements we are interested in.
@@
 
-\subsection{Hypothesis tests}
+\subsection{The runaway greenhouse hypothesis}
 We now turn to assessing the simulated surveys in terms of their ability to detect the runaway greenhouse transition and their constraining power on parameters of this trend.
@@
 For our specific problem, this implies comparing evidences for a demographic imprint of the runaway greenhouse effect to its absence.
-\subsubsection{Definition of the runaway greenhouse hypothesis}
 \begin{note}
@@
 
-\subsubsection{Bayesian model comparison}
+\subsection{Bayesian model comparison}
 We are now ready to assess the relative plausibility of $H_{\mathrm{rgh}}$ and $H_0$ given the synthetic data we have generated, assigning equal a~priori~ probabilities to these models.
@@
 We sample $x_{H_2O}$ from a log-uniform distribution to imply scale-invariant ignorance.
-Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Sect.~\ref{sec:mo_model}).
+Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Sect.~\ref{sec:met_rghmodel}).
 Finally, we adopt a broad, uniform prior for $\langle R_\mathrm{P}\rangle_\mathrm{out}$ bound by $[\prRmin, \prRmax] \, \mathrm{R_\oplus}$.

### 2023-03-08 src/tex/ms.tex
test 02

--- 2023-03-08 src/tex/ms.tex
@@
 \author{al.}
+XXX
 
@@
 
-    Here, we assess the ability of space and ground-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages contextual information from the overall planet population.
+    Here, we assess the ability of space and ground-based telescopes to test this hypothesis using Bioverse, a simulation framework that leverages contextual information from the overall planet population.
     $\ldots$
@@
 \todo[inline]{``Points for the intro:
-\bioverse\\
+Bioverse\\
 Transit survey to measure mass/radius of planets\\
@@
 
+% TEST making changes
+
+% let this be a change
+
 \begin{note}
@@
 
-...
     Venus and Earth, while having accreted from the same mass reservoir and despite their similar bulk properties, evolved into planets with very different environmental conditions on their surfaces.
@@
 We determined -- for different survey configurations -- the confidence level with which the runaway greenhouse threshold can be detected in simulated exoplanet observations.
-For this purpose we expanded the \bioverse\ framework~\citep{Bixel2020,Bixel2021}\footnote{\bioverse\ is actively maintained and documented open source software written in Python.
+For this purpose we expanded the Bioverse framework~\citep{Bixel2020,Bixel2021}\footnote{Bioverse is actively maintained and documented open source software written in Python.
 Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic stellar and planetary samples into which we injected the runaway greenhouse signal, simulated observations of this sample, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).
-\todo[inline]{\bioverse\ will need a short introduction before the modifications are mentioned. Probably should get a paragraph in the intro}
+\todo[inline]{Bioverse will need a short introduction before the modifications are mentioned. Probably should get a paragraph in the intro}
 \begin{figure*}
@@
         \includegraphics[width=\hsize]{figures/flowchart.pdf}
-        \caption{Workflow of our hypothesis testing with \bioverse. In the first block, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
+        \caption{Workflow of our hypothesis testing with Bioverse. In the first block, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
             A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:mo_model}.
@@
 %Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
-%\todo[inline]{KEVIN: describe stellar population}
-\todo[inline]{KEVIN: update reference to my paper once accepted}
-The original \bioverse\ stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function.
-Exquisite parallax and photometric data from the \textit{Gaia} mission made it possible to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \bioverse ~(Hardegree-Ullman et al., submitted).
-Here, we briefly describe how stellar parameters $T_{\mathrm{eff}}$, $L_{\star}$, $R_{\star}$, and $M_{\star}$ were derived.
-
-Hardegree-Ullman et al. (submitted) used the Gaia Catalogue of Nearby Stars \citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
-The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2021} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
-From this information, colors and absolute magnitudes were computed, and initial color-magnitude cuts were made to remove non-main sequence stars.
-Effective temperatures were computed primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
-Next, luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
-Stellar radii were then computed with the effective temperatures and luminosities using the Stefan-Boltzmann law, or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M dwarfs.
-Finally, masses were derived from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
-The derived stellar parameters from Hardegree-Ullman et al.
-(submitted) were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
-From this catalog, \bioverse\ samples stars within an isotropic distance from the Solar System as required by the sample size.
+\todo[inline]{KEVIN: describe stellar population}
+...
 
@@
 \subsection{Survey simulations}
-The survey module of \bioverse\ converts the synthetic planet sample into a set of uncertainty-laden measurements on a subset of that sample.
-This task includes selection of the targets, application of detection biases, and conducting simulated measurements, all of which are specific to the particular survey.
-For each planet-level measurement such as radius or instellation, we draw the measured value from a normal distribution centered on the true value with a standard deviation set by the survey's precision.
+The survey module of Bioverse converts the synthetic planet sample into a set of uncertainty-laden measurements on a subset of that sample.
+   This task includes selection of the targets, application of detection biases, and conducting simulated measurements.
+   All of the above are specific to the particular survey.
 
@@
             Retrieved posterior distribution of key parameters in the optimistic scenario.
-The density maps in each panel show relationships between and marginalized distributions of the threshold instellation $S_\mathrm{thresh}$, the water mass fraction $x_{H_2O}$, and the dilution factor $f_\mathrm{rgh}$ as they could be retrieved with a high-precision transit survey and a sample of \var{N_optimistic} planets.
+The density maps in each panel show relationships between and marginalized distributions of the model parameters as they could be retrieved with a high-precision transit survey and a sample of \var{N_optimistic} planets.
 True values of the parameters for the injected effect are shown in orange.
@@
 \end{figure}
-Figure~\ref{fig:statpwr_H2O-f} explores the statistical power of the hypothesis test achieved in the above scenario for different combinations of the poorly constrained parameters $x_{H_2O}$ and $f_\mathrm{rgh}$.
+Figure~\ref{fig:statpwr_H2O-f} explores the statistical power achieved in the above scenario for different combinations of the poorly constrained parameters $x_{H_2O}$ and $f_\mathrm{rgh}$.
 It is highest for large water inventories and large dilution factors.
@@
 
-\subsection{Detecting the runaway greenhouse transition with \plato}\label{sec:res_samplesize}
+\subsection{Detecting the runaway greenhouse transition with current and planned exoplanet missions}\label{sec:res_samplesize}
 %\todo{Introduce survey simulation(s).
@@
             \caption{
-            Expected delta-evidences as a function of the fraction of planets with runaway greenhouse climates for different versions of the \plato\ survey.
+            Expected delta-evidences as a function of the fraction of planets having runaway greenhouse climates for different versions of the \plato\ survey.
             The median values of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
@@
 
+What other planned missions are suited to test the runaway greenhouse hypothesis?
+The ongoing \cheops\ mission was designed as a follow-up mission to search for transits of planets discovered with other techniques, in particular with radial velocity measurements~\citep{Benz2021}.
+As such, it will provide precise radius constraints on a sample of small planets; however, only a small number of planets with periods \SI{> 50}{\day} are being observed.
+This largely limits \cheops' coverage to planets within the runaway greenhouse regime, preventing a detection of the transition.
+
+As \cheops, \ariel~\citep{Puig2016} will be a follow-up mission that is not designed to provide a large number of new radius measurements.
+\ariel's primary targets are larger planets in the range of sub-Neptune to Jupiter-like planets.
+We thus do not expect a significant contribution to exploring the inner edge of the habitable zone for Earth-sized planets.
+
+While not primarily designed to detect transiting planets, the \rst~\citep{Spergel2015} is expected to yield $\sim 10^5$ transiting planets on short orbits and constrain their radii in the course of its mission~\citep{Montet2017}.
+In a fraction of the cases, planets as small as \SI{2}{\Rearth} will be detected.
+About 200 planets smaller than Neptune are expected to be found around M~dwarfs and thus may reach into the habitable zone.
+The \rst\ could thus provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.
+
 \subsection{Statistical power of different mission designs}\label{sec:statpower_missions}
-To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
+To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated several transit surveys with different designs and strategies and measured their capability to recover the trend and constrain its parameters.
 We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
@@
 \subsubsection{Spectral type of target stars}\label{sec:results_FGK_M}
-Since the receiving radiation is sensitive on the spectral type of the host star, the share of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
+Since receiving radiation is sensitive on the spectral type of the host star, the share of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
 We tested the detectability of the runaway greenhouse signal when only FGK or only M~dwarfs are considered (see Fig.~\ref{fig:plato_fgrid}).
@@
 If, on the other hand, a follow-up campaign adds planetary mass estimates and we fit the signal in bulk density space, useful constraints are possible under more pessimistic conditions.
-A sample containing only planets around M~dwarf yields a performance comparable to that of the overall sample.
+A sample containing only planets around M~dwarf yields a comparable performance as the overall sample.
 
@@
 It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019}.
-Using \bioverse\ to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we showed that the population of greenhouse-inflated planets create a signature in the radius and density distribution that is well in the detectable range.
+Using Bioverse to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we showed that the population of greenhouse-inflated planets create a signature in the radius and density distribution that is well in the detectable range.
 
@@
         The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
-    \item Size and composition of the planetary sample: Evidently, the significance of a statistical trend increases with a larger sample size.
-In addition, the sample must include planets on both sides of the instellation threshold, which is only sufficiently given for low-mass host stars.
+    \item Sample size: Since we are looking for a population-level trend, the statistical significance increases with a larger sample size.
     \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be.
@@
 The fidelity of a future detection or falsification of a runaway greenhouse signature will depend on the significance with which the null hypothesis can be excluded.
-As a consequence, instrumentation and survey strategy aiming at probing the runaway greenhouse threshold should aim at maximizing the probability of a true positive detection.
+As a consequence, instrumentation and survey strategy aiming at testing the hypothesis should aim at maximizing the probability of a true positive detection.
+\begin{note}
+In Sect.~\ref{sec:res_samplesize} we explore the diagnostic power of space missions that are currently operating or are in the planning stage.
+We focus here on transit survey missions, as they provide the most direct radius measurements of a practical planet sample.
 
-As discussed above, key drivers from a mission design perspective are sample size, photometric precision, and the availability of planets around low-mass host stars.
+As discussed above, the key drivers from a mission design perspective are sample size and photometric precision.
 We found that along these axes, \plato\ will be the most favorable among the upcoming transit missions.
+%The required sample size and precision are well in reach of near-future exoplanet missions such as PLATO.
 With an expected transit radius precision of \SI{3}{\percent}~\citep{plato2017} and a yield of $> 100$ planets characterized with such high precision~\citep{Rauer2021}, \plato\ is comparable to the optimistic survey we introduced in Sect.~\ref{sec:res_testability}.
 If successful, it should readily detect the predicted statistical imprint or, in case of a non-detection, provide strong upper limits on the occurrence rate of runaway greenhouse planets.
+
 The latter depends on the lifetimes of runaway greenhouse phases, which are a function of the initial water inventory of the planets~\citep{Hamano2015}.
-Overall, it seems feasible to derive the typical water content of rocky planets from these occurrence estimates.
-
-What other planned missions are suited to test the runaway greenhouse hypothesis?
-The ongoing \cheops\ mission was designed as a follow-up mission to search for transits of planets discovered with other techniques, in particular with radial velocity measurements~\citep{Benz2021}.
-As such, it will provide precise radius constraints on a sample of small planets; however, only a small number of planets with periods \SI{> 50}{\day} are being observed.
-This largely limits \cheops' coverage to planets within the runaway greenhouse regime, preventing a detection of the transition.
-
-As \cheops, \ariel~\citep{Puig2016} will be a follow-up mission that is not designed to provide a large number of new radius measurements.
-\ariel's primary targets are larger planets in the range of sub-Neptune to Jupiter-like planets.
-We thus do not expect a significant contribution to exploring the inner edge of the habitable zone for Earth-sized planets.
-
-While not primarily designed to detect transiting planets, the \rst~\citep{Spergel2015} is expected to yield $\sim 10^5$ transiting planets on short orbits and constrain their radii in the course of its mission~\citep{Montet2017}.
-In a fraction of the cases, planets as small as \SI{2}{\Rearth} will be detected.
-About 200 planets smaller than Neptune are expected to be found around M~dwarfs and thus may reach into the habitable zone.
-The \rst\ could thus provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.
-
-Taken together, \plato\ appears to be the most promising exoplanet mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.
+It seems thus feasible to derive the typical water content of rocky planets from these occurrence estimates.
+\end{note}
 
@@
 Our calculations with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
-Besides their large number in a volume/magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular on orbits further out, i.e., in the optimistic habitable zone.
-In fact, we showed that the FGK part of the planet sample barely contributes to the statistical power.
-Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars][]{Turbet2019}, enhancing the demographic signal it leaves.
+Besides their large number in a volume/magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular beyond it, i.e., in the optimistic habitable zone.
+The transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars][]{Turbet2019}, enhancing the demographic signal it leaves.
 An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of \SI{100}{\mega\year}~\citep{Luger2015}.
@@
 This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the statistical radius inflation trend, also since the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
-Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
+Cases of pure rocky composition and very low volatile contants can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
 Assuming a distribution of water mass fractions instead of a fixed value would thus not significantly change our results.
@@
 Runaway greenhouse climates on rocky planets cause significant inflation of their radii.
-Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius discontinuity resulting from this inflation in the exoplanet population.
+Using Bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius discontinuity resulting from this inflation in the exoplanet population.
 Our key findings are as follows:
@@
 %K.H.-U. provided the star generator based on the Gaia catalog.
-%G.B. implemented the planet generator in the \bioverse\ framework.
+%G.B. implemented the planet generator in the Bioverse framework.
 %All authors provided comments and suggestions on the manuscript.
@@
 \software{
-\bioverse ~\citep{Bixel2021},
+Bioverse~\citep{Bixel2021},
 Astropy~\citep[][]{AstropyCollaboration2018},
@@
 \bibliographystyle{aasjournal}
-\bibliography{bib,PhD,coauthors}
+\bibliography{bib,PhD}
 \end{document}

### 2023-03-08 src/tex/ms.tex
various small changes

--- 2023-03-08 src/tex/ms.tex
@@
 \section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
+\todo[inline]{TIM: please review}
 %\todo[inline]{
@@
 The climate state of a planet has a direct influence on its size as measured by transit photometry~\citep{Turbet2019}.
-Here, we use predictions on transit atmospheric thickness from geophysical models to derive the transit radii of planets with temperate climates and those with instellation-induced runaway greenhouse climates, respectively.
+We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius that planets with instellation-induced runaway greenhouse climates experience.
 ...
@@
     \end{equation}
-    Here, we assume global redistribution of incoming flux and a fixed albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
+    We assume global redistribution of incoming flux and a fixed albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
     The absolute value of $S_\oplus$, by which all synthetic planets are scaled, is then \SI{238}{\watt\per\square\meter}.
@@
 %The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
-This parametrization introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
-In the following, we test if and under what conditions this trend is large enough to be detected with high significance.
+In the following, we test if and under what conditions this parametrization causes a demographic trend that is large enough to be detected with high significance.
 
@@
 The runaway greenhouse transition is expected to occur at orbits sufficiently close to let mission duration not be a limiting factor: $S_\mathrm{thresh}$ occurs close to Earth's orbital period for \mbox{G-type} stars and on significantly shorter periods for the more prevalent M~dwarf hosts~\citep{Goldblatt2012}.
-However, for a transit signal to be detected in a transit survey, it must reach a sufficient signal-to-noise ratio (S/N).
-This S/N is sensitive on the photometric precision that can be achieved~\citep[e.g.,][]{Burke2015,Hardegree-Ullman2019}.
+However, for a transit signal to be detected in a transit survey, it must reach a sufficient signal-to-noise ratio (S/N) that is sensitive on the achieved photometric precision~\citep[e.g.,][]{Burke2015,Hardegree-Ullman2019}.
 In its in-orbit commissioning phase, the \cheops\ mission demonstrated a photometric precision of \SI{75}{\ppm} for faint ($V\approx 12$) stars~\citep{Benz2021}.
@@
 When ordered in orbital period space, the different planet types overlap, diluting the demographic imprint.
-With net instellation as an independent variable, planets above and below the runaway greenhouse threshold separate.
+With net instellation as an independent variable, planets above and below the runaway greenhouse threshold separate: The runaway greenhouse-induced radius inflation introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
 We also show the predictions of observable average planet radii from the statistical hypotheses defined above.
@@
 \section{Discussion}\label{sec:discussion}
-\todo[inline]{\textbf{Potential discussion points: }\\
+\todo[inline]{\textbf{Potential additional discussion points: }\\
 (How close is Earth to runaway? (-> \citet{Goldblatt2012,Popp2016}; what is the max. expected change in Earth's albedo due to deglaciation?)}
-\todo[inline]{Caveats: see \citet{Turbet2019} Sect. 3, \citet{Bower2019}. Mention we only consider instellation-induced rgh.\\
-- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
-- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
-- assumption of a single albedo. ``if there is a range of albedos, will that wash away the signal?''
-- Impact of assumptions on findings, e.g.
-M-R relation.
-``We should have a discussion section dedicated to the impact of assumptions on the findings - including mass-radius.
-A good way to go about this is to present bracketing cases; or to discuss why the discountinuity in the results would not be affected by changes that led to continuous offsets in the results.
-Or both.
-But it should not be in the methods section.''}
 
 \subsection{Statistical imprint of exoplanet climates}
-\begin{note}
 The runaway greenhouse transition is a robust prediction from climate models, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
@@
 Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
-
-\end{note}
+Overall, searching for the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.
 
 \subsection{Detectability of the runaway greenhouse transition}
-\todo[inline]{the demographic imprint is detectable. Under what conditions?\\
-Mention multiplanet systems.}
+\todo[inline]{Mention multiplanet systems.}
 We showed in Sect.~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
@@
 \subsubsection{OPTIONAL...: Atmospheric signatures}
-\todo[inline]{discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (Tim's talk)}
+\todo[inline]{TIM: Ideas?\\
+discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk)}
 \todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
@@
 The \rst\ could thus provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.
+\todo[inline]{what about TESS?}
 
@@
 
-\subsubsection{Target star spectral types}
+\subsubsection{The importance of M~dwarfs in the target list}
 To date, the majority of planets with radius measurements orbit FGK~dwarfs, and, based on the instellation they receive, most of them lie in the runaway greenhouse regime~\citep{Thompson2018}.
-Obviously, a step-function-like signal in the exoplanet demographics like the one illustrated above can not be constrained well if only one side of the discontinuity is being sampled.
+Obviously, a step-function-like signal in the exoplanet demographics like the one investigated above can not be constrained well if only one side of the discontinuity is being sampled.
 This, however, is the situation for planetary systems around Sun-like stars -- their habitable zones are so distant that transiting planets within them are very rare due to pure geometrical reasons.
@@
 In fact, we showed that the FGK part of the planet sample barely contributes to the statistical power.
-Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars][]{Turbet2019}, enhancing the demographic signal it leaves.
+Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars,][]{Turbet2019}, enhancing the demographic signal it leaves.
 An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of \SI{100}{\mega\year}~\citep{Luger2015}.
@@
 \subsection{Impact of assumptions on our findings}
+\todo[inline]{Caveats\\
+context: \citet{Turbet2019} Sect. 3, \citet{Bower2019}.\\
+''present bracketing cases; or to discuss why the discountinuity in the results would not be affected by changes that led to continuous offsets in the results. Or both.''}
 \subsubsection{Baseline mass-radius relationship}
@@
 \todo[inline]{if feasible: run a test with a different M-R relation. do we expect that the choice of M-R relation impacts the results? Perhaps have an appendix section.}
+
+\subsubsection{Simplifications in the atmosphere model}
+\todo[inline]{TIM: your expertise will be valuable here.\\
+Mention we only consider instellation-induced rgh.\\
+- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
+- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
+- assumption of a single albedo. ``if there is a range of albedos, will that wash away the signal?''
+}
 

### 2023-03-08 src/tex/ms.tex
ensure section describes results without discussion

@@
 \subsection{Synthetic planet sample}\label{sec:syn_planets}
+
 Next, we assigned to the stellar sample planetary systems with frequencies, orbital parameters, and bulk properties derived from the \kepler\ mission.
-We adopted the planetary occurrence rates in orbital period and radius recently derived in \citet{Bergsten2022}.
-Following \citet{Youdin2011a}, their inferred occurrence rate density can be expressed in the form
+We adopted the planetary occurrence rate model from \citet{Bergsten2022}, which defines the occurrence of small planets in radius and orbital period.
+Following \citet{Youdin2011a}, their inferred occurrence rate density can be expressed in the form:
 \begin{equation}
@@
 \end{equation}
-where $F_0$
-...
-\todo[inline]{GALEN: Briefly describe underlying planet occurrence rates of synthetic population, its Mstar dependency, any caveats or assumptions.}
+where $F_0$ represents the average number of planets per star and $C_n$ is a normalization constant.
+The shape function $g(R,P)$ describing the occurrence distribution is the same as in \citet{Bergsten2022}.
+The functional form and stellar-mass dependent parameters (including a new fit to the sample of \kepler\ planets with hosts below $0.56\,M_{\odot}$) are included in \bioverse\ as the prominent planet-generation model; the original model based on SAG13 occurrence rates is preserved as a legacy feature.
+
+With this updated distribution function, planets are generated as in the original \bioverse\ paper, but with an additional stellar mass dependence (for a star of a given mass, the relevant model parameters from the bin containing that mass are used).
+In summary, stars are assigned a number of planets in accordance with $F_0$, with planet parameters randomly drawn in proportion to the occurrence distribution function evaluated over a grid of radii and orbital periods.
+% \todo[inline]{GALEN: Briefly describe underlying planet occurrence rates of synthetic population, its Mstar dependency, any caveats or assumptions.}
+\todo[inline]{GALEN: The most relevant assumption of using our model is that, at the relevant fluxes (I think), we predict that super-Earths are pretty common.
+I think the "drop" in our model where super-Earths become less common happens at later fluxes than what is relevant in this paper...
+Let's double check this, and add a note in the first paragraph about it?.}
 
@@
 
-Accurate retrievals from radius measurements alone require high fractions of greenhouse climate-bearing planets to achieve an accurate constraint on $S_\mathrm{thresh}$.
-If, on the other hand, a follow-up campaign adds planetary mass estimates and we fit the signal in bulk density space, useful constraints are possible under more pessimistic conditions.
-A sample containing only planets around M~dwarf yields a performance comparable to that of the overall sample.
+We find that retrievals from radius measurements alone require high fractions of greenhouse climate-bearing planets to achieve an accurate constraint on $S_\mathrm{thresh}$:
+Dilution factors $f_\mathrm{rgh} \gtrsim 0.5$ yield posterior probability distributions that are condensed at the order of magnitude of the true value; accurate constraints to within $\pm 0.25\,\mathrm{dex}$ of the truth are reached only from $f_\mathrm{rgh} \approx 0.8$.
+
+In contrast, if planet masses are available and the hypothesis test is conducted in bulk density space, useful constraints emerge already from about $f_\mathrm{rgh} \approx 0.35$.
+Accuracy and precision of the retrievals are improved.
+
+A sample containing only planets around M~dwarf yields a very similar performance, which is because these planets contribute most to the statistical power of the retrieval.
 

### 2023-03-08 src/tex/ms.tex
discuss the value of follow-up campaigns

@@
 \subsubsection{The value of follow-up campaigns}
-\begin{note}
 As runaway greenhouse phases leave a stronger imprint in bulk density than in planet radius (see Sect.~\ref{sec:res_followup}), it would be beneficial to obtain constraints on planetary masses and test the runaway greenhouse hypothesis in density space instead of radius space.
-Precise ground-based radial velocity measurements are expensive, and therefore such follow-up efforts may only be available for a subsample of a mission's targets.
-For a mission design similar to PLATO, a density-based test on about a third of the overall sample is equivalent to a pure radius-based analysis.
+This way, useful constraints can be obtained under more pessimistic conditions, e.g., a low predominant water content of planetary surfaces and atmospheres or a smaller available planet sample.
+For a mission design similar to PLATO, a density-based hypothesis test on about a third of the overall sample is equivalent to a pure radius-based analysis.
+At a fixed sample size, key parameters of the runaway greenhouse models can be more narrowly constrained when additional mass measurements are available.
+
+However, precise ground-based radial velocity measurements are expensive, and therefore such follow-up efforts may only be available for a subsample of a mission's targets.
 The diagnostic power of such a test may be improved by simultaneously fitting for the trend in radius space in the subsample without RV follow-up.
 An optimized mission in search for the inner edge of the habitable zone will further enhance its information content via a wise selection of follow-up targets, i.e., favoring those planets residing close to the expected instellation threshold $S_\mathrm{thresh}$.
-\end{note}
 
@@
 context: \citet{Turbet2019} Sect. 3, \citet{Bower2019}.\\
-''present bracketing cases; or to discuss why the discountinuity in the results would not be affected by changes that led to continuous offsets in the results. Or both.''}
+''present bracketing cases; or to discuss why the discountinuity in the results would not be affected by changes that led to continuous offsets in the results. Or both.''\\
+- not yet explored: how sensitive are we on the assumed planet population?
+$\exists$ danger of false positive detection on occurrence rate ``ridges'' like the Neptune desert?
+Currently we have a relatively smooth occurrence rate density in $R_P - S$ space based on the power-law fits in \citet{Bergsten2022}.
+We could, with relatively moderate effort, run a sanity check on an alternative sample (e.g., the more structured Bern model)}
 \subsubsection{Baseline mass-radius relationship}

### 2023-03-09 src/tex/ms.tex
describe overall method earlier

@@
         \includegraphics[width=\hsize]{figures/flowchart.pdf}
-        \caption{Workflow of our hypothesis testing with \bioverse. In the first block, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
+        \caption{Workflow of our hypothesis testing with \bioverse. First, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
             A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:met_rghmodel}.
-        The second block simulates an exoplanet survey, whereby selection effects and detection biases are introduced.
-Finally, the third block deals with testing the runaway greenhouse hypothesis based on data from the survey simulation.
-        By iterating through these steps, we compute the statistical power of testing the hypothesis given the assumed survey design.}
+        We then simulate an exoplanet survey, whereby selection effects and detection biases are introduced.
+Finally, we test the runaway greenhouse hypothesis based on data from the survey simulation.
+        By iterating through these steps, we compute the statistical power of testing the hypothesis for different survey designs.}
         \label{fig:flowchart}
@@
 \end{figure*}
+Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
+To enable a quantitative assessment of this detectability, we injected the signal into a simulated planet sample before observing it with simulated surveys.
 In the following, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, as well as the orbital parameters of the planets.
@@
 
-Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
-To enable a quantitative assessment of this detectability, we injected the signal into the simulated planet sample before observing it with simulated surveys.
 While we search for the runaway greenhouse pattern in demographic quantities such as average planet radii, the injected changes happen on the planetary level: We changed each planet's radius based on its individual set of properties and the associated predictions from steam atmosphere and water incorporation models.
@@
 We consider the following cases (see Fig.~\ref{fig:radiusevolution}):
-%\begin{itemize}
-%\item[A.] No radius change.
-%This case serves as our null hypothesis, where planetary radii are not affected by the amount of irradiation the planets receive.
-%A solid planetary surface is assumed and no steam atmosphere is present.
-%\item[B.] Radius inflation due to a steam atmosphere.
-%Atmospheric scale heights are significantly increased.
-%\item[C.] As B, but including radius decrease of inflated planets due to water partitioning in the molten mantle.
-%\end{itemize}
 
@@
 
-\subsubsection{Spectral type of target stars}\label{sec:results_FGK_M}
+\subsubsection{Dependence on host star spectral type}\label{sec:results_FGK_M}
 Since the receiving radiation is sensitive on the spectral type of the host star, the share of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.

### 2023-03-09 src/tex/ms.tex
explain planet sample
Added details for generating the synthetic planet sample.

--- 2023-03-08 src/tex/ms.tex
@@
 
-We find that retrievals from radius measurements alone require high fractions of greenhouse climate-bearing planets to achieve an accurate constraint on $S_\mathrm{thresh}$:
-Dilution factors $f_\mathrm{rgh} \gtrsim 0.5$ yield posterior probability distributions that are condensed at the order of magnitude of the true value; accurate constraints to within $\pm 0.25\,\mathrm{dex}$ of the truth are reached only from $f_\mathrm{rgh} \approx 0.8$.
-
-In contrast, if planet masses are available and the hypothesis test is conducted in bulk density space, useful constraints emerge already from about $f_\mathrm{rgh} \approx 0.35$.
-Accuracy and precision of the retrievals are improved.
-
-A sample containing only planets around M~dwarf yields a very similar performance, which is because these planets contribute most to the statistical power of the retrieval.
+Accurate retrievals from radius measurements alone require high fractions of greenhouse climate-bearing planets to achieve an accurate constraint on $S_\mathrm{thresh}$.
+If, on the other hand, a follow-up campaign adds planetary mass estimates and we fit the signal in bulk density space, useful constraints are possible under more pessimistic conditions.
+A sample containing only planets around M~dwarf yields a performance comparable to that of the overall sample.
 

### 2023-03-09 src/tex/ms.tex
move discussion of background population

--- 2023-03-09 src/tex/ms.tex
@@
 This removes it from the steam atmosphere, which in turn decreases the radius inflation.
-The effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes typically reach measurable values~\citep{Dorn2021}.
+The effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes are predicted to reach measurable values~\citep{Dorn2021}.
 The magnitude of this effect is again dependent on the planet mass and on the water mass fraction.
@@
 Blue: planets with steam atmospheres and including the effect of water incorporation in the melt.
+                Steam atmospheres cause a significant radius increase, which is slightly reduced when water incorporation in the melt is considered.
             \textit{Bottom}: Radius evolution of different planet types, illustrating degeneracies and potential for confusion among planet classes.
@@
 \end{figure}
-Figure~\ref{fig:radiusevolution} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of $x_{H_2O}= \var{wrr}$.
-Steam atmospheres cause a significant radius increase, which is slightly reduced when water incorporation in the melt is considered.
+We illustrate the mass-radius relations of the three cases in Figure~\ref{fig:radiusevolution} where a fiducial water mass fraction of $x_{H_2O}= \var{wrr}$ is assumed.
 In the following, we only distinguish between the non-runaway greenhouse and wet melt scenarios.
 
-As illustrated in Fig.~\ref{fig:radiusevolution}~(bottom panel), different processes can lead to the classification of a planet into either of these categories.
-For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature.
-Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the non-runaway greenhouse category on a timescale of $10^8\SI{}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
-To account for these effects, we introduce a dilution parameter $f_\mathrm{rgh}$, which represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
+To account for planets unable to sustain a steam atmosphere as well as evolutionary effects such as desiccation (see bottem panel of Fig.~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
+It represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
 Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the non-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
@@
 }
+As illustrated in Fig.~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the runaway or non-runaway category.
+For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature.
+Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the non-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
+We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.
 

### 2023-03-09 src/tex/ms.tex
state goal and overview of steps

@@
 %The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
-\todo[inline]{Start methods description with brief statement of goal and overview of the steps. The first sentence is similar, but would be good to separate goal from method here.
-}
-We determined -- for different survey configurations -- the confidence level with which the runaway greenhouse threshold can be detected in simulated exoplanet observations.
-For this purpose we expanded the \bioverse\ framework~\citep{Bixel2020,Bixel2021}\footnote{\bioverse\ is actively maintained and documented open source software written in Python.
-Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic stellar and planetary samples into which we injected the runaway greenhouse signal, simulated observations of this sample, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).
+%Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
+%To enable a quantitative assessment of this detectability, we injected the signal into a simulated planet sample before observing it with simulated surveys.
+The goal of this study is to determine -- for different configurations of near-future exoplanet surveys -- the confidence level with which the runaway greenhouse threshold can be detected statistically.
+%in simulated exoplanet observations.
+Our basic methodology was as follows:
+We expanded the \bioverse\ framework~\citep{Bixel2020,Bixel2021}\footnote{\bioverse\ is actively maintained and documented open source software written in Python.
+Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic samples of stars that host planets according to the observed exoplanet demographics.
+We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).\todo[inline]{make sure rgh hypothesis is mentioned before, or rephrase}
 \todo[inline]{\bioverse\ will need a short introduction before the modifications are mentioned. Probably should get a paragraph in the intro}
@@
 \end{figure*}
-Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
-To enable a quantitative assessment of this detectability, we injected the signal into a simulated planet sample before observing it with simulated surveys.
-In the following, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, as well as the orbital parameters of the planets.
+In this section, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, as well as the orbital parameters of the planets.
 
@@
 Currently we have a relatively smooth occurrence rate density in $R_P - S$ space based on the power-law fits in \citet{Bergsten2022}.
-We could, with relatively moderate effort, run a sanity check on an alternative sample (e.g., the more structured Bern model)}
+We could, with relatively moderate effort, run a sanity check on an alternative sample (e.g., the more structured Bern model populations, or the population in the Genesis database)}
 \subsubsection{Baseline mass-radius relationship}

### 2023-03-09 src/tex/ms.tex
remove mission duration

@@
 
-The runaway greenhouse transition is expected to occur at orbits sufficiently close to let mission duration not be a limiting factor: $S_\mathrm{thresh}$ occurs close to Earth's orbital period for \mbox{G-type} stars and on significantly shorter periods for the more prevalent M~dwarf hosts~\citep{Goldblatt2012}.
-However, for a transit signal to be detected in a transit survey, it must reach a sufficient signal-to-noise ratio (S/N) that is sensitive on the achieved photometric precision~\citep[e.g.,][]{Burke2015,Hardegree-Ullman2019}.
+For a transit signal to be detected in a transit survey, it must reach a sufficient signal-to-noise ratio (S/N) that is sensitive on the achieved photometric precision~\citep[e.g.,][]{Burke2015,Hardegree-Ullman2019}.
 In its in-orbit commissioning phase, the \cheops\ mission demonstrated a photometric precision of \SI{75}{\ppm} for faint ($V\approx 12$) stars~\citep{Benz2021}.

### 2023-03-09 src/tex/ms.tex
fix paragraph on photometric precision/transit depth

@@
 
-For a transit signal to be detected in a transit survey, it must reach a sufficient signal-to-noise ratio (S/N) that is sensitive on the achieved photometric precision~\citep[e.g.,][]{Burke2015,Hardegree-Ullman2019}.
-In its in-orbit commissioning phase, the \cheops\ mission demonstrated a photometric precision of \SI{75}{\ppm} for faint ($V\approx 12$) stars~\citep{Benz2021}.
-In comparison, \plato's expected precision for similar stars was estimated to be \SI{\sim 50}{\ppm}~\citep[][Matuszewski et al., in prep.]{plato2017}.
-To represent these photometric limits, we chose a minimum transit depth of \SI{75}{\ppm} as a detection limit and consider only measurements of planets exceeding this threshold.
+A successful transit detection requires a sufficient signal-to-noise ratio (S/N), which is sensitive on the achieved photometric precision.
+%~\citep[e.g.,][]{Koch2010,Ricker2014,Burke2015,Hardegree-Ullman2019} % Cabrera+ in prep for PLATO
+%In its in-orbit commissioning phase, the \cheops\ mission demonstrated a photometric precision of \SI{75}{\ppm} for faint ($V\approx 12$) stars~\citep{Benz2021}.
+%In comparison, \plato's expected precision for similar stars was estimated to be \SI{\sim 50}{\ppm}~\citep[][Matuszewski et al., in prep.]{plato2017}.
+In line with its requirement to detect Earth-sized planets in a \SI{1}{\year} orbit around Sun-like stars, \plato\ is designed to enable the detection of a \SI{80}{\ppm} transit signal~\citep[][Matuszewski et al., in prep.]{plato2017}.
+To reflect its sensitivity, we chose a minimum transit depth of \SI{80}{\ppm} as a detection limit and consider only measurements of planets exceeding this threshold.
 For uncertainties in bulk density measurements, we propagate normally distributed errors of mass measurements assuming $\sigma_\mathrm{M_\mathrm{P}} = \SI{10}{\percent}$, as expected for \plato.

### 2023-03-09 src/tex/ms.tex
move description of priors

@@
 
+A sensible choice of priors is central for evidence estimation via nested sampling.
+As the parameters of interest are poorly constrained by previous data, we used relatively uninformative priors to sample the entire physically plausible parameter space.
+For $S_\mathrm{thresh}$, we chose a uniform prior in $[\prSmin,\prSmax]\, \SI{}{\watt\per\meter\squared}$.
+We sample $x_{H_2O}$ from a log-uniform distribution to imply scale-invariant ignorance.
+Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Sect.~\ref{sec:met_rghmodel}).
+For $f_\mathrm{rgh}$, we chose a uniform prior in $[0, 1]$.
+Finally, we adopt a broad, uniform prior for $\langle R_\mathrm{P}\rangle_\mathrm{out}$ bound by $[\prRmin, \prRmax] \, \mathrm{R_\oplus}$.
+
 The measured radii $R_\mathrm{P, i}$ cannot be directly used for the hypothesis tests as they include intrinsic scatter that is not caused by measurement errors.
@@
 We are now ready to assess the relative plausibility of $H_{\mathrm{rgh}}$ and $H_0$ given the synthetic data we have generated, assigning equal a~priori~ probabilities to these models.
-This is done by comparing the Bayesian evidence $\mathcal{Z}$ of the models, which we estimate with the nested sampling~\citep{Skilling2004} algorithm \emph{dynesty}~\citep{Speagle2020}.
+This is done by comparing the Bayesian evidence $\mathcal{Z}$ of the models, which we estimated with the nested sampling~\citep{Skilling2004} algorithm \emph{dynesty}~\citep{Speagle2020}.
+We initialized the sampler with the priors defined above to let it estimate the evidence and sample the posterior distributions.
 Our criterion to reject the null hypothesis is
@@
 %The chosen threshold roughly corresponds to the $p = 0.05$ threshold for statistical significance popular in frequentist approaches.\todo{CITATION}
-
-A sensible choice of priors is central for evidence estimation via nested sampling.
-As the parameters we are interested in recovering, $S_\mathrm{thresh}$ and $x_{H_2O}$, are poorly constrained by previous data, we used relatively uninformative priors to sample the entire physically plausible parameter space.
-For $S_\mathrm{thresh}$, we chose a uniform prior in $[\prSmin,\prSmax]\, \SI{}{\watt\per\meter\squared}$.
-We sample $x_{H_2O}$ from a log-uniform distribution to imply scale-invariant ignorance.
-Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Sect.~\ref{sec:met_rghmodel}).
-Finally, we adopt a broad, uniform prior for $\langle R_\mathrm{P}\rangle_\mathrm{out}$ bound by $[\prRmin, \prRmax] \, \mathrm{R_\oplus}$.
-We then ran the nested sampler to compute the evidence and sample the posterior distributions.
 

### 2023-03-09 src/tex/ms.tex
PLATO: separate methods, results, discussion

@@
 
+\subsection{Measurement uncertainties}
+Under real-world conditions, the planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
+%\subsubsection{PLATO}
+\plato\ (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
+\plato's definition study report~\citep{plato2017} states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity~(RV) follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).
+For uncertainties in bulk density measurements, we propagate the errors of the mass measurements assuming $\sigma_\mathrm{M_\mathrm{P}} = \SI{10}{\percent}$.
+We base our assessment of \plato's diagnostic power on these estimates.
+
 \subsection{Detection bias, sample selection, and sensitivity}\label{sec:sensitivity} % Turbet2019 lists precision for CHEOPS, PLATO?, TESS?
@@
 To reflect its sensitivity, we chose a minimum transit depth of \SI{80}{\ppm} as a detection limit and consider only measurements of planets exceeding this threshold.
-For uncertainties in bulk density measurements, we propagate normally distributed errors of mass measurements assuming $\sigma_\mathrm{M_\mathrm{P}} = \SI{10}{\percent}$, as expected for \plato.
-\todo[inline]{20230307 after restructuring, check where the above sentence should go}
 We further exclude target stars with \gaia\ magnitudes $M_\mathrm{G} > \minMagnitude$.
@@
 =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
-Under real-world conditions, the planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
-%\subsubsection{PLATO}
-\plato\ (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
-The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).
-In the FGK sample of the mission, several hundred Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
-\plato's definition study report~\citep{plato2017} further states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity~(RV) follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).
-We base our assessment of \plato's diagnostic power on these estimates.
 
@@
 We found that along these axes, \plato\ will be the most favorable among the upcoming transit missions.
-With an expected transit radius precision of \SI{3}{\percent}~\citep{plato2017} and a yield of $> 100$ planets characterized with such high precision~\citep{Rauer2021}, \plato\ is comparable to the optimistic survey we introduced in Sect.~\ref{sec:res_testability}.
+The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).
+%In the FGK sample of the mission, several hundred Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
+With an expected transit radius precision of \SI{3}{\percent}~\citep{plato2017} and a yield of $> 100$ planets studied with such high precision~\citep{Rauer2021}, \plato\ is comparable to the optimistic survey we introduced in Sect.~\ref{sec:res_testability}.
 If successful, it should readily detect the predicted statistical imprint or, in case of a non-detection, provide strong upper limits on the occurrence rate of runaway greenhouse planets.

### 2023-03-10 src/tex/ms.tex
clean up density measurements

@@
 
-\subsection{Measurement uncertainties}
+\subsection{Measurements and their uncertainties}
 Under real-world conditions, the planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
@@
 \plato's definition study report~\citep{plato2017} states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity~(RV) follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).
-For uncertainties in bulk density measurements, we propagate the errors of the mass measurements assuming $\sigma_\mathrm{M_\mathrm{P}} = \SI{10}{\percent}$.
-We base our assessment of \plato's diagnostic power on these estimates.
+We adopted these estimates and assumed a \SI{10}{\percent} error on instellation measurements.
 
-\subsection{Detection bias, sample selection, and sensitivity}\label{sec:sensitivity} % Turbet2019 lists precision for CHEOPS, PLATO?, TESS?
+Since planetary bulk density $\rho \propto R_\mathrm{P}^{-3}$, we expect a stronger runaway greenhouse signal when measured through bulk density instead of radius.
+We thus simulated measurements of planetary densities assuming the mass-radius relation defined above.
+For uncertainties in bulk density measurements, we propagated the errors of the mass measurements assuming $\sigma_\mathrm{M_\mathrm{P}} = \SI{10}{\percent}$.
+
+\subsection{Detection bias, target selection, and sensitivity}\label{sec:sensitivity} % Turbet2019 lists precision for CHEOPS, PLATO?, TESS?
 Not all transiting planets are detectable with the same likelihood and detection biases have an impact on the demographic measurements we are interested in.
@@
 \subsubsection{Additional planet mass measurements}\label{sec:res_followup}
-The dependencies of a detection change when additional information beyond planet radii is available for the characterized planet population.
-Since the planet bulk density $\rho \propto R^{-3}$, the signal is stronger when measured in bulk density instead of radius:
+Comparing a measurement of the signal in radius space with a measurement in density space (which requires planetary mass measurements), a stronger detection occurs in the latter case:
 With an optimistic choice of geophysical parameters (see above), the average measured radius change is \SI{+15}{\percent} whereas the average density change is \SI{-33}{\percent}.
-Constraining bulk densities requires planetary mass measurements.
-We now examine the value of a follow-up campaign to constrain planet masses with respect to the detectability of the runaway greenhouse threshold.
 
+%We now examine the value of a follow-up campaign to constrain planet masses with respect to the detectability of the runaway greenhouse threshold.
 As can be seen in Fig.~\ref{fig:plato_fgrid}, the detectability is enhanced when radius constraints can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
@@
 \subsubsection{The value of follow-up campaigns}
+The dependencies of a detection change when additional information beyond planet radii is available for the characterized planet population.
 As runaway greenhouse phases leave a stronger imprint in bulk density than in planet radius (see Sect.~\ref{sec:res_followup}), it would be beneficial to obtain constraints on planetary masses and test the runaway greenhouse hypothesis in density space instead of radius space.

### 2023-03-10 src/tex/ms.tex
incorporate various comments

@@
 (submitted) were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
-From this catalog, \bioverse\ samples stars within an isotropic distance from the Solar System as required by the sample size.
+From this catalog, \bioverse\ samples stars within an isotropic distance from the Solar System as required by the planetary sample size.
 
@@
 \subsection{Transit probability}
-We model the occurrence of transits by assuming isotropic orientations of planetary orbits, calculating the impact parameters $b = \cos(i)/R_\star$ and considering only observable planets with $|b| < 1$.
+We model the occurrence of transits by assuming isotropic orientations of planetary orbits and calculating the impact parameters $b = \cos(i)/R_\star$.
+Following the approach in \citet{Bixel2021}, we further consider only planets with $|b| < 1$.
 For these cases we calculate the transit depth
@@
 For each planet-level measurement such as radius or instellation, we draw the measured value from a normal distribution centered on the true value with a standard deviation set by the survey's precision.
+We then follow a Bayesian hypothesis testing approach to assess various realizations of simulated surveys in terms of their ability to detect and characterize the runaway greenhouse transition.
 
@@
 
-\subsection{The runaway greenhouse hypothesis}
-We now turn to assessing the simulated surveys in terms of their ability to detect the runaway greenhouse transition and their constraining power on parameters of this trend.
+\subsection{Hypothesis tests}
+We now turn to quantifying the ability of the simulated surveys to detect the runaway greenhouse transition and to constrain parameters of this trend.
 To do this, we rely on a Bayesian hypothesis testing approach where we quantify the evidence of a hypothesis over another based on the (simulated) data.
@@
 For $S_\mathrm{thresh}$, we chose a uniform prior in $[\prSmin,\prSmax]\, \SI{}{\watt\per\meter\squared}$.
-We sample $x_{H_2O}$ from a log-uniform distribution to imply scale-invariant ignorance.
+We sampled $x_{H_2O}$ from a log-uniform distribution to imply scale-invariant ignorance.
 Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Sect.~\ref{sec:met_rghmodel}).
 For $f_\mathrm{rgh}$, we chose a uniform prior in $[0, 1]$.
-Finally, we adopt a broad, uniform prior for $\langle R_\mathrm{P}\rangle_\mathrm{out}$ bound by $[\prRmin, \prRmax] \, \mathrm{R_\oplus}$.
+Finally, we adopted a broad, uniform prior for $\langle R_\mathrm{P}\rangle_\mathrm{out}$ bound by $[\prRmin, \prRmax] \, \mathrm{R_\oplus}$.
+In the case of measuring bulk densities instead of radii, we drew uniformly from $[1, 6] \, \SI{}{\gram\per\centi\meter\cubed}$.
 
@@
 \subsection{Statistical power of different mission designs}\label{sec:statpower_missions}
-To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
-We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
 
@@
 \subsection{Diagnostic power of near-future exoplanet missions}\label{sec:dis_samplesize}
+\todo[inline]{discuss Nautilus}
 The fidelity of a future detection or falsification of a runaway greenhouse signature will depend on the significance with which the null hypothesis can be excluded.
@@
 \todo[inline]{not sure if subsubsections needed}
-We now turn to mission design trades that influence the statistical power for testing the runaway greenhouse hypothesis.
+To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
+We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
 
@@
 \begin{enumerate}
-    \item The predicted tipping point in planetary climates causes a discontinuity in the radius and density distribution of small exoplanets with respect to the incident host star irradiation.
+    \item The predicted runaway greenhouse transition causes a discontinuity in the radius and density distribution of small exoplanets with respect to the incident host star irradiation.
     \item The demographic imprint of the runaway greenhouse threshold should be detectable with high-precision transit measurements.

### 2023-03-10 src/tex/ms.tex
streamline discussion of statistical imprint, move caveats

@@
 \section{Results}
-\subsection{Statistical signature of the runaway greenhouse threshold}
+\subsection{Statistical signature of the runaway greenhouse threshold}\label{sec:res_signature}
 To characterize the population-level imprint of individual radius changes, we generated a generic planet population with an injected runaway greenhouse effect assuming a water fraction $x_\mathrm{H_2O} = \var{wrr}$.
@@
 \subsection{Statistical imprint of exoplanet climates}
-The runaway greenhouse transition is a robust prediction from climate models, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
-Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
-On the planetary level, however, such undertaking is hindered by degeneracies: From internal composition to atmospheric loss to volatile content, many factors influence a planet's size.
-It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019}.
-Using \bioverse\ to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we showed that the population of greenhouse-inflated planets create a signature in the radius and density distribution that is well in the detectable range.
-
-However, the idealized step-function central to our hypothesis tests is blurred by several factors.
-Its amplitude is affected by ``sub-Neptunes'' with significant Hydrogen/Helium envelopes, which can disguise as inflated rocky planets.
-They might not contribute to the demographic signal we are looking for, although a similar transition was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
-A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, turning them from planets appearing inflated into much smaller \mbox{(super-)Earths}.
-Further dilution comes from the population of planets that lack an atmosphere or sufficient volatiles that can create a steam atmosphere.
-There is also a time dependence, both because stellar luminosities change with age and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.
-Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, or surface gravity~\citep{Pierrehumbert2022}.
-
-All of the above mechanisms dilute the demographic imprint of the runaway greenhouse transition with contributions that are challenging to quantify.
-The ``dilution factor" we introduced embodies our ignorance of these contributions.
-Importantly, while this may offset our estimates of the signal's amplitude, its shape remains the same.
+We showed in Sect.~\ref{sec:res_signature} that injecting the theoretically predicted radius inflation effect into a synthetic planet population following the currently known demographics leaves a distinct pattern in the radius and density distribution as a function of instellation.
+Such a statistical imprints of the runaway greenhouse transition was suggested before~\citep{Turbet2019}, and our simulations show quantitatively how the contributions of the individual radius changes combine to create a significant demographic signal.
+Its strength depends on largely unknown factors such as the planetary volatile content, but turned out to be well in the detectable range under reasonable assumptions of these factors.
 
 Our finding that the expected trend is strong except for extremely low water mass fractions not only gives reason for optimism regarding its detection, it also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
-Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
+%Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
+%Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
 Overall, searching for the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.
@@
 We could, with relatively moderate effort, run a sanity check on an alternative sample (e.g., the more structured Bern model populations, or the population in the Genesis database)}
+
 \subsubsection{Baseline mass-radius relationship}
@@
 
+\subsubsection{``Sharpness'' of the demographic trend}
+The real counterpart to our idealized step-function central to our hypothesis tests could be blurred by several factors.
+Its amplitude is affected by ``sub-Neptunes'' with significant Hydrogen/Helium envelopes, which can disguise as inflated rocky planets.
+They might not contribute to the demographic signal we are looking for, although a similar transition was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
+A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, turning them from planets appearing inflated into much smaller \mbox{(super-)Earths}.
+Further dilution comes from the population of planets that lack an atmosphere or sufficient volatiles that can create a steam atmosphere.
+There is also a time dependence, both because stellar luminosities change with age and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.
+Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, or surface gravity~\citep{Pierrehumbert2022}.
+
+All of the above mechanisms dilute the demographic imprint of the runaway greenhouse transition with contributions that are challenging to quantify.
+The ``dilution factor" we introduced embodies our ignorance of these contributions.
+Importantly, while this may offset our estimates of the signal's amplitude, its shape remains the same.
+
 \section{Conclusions}
-
 Runaway greenhouse climates on rocky planets cause significant inflation of their radii.

### 2023-03-10 src/tex/ms.tex
drop content for intro

@@
     ...
+
+    The runaway greenhouse transition is a robust prediction from climate models, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
+Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
+On the planetary level, however, such undertaking is hindered by degeneracies: From internal composition to atmospheric loss to volatile content, many factors influence a planet's size.
+It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019}.
+ Using \bioverse\ to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we...
+
     Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.
+
 \end{note}

### 2023-03-10 src/tex/ms.tex
mention EPRV initiative

@@
 As runaway greenhouse phases leave a stronger imprint in bulk density than in planet radius (see Sect.~\ref{sec:res_followup}), it would be beneficial to obtain constraints on planetary masses and test the runaway greenhouse hypothesis in density space instead of radius space.
-This way, useful constraints can be obtained under more pessimistic conditions, e.g., a low predominant water content of planetary surfaces and atmospheres or a smaller available planet sample.
+This way, useful results can be obtained under more pessimistic conditions, e.g., a low predominant water content of planetary surfaces and atmospheres or a smaller available planet sample.
 For a mission design similar to PLATO, a density-based hypothesis test on about a third of the overall sample is equivalent to a pure radius-based analysis.
@@
 
-However, precise ground-based radial velocity measurements are expensive, and therefore such follow-up efforts may only be available for a subsample of a mission's targets.
-The diagnostic power of such a test may be improved by simultaneously fitting for the trend in radius space in the subsample without RV follow-up.
-An optimized mission in search for the inner edge of the habitable zone will further enhance its information content via a wise selection of follow-up targets, i.e., favoring those planets residing close to the expected instellation threshold $S_\mathrm{thresh}$.
+Precise ground-based radial velocity measurements will be needed to provide these data, and a number of instruments are already successfully employed in characterizing terrestrial-sized exoplanets~\citep[e.g.,][]{Queloz2001a,Pepe2010,Johnson2010b,Ribas2023}.
+Recently, the National Aeronautics and Space Administration (NASA) and the National Science Foundation (NSF) commissioned an ``Extreme Precision Radial Velocity (EPRV) Initiative''~\citep{Crass2001} to develop methods and facilities for precise mass measurements of temperate terrestrial planets.
+Their findings highlight that such measurements are costly, and therefore follow-up efforts may only be available for a subsample of the targets of a mission of \plato's scale.
+The diagnostic power of the hypothesis tests we demonstrated here may be improved by simultaneously fitting for the trend in radius space in the subsample without RV follow-up.
+An optimized mission in search for the inner edge of the habitable zone will further enhance its information content via a wise selection of follow-up targets, i.e., balancing objects located on either side of the expected instellation threshold $S_\mathrm{thresh}$.
 

### 2023-03-11 src/tex/ms.tex
swap detection bias and measurements

@@
 
-\subsection{Measurements and their uncertainties}
-Under real-world conditions, the planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
-%\subsubsection{PLATO}
-\plato\ (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
-\plato's definition study report~\citep{plato2017} states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity~(RV) follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).
-We adopted these estimates and assumed a \SI{10}{\percent} error on instellation measurements.
-
-Since planetary bulk density $\rho \propto R_\mathrm{P}^{-3}$, we expect a stronger runaway greenhouse signal when measured through bulk density instead of radius.
-We thus simulated measurements of planetary densities assuming the mass-radius relation defined above.
-For uncertainties in bulk density measurements, we propagated the errors of the mass measurements assuming $\sigma_\mathrm{M_\mathrm{P}} = \SI{10}{\percent}$.
-
 \subsection{Detection bias, target selection, and sensitivity}\label{sec:sensitivity} % Turbet2019 lists precision for CHEOPS, PLATO?, TESS?
@@
 %In comparison, \plato's expected precision for similar stars was estimated to be \SI{\sim 50}{\ppm}~\citep[][Matuszewski et al., in prep.]{plato2017}.
-In line with its requirement to detect Earth-sized planets in a \SI{1}{\year} orbit around Sun-like stars, \plato\ is designed to enable the detection of a \SI{80}{\ppm} transit signal~\citep[][Matuszewski et al., in prep.]{plato2017}.
+\plato\ (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
+In line with this requirement, \plato\ is designed to enable the detection of a \SI{80}{\ppm} transit signal~\citep[][Matuszewski et al., in prep.]{plato2017}.
 To reflect its sensitivity, we chose a minimum transit depth of \SI{80}{\ppm} as a detection limit and consider only measurements of planets exceeding this threshold.
@@
 We further consider only rocky planets with masses below \SI{2}{\Mearth}. % (Turbet+2020 covers only 0.1 to 2 Mearth)
+
+\subsection{Measurements and their uncertainties}
+Under real-world conditions, the planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
+%\subsubsection{PLATO}
+\plato's definition study report~\citep{plato2017} states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity~(RV) follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).
+We adopted these estimates and assumed a \SI{10}{\percent} error on instellation measurements.
+
+Since planetary bulk density $\rho \propto R_\mathrm{P}^{-3}$, we expect a stronger runaway greenhouse signal when measured through bulk density instead of radius.
+We thus simulated measurements of planetary densities assuming the mass-radius relation defined above.
+For uncertainties in bulk density measurements, we propagated the errors of the mass measurements assuming $\sigma_\mathrm{M_\mathrm{P}} = \SI{10}{\percent}$.
 

### 2023-03-11 src/tex/ms.tex
discuss Nautilus Observatory concept

@@
 
-Taken together, \plato\ appears to be the most promising exoplanet mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.
+Looking further ahead, the Nautilus Space Observatory concept~\citep{Apai2019} represents a statistical mission able to provide precise radius measurements of a large sample of small exoplanets.
+It employs a swarm of large telescopes using ultralight diffractive-refractive optical elements~\citep{Milster2020} with the primary goal to study the atmospheres of transiting exoplanets.
+Operating in an array mode, Nautilus would achieve the equivalent light-collecting area of a \SI{50}{\meter} telescope.
+Its expected \SI{1}{\ppm} photometric precision~\citep{Apai2022} would enable precise radius measurements of a large sample, also through a low number of required visits per object.
+If realized, Nautilus will be a valuable instrument for characterizing the runaway greenhouse transition.
+
+As for exoplanets missions in their implementation phase, \plato\ overall appears to be the most promising mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.
 

### 2023-03-13 src/tex/ms.tex
occurrence model caveats

--- 2023-03-11 src/tex/ms.tex
@@
 % Define document class
-%\documentclass[twocolumn]{aastex631}
-\documentclass[modern]{aastex631}
+\documentclass[twocolumn]{aastex631}
+%\documentclass[modern]{aastex631}
 
@@
 ...just dropping material and references for now...
-\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
+\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
 \todo[inline]{``Points for the intro:
@@
 ~\\
-In contrast, PLATO and Nautilus are \\
+In contrast, PLATO and \nautilus\ are \\
 Possibility of follow-up to measureme masses
@@
 In this section, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, as well as the orbital parameters of the planets.
+An overview of our key assumptions and model parameters can be found in Table~\ref{tab:params_table}.
 
@@
 \subsection{Synthetic planet sample}\label{sec:syn_planets}
-
 Next, we assigned to the stellar sample planetary systems with frequencies, orbital parameters, and bulk properties derived from the \kepler\ mission.
-We adopted the planetary occurrence rate model from \citet{Bergsten2022}, which defines the occurrence of small planets in radius and orbital period.
+We adopted the planetary occurrence rate model from~\citet{Bergsten2022}, which defines the occurrence of small planets in radius and orbital period.
 Following \citet{Youdin2011a}, their inferred occurrence rate density can be expressed in the form:
@@
 In summary, stars are assigned a number of planets in accordance with $F_0$, with planet parameters randomly drawn in proportion to the occurrence distribution function evaluated over a grid of radii and orbital periods.
-% \todo[inline]{GALEN: Briefly describe underlying planet occurrence rates of synthetic population, its Mstar dependency, any caveats or assumptions.}
 \todo[inline]{GALEN: The most relevant assumption of using our model is that, at the relevant fluxes (I think), we predict that super-Earths are pretty common.
@@
 
-\subsection{Orbit parameters and planet masses}
+\subsection{Orbit parameters and planet masses}\label{sec:met-orbits_masses}
 Eccentric orbits alter the probability of a planet to transit \citep[e.g.,][]{Barnes2007a}.
@@
 
-For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$, we assume an inflated radius due to a steam atmosphere.
+For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
 While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013,Boukrouche2021}.
@@
 
-\section{Exoplanet Survey Simulations and Hypothesis Testing}
+\section{Exoplanet Survey Simulations and Hypothesis Testing}\label{sec:met_surveys-hypotests}
 The survey module of \bioverse\ converts the synthetic planet sample into a set of uncertainty-laden measurements on a subset of that sample.
@@
 %    We thus chose a broad selection function centered on general detectability of transits with state-of-the-art technology.
-The runaway greenhouse effect becomes obsolete where no atmosphere can be maintained due to proximity to the host star and resulting atmospheric erosion.
-Ensuring to stay well clear of such regions, we clear our sample from all planets with a net instellation $S > \SI{2000}{\watt\per\square\meter}$.
+The runaway greenhouse effect becomes obsolete both for very small instellations and where no atmosphere can be maintained due to proximity to the host star and resulting atmospheric erosion.
+Ensuring to stay well clear of such regions, we clear our sample from all planets with a net instellation $S < \SI{10}{\watt\per\square\meter}$ or $S > \SI{2000}{\watt\per\square\meter}$.
 %\todo[inline]{use criterion of Zahnle \& Caitling 2017 instead? $0.8 S^{0.25} < R$}
@@
 For our specific problem, this implies comparing evidences for a demographic imprint of the runaway greenhouse effect to its absence.
-\begin{note}
+%\begin{note}
 %    Bayesian model comparison requires that an alternative mode be specified against which the comparison is made.
-\end{note}
+%\end{note}
     As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
@@
 $H_{\mathrm{rgh}}$ and $H_0$ should thus be tested against a statistical estimator that represents the population mean.
-To avoid binning and the artificial patterns it may introduce, we chose to test our hypotheses against a simple moving average $SMA$ along the instellation axis with a window of size \windowsize\ centered around each measurement.
+To avoid binning and the artificial patterns it may introduce, we chose to test our hypotheses against a simple moving average $SMA$ along the instellation axis with a window of size \windowsize\ centered around each measurement.\footnote{See Appendix~\ref{app:binnedstats} for a robustness test using a different estimator.}
 We further computed the uncertainty of this moving average by propagating the individual measurement errors and applying a rolling standard error of the mean. \todo[inline]{specify error propagation}
@@
 
+% include table with key parameters
+\input{table_parameters}
+
 \section{Results}
@@
         \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
-        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation.
-Only the planets marked as transiting are observable.
-Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, the radii of a fraction of planets are inflated by the amount indicated with gray lines.
-The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue).
-This runaway greenhouse hypothesis can be tested against the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
+        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation. Only the planets marked as transiting are observable.
+            Above the runaway greenhouse threshold $S_\mathrm{thresh} = \var{injectedSthresh}$, some planets maintain their original radii (green crosses) whereas some have their radii inflated (blue crosses) by the amount indicated with gray lines.
+        The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue line).
+            This runaway greenhouse hypothesis can be tested against the null hypothesis $H_0$ (green line), where average radii are independent of instellation.}
         \label{fig:HnullHmo}
@@
 %\todo{Introduce survey simulation(s).
-PLATO, Cheops, (Ariel, LIFE/LIFE precursor, Nautilus)\\ What are threshold missions that are able to detect this signal?
+PLATO, Cheops, (Ariel, LIFE/LIFE precursor, \nautilus\)\\ What are threshold missions that are able to detect this signal?
 =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
@@
 
-Our finding that the expected trend is strong except for extremely low water mass fractions not only gives reason for optimism regarding its detection, it also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
+Our finding that the expected trend is strong except for extremely low water mass fractions gives reason for optimism regarding its detection.
+It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
+Such constraints will shed light on whether the initial water content is variable~\citep{Lichtenberg2019} or there is a predominant quantity.
+Ultimately, such measurements will shed light on the question of if planets like Earth received their water from primordial sources during their formation or if it was supplied later~\citep[e.g.,][]{Raymond2004}.
 %Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
@@
 
-\subsubsection{False positive scenarios}
-\todo{what other processes could be confused with a magma ocean signal?\\
-If there's no other subsubsection here, make this a (the final?) subsection in ``Discussion''.}
-\begin{note}
+\subsubsection{False positive scenarios}\label{sec:dis_falsepositive}
 Runaway greenhouse climates are not the only physical mechanism that may cause a change in transit radius for a subset of planets.
@@
 Although an abrupt pattern at the expected runaway greenhouse transition seems unlikely, examples of steep occurrence rate density changes exist.
-    An example is the "Neptune desert``, a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
+An example is the ``Neptune desert'', a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
 The shape of this region is such that smaller planets become less frequent the closer to the star they are, which to some degree resembles the pattern introduced by the instellation dependency of the runaway greenhouse transition.
@@
 \todo{is there overlap/confusion with the radius/density trend in \citet{Luque2022}?}
-\end{note}
 
-%\todo[inline]{Maps to Sect. \ref{sec:statpower_missions}.:}
-%\begin{note}
-%    \citet{Penny2019} mention that a significant increase in planet yield could be achieved if the telescope's slew speed was increased.
-%\end{note}
-
-\subsubsection{OPTIONAL...: Atmospheric signatures}
+\subsubsection{OPTIONAL...: Atmospheric signatures}\label{sec:dis_atmospheres}
 \todo[inline]{TIM: Ideas?\\
@@
 \subsection{Diagnostic power of near-future exoplanet missions}\label{sec:dis_samplesize}
-\todo[inline]{discuss Nautilus}
 The fidelity of a future detection or falsification of a runaway greenhouse signature will depend on the significance with which the null hypothesis can be excluded.
@@
 What other planned missions are suited to test the runaway greenhouse hypothesis?
+\kepler\ and \ktwo\ have contributed a large number of discovered terrestrial-sized planets, but few of them are in the habitable zone and their host stars are typically too faint for RV follow-up~\citep{Dressing2015}.
+
+Similarly, the Transiting Exoplanet Survey Satellite~\citep[\tess,][]{Ricker2014a} planet sample lacks temperate, small planets around bright host stars~\citep{Ment2023}, as was expected from planet yield calculations~\citep{Barclay2018}.
+
 The ongoing \cheops\ mission was designed as a follow-up mission to search for transits of planets discovered with other techniques, in particular with radial velocity measurements~\citep{Benz2021}.
@@
 
-As \cheops, \ariel~\citep{Puig2016} will be a follow-up mission that is not designed to provide a large number of new radius measurements.
+As \cheops, the Atmospheric Remote sensing Infrared Exoplanet Large survey~\citep[\ariel,][]{Puig2016} will be a follow-up mission that is not designed to provide a large number of new radius measurements.
 \ariel's primary targets are larger planets in the range of sub-Neptune to Jupiter-like planets.
-We thus do not expect a significant contribution to exploring the inner edge of the habitable zone for Earth-sized planets.
+We thus do not expect a significant contribution to statistically exploring the inner edge of the habitable zone for Earth-sized planets.
 
@@
 The \rst\ could thus provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.
-\todo[inline]{what about TESS?}
 
-Taken together, \plato\ appears to be the most promising exoplanet mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.
+Looking further ahead, the \nautilus\ Space Observatory concept~\citep{Apai2019} represents a statistical mission able to provide precise radius measurements of a large sample of small exoplanets.
+It employs a swarm of large telescopes using ultralight diffractive-refractive optical elements~\citep{Milster2020} with the primary goal to study the atmospheres of transiting exoplanets.
+Operating in an array mode, \nautilus\ would achieve the equivalent light-collecting area of a \SI{50}{\meter} telescope.
+Its expected \SI{1}{\ppm} photometric precision~\citep{Apai2022} would enable precise radius measurements of a large sample, also through a low number of required visits per object.
+If realized, \nautilus\ will be a valuable instrument for characterizing the runaway greenhouse transition.
+
+Other mission concepts exist that focus on characterizing exoplanet habitability, most notably the \hwo~\citep{Gaudi2020c,LUVOIR2019} and the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2021}.
+Such direct imaging surveys, however, do not directly measure planetary radii and are therefore most useful for providing context through atmospheric measurements of individual planets.
+\todo[inline]{TIM: feel free to add something about LIFE if you see fit}
+
+As for exoplanets missions in their implementation phase, \plato\ overall remains to be the most promising mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.
 
@@
 Precise ground-based radial velocity measurements will be needed to provide these data, and a number of instruments are already successfully employed in characterizing terrestrial-sized exoplanets~\citep[e.g.,][]{Queloz2001a,Pepe2010,Johnson2010b,Ribas2023}.
-Recently, the National Aeronautics and Space Administration (NASA) and the National Science Foundation (NSF) commissioned an ``Extreme Precision Radial Velocity (EPRV) Initiative''~\citep{Crass2001} to develop methods and facilities for precise mass measurements of temperate terrestrial planets.
+Recently, the National Aeronautics and Space Administration (NASA) and the National Science Foundation (NSF) commissioned an ``Extreme Precision Radial Velocity (EPRV) Initiative''~\citep{Crass2021} to develop methods and facilities for precise mass measurements of temperate terrestrial planets.
 Their findings highlight that such measurements are costly, and therefore follow-up efforts may only be available for a subsample of the targets of a mission of \plato's scale.
@@
 influences the planet's climate (greenhouse effect).}
-A potential for liquid water on the surface of a planet is commonly used as an environmental marker to assess its habitability. REF
+A potential for liquid water on the surface of a planet is commonly used as an environmental marker to assess its habitability~\citep{Huang1959,Hart1978,Kasting1993,Kaltenegger2011,Kopparapu2013}.
 The runaway greenhouse transition represents an upper bound on received irradiation for this condition.
@@
 In Sect.~\ref{sec:res_constrain-S_thresh}, we show that the threshold instellation at which the runaway greenhouse transition occurs can be reasonably constrained without imposing overly optimistic conditions on the underlying planet population, instrumentation, or survey strategy.
-\todo[inline]{elaborate based on Sect.~\ref{sec:res_constrain-S_thresh}}
+A mission like \plato\ is well equipped to perform this measurement; the chances for success are directly proportional to the proportion of characterized planets around M dwarfs and to the number of planets for which masses can be determined.
 
@@
 Since these parameters are degenerate, they cannot be well constrained without independent measurements.
-\todo[inline]{discuss additional measurements that could lift the degeneracy. e.g., adding atmospheric compositional information i.e., PLATO + ARIEL or similar}
+This degeneracy could be lifted if independent measurements of atmospheric compositions can be made.
+For example, detections of water vapor in planets above the threshold instellation, combined with precise radius measurements, would constrain the predominant water content of terrestrial planets.
+A follow-up campaign with the \ariel\ mission of \plato\ planets could be considered to make this contribution.
+
+Once a runaway greenhouse region is identified in the parameter space, the community will will have a tool at hand to discern potentially habitable planets from Venusian worlds on an empirical basis.
+Together with atmospheric measurements (see Sect.~\ref{sec:dis_atmospheres}), we will be able to put a number on the probability of an individual planet to harbor sufficient surface water to sustain life.
 
@@
 
+\subsubsection{Underlying occurrence rate distributions}
+
+Our adopted occurrence rate model \citep{Bergsten2022} changes forms at larger separations ($> \SI{50}{\watt\per\meter\squared}$) and radii ($> 1.6\,R_\oplus$) than what is relevant for the runaway greenhouse threshold, where the adopted occurrence rate density is relatively smooth.
+Thus we argue that the proposed trend is not an artifact of the underlying occurrence distribution.
+However, this may be different for alternative models that have sharp features in relevant regions of $R_P - S$ space. 
+\todo[inline]{More discussion goes here?}
+
 \subsubsection{Baseline mass-radius relationship}
@@
 We thus consider our mass-radius relationship a conservative case and do not expect a negative impact on the detectability of the runaway greenhouse transition due to deviations from it.
-\todo[inline]{if feasible: run a test with a different M-R relation. do we expect that the choice of M-R relation impacts the results? Perhaps have an appendix section.}
+Nevertheless, we performed a sanity check to assess the influence varying our baseline model on our results (see Appendix~\ref{app:MR_relation}), and found general agreement between different interior composition models.
 
@@
 There is also a time dependence, both because stellar luminosities change with age and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.
+
 Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, or surface gravity~\citep{Pierrehumbert2022}.
+The choice of statistical estimator for the hypothesis tests may further blur the recovered discontinuity.
+We compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.
 
@@
           Only an adequate sample of planets orbiting M~dwarfs will ensure sufficient targets on both sides of the expected threshold instellation.
-    \item Testing the runaway greenhouse hypothesis will constrain the water inventory of rocky exoplanets, providing important context in the assessment of their habitability.
+    \item Testing the runaway greenhouse hypothesis on a population level can provide constraints on the water inventory of rocky exoplanets and thus make an important contribution to assessing their habitability.
          A detection will provide an empirical confirmation of the habitable zone concept and localize its inner edge.
@@
 
+\appendix
+\section{Robustness tests}
+\subsection{Alternative statistics for the average radius or bulk density}\label{app:binnedstats}
+The hypothesis tests introduced in Sect.~\ref{sec:met_surveys-hypotests} rely on a statistical estimator for the variation of planetary radii or bulk densities as a function of net instellation, and we chose a moving average for this estimator in our nominal setup.
+Here, we explore how robust our results are against this choice by demonstrating the recovery of the runaway greenhouse signal in the case of the optimistic survey (Sect.~\ref{sec:res_testability}) with an alternative estimator: instead of computing moving averages, we used a binned statistic.
+
+We first binned the data of the simulated survey in instellation space, choosing the number of bins via the rule of \citet{Freedman1981} and using logarithmic binning.
+In each bin, we computed the arithmetic mean of the planet radius and its standard deviation.
+Then, we assigned each planet the mean radius according to the instellation bin it occupies and used this as the measure for testing the runaway greenhouse hypothesis.
+
+\begin{figure}[ht!]
+%    \script{figurescript.py}
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/optimistic_R-S_binned.pdf}
+        \caption{
+            Detection of the runaway greenhouse threshold with a binned statistic.
+            Using simulated data from the optimistic survey case ($N = \var{N_optimistic}$, compare Fig.~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis.
+Instead of using a running mean, we computed a binned statistic (arithmetic mean and standard deviation in blue) to assign planets the average radius or bulk density in their neighborhood in instellation space.
+Random draws from the posterior of the runaway greenhouse hypothesis (Eqn.~\ref{eq:rgh_hypo}) are shown in red.
+            As in the nominal case, the pattern is detected with high significance.
+        }
+        \label{fig:optimistic_R-S_binned}
+    \end{centering}
+\end{figure}
+Figure~\ref{fig:optimistic_R-S_binned} shows the simulated data together with binned, average planet radii and draws from the posterior of the hypothesis test.
+A clear detection resulted, although with somewhat lower significance ($\Delta \ln \mathcal{Z} \approx 30$) compared to the nominal setup.
+The accuracy of the recovered instellation threshold is comparable.
+This test demonstrates that our results are not sensitive on the choice of statistical estimator to test the hypotheses against.
+
+It is conceivable that with very large sample sizes and a very sharp runaway greenhouse transition a binned solution would perform better.
+For a search with real data, both approaches, and possibly other alternatives, should be considered.
+
+\subsection{Influence of different mass-radius relationships on our results}\label{app:MR_relation}
+We assessed how a different choice of baseline mass-radius relation influences our results.
+Focussing on the detectability of the statistical runaway greenhouse signal and the ability to constrain the threshold instellation, we repeated the hypothesis test in Sect.~\ref{sec:res_testability} with alternative mass-radius relationships.
+Instead of assuming a pure $\mathrm{MgSiO_3}$ composition, we assigned planet masses using either the probabilistic relationship in \citet{Wolfgang2016} or a semi-empirical, two-layer relation assuming an Earth-like (\SI{32.5}{\percent} Fe + \SI{67.5}{\percent} MgSiO$_3$) composition~\citep{Zeng2016}.
+
+\begin{figure}[ht!]
+%    \script{figurescript.py}
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/MR-violins.pdf}
+        \caption{
+        Posterior threshold instellations as a function of dilution factor for two alternative baseline mass-radius relations.
+        For each grid step in $f_\mathrm{rgh}$, we show kernel density estimates of retrieved posteriors (averaged over 50 iterations) assuming an Earth-like (\SI{32.5}{\percent} Fe + \SI{67.5}{\percent} MgSiO$_3$) composition or the probabilistic relation from \citet{Wolfgang2016} and otherwise following the optimistic scenario in Sect.~\ref{sec:res_testability}.
+        Lines within the violins show quartiles of the distributions, and the gray line indicates the injected threshold instellation of \var{injectedSthresh}.
+        Higher log-evidence differences $\Delta \ln Z$ correspond to more significant rejections of the null hypothesis.
+        In the regime of strong detections, both mass-radius relations lead to similar, accurate constraints on $S_\mathrm{thresh}$.
+        Differences occur at low dilution factors, where the Earth-like relation leads to narrower estimates.
+        }
+        \label{fig:MR-violins}
+    \end{centering}
+\end{figure}
+Figure~\ref{fig:MR-violins} shows how the two alternative baseline mass-radius relations influence the significance of a detection and the ability to constrain the threshold instellation.
+The relation of \citet{Wolfgang2016} includes intrinsic scatter, which impedes a detection at low dilution factors.
+In this regime, a layered, Earth-like composition leads to more significant detections and a narrower, although biased, constraint on $S_\mathrm{thresh}$.
+Both mass-radius relations agree and recover the injected value where the null hypothesis can be rejected with high significance.
+This is consistent with our nominal mass-radius relation (compare Sect.~\ref{sec:met-orbits_masses}).
+We conclude that the underlying core and mantle composition of planets may affect the detectability of the transition if the fraction of planets with runaway greenhouse climates is low; however, the overall trends that our experiments revealed appear robust.
+
+We caution that this analysis may serve only as a sanity check and should not be taken as a result in itself: The atmospheric model from \citet{Turbet2020} we adopted relies on a silicate interior composition for its transit radius prediction.
+Therefore, only our nominal procedure throughout the main body of the paper represents a self-consistent treatment.
+
 \bibliographystyle{aasjournal}
 \bibliography{bib,PhD,coauthors}
+
 \end{document}

### 2023-03-14 src/tex/ms.tex
Edit, provide context to occurrence rate discussion

@@
 
-\subsubsection{Underlying occurrence rate distributions}
-
-Our adopted occurrence rate model \citep{Bergsten2022} changes forms at larger separations ($> \SI{50}{\watt\per\meter\squared}$) and radii ($> 1.6\,R_\oplus$) than what is relevant for the runaway greenhouse threshold, where the adopted occurrence rate density is relatively smooth.
-Thus we argue that the proposed trend is not an artifact of the underlying occurrence distribution.
-However, this may be different for alternative models that have sharp features in relevant regions of $R_P - S$ space. 
-\todo[inline]{More discussion goes here?}
+\subsubsection{Structures in the planet occurrence rate density}
+In principle, the baseline occurrence rate density in radius-period space that governs the generation of our synthetic planet sample might influence our findings, especially if it contains any features that coincide with the injected demographic feature.
+This is not the case in the model from \citet{Bergsten2022} that we adopted: Its occurrence rate density is smooth in the domain relevant for the runaway greenhouse transition; discontinuities only occur at smaller instellations ($< \SI{50}{\watt\per\meter\squared}$) and larger radii ($> 1.6\,R_\oplus$).
+Due to the flat occurrence rate in the region of interest, we do not expect the model underlying our planet sample to affect the recovery of the transition.
+However, it should be kept in mind that any unknown sharp features in the true distribution of terrestrial planets could complicate a detection or lead to a false positive result (see Sect.~\ref{sec:dis_falsepositive}).
 

### 2023-03-14 src/tex/ms.tex
discuss Kepler, K2, TESS

--- 2023-03-11 src/tex/ms.tex
@@
 
-\subsubsection{False positive scenarios}
+\subsubsection{False positive scenarios}\label{sec:dis_falsepositive}
 \todo{what other processes could be confused with a magma ocean signal?\\
@@
 What other planned missions are suited to test the runaway greenhouse hypothesis?
+\kepler\ and \ktwo\ have contributed a large number of discovered terrestrial-sized planets, but few of them are in the habitable zone and their host stars are typically too faint for RV follow-up~\citep{Dressing2015}.
+
+Similarly, the \tess\ planet sample lacks temperate, small planets around bright host stars~\citep{Ment2023}, as was expected from planet yield calculations~\citep{Barclay2018}.
+
 The ongoing \cheops\ mission was designed as a follow-up mission to search for transits of planets discovered with other techniques, in particular with radial velocity measurements~\citep{Benz2021}.
@@
 The \rst\ could thus provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.
-\todo[inline]{what about TESS?}
 
-Looking further ahead, the Nautilus Space Observatory concept~\citep{Apai2019} represents a statistical mission able to provide precise radius measurements of a large sample of small exoplanets.
+Looking further ahead, the \nautilus\ Space Observatory concept~\citep{Apai2019} represents a statistical mission able to provide precise radius measurements of a large sample of small exoplanets.
 It employs a swarm of large telescopes using ultralight diffractive-refractive optical elements~\citep{Milster2020} with the primary goal to study the atmospheres of transiting exoplanets.
-Operating in an array mode, Nautilus would achieve the equivalent light-collecting area of a \SI{50}{\meter} telescope.
+Operating in an array mode, \nautilus\ would achieve the equivalent light-collecting area of a \SI{50}{\meter} telescope.
 Its expected \SI{1}{\ppm} photometric precision~\citep{Apai2022} would enable precise radius measurements of a large sample, also through a low number of required visits per object.
-If realized, Nautilus will be a valuable instrument for characterizing the runaway greenhouse transition.
+If realized, \nautilus\ will be a valuable instrument for characterizing the runaway greenhouse transition.
 
-As for exoplanets missions in their implementation phase, \plato\ overall appears to be the most promising mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.
+As for exoplanets missions in their implementation phase, \plato\ overall remains to be the most promising mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.
 

### 2023-03-14 src/tex/ms.tex
mention HWO, LIFE

@@
     Although an abrupt pattern at the expected runaway greenhouse transition seems unlikely, examples of steep occurrence rate density changes exist.
-    An example is the "Neptune desert``, a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
+    An example is the ``Neptune desert'', a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
     The shape of this region is such that smaller planets become less frequent the closer to the star they are, which to some degree resembles the pattern introduced by the instellation dependency of the runaway greenhouse transition.
@@
 
-Similarly, the \tess\ planet sample lacks temperate, small planets around bright host stars~\citep{Ment2023}, as was expected from planet yield calculations~\citep{Barclay2018}.
+Similarly, the Transiting Exoplanet Survey Satellite~\citep[\tess,][]{Ricker2014a} planet sample lacks temperate, small planets around bright host stars~\citep{Ment2023}, as was expected from planet yield calculations~\citep{Barclay2018}.
 
@@
 
-As \cheops, \ariel~\citep{Puig2016} will be a follow-up mission that is not designed to provide a large number of new radius measurements.
+As \cheops, the Atmospheric Remote sensing Infrared Exoplanet Large survey~\citep[\ariel,][]{Puig2016} will be a follow-up mission that is not designed to provide a large number of new radius measurements.
 \ariel's primary targets are larger planets in the range of sub-Neptune to Jupiter-like planets.
@@
 If realized, \nautilus\ will be a valuable instrument for characterizing the runaway greenhouse transition.
+
+Other mission concepts exist that focus on characterizing exoplanet habitability, most notably the \hwo~\citep{Gaudi2020c,LUVOIR2019} and the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2021}.
+Such direct imaging surveys, however, do not directly measure planetary radii and are therefore most useful for providing context through atmospheric measurements of individual planets.
+\todo[inline]{TIM: feel free to add something about LIFE if you see fit}
 

### 2023-03-14 src/tex/ms.tex
discuss additional atmospheric measurements

@@
 
-\subsubsection{OPTIONAL...: Atmospheric signatures}
+\subsubsection{OPTIONAL...: Atmospheric signatures}\label{sec:dis_atmospheres}
 \todo[inline]{TIM: Ideas?\\
@@
 \ariel's primary targets are larger planets in the range of sub-Neptune to Jupiter-like planets.
-We thus do not expect a significant contribution to exploring the inner edge of the habitable zone for Earth-sized planets.
+We thus do not expect a significant contribution to statistically exploring the inner edge of the habitable zone for Earth-sized planets.
 

### 2023-03-14 src/tex/ms.tex
constrain initial water content

@@
 
-Our finding that the expected trend is strong except for extremely low water mass fractions not only gives reason for optimism regarding its detection, it also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
+Our finding that the expected trend is strong except for extremely low water mass fractions gives reason for optimism regarding its detection.
+It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
+Such constraints will shed light on whether the initial water content is variable~\citep{Lichtenberg2019} or there is a predominant quantity.
+Ultimately, such measurements will shed light on the question of whether planets like Earth received their water from primordial sources during their formation or whether it was supplied later~\citep[e.g.,][]{Raymond2004}.
 %Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.

### 2023-03-14 src/tex/ms.tex
update synthetic planets/hypotheses figure

@@
         \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
-        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation.
-Only the planets marked as transiting are observable.
-Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, the radii of a fraction of planets are inflated by the amount indicated with gray lines.
-The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue).
-This runaway greenhouse hypothesis can be tested against the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
+        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation. Only the planets marked as transiting are observable.
+            Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, some planets maintain their original radii (green crosses) whereas some have their radii inflated (blue crosses) by the amount indicated with gray lines.
+        The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue line).
+            This runaway greenhouse hypothesis can be tested against the null hypothesis $H_0$ (green line), where average radii are independent of instellation.}
         \label{fig:HnullHmo}

### 2023-03-15 src/tex/ms.tex
fill gaps in Discussion

@@
 Such constraints will shed light on whether the initial water content is variable~\citep{Lichtenberg2019} or there is a predominant quantity.
-Ultimately, such measurements will shed light on the question of whether planets like Earth received their water from primordial sources during their formation or whether it was supplied later~\citep[e.g.,][]{Raymond2004}.
+Ultimately, such measurements will shed light on the question of if planets like Earth received their water from primordial sources during their formation or if it was supplied later~\citep[e.g.,][]{Raymond2004}.
 %Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
@@
 \subsubsection{False positive scenarios}\label{sec:dis_falsepositive}
-\todo{what other processes could be confused with a magma ocean signal?\\
-If there's no other subsubsection here, make this a (the final?) subsection in ``Discussion''.}
-\begin{note}
 Runaway greenhouse climates are not the only physical mechanism that may cause a change in transit radius for a subset of planets.
@@
 \todo{is there overlap/confusion with the radius/density trend in \citet{Luque2022}?}
-\end{note}
-
-%\todo[inline]{Maps to Sect. \ref{sec:statpower_missions}.:}
-%\begin{note}
-%    \citet{Penny2019} mention that a significant increase in planet yield could be achieved if the telescope's slew speed was increased.
-%\end{note}
 

### 2023-03-16 src/tex/ms.tex
add a robustness test for the statistical estimator

@@
 
+\appendix
+\section{Robustness tests}
+\subsection{Alternative statistics for the average radius or bulk density}\label{app:binnedstats}
+The hypothesis tests introduced in Sect.~\ref{sec:met_surveys-hypotests} rely on a statistical estimator for the variation of planetary radii or bulk densities as a function of net instellation, and we chose a moving average for this estimator in our nominal setup.
+Here, we explore how robust our results are against this choice by demonstrating the recovery of the runaway greenhouse signal in the case of the optimistic survey (Sect.~\ref{sec:res_testability}) with an alternative estimator: instead of computing moving averages, we used a binned statistic.
+
+We first binned the data of the simulated survey in instellation space, choosing the number of bins via the rule of \citet{Freedman1981} and using logarithmic binning.
+In each bin, we computed the arithmetic mean of the planet radius and its standard deviation.
+Then, we assigned each planet the mean radius according to the instellation bin it occupies and used this as the measure for testing the runaway greenhouse hypothesis.
+
+\begin{figure}[ht!]
+%    \script{figurescript.py}
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/optimistic_R-S_binned.pdf}
+        \caption{
+            Detection of the runaway greenhouse threshold with a binned statistic.
+            Using simulated data from the optimistic survey case ($N = \var{N_optimistic}$, compare Fig.~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis.
+Instead of using a running mean, we computed a binned statistic (arithmetic mean and standard deviation in blue) to assign planets the average radius or bulk density in their neighborhood in instellation space.
+Random draws from the posterior of the runaway greenhouse hypothesis (Eqn.~\ref{eq:rgh_hypo}) are shown in red.
+            As in the nominal case, the pattern is detected with high significance.
+        }
+        \label{fig:optimistic_R-S_binned}
+    \end{centering}
+\end{figure}
+Figure~\ref{fig:optimistic_R-S_binned} shows the simulated data together with binned, average planet radii and draws from the posterior of the hypothesis test.
+A clear detection resulted, although with somewhat lower significance ($\Delta \ln \mathcal{Z} \approx 30$) compared to the nominal setup.
+The accuracy of the recovered instellation threshold is comparable.
+This test demonstrates that our results are not sensitive on the choice of statistical estimator to test the hypotheses against.
+
+It is conceivable that with very large sample sizes and a very sharp runaway greenhouse transition a binned solution would perform better.
+For a search with real data, both approaches, and possibly other alternatives, should be considered.
+
 \bibliographystyle{aasjournal}
 \bibliography{bib,PhD,coauthors}
+
 \end{document}

### 2023-03-17 src/tex/ms.tex
test different M-R relations, add plot and discuss differences

@@
         \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation. Only the planets marked as transiting are observable.
-            Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, some planets maintain their original radii (green crosses) whereas some have their radii inflated (blue crosses) by the amount indicated with gray lines.
+            Above the runaway greenhouse threshold $S_\mathrm{thresh} = \var{injectedSthresh}$, some planets maintain their original radii (green crosses) whereas some have their radii inflated (blue crosses) by the amount indicated with gray lines.
         The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue line).
@@
 
+\subsection{Influence of different mass-radius relationships on our results}\ref{app:MR_relation}
+We assessed how a different choice of baseline mass-radius relation influences our results.
+Focussing on the detectability of the statistical runaway greenhouse signal and the ability to constrain the threshold instellation, we repeated the hypothesis test in Sect.~\ref{sec:res_testability} with alternative mass-radius relationships.
+Instead of assuming a pure $\mathrm{MgSiO_3}$ composition, we assigned planet masses using either the probabilistic relationship in \citet{Wolfgang2016} or a semi-empirical, two-layer relation assuming an Earth-like (\SI{32.5}{\percent} Fe + \SI{67.5}{\percent} MgSiO$_3$) composition~\citep{Zeng2016}.
+
+\begin{figure}[ht!]
+%    \script{figurescript.py}
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/MR-violins.pdf}
+        \caption{
+        Posterior threshold instellations as a function of dilution factor for two alternative baseline mass-radius relations.
+        For each grid step in $f_\mathrm{rgh}$, we show kernel density estimates of retrieved posteriors (averaged over 50 iterations) assuming an Earth-like (\SI{32.5}{\percent} Fe + \SI{67.5}{\percent} MgSiO$_3$) composition or the probabilistic relation from \citet{Wolfgang2016} and otherwise following the optimistic scenario in Sect.~\ref{sec:res_testability}.
+        Lines within the violins show quartiles of the distributions, and the gray line indicates the injected threshold instellation of \var{injectedSthresh}.
+        Higher log-evidence differences $\Delta \ln Z$ correspond to more significant rejections of the null hypothesis.
+        In the regime of strong detections, both mass-radius relations lead to similar, accurate constraints on $S_\mathrm{thresh}$.
+        Differences occur at low dilution factors, where the Earth-like relation leads to narrower estimates.
+        }
+        \label{fig:MR-violins}
+    \end{centering}
+\end{figure}
+Figure~\ref{fig:MR-violins} shows how the two alternative baseline mass-radius relations influence the significance of a detection and the ability to constrain the threshold instellation.
+The relation of \citet{Wolfgang2016} includes intrinsic scatter, which impedes a detection at low dilution factors.
+In this regime, a layered, Earth-like composition leads to more significant detections and a narrower, although biased, constraint on $S_\mathrm{thresh}$.
+Both mass-radius relations agree and recover the injected value where the null hypothesis can be rejected with high significance.
+This is consistent with our nominal mass-radius relation (compare Sect.~\ref{sec:met-orbits_masses}).
+We conclude that the underlying core and mantle composition of planets may affect the detectability of the transition if the fraction of planets with runaway greenhouse climates is low; however, the overall trends that our experiments revealed appear robust.
+
+We caution that this analysis may serve only as a sanity check and should not be taken as a result in itself: The atmospheric model from \citet{Turbet2020} we adopted relies on a silicate interior composition for its transit radius prediction.
+Therefore, only our nominal procedure throughout the main body of the paper represents a self-consistent treatment.
+
 \bibliographystyle{aasjournal}

### 2023-03-17 src/tex/ms.tex
add a table with all key parameters and assumptions

@@
 %The chosen threshold roughly corresponds to the $p = 0.05$ threshold for statistical significance popular in frequentist approaches.\todo{CITATION}
+
+% include table with key parameters
+\input{table_parameters}
 
@@
 
-\subsection{Influence of different mass-radius relationships on our results}\ref{app:MR_relation}
+\subsection{Influence of different mass-radius relationships on our results}\label{app:MR_relation}
 We assessed how a different choice of baseline mass-radius relation influences our results.

### 2023-03-17 src/tex/table_parameters.tex
add a table with all key parameters and assumptions

\begin{deluxetable*}{lrll}
\tablecaption{Key assumptions and model parameters used in our simulation setup}
\tablehead{\colhead{Parameter} & \colhead{Value} & \colhead{Unit} & \colhead{Description}}
\startdata
\textbf{Stellar sample} &  &  &  \\
$G_\mathrm{max}$ & 16 &  & Maximum Gaia magnitude \\
$M_\mathrm{\star, max}$ & 1.5 & $M_\odot$ & Maximum stellar mass \\
Luminosity evolution &  &  & \citet{Baraffe1998} \\
~\\ \textbf{Planetary parameters} &  &  &  \\
$M_\mathrm{P}$ & 0.1 -- 2.0 & M$_\oplus$ & Planetary mass range \\
$R_\mathrm{P, min}$ & 0.75 & $R_\oplus$ & Minimum planet radius \\
Baseline mass-radius relation &  &  & \citet{Zeng2016} \SI{100}{\percent} $\mathrm{MgSiO_3}$\tablenotemark{a} \\
$\delta_\mathrm{min}$ & 80 & ppm & Minimum transit depth \\
$P_\mathrm{max}$ & 500 & d & Maximum orbital period [d] \\
$S$ & 10 -- 2000 & \SI{}{\watt\per\meter\squared} & Net instellation range \\
$S_\mathrm{thresh}$ & 280 & \SI{}{\watt\per\meter\squared} & Threshold instellation for runaway greenhouse \\
~\\ \textbf{Runaway greenhouse model} &  &  &  \\
Runaway greenhouse atmospheric models &  &  & \citet{Turbet2020,Dorn2021} \\
$x_{H_2O}$ & \SIrange{e-5}{0.1}{} &  & Water mass fraction (fiducial case: 0.005) \\
$f_\mathrm{rgh}$ & \SIrange{0}{1}{} &  & Dilution factor (fiducial case: 0.8) \\
~\\ \textbf{Priors} &  &  &  \\
$\Pi(S_\mathrm{thresh}$) & [10, 1000] & \SI{}{\watt\per\meter\squared} & uniform \\
$\Pi(x_{H_2O})$ & [\SI{e-5}{}, \SI{0.1}{}] &  & log-uniform \\
$\Pi(f_\mathrm{rgh})$ & [0, 1] &  & uniform \\
$\Pi(\langle R_\mathrm{P}\rangle_\mathrm{out})$ & [0, 15] & $R_\oplus$ & Mean radius of non-runaway planets, uniform
\enddata
\tablenotetext{a}{For a comparison with alternative interior compositions, see Appendix~\ref{app:MR_relation}.}
\label{tab:params_table}
\end{deluxetable*}

### 2023-03-20 src/tex/ms.tex
discuss occurrence model caveats
* occurrence model caveats

* Edit, provide context to occurrence rate discussion

* change wording after review & discussion

---------
merges PR  #9

Co-authored-by: Martin Schlecker <martin.schlecker@mailbox.org>

@@
 
+\subsubsection{Structures in the planet occurrence rate density}
+In principle, the baseline occurrence rate density in radius-period space that governs the generation of our synthetic planet sample might influence our findings, especially if it contains any features that coincide with the injected demographic feature.
+This is not the case in the model from \citet{Bergsten2022} that we adopted: Its occurrence rate density varies smoothly in the domain relevant for the runaway greenhouse hypothesis; transitions only occur at smaller instellations ($< \SI{50}{\watt\per\meter\squared}$) and larger radii ($> 1.6\,R_\oplus$).
+We thus do not expect the model underlying our planet sample to affect the recovery of the transition.
+However, it should be kept in mind that any unknown sharp features in the true distribution of terrestrial planets could complicate a detection or lead to a false positive result (see Sect.~\ref{sec:dis_falsepositive}).
+
 \subsubsection{Baseline mass-radius relationship}

### 2023-03-20 src/tex/ms.tex
avoid bullet points within subsubsections

@@
 What influences the probability of correctly rejecting a false null hypothesis?
-We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey:
-\todo[inline]{avoid bullet points within sub-subsections}
-\begin{itemize}
-    \item Occurrence rate of planets forming steam atmospheres: The runaway greenhouse climate relies on the production of large amounts of water vapor that then acts as a greenhouse gas.
+We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey.
+
+\textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies on the production of large amounts of water vapor that then acts as a greenhouse gas.
 This process works only on planets with an atmosphere and of sufficiently volatile-rich composition\todo[inline]{TIM: how much?}.
 The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
-    \item Planetary evolution and duration of the steam atmosphere phase: An inflated steam atmosphere can only be sustained until the planet has lost its water.
+
+\textit{Planetary evolution and duration of the steam atmosphere phase}: An inflated steam atmosphere can only be sustained until the planet has lost its water.
 Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the signal.
-    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is sensitive on the water mass fraction.
+
+\textit{Prevalent water inventory}: The magnitude of radius change at the runaway greenhouse threshold is sensitive on the water mass fraction.
 As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern:
         The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
-    \item Size and composition of the planetary sample: Evidently, the significance of a statistical trend increases with a larger sample size.
+
+\textit{Size and composition of the planetary sample}: Evidently, the significance of a statistical trend increases with a larger sample size.
 In addition, the sample must include planets on both sides of the instellation threshold, which is only sufficiently given for low-mass host stars.
-    \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be.
+
+\textit{Radius measurement precision}: The more precise individual planet radii can be determined, the less smeared out the pattern will be.
 Good \textit{accuracy} is less important, as long as it does not have a systematic error scaling with stellar irradiance.
-    \item Availability and precision of mass measurements: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space.
+
+\textit{Availability and precision of mass measurements}: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space.
 If transiting planets can be followed up to obtain mass measurements, the statistical significance increases.
-\end{itemize}
+
 Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.

### 2023-03-20 src/tex/ms.tex
introduce caveats, update ToDos

@@
 % Define document class
-\documentclass[twocolumn]{aastex631}
+\documentclass[twocolumn,twocolappendix]{aastex631}
 %\documentclass[modern]{aastex631}
@@
 \section{Introduction}
-...just dropping material and references for now...
 \todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
@@
 In summary, stars are assigned a number of planets in accordance with $F_0$, with planet parameters randomly drawn in proportion to the occurrence distribution function evaluated over a grid of radii and orbital periods.
-\todo[inline]{GALEN: The most relevant assumption of using our model is that, at the relevant fluxes (I think), we predict that super-Earths are pretty common.
-I think the "drop" in our model where super-Earths become less common happens at later fluxes than what is relevant in this paper...
-Let's double check this, and add a note in the first paragraph about it?.}
 
@@
 To avoid binning and the artificial patterns it may introduce, we chose to test our hypotheses against a simple moving average $SMA$ along the instellation axis with a window of size \windowsize\ centered around each measurement.\footnote{See Appendix~\ref{app:binnedstats} for a robustness test using a different estimator.}
-We further computed the uncertainty of this moving average by propagating the individual measurement errors and applying a rolling standard error of the mean. \todo[inline]{specify error propagation}
+We further computed the uncertainty of this moving average by propagating the individual measurement errors and applying a rolling standard error of the mean. %\todo[inline]{specify error propagation}
 
@@
 \subsection{Mission design trades}\label{sec:mission-design-trades}
-\todo[inline]{not sure if subsubsections needed}
 To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
@@
 \subsection{Constraining planetary habitability}\label{sec:habitability}
-\todo[inline]{Magma oceans influence the amount of water available at the surface and atmosphere, which is 1.
+%\todo[inline]{Magma oceans influence the amount of water available at the surface and atmosphere, which is 1.
 commonly used as environmental marker to assess habitability, and 2.
@@
 We could, with relatively moderate effort, run a sanity check on an alternative sample (e.g., the more structured Bern model populations, or the population in the Genesis database)}
+The prospects for detecting the runaway greenhouse transition depends on astro- and geophysical factors, as well as on the specific instrumentation and survey strategy of a particular mission.
+We can only approximate all this here, and we therefore seek to assess the sensitivity of our results on the assumptions we have made.
 

### 2023-03-21 src/tex/ms.tex
get equation into column
weave in bulk densities in methods

@@
 The climate state of a planet has a direct influence on its size as measured by transit photometry~\citep{Turbet2019}.
-We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius that planets with instellation-induced runaway greenhouse climates experience.
+We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience.
 ...
@@
 The functional form of the runaway greenhouse hypothesis is more complex: Besides the mean radius of planets outside the threshold $\langle R_\mathrm{P}\rangle_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$, the planetary water mass fraction~$x_{H_2O}$, and the dilution factor $f_\mathrm{rgh}$.
+For hypothesis tests based on bulk density instead of radius, we proceeded in the same way and substituted $R_\mathrm{P}$ by the bulk density $\rho$.
 
@@
 
-The measured radii $R_\mathrm{P, i}$ cannot be directly used for the hypothesis tests as they include intrinsic scatter that is not caused by measurement errors.
+The measured radii $R_\mathrm{P, i}$ or bulk densities $\rho_\mathrm{i}$ cannot be directly used for the hypothesis tests as they include intrinsic scatter that is not caused by measurement errors.
 $H_{\mathrm{rgh}}$ and $H_0$ should thus be tested against a statistical estimator that represents the population mean.
@@
 We assumed here that the individual moving averages $SMA_i$ are measured with a normally distributed uncertainty $\sigma_{SMA_i}$ and adopted a normal distribution
-\begin{equation}
-    \mathcal{L}(SMA \mid \boldsymbol{\theta})=\prod_{i}^{N} \frac{1}{\sqrt{2 \pi \sigma_{SMA_i}^{2}}} \exp \left(-\frac{\left(SMA_i - H\left(\boldsymbol{\theta}, S_i\right)\right)^{2}}{2 \sigma_{SMA_i}^{2}}\right).
-\end{equation}
+\begin{eqnarray}
+    \mathcal{L}(SMA \mid \boldsymbol{\theta})= & \prod_{i}^{N} \frac{1}{\sqrt{2 \pi \sigma_{SMA_i}^{2}}} \\
+    & \times \exp \left(-\frac{\left(SMA_i - H\left(\boldsymbol{\theta}, S_i\right)\right)^{2}}{2 \sigma_{SMA_i}^{2}}\right).
+\end{eqnarray}
 Here, $H\left(\boldsymbol{\theta}, S_i\right)$ corresponds to the functional form of the runaway greenhouse or null hypothesis.

### 2023-03-22 src/tex/ms.tex
assess TESS sample size
include new Tamburo+23 yields for Roman transiters

@@
 Similarly, the Transiting Exoplanet Survey Satellite~\citep[\tess,][]{Ricker2014a} planet sample lacks temperate, small planets around bright host stars~\citep{Ment2023}, as was expected from planet yield calculations~\citep{Barclay2018}.
+As of 2023 March 22, the NASA Exoplanet Archive\footnote{\url{https://exoplanetarchive.ipac.caltech.edu}} lists 40 \tess\ candidate or confirmed planets smaller than \SI{4}{\rEarth} with lower estimated instellation than Earth's.
 
@@
 
-While not primarily designed to detect transiting planets, the \rst~\citep{Spergel2015} is expected to yield $\sim 10^5$ transiting planets on short orbits and constrain their radii in the course of its mission~\citep{Montet2017}.
-In a fraction of the cases, planets as small as \SI{2}{\Rearth} will be detected.
-About 200 planets smaller than Neptune are expected to be found around M~dwarfs and thus may reach into the habitable zone.
-The \rst\ could thus provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.
+While not primarily designed to detect transiting planets, the Galactic Bulge Time Domain Survey of the \rst~\citep{Spergel2015} is expected to yield $\sim 10^5$ transiting planets on short orbits and constrain their radii in the course of its mission~\citep{Montet2017}.
+%In a fraction of the cases, planets as small as \SI{2}{\Rearth} will be detected.
+$\mathcal{O} (1000)$ planets smaller than Neptune could be found around early to mid-M~dwarfs, however, only a small fraction of them will reach into the habitable zone~\citep{Tamburo2023}.
+We thus conclude that the \rst\ could provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.
 

### 2023-03-22 src/tex/ms.tex
compare predicted and measured radius/density changes

@@
 Within the runaway greenhouse regime, an average radius change of \avgRadiusChange\ occurs.
+This pattern is consistent with the injected radius inflation as predicted from the atmospheric models (see Appendix~\ref{app:model-pop_comparison} for an investigation of the interplay between model predictions and our synthetic planet population).
 
@@
 
+\subsection{Statistical imprint of runaway greenhouse atmospheres}\label{app:model-pop_comparison}
+The predicted runaway greenhouse-induced planet radius changes are a function of instellation, planet mass, and water mass fraction~(compare Sect.~\ref{sec:met_rghmodel}).
+In order to better understand the interaction between the planetary populations underlying our simulations and these predictions, we compared the latter to the average radius and bulk density changes we measured in the synthetic population.
+We used the ``optimistic'' scenario with a sample size of \var{N_optimistic}.
+
+\begin{figure*}[ht!]
+%    \script{figurescript.py}
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/model_pop_comparison.pdf}
+        \caption{
+        Comparison of radius and bulk density changes predicted by the atmospheric models to the average changes measured in the synthetic planet population.
+        Model predictions $\Delta R_\mathrm{model}$ and $\Delta \rho_\mathrm{model}$ depend on individual planet masses; $\langle \Delta R_\mathrm{population} \rangle$ and $\langle \Delta \rho_\mathrm{population} \rangle$ are averaged measurements of the overall population.
+        Significant differences are thus expected.
+        }
+        \label{fig:model_pop_comparison.pdf}
+    \end{centering}
+\end{figure*}
+Figure~\ref{fig:model_pop_comparison.pdf} shows this comparison for a range of planetary masses and water mass fractions.
+The complex dependence of the radius inflation on these parameters is evident, but no significant abrupt changes capable of causing spurious signals occur.
+Differences between model prediction and population can be explained by the wide dispersion in planet mass in the population.
+
 \subsection{Influence of different mass-radius relationships on our results}\label{app:MR_relation}

### 2023-03-24 src/tex/ms.tex
minor edits about instellation

@@
 %}
-The climate state of a planet has a direct influence on its size as measured by transit photometry~\citep{Turbet2019}.
+The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019}.
 We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience.
@@
 \begin{note}
-    The power from the host star per unit area at the position of a planet or net instellation $S$ in units of Earth's insolation is given by
+    The power absorbed by the planet from the host star per unit area at its orbital distance or net instellation $S$ in units of Earth's insolation is given by
     \begin{equation}
@@
     \end{equation}
-    We assume global redistribution of incoming flux and a fixed albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
+    We assume global redistribution of incoming flux and a wavelength and spherically integrated albedo, or Bond albedo, of $0.3$, comparable to Earth's~\citep{Haar1971}.
     The absolute value of $S_\oplus$, by which all synthetic planets are scaled, is then \SI{238}{\watt\per\square\meter}.

### 2023-03-24 src/tex/ms.tex
discuss multi-planet systems

@@
 We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).\todo[inline]{make sure rgh hypothesis is mentioned before, or rephrase}
-\todo[inline]{\bioverse\ will need a short introduction before the modifications are mentioned. Probably should get a paragraph in the intro}
 \begin{figure*}
@@
 \subsection{Detectability of the runaway greenhouse transition}
-\todo[inline]{Mention multiplanet systems.}
 We showed in Sect.~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
@@
 \todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
+\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population).
+These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019})}
+
+\subsubsection{Detection in multi-planet systems}
+As suggested by \citet{Turbet2019}, an alternative approach for detecting the runaway greenhouse-induced radius inflation is to search for its ``local'' imprints in multi-planet systems.
+Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
+While degeneracies remain in interpreting bulk density fluctuations within individual systems~\citep[e.g.,][]{Turbet2020,Dorn2021}, the detection of a consistent pattern in several such systems could be a convincing statistical evidence of the runaway greenhouse transition.
 
@@
 context: \citet{Turbet2019} Sect. 3, \citet{Bower2019}.\\
-''present bracketing cases; or to discuss why the discountinuity in the results would not be affected by changes that led to continuous offsets in the results. Or both.''\\
 - not yet explored: how sensitive are we on the assumed planet population?

### 2023-03-24 src/tex/ms.tex
explore CKS multis

@@
 While degeneracies remain in interpreting bulk density fluctuations within individual systems~\citep[e.g.,][]{Turbet2020,Dorn2021}, the detection of a consistent pattern in several such systems could be a convincing statistical evidence of the runaway greenhouse transition.
+The current sample of suitable systems is sparse: The California-Kepler Survey catalog~\citep{Fulton2018} contains only six planets with instellations $< 2\, S_\oplus$ and smaller than \SI{2}{\rEarth} in five systems.
+Future additions to the multi-planet sample through missions such as \plato\ are needed.
 

### 2023-03-27 src/tex/ms.tex
new Plato runs, redraw plots, update numbers

@@
 The minimum needed fraction rises to \var{platoMinfrghHundred} for $N = 100$.
-For much smaller samples, even an optimistically strong signal can no longer be detected.
+For much smaller samples, only a very optimistically strong signal is likely to be detected.
 
@@
 A density-based survey of 100 targets is roughly equivalent to a radius-based survey of \var{N_plato} targets.
-At $N=100$, pure radius measurements require $f_\mathrm{rgh} \gtrsim 0.4$ whereas bulk density measurements enable a detection from $f_\mathrm{rgh} \gtrsim 0.1$.
+At $N=100$, pure radius measurements require $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred}$ whereas bulk density measurements enable a detection from $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred_rho}$.
 

### 2023-03-28 src/tex/ms.tex
rerun S_thresh grid, update plot

@@
             Accuracy and precision of the constraint on $S_\mathrm{thresh}$ generally improve with higher $f_\mathrm{rgh}$.
-            Bulk density-based inferences improve the constraints, and M~dwarf samples yield the highest accuracy.
+            Bulk density-based inferences improve the constraints, and M~dwarf samples yield the highest accuracy and precision.
         }
@@
 
-A sample containing only planets around M~dwarf yields a very similar performance, which is because these planets contribute most to the statistical power of the retrieval.
-
-% Additional \subsection for habitability? To reflect structure in Discussion, Conclusions?
+A sample containing only planets around M~dwarf yields still better performance and result in the overall best accuracy and precision.
+The reason is that these planets contribute most to the statistical power of the retrieval.
 

### 2023-03-29 src/tex/ms.tex
clarify why M dwarfs are key

@@
 \textit{Size and composition of the planetary sample}: Evidently, the significance of a statistical trend increases with a larger sample size.
-In addition, the sample must include planets on both sides of the instellation threshold, which is only sufficiently given for low-mass host stars.
+In addition, the sample must include planets on both sides of the instellation threshold, which, due to the strongly distance-dependent detection bias associated with the transit method, is only sufficiently given for low-mass host stars.
 

### 2023-03-29 src/tex/ms.tex
discuss false positive potential of M dwarf planet density dichotomy
(Luque&Palle 2022)

@@
 However, the Neptune desert occurs at smaller orbital periods and its location depends on a planet's size~\citep{Szabo2011}, which is not expected for the runaway greenhouse transition.
-\todo{is there overlap/confusion with the radius/density trend in \citet{Luque2022}?}
+
+\citet{Luque2022} found that small planets around orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
+This trend does not represent a false positive scenario for the runaway greenhouse transition, since no strong dependency on instellation has been found or is expected.
+However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
+If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{Venturini2020,Burn2021,Schlecker2021,Schlecker2021b}, systems of all ages can be affected by this dilution.
 
@@
 The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
-Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent weight fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021}.
+Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021}.
 Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.

### 2023-03-30 src/tex/ms.tex
mention caveats regarding phase changes, interior-atmosphere interaction

@@
 For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
-This scenario does not take into account the expected additional radius change due to interiors that are molten due to greenhouse forcing~\citep{Bower2019}.
+This scenario does not take into account the expected additional radius change due to molten interiors~\citep{Bower2019}.
 It further assumes a \textit{dry melt} without dissolved volatiles.
@@
 We illustrate the mass-radius relations of the three cases in Figure~\ref{fig:radiusevolution} where a fiducial water mass fraction of $x_{H_2O}= \var{wrr}$ is assumed.
+% Earth: ~0.0002 PERCENT or 218 ppm by mass (Turbet+2019)
 In the following, we only distinguish between the non-runaway greenhouse and wet melt scenarios.
@@
 \todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population).
-These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019})}
+These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Sect.
+4]{Turbet2019}.}
 
@@
 \todo[inline]{Caveats\\
-context: \citet{Turbet2019} Sect. 3, \citet{Bower2019}.\\
 - not yet explored: how sensitive are we on the assumed planet population?
@@
 Nevertheless, we performed a sanity check to assess the influence varying our baseline model on our results (see Appendix~\ref{app:MR_relation}), and found general agreement between different interior composition models.
+
+The models in this study do not into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in runaway greenhouse planets~\citep{Ikoma2018}.
+Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{Turbet2019}.
 

### 2023-03-30 src/tex/ms.tex
outline story for introduction

@@
 
-\begin{note}
-For observations of rocky exoplanets, the currently best-probed regime is that of warm, close-in planets.
-These bodies experience climatic conditions that are similar to the environment of the inner Solar System bodies at early stages of their formation.
-Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.
+\begin{itemize}
+    \item hot/warm planets are best studied
+    \item runaway greenhouse climate (Ingersoll1969,Kasting1988,Nakajima1992)
+    \item HZ concept.''(Kasting et al.
+1993; Kopparapu et al.
+2013, 2014)'' Relevance in the search for habitable planets and life.
+(Science gap:) Does it exist?
+Can it be tested?
+=> statistically detect inner edge
+    \item Bioverse
+    \item Transit survey => PLATO
+    \item Using PLATO as a baseline survey, we do A, B, C.
+\end{itemize}
 
-...
-    Venus and Earth, while having accreted from the same mass reservoir and despite their similar bulk properties, evolved into planets with very different environmental conditions on their surfaces.
-    At the formation of the Solar System, both planets underwent a giant impact phase [CITE!] that melted their mantles, leading to a magma ocean stage.
-    Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
-    \citet{Hamano2013} showed that such a late desiccation process is not needed to explain the stark differences: for a planet rich in volatiles and receiving high enough radiation levels from its host star, a steam atmosphere limits the outgoing radiation from the (molten) planetary surface.
-    This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}, enough to remove the entire water reservoir from a rocky planet by hydrodynamic escape.
-    \todo[inline]{important for motivating the existence of the pattern: ``the solidification timescale can become comparable to the main-sequence lifetime of the star (Hamano et al.
-2013, 2015)'' \citep{Lichtenberg2022} }
-    Consequently, it is aptly considered to be the inner boundary of the habitable zone~\citep{Kasting1993,Kopparapu2013}.
-
-    (Under the right conditions, different orbital distances alone can thus decide about a planet's water content and ultimately its habitability.)
-    If received flux makes the difference between our own habitable planet and dry, dead Venus, there is good reason to believe that such patterns exist in exoplanet systems as well.
-
-    ...long-term goal/overarching objective: derive the geophysical history of rocky extrasolar planets.
-More concrete: Constrain the limits of runaway greenhouse transitions, and thus the inner edges of the habitable zone.
-Also: How close is Earth from this runaway greenhouse irradiation limit?
-   ...
-\todo[inline]{introduce magma oceans and their influence on planetary radii~\citep{Dorn2021}.}
-    ...current/future observations of planets that are currently in the runaway greenhouse phase may constrain properties of their planetary mantles...
-make connection between interior and atmospheres.
-    ...
-
-    The runaway greenhouse transition is a robust prediction from climate models, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
-Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
-On the planetary level, however, such undertaking is hindered by degeneracies: From internal composition to atmospheric loss to volatile content, many factors influence a planet's size.
-It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019}.
- Using \bioverse\ to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we...
-
-    Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.
-
-\end{note}
+%\begin{note}
+%Recent years have brought tremendous advances in our understanding of terrestrial-sized planets, mainly due to space-based transit surveys such as \kepler\ \citep{Borucki2010} and \tess\ \citep{Ricker2014a} that pushed detection sensitivities into the terrestrial regime.
+%Due to the inherent detection biases of the transit technique, most current well-probed small planets are much closer to their host star than Earth is to the Sun. CITE
+%These bodies experience climatic conditions that are similar to the environment of the inner Solar System planets at early stages of their formation. CITE
+%Studies of the geophysical state and evolution of warm exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.
+%
+%A much regarded object of study are the divergent atmospheric evolutions of Earth and Venus.
+%While both have accreted from the same mass reservoir and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces.
+"(Donahue et al.
+1982; Kasting 1988; Hamano et al.
+2013; Way et al.
+2016; Kane et al.
+2019; Way \& Del Genio 2020; Turbet et al.
+2021)."
+%
+%
+%
+%
+%
+%
+%
+%
+%
+%\end{note}
 
@@
 \section{Discussion}\label{sec:discussion}
-\todo[inline]{\textbf{Potential additional discussion points: }\\
-(How close is Earth to runaway? (-> \citet{Goldblatt2012,Popp2016}; what is the max. expected change in Earth's albedo due to deglaciation?)}
+%\todo[inline]{\textbf{Potential additional discussion points: }\\
+%(How close is Earth to runaway? (-> \citet{Goldblatt2012,Popp2016}; what is the max. expected change in Earth's albedo due to deglaciation?)}
 
@@
 While degeneracies remain in interpreting bulk density fluctuations within individual systems~\citep[e.g.,][]{Turbet2020,Dorn2021}, the detection of a consistent pattern in several such systems could be a convincing statistical evidence of the runaway greenhouse transition.
-The current sample of suitable systems is sparse: The California-Kepler Survey catalog~\citep{Fulton2018} contains only six planets with instellations $< 2\, S_\oplus$ and smaller than \SI{2}{\rEarth} in five systems.
+The current sample of suitable systems is sparse: The California-Kepler Survey catalog~\citep{Fulton2018} contains only six planets with instellations $< 2\, S_\oplus$ and smaller than \SI{2}{\rEarth} in five multi-planet systems.
 Future additions to the multi-planet sample through missions such as \plato\ are needed.

### 2023-03-30 src/tex/ms.tex
outline story for introduction

@@
 \end{figure}
-Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the runaway greenhouse transition.
-For the sake of clarity, we chose an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
-We assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
+Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the HZ Inner Edge Discontinuity.
+Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
+Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
 In this case, the pattern was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).
@@
  For all but very low water fractions, $f_\mathrm{rgh}$ dominates this trend: It enters linearly into the average planet radius, whereas the contribution of $x_{H_2O}$ - as predicted by the geophysical models - is sublinear with a power-law exponent of $\sim 0.3$.
-As long as $f_\mathrm{rgh}$ is larger than $\sim 0.2$, a sample size of \var{N_optimistic} is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $10^{-3}$.
+Within the framework of our model and as long as $f_\mathrm{rgh}$ is larger than $\sim 0.2$, a sample size of \var{N_optimistic} is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $10^{-3}$.
 
@@
     \end{figure*}
-
-For a volume-limited sample, considering all host star spectral types, we find that a yield of \var{N_plato} planets enables a significant detection if the fraction of runaway greenhouse planets is larger than $\sim \var{platoMinfrgh}$ (see Fig.~\ref{fig:plato_fgrid}).
+To simulate the transit survey of ESA's \plato\ mission, we considered a volume-limited sample with a size according to projections and including all stellar spectral types.
+Given \plato's expected radius precision, we find that a yield of \var{N_plato} planets enables a significant detection if the fraction of runaway greenhouse planets is larger than $ f_\mathrm{rgh} \sim \var{platoMinfrgh}$ (see Fig.~\ref{fig:plato_fgrid}).
 The minimum needed fraction rises to \var{platoMinfrghHundred} for $N = 100$.
@@
 Comparing a measurement of the signal in radius space with a measurement in density space (which requires planetary mass measurements), a stronger detection occurs in the latter case:
-With an optimistic choice of geophysical parameters (see above), the average measured radius change is \SI{+15}{\percent} whereas the average density change is \SI{-33}{\percent}.
+With an optimistic choice of geophysical parameters (see Sect.~\ref{sec:res_testability}), the average measured radius change is \SI{+15}{\percent} whereas the average density change is \SI{-33}{\percent}.
 
 %We now examine the value of a follow-up campaign to constrain planet masses with respect to the detectability of the runaway greenhouse threshold.
-As can be seen in Fig.~\ref{fig:plato_fgrid}, the detectability is enhanced when radius constraints can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
+Figure~\ref{fig:plato_fgrid} shows that investigating the HZ Inner Edge Discontinuity is more constraining when radius measurements can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
 Consequently, we achieve a statistically significant detection with smaller samples or lower dilution factors $f_\mathrm{rgh}$.

### 2023-04-01 src/tex/ms.tex
flesh out background, introduce Bioverse

--- 2023-03-30 src/tex/ms.tex
@@
 \section{Introduction}
-\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
+\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021}.}
 \todo[inline]{``Points for the intro:
@@
     \item hot/warm planets are best studied
-    \item runaway greenhouse climate (Ingersoll1969,Kasting1988,Nakajima1992)
+    \item runaway greenhouse climate (Ingersoll1969,Kasting1988,Nakajima1992). Transition is sharp in instellation.
     \item HZ concept.''(Kasting et al.
@@
 
-%\begin{note}
-%Recent years have brought tremendous advances in our understanding of terrestrial-sized planets, mainly due to space-based transit surveys such as \kepler\ \citep{Borucki2010} and \tess\ \citep{Ricker2014a} that pushed detection sensitivities into the terrestrial regime.
-%Due to the inherent detection biases of the transit technique, most current well-probed small planets are much closer to their host star than Earth is to the Sun. CITE
-%These bodies experience climatic conditions that are similar to the environment of the inner Solar System planets at early stages of their formation. CITE
-%Studies of the geophysical state and evolution of warm exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.
-%
-%A much regarded object of study are the divergent atmospheric evolutions of Earth and Venus.
-%While both have accreted from the same mass reservoir and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces.
+\begin{note}
+Recent years have brought tremendous advances in our understanding of terrestrial-sized planets, mainly due to space-based transit surveys such as \kepler\ \citep{Borucki2010} and \tess\ \citep{Ricker2014a} that pushed detection sensitivities into the terrestrial regime.
+Due to the inherent detection biases of the transit technique, most current well-probed small planets are much closer to their host star than Earth is to the Sun. CITE
+These bodies experience climatic conditions that are similar to the environment of the inner Solar System planets at early stages of their formation. CITE
+Studies of the geophysical state and evolution of warm exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.
+
+A much regarded object of study are the divergent atmospheric evolutions of Earth and Venus.
+While both have accreted from a similar mass reservoir~\citep{Morbidelli2012} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{Kasting1988,Hamano2013,Kane2014,Turbet2021}.
 "(Donahue et al.
-1982; Kasting 1988; Hamano et al.
-2013; Way et al.
-2016; Kane et al.
-2019; Way \& Del Genio 2020; Turbet et al.
-2021)."
-%
-%
-%
-%
-%
-%
-%
-%
-%
-%\end{note}
+1982; Bézard, B.
+\& de Bergh, 2007; Way et al.
+2016; Way \& Del Genio 2020; )."
+Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
+%magma oceans...
+However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet.
+Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits outgoing radiation and thus the release of excess heat~\citep{Ingersoll1969,Kasting1988,Wordsworth2022}.
+This runaway greenhouse state can not only extend the magma ocean stage of a rocky planet to hundreds of \SI{}{\mega\year}, it can also lead to a loss of its entire water reservoir by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen.
+CITE
+Venus' atmospheric composition is consistent with the planet having gone through this phase, although the past presence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016}.
+A water-dominated atmosphere of a planet in the runaway greenhouse state is significantly inflated~\citep{Turbet2019,Mousis2020}, increasing its transit radius by an amount mainly depending on the surface water content.
+An Earth-like planet experiences a \SI{\sim 500}{\kilo\meter} radius increase upon crossing the threshold instellation, and its transit depth increases by \SI{10}{\ppm} to \SI{1000}{\ppm} depending on the spectral type of the host star~\citep{Turbet2019}.
+... water incorporation...
+
+The radiation-induced transition between a runaway greenhouse state and a temperate climate has been predicted to be a sharp threshold in instellation~\citep{Hamano2013}. CITE
+Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a CO$_2$/H$_2$O/N$_2$ atmosphere~(\citealp[][but other definitions have been put forward, e.g., ]{Kasting1993,Kopparapu2013,Kopparapu2014} \citealp{Hall2023,Spinelli2023}).
+The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1853,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
+Investigating its boundaries in parameter space has helped improve our understanding of habitability. CITE
+Nevertheless, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017}.
+The question naturally emerges if a habitable zone -- in its common form as boundaries in receiving instellation -- indeed exists and how the habitable zone hypothesis can be tested.
+Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
+A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
+Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos ~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
+All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.
+
+Here, we explore the feasibility of a statistical test of the habitable zone hypothesis by surveying planetary bulk properties close to its inner edge, the runaway greenhouse transition.
+Specifically, we assess the ability of near-future transit surveys to test the hypothesis that the runaway greenhouse effect causes an abrupt change of planetary radii and bulk densities when ordered by receiving instellation.
+Our main tool for this is \bioverse, a simulation framework for assessing the statistical power of exoplanet surveys~\citep{Bixel2021}.
+It consists of a sample generator that populates stars from the \gaia\ catalog CITE with planetary systems based on state-of-the-art occurrence rates, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
+Trying different instrumentation and survey designs, we try to recover runaway greenhouse-induced effects based on model predictions~\citep{Turbet2020,Dorn2021} that we inject into a baseline planet population.
+
+Our goal is to...
+We then perform a parameter study to explore which trade-offs in the survey design of a \plato-like mission optimize the chance to detect and characterize the runaway greenhouse transition.
+
+\end{note}
 

### 2023-04-01 src/tex/ms.tex
add PLATO, water redistribution, and goal.

@@
 \section{Introduction}
-\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021}.}
-\todo[inline]{``Points for the intro:
-\bioverse\\
-Transit survey to measure mass/radius of planets\\
-Does the HZ have an inner edge?\\
-Run-away GH\\
-Magma oceans\\
-Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman.
+%\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021}.}
+%\todo[inline]{``Points for the intro:
+%\bioverse\\
+%Transit survey to measure mass/radius of planets\\
+%Does the HZ have an inner edge?\\
+%Run-away GH\\
+%Magma oceans\\
+%Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman.
 Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
-~\\
-In contrast, PLATO and \nautilus\ are \\
-Possibility of follow-up to measureme masses
-''}
+%~\\
+%In contrast, PLATO and \nautilus\ are \\
+%Possibility of follow-up to measureme masses
+%''}
 
@@
 An Earth-like planet experiences a \SI{\sim 500}{\kilo\meter} radius increase upon crossing the threshold instellation, and its transit depth increases by \SI{10}{\ppm} to \SI{1000}{\ppm} depending on the spectral type of the host star~\citep{Turbet2019}.
-... water incorporation...
+It was further found that the redistribution of water between the steam atmosphere and the molten mantle leads to additional, non-negligible deviations in atmospheric scale-heights and transit radii~\citep{Dorn2021}.
 
-The radiation-induced transition between a runaway greenhouse state and a temperate climate has been predicted to be a sharp threshold in instellation~\citep{Hamano2013}. CITE
+Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{Hamano2013}.
+CITE
 Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a CO$_2$/H$_2$O/N$_2$ atmosphere~(\citealp[][but other definitions have been put forward, e.g., ]{Kasting1993,Kopparapu2013,Kopparapu2014} \citealp{Hall2023,Spinelli2023}).
@@
 Here, we explore the feasibility of a statistical test of the habitable zone hypothesis by surveying planetary bulk properties close to its inner edge, the runaway greenhouse transition.
-Specifically, we assess the ability of near-future transit surveys to test the hypothesis that the runaway greenhouse effect causes an abrupt change of planetary radii and bulk densities when ordered by receiving instellation.
+Our goal is to assess the ability of near-future transit surveys to test the hypothesis that the runaway greenhouse effect causes an abrupt change of planetary radii and bulk densities when ordered by receiving instellation~\citep{Turbet2019}.
 Our main tool for this is \bioverse, a simulation framework for assessing the statistical power of exoplanet surveys~\citep{Bixel2021}.
-It consists of a sample generator that populates stars from the \gaia\ catalog CITE with planetary systems based on state-of-the-art occurrence rates, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
-Trying different instrumentation and survey designs, we try to recover runaway greenhouse-induced effects based on model predictions~\citep{Turbet2020,Dorn2021} that we inject into a baseline planet population.
+It consists of a sample generator that populates stars from the \gaia\ catalog~\citep{Gaia_mission,Gaia_DR3}  with planetary systems based on state-of-the-art occurrence rates, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
+Trying different instrumentation and survey designs, we utilize \bioverse\ to recover runaway greenhouse-induced effects based on model predictions~\citep{Turbet2020,Dorn2021} that we inject into a baseline planet population.
 
-Our goal is to...
+In particular, we test the capability of the \plato\ \citep[PLAnetary Transits and Oscillation of stars, ][]{Rauer2016} mission, which will measure the radii of a large number of terrestrial-sized planets, to detect the runaway greenhouse transition and identify the astrophysical unknowns on which this depends.
 We then perform a parameter study to explore which trade-offs in the survey design of a \plato-like mission optimize the chance to detect and characterize the runaway greenhouse transition.
@@
 For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
-While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013,Boukrouche2021}.
+While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013,Boukrouche2021}.
 To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
@@
 \todo[inline]{TIM: Ideas?\\
-discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk)}
+discuss potential atmospheric signatures of magma oceans.
+e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g.
+Tim's recent talk).
+Or: high O2 abundance during H loss (Wordsworth \& Pierrehumbert 2013b; Luger \& Barnes 2015, Schaefer et al.
+2016)}
 \todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
@@
 
-Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, or surface gravity~\citep{Pierrehumbert2022}.
+Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Turbet2021,Pierrehumbert2022}.
 The choice of statistical estimator for the hypothesis tests may further blur the recovered discontinuity.
@@
 %We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
+%This work has made use of data from the European Space Agency (ESA) mission {\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia} Data Processing and Analysis Consortium (DPAC, \url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}).
+Funding for the DPAC has been provided by national institutions, in particular the institutions participating in the {\it Gaia} Multilateral Agreement.
 %\end{acknowledgments}

### 2023-04-01 src/tex/ms.tex
add PLATO, water redistribution, and goal.

@@
 \end{figure}
-Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the HZ Inner Edge Discontinuity.
-Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
-Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
+Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the runaway greenhouse transition.
+For the sake of clarity, we chose an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
+We assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
 In this case, the pattern was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).
@@
  For all but very low water fractions, $f_\mathrm{rgh}$ dominates this trend: It enters linearly into the average planet radius, whereas the contribution of $x_{H_2O}$ - as predicted by the geophysical models - is sublinear with a power-law exponent of $\sim 0.3$.
-Within the framework of our model and as long as $f_\mathrm{rgh}$ is larger than $\sim 0.2$, a sample size of \var{N_optimistic} is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $10^{-3}$.
+As long as $f_\mathrm{rgh}$ is larger than $\sim 0.2$, a sample size of \var{N_optimistic} is sufficient for a $\SI{50}{\percent}$ detection rate even for water ratios as low as $10^{-3}$.
 
@@
     \end{figure*}
-To simulate the transit survey of ESA's \plato\ mission, we considered a volume-limited sample with a size according to projections and including all stellar spectral types.
-Given \plato's expected radius precision, we find that a yield of \var{N_plato} planets enables a significant detection if the fraction of runaway greenhouse planets is larger than $ f_\mathrm{rgh} \sim \var{platoMinfrgh}$ (see Fig.~\ref{fig:plato_fgrid}).
+
+For a volume-limited sample, considering all host star spectral types, we find that a yield of \var{N_plato} planets enables a significant detection if the fraction of runaway greenhouse planets is larger than $\sim \var{platoMinfrgh}$ (see Fig.~\ref{fig:plato_fgrid}).
 The minimum needed fraction rises to \var{platoMinfrghHundred} for $N = 100$.
@@
 Comparing a measurement of the signal in radius space with a measurement in density space (which requires planetary mass measurements), a stronger detection occurs in the latter case:
-With an optimistic choice of geophysical parameters (see Sect.~\ref{sec:res_testability}), the average measured radius change is \SI{+15}{\percent} whereas the average density change is \SI{-33}{\percent}.
+With an optimistic choice of geophysical parameters (see above), the average measured radius change is \SI{+15}{\percent} whereas the average density change is \SI{-33}{\percent}.
 
 %We now examine the value of a follow-up campaign to constrain planet masses with respect to the detectability of the runaway greenhouse threshold.
-Figure~\ref{fig:plato_fgrid} shows that investigating the HZ Inner Edge Discontinuity is more constraining when radius measurements can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
+As can be seen in Fig.~\ref{fig:plato_fgrid}, the detectability is enhanced when radius constraints can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
 Consequently, we achieve a statistically significant detection with smaller samples or lower dilution factors $f_\mathrm{rgh}$.

### 2023-04-01 src/tex/ms.tex
add paper outline

--- 2023-04-01 src/tex/ms.tex
@@
 
+We organize the paper as follows:
+Section~\ref{sec:met_baseline} introduces our baseline model produce synthetic star and planet samples.
+    In Sect.~\ref{sec:met_rghmodel}, we describe the model component that produces runaway greenhouse-induced radius changes.
+    Section~\ref{sec:met_surveys-hypotests} explains our survey simulations and hypothesis tests.
+    We present our results in Sect.~\ref{sec:results} before interpreting them in Sect.~\ref{sec:discussion}.
+    Finally, we summarize our findings in Sect.~\ref{sec:conclusions}.
 \end{note}
@@
 
-\section{Results}
+\section{Results}\label{sec:results}
 \subsection{Statistical signature of the runaway greenhouse threshold}\label{sec:res_signature}
@@
 
-\section{Conclusions}
+\section{Conclusions}\label{sec:conclusions}
 Runaway greenhouse climates on rocky planets cause significant inflation of their radii.

### 2023-04-03 src/tex/ms.tex
remove bullets
edit optimistic case and PLATO

--- 2023-04-01 src/tex/ms.tex
@@
 \section{Introduction}
-%\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021}.}
-%\todo[inline]{``Points for the intro:
-%\bioverse\\
-%Transit survey to measure mass/radius of planets\\
-%Does the HZ have an inner edge?\\
-%Run-away GH\\
-%Magma oceans\\
-%Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman.
+\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
+\todo[inline]{``Points for the intro:
+\bioverse\\
+Transit survey to measure mass/radius of planets\\
+Does the HZ have an inner edge?\\
+Run-away GH\\
+Magma oceans\\
+Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman.
 Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
-%~\\
-%In contrast, PLATO and \nautilus\ are \\
-%Possibility of follow-up to measureme masses
-%''}
+~\\
+In contrast, PLATO and \nautilus\ are \\
+Possibility of follow-up to measureme masses
+''}
 
-\begin{itemize}
-    \item hot/warm planets are best studied
-    \item runaway greenhouse climate (Ingersoll1969,Kasting1988,Nakajima1992). Transition is sharp in instellation.
-    \item HZ concept.''(Kasting et al.
-1993; Kopparapu et al.
-2013, 2014)'' Relevance in the search for habitable planets and life.
-(Science gap:) Does it exist?
-Can it be tested?
-=> statistically detect inner edge
-    \item Bioverse
-    \item Transit survey => PLATO
-    \item Using PLATO as a baseline survey, we do A, B, C.
-\end{itemize}
+\todo[inline]{introduce the HZ Inner Edge Discontinuity\textsuperscript{\texttrademark}}
 
 \begin{note}
-Recent years have brought tremendous advances in our understanding of terrestrial-sized planets, mainly due to space-based transit surveys such as \kepler\ \citep{Borucki2010} and \tess\ \citep{Ricker2014a} that pushed detection sensitivities into the terrestrial regime.
-Due to the inherent detection biases of the transit technique, most current well-probed small planets are much closer to their host star than Earth is to the Sun. CITE
-These bodies experience climatic conditions that are similar to the environment of the inner Solar System planets at early stages of their formation. CITE
-Studies of the geophysical state and evolution of warm exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.
+For observations of rocky exoplanets, the currently best-probed regime is that of warm, close-in planets.
+These bodies experience climatic conditions that are similar to the environment of the inner Solar System bodies at early stages of their formation.
+Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.
 
-A much regarded object of study are the divergent atmospheric evolutions of Earth and Venus.
-While both have accreted from a similar mass reservoir~\citep{Morbidelli2012} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{Kasting1988,Hamano2013,Kane2014,Turbet2021}.
-"(Donahue et al.
-1982; Bézard, B.
-\& de Bergh, 2007; Way et al.
-2016; Way \& Del Genio 2020; )."
+...
+    Venus and Earth, while having accreted from the same mass reservoir and despite their similar bulk properties, evolved into planets with very different environmental conditions on their surfaces.
+    At the formation of the Solar System, both planets underwent a giant impact phase [CITE!] that melted their mantles, leading to a magma ocean stage.
     Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
-%magma oceans...
-However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet.
-Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits outgoing radiation and thus the release of excess heat~\citep{Ingersoll1969,Kasting1988,Wordsworth2022}.
-This runaway greenhouse state can not only extend the magma ocean stage of a rocky planet to hundreds of \SI{}{\mega\year}, it can also lead to a loss of its entire water reservoir by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen.
-CITE
-Venus' atmospheric composition is consistent with the planet having gone through this phase, although the past presence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016}.
-A water-dominated atmosphere of a planet in the runaway greenhouse state is significantly inflated~\citep{Turbet2019,Mousis2020}, increasing its transit radius by an amount mainly depending on the surface water content.
-An Earth-like planet experiences a \SI{\sim 500}{\kilo\meter} radius increase upon crossing the threshold instellation, and its transit depth increases by \SI{10}{\ppm} to \SI{1000}{\ppm} depending on the spectral type of the host star~\citep{Turbet2019}.
-It was further found that the redistribution of water between the steam atmosphere and the molten mantle leads to additional, non-negligible deviations in atmospheric scale-heights and transit radii~\citep{Dorn2021}.
+    \citet{Hamano2013} showed that such a late desiccation process is not needed to explain the stark differences: for a planet rich in volatiles and receiving high enough radiation levels from its host star, a steam atmosphere limits the outgoing radiation from the (molten) planetary surface.
+    This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}, enough to remove the entire water reservoir from a rocky planet by hydrodynamic escape.
+    \todo[inline]{important for motivating the existence of the pattern: ``the solidification timescale can become comparable to the main-sequence lifetime of the star (Hamano et al.
+2013, 2015)'' \citep{Lichtenberg2022} }
+    Consequently, it is aptly considered to be the inner boundary of the habitable zone~\citep{Kasting1993,Kopparapu2013}.
 
-Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{Hamano2013}.
-CITE
-Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a CO$_2$/H$_2$O/N$_2$ atmosphere~(\citealp[][but other definitions have been put forward, e.g., ]{Kasting1993,Kopparapu2013,Kopparapu2014} \citealp{Hall2023,Spinelli2023}).
-The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1853,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
-Investigating its boundaries in parameter space has helped improve our understanding of habitability. CITE
-Nevertheless, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017}.
-The question naturally emerges if a habitable zone -- in its common form as boundaries in receiving instellation -- indeed exists and how the habitable zone hypothesis can be tested.
-Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
-A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
-Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos ~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
-All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.
+    (Under the right conditions, different orbital distances alone can thus decide about a planet's water content and ultimately its habitability.)
+    If received flux makes the difference between our own habitable planet and dry, dead Venus, there is good reason to believe that such patterns exist in exoplanet systems as well.
 
-Here, we explore the feasibility of a statistical test of the habitable zone hypothesis by surveying planetary bulk properties close to its inner edge, the runaway greenhouse transition.
-Our goal is to assess the ability of near-future transit surveys to test the hypothesis that the runaway greenhouse effect causes an abrupt change of planetary radii and bulk densities when ordered by receiving instellation~\citep{Turbet2019}.
-Our main tool for this is \bioverse, a simulation framework for assessing the statistical power of exoplanet surveys~\citep{Bixel2021}.
-It consists of a sample generator that populates stars from the \gaia\ catalog~\citep{Gaia_mission,Gaia_DR3}  with planetary systems based on state-of-the-art occurrence rates, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
-Trying different instrumentation and survey designs, we utilize \bioverse\ to recover runaway greenhouse-induced effects based on model predictions~\citep{Turbet2020,Dorn2021} that we inject into a baseline planet population.
+    ...long-term goal/overarching objective: derive the geophysical history of rocky extrasolar planets.
+More concrete: Constrain the limits of runaway greenhouse transitions, and thus the inner edges of the habitable zone.
+Also: How close is Earth from this runaway greenhouse irradiation limit?
+   ...
+\todo[inline]{introduce magma oceans and their influence on planetary radii~\citep{Dorn2021}.}
+    ...current/future observations of planets that are currently in the runaway greenhouse phase may constrain properties of their planetary mantles...
+make connection between interior and atmospheres.
+    ...
 
-In particular, we test the capability of the \plato\ \citep[PLAnetary Transits and Oscillation of stars, ][]{Rauer2016} mission, which will measure the radii of a large number of terrestrial-sized planets, to detect the runaway greenhouse transition and identify the astrophysical unknowns on which this depends.
-We then perform a parameter study to explore which trade-offs in the survey design of a \plato-like mission optimize the chance to detect and characterize the runaway greenhouse transition.
+    The runaway greenhouse transition is a robust prediction from climate models, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
+Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
+On the planetary level, however, such undertaking is hindered by degeneracies: From internal composition to atmospheric loss to volatile content, many factors influence a planet's size.
+It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019}.
+ Using \bioverse\ to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we...
 
-We organize the paper as follows:
-Section~\ref{sec:met_baseline} introduces our baseline model produce synthetic star and planet samples.
-    In Sect.~\ref{sec:met_rghmodel}, we describe the model component that produces runaway greenhouse-induced radius changes.
-    Section~\ref{sec:met_surveys-hypotests} explains our survey simulations and hypothesis tests.
-    We present our results in Sect.~\ref{sec:results} before interpreting them in Sect.~\ref{sec:discussion}.
-    Finally, we summarize our findings in Sect.~\ref{sec:conclusions}.
+    Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.
+
 \end{note}
@@
 For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
-While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013,Boukrouche2021}.
+While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013,Boukrouche2021}.
 To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
@@
 
-\section{Results}\label{sec:results}
+\section{Results}
 \subsection{Statistical signature of the runaway greenhouse threshold}\label{sec:res_signature}
@@
 \todo[inline]{TIM: Ideas?\\
-discuss potential atmospheric signatures of magma oceans.
-e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g.
-Tim's recent talk).
-Or: high O2 abundance during H loss (Wordsworth \& Pierrehumbert 2013b; Luger \& Barnes 2015, Schaefer et al.
-2016)}
+discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk)}
 \todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
@@
 
-Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Turbet2021,Pierrehumbert2022}.
+Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, or surface gravity~\citep{Pierrehumbert2022}.
 The choice of statistical estimator for the hypothesis tests may further blur the recovered discontinuity.
@@
 
-\section{Conclusions}\label{sec:conclusions}
+\section{Conclusions}
 Runaway greenhouse climates on rocky planets cause significant inflation of their radii.
@@
 %We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
-%This work has made use of data from the European Space Agency (ESA) mission {\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia} Data Processing and Analysis Consortium (DPAC, \url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}).
-Funding for the DPAC has been provided by national institutions, in particular the institutions participating in the {\it Gaia} Multilateral Agreement.
 %\end{acknowledgments}

### 2023-04-06 src/tex/ms.tex
Additions to abstract and intro, 230406

@@
 \affiliation{Department of Astronomy/Steward Observatory, The University of Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
+%\author[0000-0002-3286-7683]{Tim Lichtenberg}
+%\affiliation{Kapteyn Astronomical Institute, University of Groningen, PO Box 800, 9700 AV Groningen, The Netherlands}
 \author{al.}
@@
 \begin{abstract}
-    $\ldots$ magma oceans $\ldots$
-
-    Here, we assess the ability of space and ground-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages contextual information from the overall planet population.
-    $\ldots$
-
-%    We argue that in the near future, ESA's \plato\ mission will be the most promising endeavors to constrain this demographic feature.
-%    We identify the key mission design drivers that enable a statistically sound detection.
-%    We also show what survey strategy optimizes the statistical power of the combined dataset.
-%    $\ldots$ its measurement will also provide insights into which stars harbor the nearest habitable worlds.
+    Key atmospheric volatiles are distributed between terrestrial planetary interiors and atmospheres during magma ocean epochs in the course of planetary accretion.
+Long-term magma ocean regimes on rocky exoplanets inside the orbital runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may thus offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
+Thermal stratification of super-runaway planets is expected to inflate the transit radii relative to sub-runaway planets, potentially rendering the climatic transition observationally accessible.
+Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages contextual information from the overall planet population.
+We find that the demographic imprint of super-runaway climates should be detectable with high-precision transit photometry for planet populations $\gtrsim$ 100 planets if at least 10\% of these planets undergo a super-runaway climate.
+We argue that in the near future, ESA's \plato\ mission will be the most promising survey to constrain this demographic feature.
+We identify the key mission design drivers that enable a statistically sound detection, and discuss survey strategies that optimize the statistical power of combined datasets.
+Observational detection of the runaway climate transition will provide crucial insights into the distribution of atmospheric volatiles among low-mass exoplanets, and inform us about which stars harbor the nearest potentially habitable worlds.
 \end{abstract}
@@
 \section{Introduction}
-\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
-\todo[inline]{``Points for the intro:
-\bioverse\\
-Transit survey to measure mass/radius of planets\\
-Does the HZ have an inner edge?\\
-Run-away GH\\
-Magma oceans\\
-Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman.
+%\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021}.}
+%\todo[inline]{``Points for the intro:
+%\bioverse\\
+%Transit survey to measure mass/radius of planets\\
+%Does the HZ have an inner edge?\\
+%Run-away GH\\
+%Magma oceans\\
+%Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman.
 Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
-~\\
-In contrast, PLATO and \nautilus\ are \\
-Possibility of follow-up to measureme masses
-''}
+%~\\
+%In contrast, PLATO and \nautilus\ are \\
+%Possibility of follow-up to measureme masses
+%''}
 
-\todo[inline]{introduce the HZ Inner Edge Discontinuity\textsuperscript{\texttrademark}}
+The currently best-probed regime of rocky exoplanets with a bulk density approaching that of Earth is that of hot ot warm, close-in planets \citep{2019AREPS..47..141J,2021JGRE..12606639B}.
+These bodies experience climatic conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution \citep{2018SSRv..214...76I,2021ChEG...81l5735C}, which sensitively affected the distribution of atmospheric volatiles between planetary core, mantle, and atmosphere.
+Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds \citep{Lichtenberg2022,2022arXiv220310056K}.
 
-\begin{note}
-For observations of rocky exoplanets, the currently best-probed regime is that of warm, close-in planets.
-These bodies experience climatic conditions that are similar to the environment of the inner Solar System bodies at early stages of their formation.
-Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.
-
-...
-    Venus and Earth, while having accreted from the same mass reservoir and despite their similar bulk properties, evolved into planets with very different environmental conditions on their surfaces.
-    At the formation of the Solar System, both planets underwent a giant impact phase [CITE!] that melted their mantles, leading to a magma ocean stage.
+An example from the Solar System for the potential significance of the magma ocean -- or runaway greenhouse phase -- are the different climate states of Venus and Earth \citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
+While having accreted from the same mass reservoir and despite their similar bulk properties, evolved into planets with very different environmental conditions on their surfaces.
+At the formation of the Solar System, both planets underwent a giant impact phase \citep{2020NatGe..13..265G} that melted their mantles, leading to a magma ocean stage \citep{2012AREPS..40..113E,2018RSPTA.37680109S}.
 Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
-    \citet{Hamano2013} showed that such a late desiccation process is not needed to explain the stark differences: for a planet rich in volatiles and receiving high enough radiation levels from its host star, a steam atmosphere limits the outgoing radiation from the (molten) planetary surface.
+However, \citet{Hamano2013} demonstrated theoretically that such a late desiccation process is not needed to explain the stark differences: for a planet rich in volatiles and receiving high enough radiation levels from its host star, a steam atmosphere limits the outgoing radiation from the (molten) planetary surface \citep{2013NatGe...6..661G,2015ExA....40..449L}.
 This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}, enough to remove the entire water reservoir from a rocky planet by hydrodynamic escape.
-    \todo[inline]{important for motivating the existence of the pattern: ``the solidification timescale can become comparable to the main-sequence lifetime of the star (Hamano et al.
-2013, 2015)'' \citep{Lichtenberg2022} }
+Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory \citep{2013Natur.497..607H,2015ApJ...806..216H,Lichtenberg2022}.
 Consequently, it is aptly considered to be the inner boundary of the habitable zone~\citep{Kasting1993,Kopparapu2013}.
-
-    (Under the right conditions, different orbital distances alone can thus decide about a planet's water content and ultimately its habitability.)
+Under the right conditions, different orbital distances alone can thus decide about a planet's water content and ultimately its habitability.
 If received flux makes the difference between our own habitable planet and dry, dead Venus, there is good reason to believe that such patterns exist in exoplanet systems as well.
 
-    ...long-term goal/overarching objective: derive the geophysical history of rocky extrasolar planets.
-More concrete: Constrain the limits of runaway greenhouse transitions, and thus the inner edges of the habitable zone.
-Also: How close is Earth from this runaway greenhouse irradiation limit?
-   ...
-\todo[inline]{introduce magma oceans and their influence on planetary radii~\citep{Dorn2021}.}
-    ...current/future observations of planets that are currently in the runaway greenhouse phase may constrain properties of their planetary mantles...
-make connection between interior and atmospheres.
-    ...
+A long-term goal of exoplanet science is to measure the diversity of exoplanets and the distribution of their climate states through assessment of their geophysical history \citep{2021JGRE..12606711L}.
+This inludes deriving observationally motivated constraints on the limits of the runaway greenhouse transition, and thus the inner edges of the habitable zone.
+Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets, because atmospheric volatiles are highly soluble in magmatic liquids \citep{2021ApJ...909L..22K,Dorn2021}.
+Planets in a super-runaway climate state (actively undergoing a runaway greenhouse) are expected to be thermally inflated \citep{Turbet2019,Turbet2020,Mousis2020b}, which directly increases their tranit radii.
+However, dissolution in the magma may decrease this effect \citep{Dorn2021}, and variations in interior chemical exchange between core and mantle manterial may influence the outgassed volatiles that are visible in the atmosphere by transmission spectroscopy \citep{2021ApJ...914L...4L,2022PSJ.....3..127S}.
+Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their planetary mantles and establish and obversational connection between exoplanetary interior and atmospheres \citep{Lichtenberg2022,Wordsworth2022}.
 
-    The runaway greenhouse transition is a robust prediction from climate models, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
+The runaway greenhouse transition is a robust prediction from climate models \citep{Forget2014,2021ApJ...919..130B,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
 Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
 On the planetary level, however, such undertaking is hindered by degeneracies: From internal composition to atmospheric loss to volatile content, many factors influence a planet's size.
-It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019}.
+It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019,Lichtenberg2019}.
+
 Using \bioverse\ to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we...
@@
 Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.
-
-\end{note}
 
@@
 \bibliographystyle{aasjournal}
-\bibliography{bib,PhD,coauthors}
+\bibliography{bib,PhD,coauthors,refs_tim}
 

### 2023-04-06 src/tex/ms.tex
Additions to abstract and intro, 230406

@@
 \affiliation{Department of Astronomy/Steward Observatory, The University of Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
-%\author[0000-0002-3286-7683]{Tim Lichtenberg}
-%\affiliation{Kapteyn Astronomical Institute, University of Groningen, PO Box 800, 9700 AV Groningen, The Netherlands}
+\author[0000-0002-3286-7683]{Tim Lichtenberg}
+\affiliation{Kapteyn Astronomical Institute, University of Groningen, PO Box 800, 9700 AV Groningen, The Netherlands}
 \author{al.}
@@
 \section{Introduction}
-%\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021}.}
-%\todo[inline]{``Points for the intro:
-%\bioverse\\
-%Transit survey to measure mass/radius of planets\\
-%Does the HZ have an inner edge?\\
-%Run-away GH\\
-%Magma oceans\\
-%Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman.
+\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
+\todo[inline]{``Points for the intro:
+\bioverse\\
+Transit survey to measure mass/radius of planets\\
+Does the HZ have an inner edge?\\
+Run-away GH\\
+Magma oceans\\
+Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman.
 Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
-%~\\
-%In contrast, PLATO and \nautilus\ are \\
-%Possibility of follow-up to measureme masses
-%''}
+~\\
+In contrast, PLATO and \nautilus\ are \\
+Possibility of follow-up to measureme masses
+''}
 

### 2023-04-06 src/tex/ms.tex
Merge Tim's and Martin's Introduction drafts

--- 2023-04-06 src/tex/ms.tex
@@
 
-The currently best-probed regime of rocky exoplanets with a bulk density approaching that of Earth is that of hot ot warm, close-in planets \citep{2019AREPS..47..141J,2021JGRE..12606639B}.
+%\begin{itemize}
+%    \item hot/warm planets are best studied
+%    \item runaway greenhouse climate (Ingersoll1969,Kasting1988,Nakajima1992). Transition is sharp in instellation.
+%    \item HZ concept.''(Kasting et al.
+1993; Kopparapu et al.
+2013, 2014)'' Relevance in the search for habitable planets and life.
+(Science gap:) Does it exist?
+Can it be tested?
+=> statistically detect inner edge
+%    \item Bioverse
+%    \item Transit survey => PLATO
+%    \item Using PLATO as a baseline survey, we do A, B, C.
+%\end{itemize}
+
+\begin{note}
+Recent years have brought tremendous advances in our understanding of terrestrial-sized planets, mainly due to space-based transit surveys such as \kepler\ \citep{Borucki2010} and \tess\ \citep{Ricker2014a} that pushed detection sensitivities into the terrestrial regime.
+Due to the inherent detection biases of the transit technique, the currently best-probed regime of rocky exoplanets with a bulk density approaching that of Earth is that of hot or warm, close-in planets \citep{2019AREPS..47..141J,2021JGRE..12606639B}.
 These bodies experience climatic conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution \citep{2018SSRv..214...76I,2021ChEG...81l5735C}, which sensitively affected the distribution of atmospheric volatiles between planetary core, mantle, and atmosphere.
@@
 
-An example from the Solar System for the potential significance of the magma ocean -- or runaway greenhouse phase -- are the different climate states of Venus and Earth \citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
-While having accreted from the same mass reservoir and despite their similar bulk properties, evolved into planets with very different environmental conditions on their surfaces.
-At the formation of the Solar System, both planets underwent a giant impact phase \citep{2020NatGe..13..265G} that melted their mantles, leading to a magma ocean stage \citep{2012AREPS..40..113E,2018RSPTA.37680109S}.
+An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth \citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
+While having accreted from the same mass reservoir~\citep{Morbidelli2012} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{Kasting1988,Hamano2013,Kane2014,Turbet2021} (Donahue et al.
+1982; Bézard, B.
+\& de Bergh, 2007; Way et al.
+2016; Way \& Del Genio 2020; ).
+At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020NatGe..13..265G} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S}.
+Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets, because atmospheric volatiles are highly soluble in magmatic liquids \citep{2021ApJ...909L..22K,Dorn2021}.
 Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
-However, \citet{Hamano2013} demonstrated theoretically that such a late desiccation process is not needed to explain the stark differences: for a planet rich in volatiles and receiving high enough radiation levels from its host star, a steam atmosphere limits the outgoing radiation from the (molten) planetary surface \citep{2013NatGe...6..661G,2015ExA....40..449L}.
-This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}, enough to remove the entire water reservoir from a rocky planet by hydrodynamic escape.
-Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory \citep{2013Natur.497..607H,2015ApJ...806..216H,Lichtenberg2022}.
-Consequently, it is aptly considered to be the inner boundary of the habitable zone~\citep{Kasting1993,Kopparapu2013}.
-Under the right conditions, different orbital distances alone can thus decide about a planet's water content and ultimately its habitability.
-If received flux makes the difference between our own habitable planet and dry, dead Venus, there is good reason to believe that such patterns exist in exoplanet systems as well.
 
-A long-term goal of exoplanet science is to measure the diversity of exoplanets and the distribution of their climate states through assessment of their geophysical history \citep{2021JGRE..12606711L}.
-This inludes deriving observationally motivated constraints on the limits of the runaway greenhouse transition, and thus the inner edges of the habitable zone.
-Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets, because atmospheric volatiles are highly soluble in magmatic liquids \citep{2021ApJ...909L..22K,Dorn2021}.
-Planets in a super-runaway climate state (actively undergoing a runaway greenhouse) are expected to be thermally inflated \citep{Turbet2019,Turbet2020,Mousis2020b}, which directly increases their tranit radii.
-However, dissolution in the magma may decrease this effect \citep{Dorn2021}, and variations in interior chemical exchange between core and mantle manterial may influence the outgassed volatiles that are visible in the atmosphere by transmission spectroscopy \citep{2021ApJ...914L...4L,2022PSJ.....3..127S}.
-Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their planetary mantles and establish and obversational connection between exoplanetary interior and atmospheres \citep{Lichtenberg2022,Wordsworth2022}.
+However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet.
+Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2013NatGe...6..661G,2015ExA....40..449L}.
+This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen.
+CITE
+%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory \citep{2013Natur.497..607H,2015ApJ...806..216H,Lichtenberg2022}.
+Venus' atmospheric composition is consistent with the planet having gone through this phase, although the past presence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016}.
 
 The runaway greenhouse transition is a robust prediction from climate models \citep{Forget2014,2021ApJ...919..130B,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
-Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
-On the planetary level, however, such undertaking is hindered by degeneracies: From internal composition to atmospheric loss to volatile content, many factors influence a planet's size.
-It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019,Lichtenberg2019}.
+In particular, planets in a super-runaway climate state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii.
+However, dissolution in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy \citep{2021ApJ...914L...4L,2022PSJ.....3..127S}.
+Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their planetary mantles and establish an observational connection between exoplanetary interior and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.
 
-Using \bioverse\ to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we...
+%An Earth-like planet experiences a \SI{\sim 500}{\kilo\meter} radius increase upon crossing the threshold instellation, and its transit depth increases by \SI{10}{\ppm} to \SI{1000}{\ppm} depending on the spectral type of the host star~\citep{Turbet2019}.
 
-Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.
+Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{Hamano2013}.
+CITE
+Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a CO$_2$/H$_2$O/N$_2$ atmosphere~(\citealp[][but other definitions have been put forward, e.g., ]{Kasting1993,Kopparapu2013,Kopparapu2014} \citealp{Hall2023,Spinelli2023}).
+The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1853,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
+Investigating its boundaries in parameter space has helped improve our understanding of habitability. CITE
+Nevertheless, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017}.
+The question naturally emerges if a habitable zone -- in its common form as boundaries in receiving instellation -- indeed exists and how the habitable zone hypothesis can be tested.
+Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
+A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
+Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos ~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
+All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.
+
+Here, we explore the feasibility of a statistical test of the habitable zone hypothesis by surveying planetary bulk properties close to its inner edge, the runaway greenhouse transition.
+Our goal is to assess the ability of near-future transit surveys to test the hypothesis that the runaway greenhouse effect causes an abrupt change of planetary radii and bulk densities when ordered by receiving instellation~\citep{Turbet2019}.
+Our main tool for this is \bioverse, a simulation framework for assessing the statistical power of exoplanet surveys~\citep{Bixel2021}.
+It consists of a sample generator that populates stars from the \gaia\ catalog~\citep{Gaia_mission,Gaia_DR3}  with planetary systems based on state-of-the-art occurrence rates, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
+Trying different instrumentation and survey designs, we utilize \bioverse\ to recover runaway greenhouse-induced effects based on model predictions~\citep{Turbet2020,Dorn2021} that we inject into a baseline planet population.
+
+In particular, we test the capability of the \plato\ \citep[PLAnetary Transits and Oscillation of stars, ][]{Rauer2016} mission, which will measure the radii of a large number of terrestrial-sized planets, to detect the runaway greenhouse transition and identify the astrophysical unknowns on which this depends.
+We then perform a parameter study to explore which trade-offs in the survey design of a \plato-like mission optimize the chance to detect and characterize the runaway greenhouse transition.
+
+We organize the paper as follows:
+Section~\ref{sec:met_baseline} introduces our baseline model produce synthetic star and planet samples.
+    In Sect.~\ref{sec:met_rghmodel}, we describe the model component that produces runaway greenhouse-induced radius changes.
+    Section~\ref{sec:met_surveys-hypotests} explains our survey simulations and hypothesis tests.
+    We present our results in Sect.~\ref{sec:results} before interpreting them in Sect.~\ref{sec:discussion}.
+    Finally, we summarize our findings in Sect.~\ref{sec:conclusions}.
+\end{note}
 

### 2023-04-07 src/tex/ms.tex
edit Introduction, fix citations

@@
 
-\begin{note}
 Recent years have brought tremendous advances in our understanding of terrestrial-sized planets, mainly due to space-based transit surveys such as \kepler\ \citep{Borucki2010} and \tess\ \citep{Ricker2014a} that pushed detection sensitivities into the terrestrial regime.
-Due to the inherent detection biases of the transit technique, the currently best-probed regime of rocky exoplanets with a bulk density approaching that of Earth is that of hot or warm, close-in planets \citep{2019AREPS..47..141J,2021JGRE..12606639B}.
-These bodies experience climatic conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution \citep{2018SSRv..214...76I,2021ChEG...81l5735C}, which sensitively affected the distribution of atmospheric volatiles between planetary core, mantle, and atmosphere.
-Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds \citep{Lichtenberg2022,2022arXiv220310056K}.
+Due to the inherent detection biases of the transit technique, the currently best studied category of rocky exoplanets is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
+These bodies experience climatic conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution~\citep{2018SSRv..214...76I,2021ChEG...81l5735C}, which sensitively affected the distribution of atmospheric volatiles between planetary core, mantle, and atmosphere.
+Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds~\citep{Lichtenberg2022,Krijt2022}.
 
-An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth \citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
+An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
 While having accreted from the same mass reservoir~\citep{Morbidelli2012} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{Kasting1988,Hamano2013,Kane2014,Turbet2021} (Donahue et al.
@@
 At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020NatGe..13..265G} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S}.
-Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets, because atmospheric volatiles are highly soluble in magmatic liquids \citep{2021ApJ...909L..22K,Dorn2021}.
+Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets, because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
 Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
@@
 Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2013NatGe...6..661G,2015ExA....40..449L}.
-This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen.
-CITE
-%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory \citep{2013Natur.497..607H,2015ApJ...806..216H,Lichtenberg2022}.
+This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year} CITE, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen.
+%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,Lichtenberg2022}.
 Venus' atmospheric composition is consistent with the planet having gone through this phase, although the past presence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016}.
 
-The runaway greenhouse transition is a robust prediction from climate models \citep{Forget2014,2021ApJ...919..130B,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
+The runaway greenhouse transition is a robust prediction from climate models~\citep{Forget2014,Boukrouche2021,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
 In particular, planets in a super-runaway climate state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii.
-However, dissolution in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy \citep{2021ApJ...914L...4L,2022PSJ.....3..127S}.
+However, dissolution in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,2022PSJ.....3..127S}.
 Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their planetary mantles and establish an observational connection between exoplanetary interior and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.
@@
 Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a CO$_2$/H$_2$O/N$_2$ atmosphere~(\citealp[][but other definitions have been put forward, e.g., ]{Kasting1993,Kopparapu2013,Kopparapu2014} \citealp{Hall2023,Spinelli2023}).
-The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1853,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
+The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
 Investigating its boundaries in parameter space has helped improve our understanding of habitability. CITE
@@
 Finally, we summarize our findings in Sect.~\ref{sec:conclusions}.
-\end{note}
 
@@
 
-Hardegree-Ullman et al. (submitted) used the Gaia Catalogue of Nearby Stars \citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
+Hardegree-Ullman et al. (submitted) used the Gaia Catalogue of Nearby Stars~\citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
 The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2021} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
@@
 \subsection{Orbit parameters and planet masses}\label{sec:met-orbits_masses}
-Eccentric orbits alter the probability of a planet to transit \citep[e.g.,][]{Barnes2007a}.
-The distribution of eccentricities $e$ of exoplanets has been found to resemble a Beta function \citep{Kipping2013b}, which we chose to draw synthetic eccentricities from.
+Eccentric orbits alter the probability of a planet to transit~\citep[e.g.,][]{Barnes2007a}.
+The distribution of eccentricities $e$ of exoplanets has been found to resemble a Beta function~\citep{Kipping2013b}, which we chose to draw synthetic eccentricities from.
 Following~\citet{Kipping2013b}, we used a Beta distribution with parameters $a=0.867$ and $b=3.03$, and truncated the distribution at $e = 0.8$.

### 2023-04-09 src/tex/ms.tex
Second iteration of introduction

@@
 An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
-While having accreted from the same mass reservoir~\citep{Morbidelli2012} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{Kasting1988,Hamano2013,Kane2014,Turbet2021} (Donahue et al.
-1982; Bézard, B.
-\& de Bergh, 2007; Way et al.
-2016; Way \& Del Genio 2020; ).
-At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020NatGe..13..265G} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S}.
-Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets, because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
+While having accreted from a similar mass reservoir~\citep{2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,2020JGRE..12506276W,Turbet2021}.
+At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020NatGe..13..265G,Liu2022} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
+Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets \citep{2020ChEG...80l5594F}, because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
 Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
 
-However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet.
-Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2013NatGe...6..661G,2015ExA....40..449L}.
-This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year} CITE, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen.
+However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet \citep{Hamano2013}, which is supported by isotopic and petrological evidence from meteoritic samples \citep{2018SSRv..214...36A,2022Natur.611..245B}.
+Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,2013NatGe...6..661G,2015ExA....40..449L,2017JGRE..122.1458S}.
+This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year} \citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen \citep{2015AsBio..15..119L,2013ApJ...778..154W,2014ApJ...785L..20W}.
 %Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,Lichtenberg2022}.
-Venus' atmospheric composition is consistent with the planet having gone through this phase, although the past presence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016}.
+Venus' atmospheric composition is consistent with the planet having gone through this phase \citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.
 
@@
 
-Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{Hamano2013}.
-CITE
-Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a CO$_2$/H$_2$O/N$_2$ atmosphere~(\citealp[][but other definitions have been put forward, e.g., ]{Kasting1993,Kopparapu2013,Kopparapu2014} \citealp{Hall2023,Spinelli2023}).
+Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{2010ppc..book.....P,2013ApJ...765..131K}.
+Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a oxidized CO$_2$/H$_2$O/N$_2$-rich atmospheres~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but can be influenced for different internal redox states and volatile compositions \citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
 The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
-Investigating its boundaries in parameter space has helped improve our understanding of habitability. CITE
-Nevertheless, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017}.
-The question naturally emerges if a habitable zone -- in its common form as boundaries in receiving instellation -- indeed exists and how the habitable zone hypothesis can be tested.
+However, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017}.
+The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predicitive theoretical concept and how the habitable zone hypothesis can be tested observationally.
 Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
@@
 Section~\ref{sec:met_baseline} introduces our baseline model produce synthetic star and planet samples.
-In Sect.~\ref{sec:met_rghmodel}, we describe the model component that produces runaway greenhouse-induced radius changes.
+In Section~\ref{sec:met_rghmodel}, we describe the model component that produces runaway greenhouse-induced radius changes.
 Section~\ref{sec:met_surveys-hypotests} explains our survey simulations and hypothesis tests.
-We present our results in Sect.~\ref{sec:results} before interpreting them in Sect.~\ref{sec:discussion}.
-Finally, we summarize our findings in Sect.~\ref{sec:conclusions}.
+We present our results in Section~\ref{sec:results} before interpreting them in Section~\ref{sec:discussion}.
+Finally, we summarize our findings in Section~\ref{sec:conclusions}.
 

### 2023-04-09 src/tex/ms.tex
Updated Section 3

@@
 \section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
-\todo[inline]{TIM: please review}
+%\todo[inline]{TIM: please review} DONE
 %\todo[inline]{
@@
 %}
-The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019}.
-We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience.
-...
+The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}.
+With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state \citep{2021JGRE..12606711L,2021ApJ...919..130B}.
+We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean \citep{Dorn2021,2023Icar..39015265S}.
 
@@
 
-\textit{Non-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
+\textit{Sub-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
 This case serves as our null hypothesis.
 
-For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
-While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013,Boukrouche2021}.
+\textit{Super-runaway} planets: For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
+While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,2013ApJ...765..131K,Leconte2013,2015ApJ...806..216H,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
 To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
-Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our non-runaway planets~\citep{Zeng2016}.
+Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our sub-runaway planets~\citep{Zeng2016}.
 For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
-This scenario does not take into account the expected additional radius change due to molten interiors~\citep{Bower2019}.
-It further assumes a \textit{dry melt} without dissolved volatiles.
+This scenario does not take into account the expected additional radius increase due to the density deficit of molten interiors~\citep{Bower2019}.
 
-\textit{Wet melt} planets follow the rules above, but we take into account a radius decrease from water partitioning in the melt:
-In the situation of a magma ocean below a steam atmosphere, part of the water will partition into the melt.
-This removes it from the steam atmosphere, which in turn decreases the radius inflation.
+Nominally, the above models assume a \textit{dry melt} without dissolved volatiles.
+Here, however, we consider a \textit{wet melt} magma ocean, take into account a radius decrease from water partitioning in the melt, following \citet{Dorn2021}: In the situation of a magma ocean below a steam atmosphere, part of the water will partition into the melt.
+This removes it from the steam atmosphere, which in turn decreases the radius inflation effect.
 The effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes are predicted to reach measurable values~\citep{Dorn2021}.
@@
 % Earth: ~0.0002 PERCENT or 218 ppm by mass (Turbet+2019)
-In the following, we only distinguish between the non-runaway greenhouse and wet melt scenarios.
+In the following, we only distinguish between the sub-runaway greenhouse and wet melt scenarios.
 
-To account for planets unable to sustain a steam atmosphere as well as evolutionary effects such as desiccation (see bottem panel of Fig.~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
+To account for planets unable to sustain a steam atmosphere over extended time spans, as well as evolutionary effects such as desiccation (see bottem panel of Fig.~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
 It represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
-Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the non-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
-%The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
+Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the sub-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
+The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
 In the following, we test if and under what conditions this parametrization causes a demographic trend that is large enough to be detected with high significance.
@@
 }
-As illustrated in Fig.~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the runaway or non-runaway category.
+As illustrated in Fig.~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the runaway or sub-runaway category.
 For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature.
-Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the non-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
+Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the sub-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
 We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.

### 2023-04-09 src/tex/ms.tex
Section 6.1

@@
 We showed in Sect.~\ref{sec:res_signature} that injecting the theoretically predicted radius inflation effect into a synthetic planet population following the currently known demographics leaves a distinct pattern in the radius and density distribution as a function of instellation.
-Such a statistical imprints of the runaway greenhouse transition was suggested before~\citep{Turbet2019}, and our simulations show quantitatively how the contributions of the individual radius changes combine to create a significant demographic signal.
+The transit radius inflation was suggested before as an observational diagnostic to probe the runaway greenhouse transition~\citep{Turbet2019}, and our simulations show quantitatively how the contributions of the individual radius changes combine to create a significant demographic signal.
 Its strength depends on largely unknown factors such as the planetary volatile content, but turned out to be well in the detectable range under reasonable assumptions of these factors.
@@
 It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
-Such constraints will shed light on whether the initial water content is variable~\citep{Lichtenberg2019} or there is a predominant quantity.
+Such constraints will shed light on whether the initial water content of rocky planetary is bifurcated by systematic inter-system effects~\citep{Lichtenberg2019,2021ApJ...913L..20L,2022ApJ...938L...3L} or if there is a predominant volatile-enrichment across planetary systems that is only modified by intra-system effects, such as migration \citep{Schlecker2021} and escape \citep{Owen2016}.
 Ultimately, such measurements will shed light on the question of if planets like Earth received their water from primordial sources during their formation or if it was supplied later~\citep[e.g.,][]{Raymond2004}.

### 2023-04-09 src/tex/ms.tex
6.2.1

@@
 It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
-Such constraints will shed light on whether the initial water content of rocky planetary is bifurcated by systematic inter-system effects~\citep{Lichtenberg2019,2021ApJ...913L..20L,2022ApJ...938L...3L} or if there is a predominant volatile-enrichment across planetary systems that is only modified by intra-system effects, such as migration \citep{Schlecker2021} and escape \citep{Owen2016}.
-Ultimately, such measurements will shed light on the question of if planets like Earth received their water from primordial sources during their formation or if it was supplied later~\citep[e.g.,][]{Raymond2004}.
+Such constraints will shed light on whether the initial water content of rocky planetary is bifurcated by systematic inter-system effects~\citep{Lichtenberg2019,2021ApJ...913L..20L,2022ApJ...938L...3L} or if there is a predominant pattern of volatile-enrichment across planetary systems that is only modified by intra-system effects, such as migration \citep{Schlecker2021} and escape \citep{Owen2016}.
+Ultimately, such measurements will shed light on the question how delivery and loss effects during rocky planet formation and evolution shape the diversity of exoplanetaty climates.
 %Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
 %Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
-Overall, searching for the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.
+Overall, the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.
 
@@
 
-\textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies on the production of large amounts of water vapor that then acts as a greenhouse gas.
-This process works only on planets with an atmosphere and of sufficiently volatile-rich composition\todo[inline]{TIM: how much?}.
+\textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies sufficient amounts of water vapor available in the atmosphere to act as greenhouse gas, but already about ~10--20 bar of water vapour -- a minor fraction of one Earth ocean (the lower limit of water on Earth) -- is enough to sustain sufficiently high surface temperatures to keep the surface molten, and thus in a magma ocean stage \citep{Boukrouche2021,2021JGRE..12606711L}.
 The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
+From a formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard outcome of formation \citep{2019PNAS..116.9723Z,2020A&A...643L...1V}.
+But while ``typically'' considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the Solar System \citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation.
+Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amount of water inside of the planet during the magma ocean phase \citep{2018SSRv..214...76I,2021ApJ...909L..22K,2022NatAs...6.1296K}.
+Enhanced equilibration between the core, mantle, and atmospehre during magma ocean evolution of rocky exoplanets further enhances this process \citep{2021ApJ...914L...4L,Schlichting2022}.
+Therefore, even nominally dry planets generate substantial amounts of water during formation and early evolution.
+Reduced heating from short-lived radionuclides in extrasolar planetary systems increases the expected water abundance in exoplanet systems further, in particular for M dwarf systems \citep{2022ApJ...938L...3L}.
 
@@
 Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the signal.
+However, the delivery uncertainty is much greater than the predicted loss rates of water by atmospheric escape, which typically is limited to in total a few tens of oceans \citep{2018AJ....155..195W} on even the most irradiated exoplanets \citep[see the discussion in][]{2022ApJ...938L...3L}.
 
-\textit{Prevalent water inventory}: The magnitude of radius change at the runaway greenhouse threshold is sensitive on the water mass fraction.
-As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern:
-        The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
+\textit{Prevalent water inventory}: The magnitude of radius change at the runaway greenhouse threshold is sensitive to the water mass fraction.
+As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern: The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
+Howver, dissolution into the magma ocean and atmospheric inflation show non-linear coupling; thus a straightforward trend is difficult to discern.
 

### 2023-04-09 src/tex/ms.tex
6.2.2.

@@
 An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
-While having accreted from a similar mass reservoir~\citep{2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,2020JGRE..12506276W,Turbet2021}.
-At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020NatGe..13..265G,Liu2022} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
+While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,2020JGRE..12506276W,Turbet2021}.
+At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
 Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets \citep{2020ChEG...80l5594F}, because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
@@
 But while ``typically'' considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the Solar System \citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation.
-Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amount of water inside of the planet during the magma ocean phase \citep{2018SSRv..214...76I,2021ApJ...909L..22K,2022NatAs...6.1296K}.
+Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amount of water inside of the planet during the magma ocean phase \citep{2018SSRv..214...76I,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}.
 Enhanced equilibration between the core, mantle, and atmospehre during magma ocean evolution of rocky exoplanets further enhances this process \citep{2021ApJ...914L...4L,Schlichting2022}.
@@
 Runaway greenhouse climates are not the only physical mechanism that may cause a change in transit radius for a subset of planets.
-Some of the most prominent alternative scenarios involve atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
+Some of the most prominent alternative scenarios involve atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{2012ApJ...753...66I,Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
 Both processes are being traded as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
@@
 Either of them leads to a decrease of planet radius for planets close to their host star, which is distinct from the radius inflation introduced by runaway greenhouse climates.
+For example, the K2-3 system shows tentative evidence for a break in the planet radii, contrary to what would be expected from atmospheric escape \citep{2022AJ....164..172D}.
 
@@
 However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
-If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{Venturini2020,Burn2021,Schlecker2021,Schlecker2021b}, systems of all ages can be affected by this dilution.
+If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{2020A&A...643L...1V,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals \citep{2019NatAs...3..307L,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.z
 

### 2023-04-09 src/tex/ms.tex
6.2.3.

@@
 However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
-If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{2020A&A...643L...1V,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals \citep{2019NatAs...3..307L,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.z
+If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{2020A&A...643L...1V,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals \citep{2019NatAs...3..307L,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.
 
-\subsubsection{OPTIONAL...: Atmospheric signatures}\label{sec:dis_atmospheres}
+\subsubsection{Atmospheric spectral signatures}\label{sec:dis_atmospheres}
 \todo[inline]{TIM: Ideas?\\
@@
 4]{Turbet2019}.}
+a
+We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population.
+(i) Detecting the atmospheric windows of water vapour in the near- to mid-infrared.
+In a super-runaway state the atmosphere is dominated by the opacity of water vapour, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m \citep[e.g.][]{2021ApJ...919..130B}.
+This would require an instrument with capabilities in this wavelenght regime, for instance ELT METIS, JWST MIRI, or future missions such as LIFE \citep{2019A&A...621A.125B,2022A&A...664A..21Q}.
+Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime \citep{2020SSRv..216...98G}.
+(ii) Post-runaway planets may be detectable by O$_{3}$ absoprtion due to build-up of abiotic oxygen, left-over from photochemical dissociation and hydrogen loss \citep{2014ApJ...785L..20W,2015AsBio..15..119L}.
+(iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere \citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
+This may be detectable in high-resolution observations focusing on isotope-sensitive transitions \citep{2019AJ....158...26L,2019A&A...622A.139M}.
+(iv) Disequilibrium chemistry in tidally-locked super-runaway planets.
+It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers \citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interace \citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
+Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma \citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
+Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed.
+This has been suggested to be done via (v) eclipse photometry \citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidifed magma ocean \citep{2020ApJ...898..160E,Fortin2022}.
 

### 2023-04-09 src/tex/ms.tex
6.3.

@@
 However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
-If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{2020A&A...643L...1V,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals \citep{2019NatAs...3..307L,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.
+If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{2020A&A...643L...1V,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals \citep{Lichtenberg2019,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.
 
 \subsubsection{Atmospheric spectral signatures}\label{sec:dis_atmospheres}
-\todo[inline]{TIM: Ideas?\\
-discuss potential atmospheric signatures of magma oceans.
+%\todo[inline]{TIM: Ideas?\\
+%discuss potential atmospheric signatures of magma oceans.
 e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g.
@@
 4]{Turbet2019}.}
-a
+
 We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population.
@@
 (iv) Disequilibrium chemistry in tidally-locked super-runaway planets.
-It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers \citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interace \citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
+It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers \citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface \citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
 Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma \citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
@@
 As suggested by \citet{Turbet2019}, an alternative approach for detecting the runaway greenhouse-induced radius inflation is to search for its ``local'' imprints in multi-planet systems.
-Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
+Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, K2-3 \citep{2022AJ....164..172D}, or Kepler-138 \citep{2022NatAs.tmp..269P}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
 While degeneracies remain in interpreting bulk density fluctuations within individual systems~\citep[e.g.,][]{Turbet2020,Dorn2021}, the detection of a consistent pattern in several such systems could be a convincing statistical evidence of the runaway greenhouse transition.
@@
 The latter depends on the lifetimes of runaway greenhouse phases, which are a function of the initial water inventory of the planets~\citep{Hamano2015}.
-Overall, it seems feasible to derive the typical water content of rocky planets from these occurrence estimates.
+Overall, it seems feasible to derive the typical water content of low-mass exoplanets from these occurrence estimates.
 
@@
 Other mission concepts exist that focus on characterizing exoplanet habitability, most notably the \hwo~\citep{Gaudi2020c,LUVOIR2019} and the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2021}.
-Such direct imaging surveys, however, do not directly measure planetary radii and are therefore most useful for providing context through atmospheric measurements of individual planets.
-\todo[inline]{TIM: feel free to add something about LIFE if you see fit}
+In principle, direct imaging surveys do not directly measure planetary radii and are therefore most useful for providing context through atmospheric measurements of individual planets.
+However, because mid-infrared retrievals do not suffer from the degeneracy between cloud albedo and changes in surface area, the planet radius can be retrieved in mid-infrared wavelengths \citep{2018ExA....46..543D,2021ExA...tmp..118Q}.
+Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light \citep{Wordsworth2022}.
+Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios (Konrad+, Alei+).
+Finally, since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages \citep{2014ApJ...784...27L,2019A&A...621A.125B}.
+With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states \citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.
 
-As for exoplanets missions in their implementation phase, \plato\ overall remains to be the most promising mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.
+As for exoplanets missions in their implementation phase, however, \plato\ overall remains to be the most promising mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.
 

### 2023-04-09 src/tex/ms.tex
6.6.2.

@@
 In a super-runaway state the atmosphere is dominated by the opacity of water vapour, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m \citep[e.g.][]{2021ApJ...919..130B}.
-This would require an instrument with capabilities in this wavelenght regime, for instance ELT METIS, JWST MIRI, or future missions such as LIFE \citep{2019A&A...621A.125B,2022A&A...664A..21Q}.
+This would require an instrument with capabilities in this wavelenght regime, for instance ELT METIS, JWST MIRI, or future missions such as LIFE \citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}.
 Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime \citep{2020SSRv..216...98G}.
@@
 However, while interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure.
-The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure.\todo[inline]{TIM: do you have a reference for this statement?}
+The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure \citep[][, add Noack+ paper]{2016ApJ...819..127Z,2021JGRE..12606724B}.
 We thus consider our mass-radius relationship a conservative case and do not expect a negative impact on the detectability of the runaway greenhouse transition due to deviations from it.
@@
 
-The models in this study do not into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in runaway greenhouse planets~\citep{Ikoma2018}.
-Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{Turbet2019}.
+The models in this study do not into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions ~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
+Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{Lichtenberg2022,2022ARA&A..60..159W}.
 

### 2023-04-09 src/tex/ms.tex
6.6.3

@@
 \subsubsection{Simplifications in the atmosphere model}
-\todo[inline]{TIM: your expertise will be valuable here.\\
-Mention we only consider instellation-induced rgh.\\
-- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
-- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
-- assumption of a single albedo. ``if there is a range of albedos, will that wash away the signal?''
-}
-As illustrated in Fig.~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the runaway or sub-runaway category.
+%\todo[inline]{TIM: your expertise will be valuable here.\\
+%Mention we only consider instellation-induced rgh.\\
+%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets %within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
+%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
+%- assumption of a single albedo. ``if there is a range of albedos, will that wash away the signal?''
+%}
+% >>>>>>>>>>
+% TL: I just wrote a review article about this exact problem (magma ocean atmosphere exchange) with Yamila Miguel, hopefully soon on arXiv.
+Honestly, I think we keep it short here, otherwise the paper will get way too long and diluted.
+I referenced the latest reviews on this now, and we can cite Lichtenberg & Miguel 23 as soon as it's on arXiv for a much much longer discussion.
+% >>>>>>>>>>
+As illustrated in Fig.~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the super- or sub-runaway category.
 For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature.
 Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the sub-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
+Developing more sophisticated models of the coupled interaction between atmosphere and interior for an arbitrary planetary composition remains a long-term goal of the community \citep{2021SSRv..217...22G,Lichtenberg2022,2022ARA&A..60..159W}.
 We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.

### 2023-04-09 src/tex/ms.tex
6.6.4. till end

@@
 The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
-Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021}.
+Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,2020A&A...643L...1V,Emsenhuber2021b,Schlecker2021,Lichtenberg2022a,2022ApJ...939L..19I}.
 Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.

### 2023-04-09 src/tex/ms.tex
Some references clean-up

--- 2023-04-09 src/tex/ms.tex
@@
 These bodies experience climatic conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution~\citep{2018SSRv..214...76I,2021ChEG...81l5735C}, which sensitively affected the distribution of atmospheric volatiles between planetary core, mantle, and atmosphere.
-Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds~\citep{Lichtenberg2022,Krijt2022}.
+Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds~\citep{2022arXiv220310023L,Krijt2022}.
 
@@
 While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,2020JGRE..12506276W,Turbet2021}.
-At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
+At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,2022arXiv220310023L}.
 Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets \citep{2020ChEG...80l5594F}, because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
@@
 This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year} \citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen \citep{2015AsBio..15..119L,2013ApJ...778..154W,2014ApJ...785L..20W}.
-%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,Lichtenberg2022}.
+%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,2022arXiv220310023L}.
 Venus' atmospheric composition is consistent with the planet having gone through this phase \citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.
@@
 However, dissolution in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,2022PSJ.....3..127S}.
-Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their planetary mantles and establish an observational connection between exoplanetary interior and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.
+Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their planetary mantles and establish an observational connection between exoplanetary interior and atmospheres~\citep{2022arXiv220310023L,Wordsworth2022}.
 
@@
 The models in this study do not into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions ~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
-Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{Lichtenberg2022,2022ARA&A..60..159W}.
+Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{2022arXiv220310023L,2022ARA&A..60..159W}.
 
@@
 Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the sub-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
-Developing more sophisticated models of the coupled interaction between atmosphere and interior for an arbitrary planetary composition remains a long-term goal of the community \citep{2021SSRv..217...22G,Lichtenberg2022,2022ARA&A..60..159W}.
+Developing more sophisticated models of the coupled interaction between atmosphere and interior for an arbitrary planetary composition remains a long-term goal of the community \citep{2021SSRv..217...22G,2022arXiv220310023L,2022ARA&A..60..159W}.
 We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.
@@
 The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
-Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021}.
+Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,2020A&A...643L...1V,Emsenhuber2021b,Schlecker2021,2022ApJ...938L...3L,2022ApJ...939L..19I}.
 Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.

### 2023-04-11 src/tex/ms.tex
review introduction

@@
 At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,2022arXiv220310023L}.
-Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets \citep{2020ChEG...80l5594F}, because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
+Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets~\citep{2020ChEG...80l5594F} because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
 Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
 
-However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet \citep{Hamano2013}, which is supported by isotopic and petrological evidence from meteoritic samples \citep{2018SSRv..214...36A,2022Natur.611..245B}.
+However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet~\citep{Hamano2013}, which is supported by isotopic and petrological evidence from meteoritic samples \citep{2018SSRv..214...36A,2022Natur.611..245B}.
 Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,2013NatGe...6..661G,2015ExA....40..449L,2017JGRE..122.1458S}.
-This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year} \citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen \citep{2015AsBio..15..119L,2013ApJ...778..154W,2014ApJ...785L..20W}.
+This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year} \citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen \citep{2013ApJ...778..154W,2014ApJ...785L..20W,2015AsBio..15..119L}.
 %Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,2022arXiv220310023L}.
@@
 Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{2010ppc..book.....P,2013ApJ...765..131K}.
-Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a oxidized CO$_2$/H$_2$O/N$_2$-rich atmospheres~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but can be influenced for different internal redox states and volatile compositions \citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
+Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone.
+Its prevalent definition refers to the possibility of liquid water on an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmospheres~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but it can be influenced by different internal redox states and volatile compositions~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
 The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
-However, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017}.
-The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predicitive theoretical concept and how the habitable zone hypothesis can be tested observationally.
+However, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
+The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.
+
 Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
@@
 %- Limitation to small planets without significant atmospheric envelopes.\\
-%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
+%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{2022arXiv220310023L}.\\
 %- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
@@
 %Mention we only consider instellation-induced rgh.\\
-%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets %within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
+%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets %within the runaway greenhouse regime~\citep{2022arXiv220310023L}.\\
 %- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\

### 2023-04-11 src/tex/ms.tex
review Tim's edits of discussion

@@
 
-However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet~\citep{Hamano2013}, which is supported by isotopic and petrological evidence from meteoritic samples \citep{2018SSRv..214...36A,2022Natur.611..245B}.
+However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet~\citep{Hamano2013}, which is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
 Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,2013NatGe...6..661G,2015ExA....40..449L,2017JGRE..122.1458S}.
-This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year} \citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen \citep{2013ApJ...778..154W,2014ApJ...785L..20W,2015AsBio..15..119L}.
+This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,2015AsBio..15..119L}.
 %Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,2022arXiv220310023L}.
-Venus' atmospheric composition is consistent with the planet having gone through this phase \citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.
+Venus' atmospheric composition is consistent with the planet having gone through this phase~\citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.
 
@@
 \section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
-%\todo[inline]{TIM: please review} DONE
 %\todo[inline]{
@@
 The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}.
-With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state \citep{2021JGRE..12606711L,2021ApJ...919..130B}.
-We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean \citep{Dorn2021,2023Icar..39015265S}.
+With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state~\citep{2021JGRE..12606711L,2021ApJ...919..130B}.
+We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean~\citep{Dorn2021,2023Icar..39015265S}.
 
@@
 
-\textit{Super-runaway} planets: For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
+For \textit{super-runaway} planets, i.e., all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
 While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,2013ApJ...765..131K,Leconte2013,2015ApJ...806..216H,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
@@
 Nominally, the above models assume a \textit{dry melt} without dissolved volatiles.
-Here, however, we consider a \textit{wet melt} magma ocean, take into account a radius decrease from water partitioning in the melt, following \citet{Dorn2021}: In the situation of a magma ocean below a steam atmosphere, part of the water will partition into the melt.
+Here, however, we consider a \textit{wet melt} magma ocean and take into account a radius decrease from water partitioning in the melt~\citep{Dorn2021}: In the situation of a magma ocean below a steam atmosphere, part of the water will partition into the melt.
 This removes it from the steam atmosphere, which in turn decreases the radius inflation effect.
@@
 It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
-Such constraints will shed light on whether the initial water content of rocky planetary is bifurcated by systematic inter-system effects~\citep{Lichtenberg2019,2021ApJ...913L..20L,2022ApJ...938L...3L} or if there is a predominant pattern of volatile-enrichment across planetary systems that is only modified by intra-system effects, such as migration \citep{Schlecker2021} and escape \citep{Owen2016}.
-Ultimately, such measurements will shed light on the question how delivery and loss effects during rocky planet formation and evolution shape the diversity of exoplanetaty climates.
+Such constraints will provide insight into whether the initial water content of rocky planets is bifurcated by systematic inter-system effects~\citep{Lichtenberg2019,2021ApJ...913L..20L,2022ApJ...938L...3L} or if there is a predominant pattern of volatile-enrichment across planetary systems that is only modified by intra-system effects such as planet migration~\citep{Schlecker2021} or atmospheric escape~\citep{Owen2016}.
+Ultimately, such measurements will shed light on how delivery and loss effects during rocky planet formation and evolution shape the diversity of exoplanetary climates.
 %Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
 %Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
-Overall, the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.
+Overall, probing the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.
 
@@
 
-\textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies sufficient amounts of water vapor available in the atmosphere to act as greenhouse gas, but already about ~10--20 bar of water vapour -- a minor fraction of one Earth ocean (the lower limit of water on Earth) -- is enough to sustain sufficiently high surface temperatures to keep the surface molten, and thus in a magma ocean stage \citep{Boukrouche2021,2021JGRE..12606711L}.
+\textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies on sufficient amounts of atmospheric water vapor that can act as a greenhouse gas.
+However, already about \SIrange{\sim 10}{20}~bar of water vapor -- corresponding to a minor fraction of one Earth ocean and thus the lower limit of water on Earth -- is enough to sustain sufficiently high surface temperatures to keep the planet in a magma ocean stage~\citep{Boukrouche2021,2021JGRE..12606711L}.
 The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
-From a formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard outcome of formation \citep{2019PNAS..116.9723Z,2020A&A...643L...1V}.
-But while ``typically'' considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the Solar System \citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation.
-Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amount of water inside of the planet during the magma ocean phase \citep{2018SSRv..214...76I,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}.
-Enhanced equilibration between the core, mantle, and atmospehre during magma ocean evolution of rocky exoplanets further enhances this process \citep{2021ApJ...914L...4L,Schlichting2022}.
+From a planet formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard outcome~\citep{2019PNAS..116.9723Z,2020A&A...643L...1V}.
+But while commonly considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the Solar System~\citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation.
+Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amounts of water inside of the planet during the magma ocean phase~\citep{2018SSRv..214...76I,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}.
+Enhanced equilibration between the core, mantle, and atmosphere during magma ocean evolution of rocky exoplanets further enhances this process~\citep{2021ApJ...914L...4L,Schlichting2022}.
 Therefore, even nominally dry planets generate substantial amounts of water during formation and early evolution.
-Reduced heating from short-lived radionuclides in extrasolar planetary systems increases the expected water abundance in exoplanet systems further, in particular for M dwarf systems \citep{2022ApJ...938L...3L}.
+Reduced heating from short-lived radionuclides in extrasolar planetary systems increases the expected water abundance in exoplanet systems further, in particular for M~dwarf systems~\citep{2022ApJ...938L...3L}.
 
@@
 Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the signal.
-However, the delivery uncertainty is much greater than the predicted loss rates of water by atmospheric escape, which typically is limited to in total a few tens of oceans \citep{2018AJ....155..195W} on even the most irradiated exoplanets \citep[see the discussion in][]{2022ApJ...938L...3L}.
+However, the delivery uncertainty is much greater than the predicted loss rates of water by atmospheric escape, which typically is limited to in total a few tens of terrestrial oceans per Gyr~\citep{2018AJ....155..195W} on even the most irradiated exoplanets~\citep[see discussion in][]{2022ApJ...938L...3L}.
 
 \textit{Prevalent water inventory}: The magnitude of radius change at the runaway greenhouse threshold is sensitive to the water mass fraction.
-As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern: The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
-Howver, dissolution into the magma ocean and atmospheric inflation show non-linear coupling; thus a straightforward trend is difficult to discern.
+As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern:
+The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
+However, dissolution into the magma ocean and atmospheric inflation show non-linear coupling; thus, a straightforward trend is difficult to discern.
 
@@
 Either of them leads to a decrease of planet radius for planets close to their host star, which is distinct from the radius inflation introduced by runaway greenhouse climates.
-For example, the K2-3 system shows tentative evidence for a break in the planet radii, contrary to what would be expected from atmospheric escape \citep{2022AJ....164..172D}.
+For example, the \mbox{K2-3} system shows tentative evidence for a break in the planet radii, contrary to what would be expected from atmospheric escape~\citep{2022AJ....164..172D}.
 
@@
 However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
-If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{2020A&A...643L...1V,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals \citep{Lichtenberg2019,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.
+If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{Venturini2020,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals~\citep{Lichtenberg2019,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.
 
@@
 (i) Detecting the atmospheric windows of water vapour in the near- to mid-infrared.
-In a super-runaway state the atmosphere is dominated by the opacity of water vapour, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m \citep[e.g.][]{2021ApJ...919..130B}.
-This would require an instrument with capabilities in this wavelenght regime, for instance ELT METIS, JWST MIRI, or future missions such as LIFE \citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}.
-Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime \citep{2020SSRv..216...98G}.
-(ii) Post-runaway planets may be detectable by O$_{3}$ absoprtion due to build-up of abiotic oxygen, left-over from photochemical dissociation and hydrogen loss \citep{2014ApJ...785L..20W,2015AsBio..15..119L}.
-(iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere \citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
-This may be detectable in high-resolution observations focusing on isotope-sensitive transitions \citep{2019AJ....158...26L,2019A&A...622A.139M}.
+In a super-runaway state the atmosphere is dominated by the opacity of water vapour, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.][]{2021ApJ...919..130B}.
+This would require an instrument with capabilities in this wavelenght regime, for instance ELT METIS, JWST MIRI, or future missions such as LIFE~\citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}.
+Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime~\citep{2020SSRv..216...98G}.
+(ii) Post-runaway planets may be detectable by O$_{3}$ absoprtion due to build-up of abiotic oxygen, left-over from photochemical dissociation and hydrogen loss~\citep{2014ApJ...785L..20W,2015AsBio..15..119L}.
+(iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
+This may be detectable in high-resolution observations focusing on isotope-sensitive transitions~\citep{2019AJ....158...26L,2019A&A...622A.139M}.
 (iv) Disequilibrium chemistry in tidally-locked super-runaway planets.
-It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers \citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface \citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
-Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma \citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
+It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers~\citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface~\citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
+Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma~\citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
 Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed.
-This has been suggested to be done via (v) eclipse photometry \citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidifed magma ocean \citep{2020ApJ...898..160E,Fortin2022}.
+This has been suggested to be done via (v) eclipse photometry~\citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidifed magma ocean~\citep{2020ApJ...898..160E,Fortin2022}.
 
@@
 As suggested by \citet{Turbet2019}, an alternative approach for detecting the runaway greenhouse-induced radius inflation is to search for its ``local'' imprints in multi-planet systems.
-Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, K2-3 \citep{2022AJ....164..172D}, or Kepler-138 \citep{2022NatAs.tmp..269P}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
+Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, K2-3~\citep{2022AJ....164..172D}, or Kepler-138~\citep{2022NatAs.tmp..269P}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
 While degeneracies remain in interpreting bulk density fluctuations within individual systems~\citep[e.g.,][]{Turbet2020,Dorn2021}, the detection of a consistent pattern in several such systems could be a convincing statistical evidence of the runaway greenhouse transition.
@@
 In principle, direct imaging surveys do not directly measure planetary radii and are therefore most useful for providing context through atmospheric measurements of individual planets.
-However, because mid-infrared retrievals do not suffer from the degeneracy between cloud albedo and changes in surface area, the planet radius can be retrieved in mid-infrared wavelengths \citep{2018ExA....46..543D,2021ExA...tmp..118Q}.
-Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light \citep{Wordsworth2022}.
+However, because mid-infrared retrievals do not suffer from the degeneracy between cloud albedo and changes in surface area, the planet radius can be retrieved in mid-infrared wavelengths~\citep{2018ExA....46..543D,2021ExA...tmp..118Q}.
+Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light~\citep{Wordsworth2022}.
 Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios (Konrad+, Alei+).
-Finally, since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages \citep{2014ApJ...784...27L,2019A&A...621A.125B}.
+Finally, since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
 With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states \citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.
@@
 However, while interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure.
-The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure \citep[][, add Noack+ paper]{2016ApJ...819..127Z,2021JGRE..12606724B}.
+The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure~\citep[][, add Noack+ paper]{2016ApJ...819..127Z,2021JGRE..12606724B}.
 We thus consider our mass-radius relationship a conservative case and do not expect a negative impact on the detectability of the runaway greenhouse transition due to deviations from it.

### 2023-04-11 src/tex/ms.tex
review discussion on atmospheric spectral features
...and a few minor edits

@@
 %Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman.
-Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
+Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, \life\.\\
 %~\\
@@
 We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population.
-(i) Detecting the atmospheric windows of water vapour in the near- to mid-infrared.
-In a super-runaway state the atmosphere is dominated by the opacity of water vapour, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.][]{2021ApJ...919..130B}.
-This would require an instrument with capabilities in this wavelenght regime, for instance ELT METIS, JWST MIRI, or future missions such as LIFE~\citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}.
+(i) Detecting the atmospheric windows of water vapor in the near- to mid-infrared.
+In a super-runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.][]{2021ApJ...919..130B}.
+This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS, JWST~MIRI, or future missions such as \life\ ~\citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}.
+CITE METIS, MIRI!
 Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime~\citep{2020SSRv..216...98G}.
-(ii) Post-runaway planets may be detectable by O$_{3}$ absoprtion due to build-up of abiotic oxygen, left-over from photochemical dissociation and hydrogen loss~\citep{2014ApJ...785L..20W,2015AsBio..15..119L}.
+(ii) Post-runaway planets may be detectable by O$_{3}$ absorption due to build-up of abiotic oxygen, left-over from photochemical dissociation, and hydrogen loss~\citep{2014ApJ...785L..20W,2015AsBio..15..119L}.
 (iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
@@
 Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed.
-This has been suggested to be done via (v) eclipse photometry~\citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidifed magma ocean~\citep{2020ApJ...898..160E,Fortin2022}.
+This has been suggested to be done via (v) eclipse photometry~\citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidified magma ocean~\citep{2020ApJ...898..160E,Fortin2022}.
 
@@
 As suggested by \citet{Turbet2019}, an alternative approach for detecting the runaway greenhouse-induced radius inflation is to search for its ``local'' imprints in multi-planet systems.
-Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, K2-3~\citep{2022AJ....164..172D}, or Kepler-138~\citep{2022NatAs.tmp..269P}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
+Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, \mbox{K2-3}~\citep{2022AJ....164..172D}, or \mbox{Kepler-138}~\citep{2022NatAs.tmp..269P}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
 While degeneracies remain in interpreting bulk density fluctuations within individual systems~\citep[e.g.,][]{Turbet2020,Dorn2021}, the detection of a consistent pattern in several such systems could be a convincing statistical evidence of the runaway greenhouse transition.
@@
 Other mission concepts exist that focus on characterizing exoplanet habitability, most notably the \hwo~\citep{Gaudi2020c,LUVOIR2019} and the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2021}.
-In principle, direct imaging surveys do not directly measure planetary radii and are therefore most useful for providing context through atmospheric measurements of individual planets.
+Direct imaging surveys do not directly measure planetary radii and are primarily useful for providing context through atmospheric measurements of individual planets.
 However, because mid-infrared retrievals do not suffer from the degeneracy between cloud albedo and changes in surface area, the planet radius can be retrieved in mid-infrared wavelengths~\citep{2018ExA....46..543D,2021ExA...tmp..118Q}.
 Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light~\citep{Wordsworth2022}.
-Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios (Konrad+, Alei+).
+Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios (Konrad+, Alei+). CITE!
 Finally, since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
-With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states \citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.
+With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states~\citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.
 
@@
 The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure~\citep[][, add Noack+ paper]{2016ApJ...819..127Z,2021JGRE..12606724B}.
+CITE!
 We thus consider our mass-radius relationship a conservative case and do not expect a negative impact on the detectability of the runaway greenhouse transition due to deviations from it.
@@
 
-The models in this study do not into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions ~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
-Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{2022arXiv220310023L,2022ARA&A..60..159W}.
+The models in this study do not into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
+Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{2022arXiv220310023L,Wordsworth2022}.
 
@@
 Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the sub-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
-Developing more sophisticated models of the coupled interaction between atmosphere and interior for an arbitrary planetary composition remains a long-term goal of the community \citep{2021SSRv..217...22G,2022arXiv220310023L,2022ARA&A..60..159W}.
+Developing more sophisticated models of the coupled interaction between atmosphere and interior for an arbitrary planetary composition remains a long-term goal of the community~\citep{2021SSRv..217...22G,2022arXiv220310023L,Wordsworth2022}.
 We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.

### 2023-04-11 src/tex/ms.tex
resolve questions, remove duplicate citations

@@
 Due to the inherent detection biases of the transit technique, the currently best studied category of rocky exoplanets is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
-These bodies experience climatic conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution~\citep{2018SSRv..214...76I,2021ChEG...81l5735C}, which sensitively affected the distribution of atmospheric volatiles between planetary core, mantle, and atmosphere.
+These bodies experience climatic conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which sensitively affected the distribution of atmospheric volatiles between planetary core, mantle, and atmosphere.
 Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds~\citep{2022arXiv220310023L,Krijt2022}.
@@
 However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet~\citep{Hamano2013}, which is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
-Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,2013NatGe...6..661G,2015ExA....40..449L,2017JGRE..122.1458S}.
-This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,2015AsBio..15..119L}.
-%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,2022arXiv220310023L}.
+Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
+This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
+%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,Hamano2015,2022arXiv220310023L}.
 Venus' atmospheric composition is consistent with the planet having gone through this phase~\citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.
@@
 The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}.
-With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state~\citep{2021JGRE..12606711L,2021ApJ...919..130B}.
+With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state~\citep{2021JGRE..12606711L,Boukrouche2021}.
 We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean~\citep{Dorn2021,2023Icar..39015265S}.
@@
 For \textit{super-runaway} planets, i.e., all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
-While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,2013ApJ...765..131K,Leconte2013,2015ApJ...806..216H,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
+While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
 To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
@@
 The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
-From a planet formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard outcome~\citep{2019PNAS..116.9723Z,2020A&A...643L...1V}.
+From a planet formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard outcome~\citep{2019PNAS..116.9723Z,Venturini2020}.
 But while commonly considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the Solar System~\citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation.
-Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amounts of water inside of the planet during the magma ocean phase~\citep{2018SSRv..214...76I,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}.
+Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amounts of water inside of the planet during the magma ocean phase~\citep{Ikoma2018,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}.
 Enhanced equilibration between the core, mantle, and atmosphere during magma ocean evolution of rocky exoplanets further enhances this process~\citep{2021ApJ...914L...4L,Schlichting2022}.
@@
 Either of them leads to a decrease of planet radius for planets close to their host star, which is distinct from the radius inflation introduced by runaway greenhouse climates.
-For example, the \mbox{K2-3} system shows tentative evidence for a break in the planet radii, contrary to what would be expected from atmospheric escape~\citep{2022AJ....164..172D}.
+For example, the innermost planet in the \mbox{K2-3} system has an increased radius compared to its outer siblings, contrary to what would be expected from atmospheric escape~\citep{2022AJ....164..172D}.
 
@@
 (i) Detecting the atmospheric windows of water vapor in the near- to mid-infrared.
-In a super-runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.][]{2021ApJ...919..130B}.
+In a super-runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.][]{Boukrouche2021}.
 This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS, JWST~MIRI, or future missions such as \life\ ~\citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}.
@@
 Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime~\citep{2020SSRv..216...98G}.
-(ii) Post-runaway planets may be detectable by O$_{3}$ absorption due to build-up of abiotic oxygen, left-over from photochemical dissociation, and hydrogen loss~\citep{2014ApJ...785L..20W,2015AsBio..15..119L}.
+(ii) Post-runaway planets may be detectable by O$_{3}$ absorption due to build-up of abiotic oxygen, left-over from photochemical dissociation, and hydrogen loss~\citep{2014ApJ...785L..20W,Luger2015}.
 (iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
@@
 However, while interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure.
-The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure~\citep[][, add Noack+ paper]{2016ApJ...819..127Z,2021JGRE..12606724B}.
+The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure~\citep[][, add Noack+ paper]{Zeng2016,2021JGRE..12606724B}.
 CITE!
@@
 The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
-Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,2020A&A...643L...1V,Emsenhuber2021b,Schlecker2021,2022ApJ...938L...3L,2022ApJ...939L..19I}.
+Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021,2022ApJ...938L...3L,2022ApJ...939L..19I}.
 Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.

### 2023-04-13 src/tex/ms.tex
make HnullHmo figure two-column
edit discussion based on feedback

@@
 To characterize the population-level imprint of individual radius changes, we generated a generic planet population with an injected runaway greenhouse effect assuming a water fraction $x_\mathrm{H_2O} = \var{wrr}$.
-\begin{figure}
+\begin{figure*}
+%    \script{figurescript.py}
     \begin{centering}
@@
     \end{centering}
-\end{figure}
+\end{figure*}
 Figure~\ref{fig:HnullHmo} shows the resulting planetary radii.
@@
 \end{figure}
-Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the HZ Inner Edge Discontinuity.
+Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the habitable zone inner edge discontinuity.
 Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
@@
 %We now examine the value of a follow-up campaign to constrain planet masses with respect to the detectability of the runaway greenhouse threshold.
-Figure~\ref{fig:plato_fgrid} shows that investigating the HZ Inner Edge Discontinuity is more constraining when radius measurements can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
+Figure~\ref{fig:plato_fgrid} shows that investigating the habitable zone inner edge discontinuity is more constraining when radius measurements can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
 Consequently, we achieve a statistically significant detection with smaller samples or lower dilution factors $f_\mathrm{rgh}$.
@@
 \subsubsection{Dependence on host star spectral type}\label{sec:results_FGK_M}
-Since the receiving radiation is sensitive on the spectral type of the host star, the share of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
+Since the incident radiation depends on the spectral type of the host star, the relative number of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
 We tested the detectability of the runaway greenhouse signal when only FGK or only M~dwarfs are considered (see Fig.~\ref{fig:plato_fgrid}).
@@
 \end{figure}
-A key constraint resulting from a successful detection is a measurement of $S_{thresh}$.
+A key constraint resulting from a detection is a measurement of $S_{thresh}$.
 Here, we assess the ability of different mission concepts explored above to constrain this parameter.
@@
 It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
-Such constraints will provide insight into whether the initial water content of rocky planets is bifurcated by systematic inter-system effects~\citep{Lichtenberg2019,2021ApJ...913L..20L,2022ApJ...938L...3L} or if there is a predominant pattern of volatile-enrichment across planetary systems that is only modified by intra-system effects such as planet migration~\citep{Schlecker2021} or atmospheric escape~\citep{Owen2016}.
+Such constraints will provide insight into whether the initial water content of rocky planets is varied by systematic inter-system effects~\citep[e.g.,][]{Raymond2004,Mulders2015b,Sato2016,Lichtenberg2019,2021ApJ...913L..20L,2022ApJ...938L...3L} or if there is a predominant pattern of volatile-enrichment across planetary systems that is only modified by intra-system effects such as planet migration~\citep{Schlecker2021} or atmospheric escape~\citep{Owen2016}.
 Ultimately, such measurements will shed light on how delivery and loss effects during rocky planet formation and evolution shape the diversity of exoplanetary climates.
@@
 We showed in Sect.~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
-However, even assuming capable instrumentation and an optimal survey design, geophysical and demographic unknowns may jeopardize this undertaking.
-The detectability of the runaway greenhouse threshold is thus a function of both survey design and astrophysical circumstances.
+Of course, even assuming capable instrumentation and an optimal survey design, our current models may not capture fully the complexity of trends in the geophysics and demographics of small planets.
+The detectability of the runaway greenhouse threshold is thus a function of both survey design and the diverse outcomes of rocky planet formation and evolution.
+In the following, we will explore the key factors that may influence the emerging demographic signal.
 
-\subsubsection{Key factors influencing tests of the runaway greenhouse hypothesis}
+\subsubsection{Key factors influencing tests of the runaway greenhouse hypothesis}\ref{sec:dis_keyfactors}
 What influences the probability of correctly rejecting a false null hypothesis?
-We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey.
+We identified the following six drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey:
+\begin{itemize}
+    \item Occurrence rate of planets forming steam atmospheres
+    \item Planetary evolution and duration of the steam atmosphere phase
+    \item Prevalent water inventory
+    \item Size and composition of the planetary sample
+    \item Radius measurement precision
+    \item Availability and precision of mass measurements
+\end{itemize}
+We will now briefly explore the above drivers.
 
@@
 As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern:
-The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
+The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the likelihood that a discontinuity is detectable.
 However, dissolution into the magma ocean and atmospheric inflation show non-linear coupling; thus, a straightforward trend is difficult to discern.
@@
 \textit{Size and composition of the planetary sample}: Evidently, the significance of a statistical trend increases with a larger sample size.
-In addition, the sample must include planets on both sides of the instellation threshold, which, due to the strongly distance-dependent detection bias associated with the transit method, is only sufficiently given for low-mass host stars.
+In addition, the sample must include planets on both sides of the instellation threshold.
+Due to the strongly distance-dependent detection bias associated with the transit method and the temporal coverage of upcoming transit missions, this is only likely for low-mass host stars.
 
-\textit{Radius measurement precision}: The more precise individual planet radii can be determined, the less smeared out the pattern will be.
+\textit{Radius measurement precision}: The more precise individual planet radii can be determined, the more pronounced the discontinuity will be.
 Good \textit{accuracy} is less important, as long as it does not have a systematic error scaling with stellar irradiance.
 
-\textit{Availability and precision of mass measurements}: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space.
+\textit{Availability and precision of mass measurements}: For simple geometric reasons ($\rho \propto R^{-3}$), the expected discontinuity at the habitable zone inner edge is stronger when measured in bulk density than it is in planet radius space.
 If transiting planets can be followed up to obtain mass measurements, the statistical significance increases.
@@
 Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.
-This can be different for young host stars when their ages can not be well constrained; in particular the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).
+This can be different for young host stars when their ages cannot be well constrained; in particular, the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).
 
@@
 Some of the most prominent alternative scenarios involve atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{2012ApJ...753...66I,Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
-Both processes are being traded as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
+Both processes are being discussed as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
 %    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
-Either of them leads to a decrease of planet radius for planets close to their host star, which is distinct from the radius inflation introduced by runaway greenhouse climates.
+Either of them leads to a decrease of planet radius for planets close to their host star~\citep{Pascucci2019,Bergsten2022}, which is distinct from the radius inflation introduced by runaway greenhouse climates.
 For example, the innermost planet in the \mbox{K2-3} system has an increased radius compared to its outer siblings, contrary to what would be expected from atmospheric escape~\citep{2022AJ....164..172D}.
@@
 This trend does not represent a false positive scenario for the runaway greenhouse transition, since no strong dependency on instellation has been found or is expected.
-However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
-If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{Venturini2020,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals~\citep{Lichtenberg2019,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.
+A population of ``water worlds'' with low bulk densities on a wide range of orbits would merely attenuate the statistical runaway greenhouse imprint.
+If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{Venturini2020,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals~\citep{Lichtenberg2019,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this attenuation.
 
@@
 \subsection{Diagnostic power of near-future exoplanet missions}\label{sec:dis_samplesize}
-The fidelity of a future detection or falsification of a runaway greenhouse signature will depend on the significance with which the null hypothesis can be excluded.
-As a consequence, instrumentation and survey strategy aiming at probing the runaway greenhouse threshold should aim at maximizing the probability of a true positive detection.
-
-As discussed above, key drivers from a mission design perspective are sample size, photometric precision, and the availability of planets around low-mass host stars.
+Confirming or disproving the predicted habitable zone inner edge discontinuity will depend on the significance with which the null hypothesis can be excluded, which is a function of instrumentation and survey strategy.
+As discussed in Section~\label{sec:dis_keyfactors}, key drivers from a mission design perspective are sample size, photometric precision, and the availability of planets around low-mass host stars.
 We found that along these axes, \plato\ will be the most favorable among the upcoming transit missions.
-The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).
+The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission.% (Matuszewski, priv. comm.).
 %In the FGK sample of the mission, several hundred Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
-With an expected transit radius precision of \SI{3}{\percent}~\citep{plato2017} and a yield of $> 100$ planets studied with such high precision~\citep{Rauer2021}, \plato\ is comparable to the optimistic survey we introduced in Sect.~\ref{sec:res_testability}.
+With an expected transit radius precision of \SI{3}{\percent}~\citep{plato2017} for hundreds of planets~\citep{Rauer2021}, the \plato\ mission is comparable to the optimistic survey (see Section~\ref{sec:res_testability}) in terms of sample size and precision.
 If successful, it should readily detect the predicted statistical imprint or, in case of a non-detection, provide strong upper limits on the occurrence rate of runaway greenhouse planets.
@@
 
-What other planned missions are suited to test the runaway greenhouse hypothesis?
-\kepler\ and \ktwo\ have contributed a large number of discovered terrestrial-sized planets, but few of them are in the habitable zone and their host stars are typically too faint for RV follow-up~\citep{Dressing2015}.
+What other planned missions are suited to probe the habitable zone inner edge discontinuity?
+\kepler\ and \ktwo\ have contributed a large number of discovered terrestrial-sized planets, but few of them are in the habitable zone and their host stars are typically too faint for RV follow-up with current instrumentation~\citep{Dressing2015}.
 
@@
 
-Looking further ahead, the \nautilus\ Space Observatory concept~\citep{Apai2019} represents a statistical mission able to provide precise radius measurements of a large sample of small exoplanets.
-It employs a swarm of large telescopes using ultralight diffractive-refractive optical elements~\citep{Milster2020} with the primary goal to study the atmospheres of transiting exoplanets.
+Looking further ahead, the \nautilus\ Space Observatory concept~\citep{Apai2019} represents a statistical mission able to provide precise radius measurements of a large sample ($\sim \num{1000}$) of small exoplanets.
+It employs a constellation of $\sim \num{35}$ large-diameter ($D \sim \SI{8.5}{\meter}$) telescopes using ultralight diffractive-refractive optical elements~\citep{Milster2020} with the primary goal to study the atmospheres of transiting exoplanets.
 Operating in an array mode, \nautilus\ would achieve the equivalent light-collecting area of a \SI{50}{\meter} telescope.
@@
 
-Other mission concepts exist that focus on characterizing exoplanet habitability, most notably the \hwo~\citep{Gaudi2020c,LUVOIR2019} and the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2021}.
+Other missions have been proposed that focus on characterizing exoplanet habitability, most notably the \hwo\ concept, which will build on the two precursor direct imaging concepts LUVOIR~\citep{LUVOIR2019} and HabEx~\citep{Gaudi2020c}.
+Similar science objectives are pursued by the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2021}, a mission concept utilizing a space-based mid-infrared nulling interferometer.
 Direct imaging surveys do not directly measure planetary radii and are primarily useful for providing context through atmospheric measurements of individual planets.

### 2023-04-14 src/tex/ms.tex
address comments on Discussion

@@
 Its prevalent definition refers to the possibility of liquid water on an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmospheres~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but it can be influenced by different internal redox states and volatile compositions~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
-The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
+The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in the National Aeronautics and Space Administration (NASA) Astrophysics Data System returned $\sim\num{1700}$ results.}.
 However, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
@@
 
-\subsection{Mission design trades}\label{sec:mission-design-trades}
+\subsection{Mission design trade studies}\label{sec:mission-design-trades}
 To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
@@
 \subsubsection{The value of follow-up campaigns}
-The dependencies of a detection change when additional information beyond planet radii is available for the characterized planet population.
-As runaway greenhouse phases leave a stronger imprint in bulk density than in planet radius (see Sect.~\ref{sec:res_followup}), it would be beneficial to obtain constraints on planetary masses and test the runaway greenhouse hypothesis in density space instead of radius space.
+The constraining power changes when additional information beyond planet radii is available for the characterized planet population.
+As runaway greenhouse phases leave a stronger imprint on bulk density than on planet radius (see Sect.~\ref{sec:res_followup}), it would be beneficial to obtain constraints on planetary masses and test the runaway greenhouse hypothesis in density space instead of radius space.
 This way, useful results can be obtained under more pessimistic conditions, e.g., a low predominant water content of planetary surfaces and atmospheres or a smaller available planet sample.
@@
 Precise ground-based radial velocity measurements will be needed to provide these data, and a number of instruments are already successfully employed in characterizing terrestrial-sized exoplanets~\citep[e.g.,][]{Queloz2001a,Pepe2010,Johnson2010b,Ribas2023}.
-Recently, the National Aeronautics and Space Administration (NASA) and the National Science Foundation (NSF) commissioned an ``Extreme Precision Radial Velocity (EPRV) Initiative''~\citep{Crass2021} to develop methods and facilities for precise mass measurements of temperate terrestrial planets.
+A new generation of instruments on extremely large telescopes such as \gclef\ on the \gmt ~\citep{Szentgyorgyi2016}, \andes\ on the \elt ~\citep{Marcantonio2022}, or \modhis\ on the \tmt ~\citep{Mawet2019} will open up the discovery space even further.
+
+Recently, NASA and the National Science Foundation (NSF) commissioned an ``Extreme Precision Radial Velocity Initiative''~\citep{Crass2021} to develop methods and facilities for precise mass measurements of temperate terrestrial planets.
 Their findings highlight that such measurements are costly, and therefore follow-up efforts may only be available for a subsample of the targets of a mission of \plato's scale.
-The diagnostic power of the hypothesis tests we demonstrated here may be improved by simultaneously fitting for the trend in radius space in the subsample without RV follow-up.
-An optimized mission in search for the inner edge of the habitable zone will further enhance its information content via a wise selection of follow-up targets, i.e., balancing objects located on either side of the expected instellation threshold $S_\mathrm{thresh}$.
+The diagnostic power of the hypothesis tests we demonstrated here may be improved by simultaneously fitting for the habitable zone inner edge discontinuity in the subsample without RV follow-up.
+An optimized mission in search for the inner edge of the habitable zone will further enhance its information content via a informed selection of follow-up targets, i.e., balancing objects located on either side of the expected instellation threshold $S_\mathrm{thresh}$.
 
@@
 To date, the majority of planets with radius measurements orbit FGK~dwarfs, and, based on the instellation they receive, most of them lie in the runaway greenhouse regime~\citep{Thompson2018}.
-Obviously, a step-function-like signal in the exoplanet demographics like the one investigated above can not be constrained well if only one side of the discontinuity is being sampled.
+Obviously, a radius/density discontinuity in the exoplanet demographics like the habitable zone inner edge discontinuity cannot be constrained well if only one side of the discontinuity is being sampled.
 This, however, is the situation for planetary systems around Sun-like stars -- their habitable zones are so distant that transiting planets within them are very rare due to pure geometrical reasons.
@@
 Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars,][]{Turbet2019}, enhancing the demographic signal it leaves.
-An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of \SI{100}{\mega\year}~\citep{Luger2015}.
+An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of gigayears~\citep{Luger2015}.
 This increases the probability of observing any given planet in the sample during the runaway greenhouse phase, essentially driving $f_\mathrm{rgh}$ to higher values.
@@
 influences the planet's climate (greenhouse effect).}
-A potential for liquid water on the surface of a planet is commonly used as an environmental marker to assess its habitability~\citep{Huang1959,Hart1978,Kasting1993,Kaltenegger2011,Kopparapu2013}.
+A potential for liquid water on the surface of a planet is commonly used as an environmental marker to assess its surface habitability~\citep{Huang1959,Hart1978,Kasting1993,Kaltenegger2011,Kopparapu2013}.
 The runaway greenhouse transition represents an upper bound on received irradiation for this condition.
@@
 In Sect.~\ref{sec:res_constrain-S_thresh}, we show that the threshold instellation at which the runaway greenhouse transition occurs can be reasonably constrained without imposing overly optimistic conditions on the underlying planet population, instrumentation, or survey strategy.
-A mission like \plato\ is well equipped to perform this measurement; the chances for success are directly proportional to the proportion of characterized planets around M dwarfs and to the number of planets for which masses can be determined.
+A mission like \plato\ is well equipped to perform this measurement; the constraining power is directly proportional to the proportion of characterized planets around M dwarfs and to the number of planets for which masses can be determined.
 
@@
 For example, detections of water vapor in planets above the threshold instellation, combined with precise radius measurements, would constrain the predominant water content of terrestrial planets.
-A follow-up campaign with the \ariel\ mission of \plato\ planets could be considered to make this contribution.
+%A follow-up campaign with the \ariel\ mission of \plato\ planets could be considered to make this contribution.
 
@@
 This is not the case in the model from \citet{Bergsten2022} that we adopted: Its occurrence rate density varies smoothly in the domain relevant for the runaway greenhouse hypothesis; transitions only occur at smaller instellations ($< \SI{50}{\watt\per\meter\squared}$) and larger radii ($> 1.6\,R_\oplus$).
-We thus do not expect the model underlying our planet sample to affect the recovery of the transition.
+We thus do not expect the model underlying our planet sample to affect the investigation of the habitable zone inner edge discontinuity.
 However, it should be kept in mind that any unknown sharp features in the true distribution of terrestrial planets could complicate a detection or lead to a false positive result (see Sect.~\ref{sec:dis_falsepositive}).
@@
 
-The models in this study do not into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
+The models in this study do not take into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
 Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{2022arXiv220310023L,Wordsworth2022}.
@@
 
-\subsubsection{``Sharpness'' of the demographic trend}
+\subsubsection{``Sharpness'' of the habitable zone inner edge discontinuity}
 The real counterpart to our idealized step-function central to our hypothesis tests could be blurred by several factors.
 Its amplitude is affected by ``sub-Neptunes'' with significant Hydrogen/Helium envelopes, which can disguise as inflated rocky planets.
-They might not contribute to the demographic signal we are looking for, although a similar transition was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
+They might not contribute to the habitable zone inner edge discontinuity, although a similar transition was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
 A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, turning them from planets appearing inflated into much smaller \mbox{(super-)Earths}.
@@
 
-All of the above mechanisms dilute the demographic imprint of the runaway greenhouse transition with contributions that are challenging to quantify.
+All of the above mechanisms may dilute the demographic imprint of the runaway greenhouse transition with contributions that are challenging to quantify.
 The ``dilution factor" we introduced embodies our ignorance of these contributions.

### 2023-04-14 src/tex/ms.tex
address comments on Conclusion

@@
 \section{Conclusions}\label{sec:conclusions}
-Runaway greenhouse climates on rocky planets cause significant inflation of their radii.
-Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius discontinuity resulting from this inflation in the exoplanet population.
+A robust prediction of runaway greenhouse models is a significant inflation of rocky planet radii.
+Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius/density discontinuity resulting from this inflation in the exoplanet population.
 Our key findings are as follows:
@@
     \item The predicted runaway greenhouse transition causes a discontinuity in the radius and density distribution of small exoplanets with respect to the incident host star irradiation.
-    \item The demographic imprint of the runaway greenhouse threshold should be detectable with high-precision transit measurements.
+    \item This habitable zone inner edge discontinuity should be detectable with high-precision transit measurements.
           For a large sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.

### 2023-04-14 src/tex/ms.tex
brand the pattern as "habitable zone inner edge discontinuity"

@@
 Long-term magma ocean regimes on rocky exoplanets inside the orbital runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may thus offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
-Thermal stratification of super-runaway planets is expected to inflate the transit radii relative to sub-runaway planets, potentially rendering the climatic transition observationally accessible.
+Thermal stratification of super-runaway planets is expected to inflate the transit radii relative to sub-runaway planets, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius/density space.
 Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages contextual information from the overall planet population.
 We find that the demographic imprint of super-runaway climates should be detectable with high-precision transit photometry for planet populations $\gtrsim$ 100 planets if at least 10\% of these planets undergo a super-runaway climate.
-We argue that in the near future, ESA's \plato\ mission will be the most promising survey to constrain this demographic feature.
+We argue that in the near future, ESA's \plato\ mission will be the most promising survey to detect and constrain habitable zone inner edge discontinuity.
 We identify the key mission design drivers that enable a statistically sound detection, and discuss survey strategies that optimize the statistical power of combined datasets.
@@
 \subsection{Hypothesis tests}
-We now turn to quantifying the ability of the simulated surveys to detect the runaway greenhouse transition and to constrain parameters of this trend.
+We now turn to quantifying the ability of the simulated surveys to detect the habitable zone inner edge discontinuity and to constrain parameters of the runaway greenhouse transition.
 To do this, we rely on a Bayesian hypothesis testing approach where we quantify the evidence of a hypothesis over another based on the (simulated) data.
@@
 Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
-In this case, the pattern was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).
+In this case, the habitable zone inner edge discontinuity was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).
 
@@
             \textit{Right:} Evidences when only FGK or only M~dwarfs are considered.
-                Only M~dwarfs host enough planets on both sides of the threshold instellation to allow a reliable detection of the runaway greenhouse signal.
+                Only M~dwarfs host enough planets on both sides of the threshold instellation to allow a reliable detection of the habitable zone inner edge discontinuity.
             }
@@
 The minimum needed fraction rises to \var{platoMinfrghHundred} for $N = 100$.
-For much smaller samples, only a very optimistically strong signal is likely to be detected.
+For much smaller samples, only an optimistically strong signal is likely to be detected.
 
@@
 \subsubsection{Additional planet mass measurements}\label{sec:res_followup}
-Comparing a measurement of the signal in radius space with a measurement in density space (which requires planetary mass measurements), a stronger detection occurs in the latter case:
+Comparing a measurement of the habitable zone inner edge discontinuity in radius space with a measurement in density space (which requires planetary mass measurements), a stronger detection occurs in the latter case:
 With an optimistic choice of geophysical parameters (see Sect.~\ref{sec:res_testability}), the average measured radius change is \SI{+15}{\percent} whereas the average density change is \SI{-33}{\percent}.
@@
 %We now examine the value of a follow-up campaign to constrain planet masses with respect to the detectability of the runaway greenhouse threshold.
-Figure~\ref{fig:plato_fgrid} shows that investigating the habitable zone inner edge discontinuity is more constraining when radius measurements can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
+Figure~\ref{fig:plato_fgrid} shows that investigations of the discontinuity are more constraining when radius measurements can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
 Consequently, we achieve a statistically significant detection with smaller samples or lower dilution factors $f_\mathrm{rgh}$.
@@
 Since the incident radiation depends on the spectral type of the host star, the relative number of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
-We tested the detectability of the runaway greenhouse signal when only FGK or only M~dwarfs are considered (see Fig.~\ref{fig:plato_fgrid}).
+We tested the detectability of the habitable zone inner edge discontinuity when only FGK or only M~dwarfs are considered (see Fig.~\ref{fig:plato_fgrid}).
 %($d_\mathrm{max} = \SI{75}{\parsec}, M_\mathrm{G} < 11$)
@@
 
-Our finding that the expected trend is strong except for extremely low water mass fractions gives reason for optimism regarding its detection.
+Our finding that the expected habitable zone inner edge discontinuity is strong except for extremely low water mass fractions gives reason for optimism regarding its detection.
 It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
@@
 
-\subsection{Detectability of the runaway greenhouse transition}
+\subsection{Detectability of the habitable zone inner edge discontinuity}
 We showed in Sect.~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
@@
 \textit{Planetary evolution and duration of the steam atmosphere phase}: An inflated steam atmosphere can only be sustained until the planet has lost its water.
-Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the signal.
+Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the habitable zone inner edge discontinuity.
 However, the delivery uncertainty is much greater than the predicted loss rates of water by atmospheric escape, which typically is limited to in total a few tens of terrestrial oceans per Gyr~\citep{2018AJ....155..195W} on even the most irradiated exoplanets~\citep[see discussion in][]{2022ApJ...938L...3L}.
@@
 \citet{Luque2022} found that small planets around orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
-This trend does not represent a false positive scenario for the runaway greenhouse transition, since no strong dependency on instellation has been found or is expected.
+This trend does not represent a false positive scenario for the habitable zone inner edge discontinuity, since no strong dependency on instellation has been found or is expected.
 A population of ``water worlds'' with low bulk densities on a wide range of orbits would merely attenuate the statistical runaway greenhouse imprint.
@@
 \subsection{Mission design trade studies}\label{sec:mission-design-trades}
-To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
+To explore the impact of mission trade-offs on the detectability of the habitable zone inner edge discontinuity, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
 We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.
@@
 Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
-This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the statistical radius inflation trend, also since the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
+This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the habitable zone inner edge discontinuity, also since the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
 Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
@@
 All of the above mechanisms may dilute the demographic imprint of the runaway greenhouse transition with contributions that are challenging to quantify.
-The ``dilution factor" we introduced embodies our ignorance of these contributions.
+The ``dilution factor'' we introduced embodies our ignorance of these contributions.
 Importantly, while this may offset our estimates of the signal's amplitude, its shape remains the same.

### 2023-04-14 src/tex/ms.tex
"Fig." -> "Figure"

@@
 Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic samples of stars that host planets according to the observed exoplanet demographics.
-We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).\todo[inline]{make sure rgh hypothesis is mentioned before, or rephrase}
+We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Figure~\ref{fig:flowchart}).\todo[inline]{make sure rgh hypothesis is mentioned before, or rephrase}
 \begin{figure*}
@@
 
-To assign masses to our planets, we use the semi-empirical mass-radius relationship assuming a pure $\mathrm{MgSiO_3}$ composition from \citet{Zeng2016} (see green line in Fig.~\ref{fig:radiusevolution}).
+To assign masses to our planets, we use the semi-empirical mass-radius relationship assuming a pure $\mathrm{MgSiO_3}$ composition from \citet{Zeng2016} (see green line in Figure~\ref{fig:radiusevolution}).
 This represents the baseline bulk density before any climate-related effects are applied.
@@
 Relevant properties are a planet's mass $M$, its net instellation $S$, and its bulk water inventory expressed as a water mass fraction $x_{H_2O}$.
-We consider the following cases (see Fig.~\ref{fig:radiusevolution}):
+We consider the following cases (see Figure~\ref{fig:radiusevolution}):
 
@@
 
-To account for planets unable to sustain a steam atmosphere over extended time spans, as well as evolutionary effects such as desiccation (see bottem panel of Fig.~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
+To account for planets unable to sustain a steam atmosphere over extended time spans, as well as evolutionary effects such as desiccation (see bottem panel of Figure~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
 It represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
@@
 To simulate the transit survey of ESA's \plato\ mission, we considered a volume-limited sample with a size according to projections and including all stellar spectral types.
-Given \plato's expected radius precision, we find that a yield of \var{N_plato} planets enables a significant detection if the fraction of runaway greenhouse planets is larger than $ f_\mathrm{rgh} \sim \var{platoMinfrgh}$ (see Fig.~\ref{fig:plato_fgrid}).
+Given \plato's expected radius precision, we find that a yield of \var{N_plato} planets enables a significant detection if the fraction of runaway greenhouse planets is larger than $ f_\mathrm{rgh} \sim \var{platoMinfrgh}$ (see Figure~\ref{fig:plato_fgrid}).
 The minimum needed fraction rises to \var{platoMinfrghHundred} for $N = 100$.
@@
 Since the incident radiation depends on the spectral type of the host star, the relative number of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
-We tested the detectability of the habitable zone inner edge discontinuity when only FGK or only M~dwarfs are considered (see Fig.~\ref{fig:plato_fgrid}).
+We tested the detectability of the habitable zone inner edge discontinuity when only FGK or only M~dwarfs are considered (see Figure~\ref{fig:plato_fgrid}).
 %($d_\mathrm{max} = \SI{75}{\parsec}, M_\mathrm{G} < 11$)
@@
 Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.
-This can be different for young host stars when their ages cannot be well constrained; in particular, the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).
+This can be different for young host stars when their ages cannot be well constrained; in particular, the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Figure~\ref{fig:luminosity_tracks}).
 
@@
 We thus do not expect the model underlying our planet sample to affect the investigation of the habitable zone inner edge discontinuity.
-However, it should be kept in mind that any unknown sharp features in the true distribution of terrestrial planets could complicate a detection or lead to a false positive result (see Sect.~\ref{sec:dis_falsepositive}).
+If any currently unknown sharp features in the distribution of terrestrial planets emerge, they should be considered in future studies.
+%However, it should be kept in mind that any unknown sharp features in the true distribution of terrestrial planets could complicate a detection or lead to a false positive result (see Sect.~\ref{sec:dis_falsepositive}).
 
@@
 \subsubsection{Simplifications in the atmosphere model}
-%\todo[inline]{TIM: your expertise will be valuable here.\\
 %Mention we only consider instellation-induced rgh.\\
@@
 % >>>>>>>>>>
-As illustrated in Fig.~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the super- or sub-runaway category.
+As illustrated in Figure~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the super- or sub-runaway category.
 For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature.
@@
             Detection of the runaway greenhouse threshold with a binned statistic.
-            Using simulated data from the optimistic survey case ($N = \var{N_optimistic}$, compare Fig.~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis.
+            Using simulated data from the optimistic survey case ($N = \var{N_optimistic}$, compare Figure~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis.
 Instead of using a running mean, we computed a binned statistic (arithmetic mean and standard deviation in blue) to assign planets the average radius or bulk density in their neighborhood in instellation space.

### 2023-04-14 src/tex/ms.tex
reframe "assumptions" sections from caveats to future refinement

@@
 \subsection{Impact of assumptions on our findings}
-\todo[inline]{Caveats\\
-- not yet explored: how sensitive are we on the assumed planet population?
+%\todo[inline]{Caveats\\
+%- not yet explored: how sensitive are we on the assumed planet population?
 $\exists$ danger of false positive detection on occurrence rate ``ridges'' like the Neptune desert?
@@
 We could, with relatively moderate effort, run a sanity check on an alternative sample (e.g., the more structured Bern model populations, or the population in the Genesis database)}
-The prospects for detecting the runaway greenhouse transition depends on astro- and geophysical factors, as well as on the specific instrumentation and survey strategy of a particular mission.
-We can only approximate all this here, and we therefore seek to assess the sensitivity of our results on the assumptions we have made.
+The prospects for probing the runaway greenhouse transition depends on astro- and geophysical factors, as well as on the specific instrumentation and survey strategy of a particular mission.
+%We can only approximate all this here, and we therefore seek to assess the sensitivity of our results on the assumptions we have made.
+Our state-of-the-art models approximate the situation and offer testable predictions.
+In the following, we review a few considerations that future models may include to refine these predictions.
 
 \subsubsection{Structures in the planet occurrence rate density}
-In principle, the baseline occurrence rate density in radius-period space that governs the generation of our synthetic planet sample might influence our findings, especially if it contains any features that coincide with the injected demographic feature.
+The baseline occurrence rate density in radius-period space that governs the generation of synthetic planets might influence our findings, especially if it contains any features that coincide with the injected demographic feature.
 This is not the case in the model from \citet{Bergsten2022} that we adopted: Its occurrence rate density varies smoothly in the domain relevant for the runaway greenhouse hypothesis; transitions only occur at smaller instellations ($< \SI{50}{\watt\per\meter\squared}$) and larger radii ($> 1.6\,R_\oplus$).
@@
 Our baseline mass-radius relationship assuming pure $\mathrm{MgSiO_3}$ interiors~\citep{Zeng2016} might not be representative of the rocky planet population.
-However, while interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure.
-The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure~\citep[][, add Noack+ paper]{Zeng2016,2021JGRE..12606724B}.
-CITE!
-We thus consider our mass-radius relationship a conservative case and do not expect a negative impact on the detectability of the runaway greenhouse transition due to deviations from it.
-Nevertheless, we performed a sanity check to assess the influence varying our baseline model on our results (see Appendix~\ref{app:MR_relation}), and found general agreement between different interior composition models.
+However, while interior composition may introduce an offset to the radius habitable zone inner edge discontinuity, we do not expect a change of its structure.
+Since the magnitude of the radius inflation effect is expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure~\citep{Zeng2016,2021JGRE..12606724B},
+\todo[inline]{Tim: add Noack+ paper}
+we consider our mass-radius relationship a conservative case.
+We performed a sanity check to assess the impact of varying our baseline model (see Appendix~\ref{app:MR_relation}) and found general agreement between different interior compositions.
 
-The models in this study do not take into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
-Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{2022arXiv220310023L,Wordsworth2022}.
+The models in this study neglect a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
+Future self-consistent modeling of interior-atmosphere interactions may fill these gaps~\citep{2022arXiv220310023L,Wordsworth2022}.
 
-\subsubsection{Simplifications in the atmosphere model}
+\subsubsection{Atmosphere model}
 %Mention we only consider instellation-induced rgh.\\
@@
 Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the sub-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
+We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.
 Developing more sophisticated models of the coupled interaction between atmosphere and interior for an arbitrary planetary composition remains a long-term goal of the community~\citep{2021SSRv..217...22G,2022arXiv220310023L,Wordsworth2022}.
-We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.
 
@@
 Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
-This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the habitable zone inner edge discontinuity, also since the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
+This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the habitable zone inner edge discontinuity.
+%, and the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
 Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
@@
 \subsubsection{``Sharpness'' of the habitable zone inner edge discontinuity}
-The real counterpart to our idealized step-function central to our hypothesis tests could be blurred by several factors.
-Its amplitude is affected by ``sub-Neptunes'' with significant Hydrogen/Helium envelopes, which can disguise as inflated rocky planets.
-They might not contribute to the habitable zone inner edge discontinuity, although a similar transition was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
-A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, turning them from planets appearing inflated into much smaller \mbox{(super-)Earths}.
-Further dilution comes from the population of planets that lack an atmosphere or sufficient volatiles that can create a steam atmosphere.
-There is also a time dependence, both because stellar luminosities change with age and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.
+The real counterpart to the idealized step-function in our hypothesis tests could be blurred by several factors.
+Planets that lack an atmosphere or sufficient volatiles cannot bear a steam atmosphere.
+Those with significant Hydrogen/Helium envelopes can disguise as inflated rocky planets and might not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was recently suggested for water-dominated ``sub-Neptunes''~\citep{Pierrehumbert2022}.
+A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, reducing their transit radius.
+A time dependence is introduced through stellar luminosity evolution and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.
 
 Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Turbet2021,Pierrehumbert2022}.
-The choice of statistical estimator for the hypothesis tests may further blur the recovered discontinuity.
-We compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.
+The choice of a statistical estimator for the hypothesis tests may further influence the recovered discontinuity; we compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.
 
-All of the above mechanisms may dilute the demographic imprint of the runaway greenhouse transition with contributions that are challenging to quantify.
-The ``dilution factor'' we introduced embodies our ignorance of these contributions.
-Importantly, while this may offset our estimates of the signal's amplitude, its shape remains the same.
+All of the above mechanisms may affect the habitable zone inner edge discontinuity, and their contributions are challenging to quantify.
+However, while they may offset the signal's amplitude, they preserve its general shape.
+The ``dilution factor'' we introduced embodies our ignorance of the magnitude of this offset.
 

### 2023-04-15 src/tex/ms.tex
complete runaway greenhouse model section

@@
 
-\todo[inline]{ARNAUD: revise definition of $S$ following our discussion on 2022-12-22. Introduce net instellation and clarify what is meant by it.}
-\begin{note}
-    The power absorbed by the planet from the host star per unit area at its orbital distance or net instellation $S$ in units of Earth's insolation is given by
-    \begin{equation}
-        \frac{S}{S_\oplus} = \left(\frac{L_\star}{L_\odot}\right) \left(\frac{au}{a}\right)^2 .
-    \end{equation}
-    We assume global redistribution of incoming flux and a wavelength and spherically integrated albedo, or Bond albedo, of $0.3$, comparable to Earth's~\citep{Haar1971}.
-    The absolute value of $S_\oplus$, by which all synthetic planets are scaled, is then \SI{238}{\watt\per\square\meter}.
-%    We further define the solar-equivalent semi-major axis $a_{eff} = a (L_\star/L_\odot)^{-1/2}$, at which a planet experiences the same instellation as a Solar System planet at an orbital distance~$a$.
-   We do not take into account additional heating sources such as tidal effects~\citep[e.g.][]{Barnes2013}.
-\end{note}
+The net absorbed stellar fluxes of planets are sensitive on their albedos, which, however, are generally poorly constrained CITE.
+Here, we assume global redistribution of incoming flux and a fixed Bond albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
+We do not take into account additional heating sources such as tidal effects~\citep[e.g.,][]{Barnes2013}.
 
@@
 
-For \textit{super-runaway} planets, i.e., all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
+For \textit{super-runaway} planets, i.e., all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
 While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
@@
 Nominally, the above models assume a \textit{dry melt} without dissolved volatiles.
-Here, however, we consider a \textit{wet melt} magma ocean and take into account a radius decrease from water partitioning in the melt~\citep{Dorn2021}: In the situation of a magma ocean below a steam atmosphere, part of the water will partition into the melt.
-This removes it from the steam atmosphere, which in turn decreases the radius inflation effect.
-The effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes are predicted to reach measurable values~\citep{Dorn2021}.
-The magnitude of this effect is again dependent on the planet mass and on the water mass fraction.
+Here, however, we consider a \textit{wet melt} magma ocean and take into account a radius decrease from water partitioning in the melt, which is expected in the situation of a magma ocean below a steam atmosphere~\citep{Dorn2021}.
+The magnitude of this effect depends on the planet's mass and water content and is generally small compared to the radius inflation from a steam atmosphere.
 We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
@@
 Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the sub-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
-The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
+The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere.
 In the following, we test if and under what conditions this parametrization causes a demographic trend that is large enough to be detected with high significance.

### 2023-04-15 src/tex/ms.tex
draft a title
sort out TODOs and minor changes

@@
 % Title
-\title{The imprint of global magma oceans on exoplanet demographics}
+\title{The Habitable Zone Inner Edge Discontinuity: Probing the Imprint of Runaway Greenhouse Climates on Exoplanet Demographics}
 
@@
 \subsubsection{Atmospheric spectral signatures}\label{sec:dis_atmospheres}
-%\todo[inline]{TIM: Ideas?\\
 %discuss potential atmospheric signatures of magma oceans.
@@
 2016)}
-\todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
-\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population).
+%\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population).
 These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Sect.
@@
 (i) Detecting the atmospheric windows of water vapor in the near- to mid-infrared.
-In a super-runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.][]{Boukrouche2021}.
+In a super-runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
 This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS, JWST~MIRI, or future missions such as \life\ ~\citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}.
@@
 \section{Conclusions}\label{sec:conclusions}
-A robust prediction of runaway greenhouse models is a significant inflation of rocky planet radii.
+Significant inflation of rocky planet radii is a robust prediction of runaway greenhouse models.
 Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius/density discontinuity resulting from this inflation in the exoplanet population.

### 2023-04-18 src/tex/ms.tex
streamline caveats discussion

@@
 Runaway greenhouse climates are not the only physical mechanism that may cause a change in transit radius for a subset of planets.
-Some of the most prominent alternative scenarios involve atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{2012ApJ...753...66I,Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
+Alternatives include atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{2012ApJ...753...66I,Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
 Both processes are being discussed as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
@@
 
-Another false positive contribution may come from potential unknown variations in the occurrence rate gradients in radius-instellation space.
-They can impact the statistical inference of the transitions, especially if these variations are similar to the expected pattern.
-Although an abrupt pattern at the expected runaway greenhouse transition seems unlikely, examples of steep occurrence rate density changes exist.
+Other false positive contributions may stem from potential unknown occurrence rate gradients in radius-instellation space, especially if these variations are similar to the expected habitable zone inner edge discontinuity.
+Although an abrupt pattern at the expected location of the transition seems unlikely, examples of steep occurrence rate density changes exist.
 An example is the ``Neptune desert'', a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
@@
 
-The models in this study neglect a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
-Future self-consistent modeling of interior-atmosphere interactions may fill these gaps~\citep{2022arXiv220310023L,Wordsworth2022}.
+Future self-consistent modeling of interior-atmosphere interactions may include constraints on additional radius increases due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
 
-\subsubsection{Atmosphere model}
-%Mention we only consider instellation-induced rgh.\\
-%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets %within the runaway greenhouse regime~\citep{2022arXiv220310023L}.\\
-%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
-%- assumption of a single albedo. ``if there is a range of albedos, will that wash away the signal?''
-%}
 % >>>>>>>>>>
@@
 % >>>>>>>>>>
-As illustrated in Figure~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the super- or sub-runaway category.
-For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature.
-Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the sub-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
-We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.
-Developing more sophisticated models of the coupled interaction between atmosphere and interior for an arbitrary planetary composition remains a long-term goal of the community~\citep{2021SSRv..217...22G,2022arXiv220310023L,Wordsworth2022}.
 
@@
 \subsubsection{``Sharpness'' of the habitable zone inner edge discontinuity}
-The real counterpart to the idealized step-function in our hypothesis tests could be blurred by several factors.
-Planets that lack an atmosphere or sufficient volatiles cannot bear a steam atmosphere.
-Those with significant Hydrogen/Helium envelopes can disguise as inflated rocky planets and might not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was recently suggested for water-dominated ``sub-Neptunes''~\citep{Pierrehumbert2022}.
+%Several factors may blur the idealized step-function in our hypothesis tests:
+The habitable zone inner edge discontinuity may be affected by several processes that are challenging to quantify:
+Planets that lack an atmosphere or sufficient volatiles cannot bear steam atmospheres, and those that do eventually move to the sub-runaway greenhouse category due to desiccation~\citep[][]{Watson1981} or evolution of their host star~\citep{Luger2015}.
+Hydrogen/Helium-dominated ``sub-Neptunes'' may disguise as inflated rocky planets and not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was suggested for water-dominated ``sub-Neptunes''~\citep{Pierrehumbert2022}.
 A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, reducing their transit radius.
-A time dependence is introduced through stellar luminosity evolution and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.
-
 Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Turbet2021,Pierrehumbert2022}.
@@
 
-All of the above mechanisms may affect the habitable zone inner edge discontinuity, and their contributions are challenging to quantify.
-However, while they may offset the signal's amplitude, they preserve its general shape.
-The ``dilution factor'' we introduced embodies our ignorance of the magnitude of this offset.
+While these factors may offset the signal's amplitude, they preserve its general shape.
+The ``dilution factor'' $f_\mathrm{rgh}$ in our model embodies our ignorance of the magnitude of this offset.
+In a real survey, additional contextual information about planets in the sample may be available.
 

### 2023-04-18 src/tex/ms.tex
add missing references

@@
 A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
-Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos ~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
+Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
 All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.
@@
 The shape function $g(R,P)$ describing the occurrence distribution is the same as in \citet{Bergsten2022}.
-The functional form and stellar-mass dependent parameters (including a new fit to the sample of \kepler\ planets with hosts below $0.56\,M_{\odot}$) are included in \bioverse\ as the prominent planet-generation model; the original model based on SAG13 occurrence rates is preserved as a legacy feature.
+The functional form and stellar-mass dependent parameters (including a new fit to the sample of \kepler\ planets with hosts below $0.56\,M_{\odot}$) are included in \bioverse\ as the prominent planet-generation model; the original model based on the ExoPAG Study Analysis Group~13's occurrence rates~\citep{Kopparapu2018} is preserved as a legacy feature.
 
@@
 
-The net absorbed stellar fluxes of planets are sensitive on their albedos, which, however, are generally poorly constrained CITE.
+The net absorbed stellar fluxes of planets are sensitive on their albedos, which are generally poorly constrained for planets outside the Solar System~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019b}\todo[inline]{better/more references available?}
 Here, we assume global redistribution of incoming flux and a fixed Bond albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
@@
 While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
+
 To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
@@
 In a super-runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
-This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS, JWST~MIRI, or future missions such as \life\ ~\citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}.
-CITE METIS, MIRI!
+This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as \life\ ~\citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}.
 Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime~\citep{2020SSRv..216...98G}.
@@
 (iv) Disequilibrium chemistry in tidally-locked super-runaway planets.
-It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers~\citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface~\citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
+It has been suggested that the atmospheric depth and the presence or absence of oceans on ``sub-Neptunes'' could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers~\citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface~\citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
 Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma~\citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
@@
 Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light~\citep{Wordsworth2022}.
-Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios (Konrad+, Alei+). CITE!
+Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios.\todo[inline]{ (Konrad+, Alei+). CITE}
 Finally, since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
@@
 Planets that lack an atmosphere or sufficient volatiles cannot bear steam atmospheres, and those that do eventually move to the sub-runaway greenhouse category due to desiccation~\citep[][]{Watson1981} or evolution of their host star~\citep{Luger2015}.
-Hydrogen/Helium-dominated ``sub-Neptunes'' may disguise as inflated rocky planets and not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was suggested for water-dominated ``sub-Neptunes''~\citep{Pierrehumbert2022}.
+Hydrogen/Helium-dominated planets may disguise as inflated rocky planets and not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
 A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, reducing their transit radius.
@@
 %\begin{acknowledgments}
-%The authors thank Gabriele Cugno, Yann Fabel, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
+%The authors thank Gabriele Cugno, Yann Fabel, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
 %We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.

### 2023-04-19 src/tex/ms.tex
streamline methodology of synthetic planet sample

@@
 Next, we assigned to the stellar sample planetary systems with frequencies, orbital parameters, and bulk properties derived from the \kepler\ mission.
-We adopted the planetary occurrence rate model from~\citet{Bergsten2022}, which defines the occurrence of small planets in radius and orbital period.
-Following \citet{Youdin2011a}, their inferred occurrence rate density can be expressed in the form:
+We adopted the model from~\citet{Bergsten2022}, which defines the occurrence rate of small planets in radius and orbital period.
+Following \citet{Youdin2011a}, their inferred occurrence rate density can be expressed in the form
 \begin{equation}
-    \frac{\mathrm{d}^2n}{\mathrm{d}R \, \partial P} = F_0 C_n g(R, P),
+    \frac{\mathrm{d}^2n}{\mathrm{d}R \, \partial P} = F_0 C_n g(R, P, M_\star),
 \end{equation}
-where $F_0$ represents the average number of planets per star and $C_n$ is a normalization constant.
-The shape function $g(R,P)$ describing the occurrence distribution is the same as in \citet{Bergsten2022}.
-The functional form and stellar-mass dependent parameters (including a new fit to the sample of \kepler\ planets with hosts below $0.56\,M_{\odot}$) are included in \bioverse\ as the prominent planet-generation model; the original model based on the ExoPAG Study Analysis Group~13's occurrence rates~\citep{Kopparapu2018} is preserved as a legacy feature.
-
-With this updated distribution function, planets are generated as in the original \bioverse\ paper, but with an additional stellar mass dependence (for a star of a given mass, the relevant model parameters from the bin containing that mass are used).
-In summary, stars are assigned a number of planets in accordance with $F_0$, with planet parameters randomly drawn in proportion to the occurrence distribution function evaluated over a grid of radii and orbital periods.
+where $F_0$ represents the average number of planets per star, $C_n$ is a normalization constant, and the shape function $g(R, P, M_\star)$ describes the distribution of planets in radius, orbital period, and stellar host mass.
+\bioverse\ generates planets based on the above occurrence rate density and assigns them to the previously generated stars.
 

### 2023-04-19 src/tex/ms.tex
revise introduction
implementing comments

@@
 Due to the inherent detection biases of the transit technique, the currently best studied category of rocky exoplanets is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
-These bodies experience climatic conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which sensitively affected the distribution of atmospheric volatiles between planetary core, mantle, and atmosphere.
-Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds~\citep{2022arXiv220310023L,Krijt2022}.
+These bodies experience conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which sensitively affected the distribution of volatiles between planetary core, mantle, and atmosphere.
+Studies of the geophysical state of hot exoplanets can thus inform our understanding of the early evolutionary stages of Earth and other habitable worlds~\citep{2022arXiv220310023L,Krijt2022}.
 
 An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
-While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,2020JGRE..12506276W,Turbet2021}.
-At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,2022arXiv220310023L}.
+While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties~\citep{Smrekar2018}, they evolved into planets with very different surface conditions~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,2020JGRE..12506276W,Turbet2021}.
+Both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,2022arXiv220310023L}.
 Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets~\citep{2020ChEG...80l5594F} because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
-Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
+Due to these similar formation sequences, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Way2020}.
 
-However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet~\citep{Hamano2013}, which is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
+Yet, \citet{Hamano2013} suggested that the dry conditions on Venus may have been directly inherited from the early magma ocean stage if a strongly infrared-absorbing, condensable species such as water is present on the planet, which is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
 Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
 This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
-%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,Hamano2015,2022arXiv220310023L}.
 Venus' atmospheric composition is consistent with the planet having gone through this phase~\citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.
@@
 The runaway greenhouse transition is a robust prediction from climate models~\citep{Forget2014,Boukrouche2021,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
-In particular, planets in a super-runaway climate state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii.
-However, dissolution in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,2022PSJ.....3..127S}.
+In particular, planets in a super-runaway climate state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii by an amount that is sensitive on the water content.
+However, dissolution of water~\citep[e.g.,][]{Elkins-Tanton2008,Hier-Majumder2017,Salvador2023} in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,2022PSJ.....3..127S}.
 Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their planetary mantles and establish an observational connection between exoplanetary interior and atmospheres~\citep{2022arXiv220310023L,Wordsworth2022}.
-
-%An Earth-like planet experiences a \SI{\sim 500}{\kilo\meter} radius increase upon crossing the threshold instellation, and its transit depth increases by \SI{10}{\ppm} to \SI{1000}{\ppm} depending on the spectral type of the host star~\citep{Turbet2019}.
 
@@
 
-Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
+Observational tests being considered include searches for direct evidence of liquid water conveyed by ocean glint~\citep{Williams2008,Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
 A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
@@
 It consists of a sample generator that populates stars from the \gaia\ catalog~\citep{Gaia_mission,Gaia_DR3}  with planetary systems based on state-of-the-art occurrence rates, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
-Trying different instrumentation and survey designs, we utilize \bioverse\ to recover runaway greenhouse-induced effects based on model predictions~\citep{Turbet2020,Dorn2021} that we inject into a baseline planet population.
+Trying different instrumentation and survey designs, we use \bioverse\ to recover runaway greenhouse-induced effects based on model predictions~\citep{Turbet2020,Dorn2021} that we inject into a baseline planet population.
 
@@
 We organize the paper as follows:
-Section~\ref{sec:met_baseline} introduces our baseline model produce synthetic star and planet samples.
-In Section~\ref{sec:met_rghmodel}, we describe the model component that produces runaway greenhouse-induced radius changes.
+Section~\ref{sec:met_baseline} introduces the baseline model we use to produce synthetic star and planet samples.
+In Section~\ref{sec:met_rghmodel}, we describe the model component that produces runaway greenhouse-induced transit radius changes.
 Section~\ref{sec:met_surveys-hypotests} explains our survey simulations and hypothesis tests.

### 2023-04-19 src/tex/ms.tex
address minor comments

@@
 Planetary systems are hosted by stars of a wide range of ages, and stellar luminosities evolve with time.
-Since the emergence of a runaway greenhouse phase on a planet is highly dependent on the level of radiation it receives, and thus on the luminosity of the host star, we assigned age-dependent luminosities to our synthetic stars.
+Since the presence of a runaway greenhouse state is highly dependent on the amount of radiation received by the planet, and thus on the luminosity of the host star, we assigned age-dependent luminosities to our synthetic stars.
 
@@
 Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light~\citep{Wordsworth2022}.
-Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios.\todo[inline]{ (Konrad+, Alei+). CITE}
+Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios.\todo[inline]{TIM: what papers by Konrad+, Alei+ did you want to cite here?}
 Finally, since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.

### 2023-04-19 src/tex/ms.tex
revise abstract

@@
 \begin{abstract}
-Key atmospheric volatiles are distributed between terrestrial planetary interiors and atmospheres during magma ocean epochs in the course of planetary accretion.
-Long-term magma ocean regimes on rocky exoplanets inside the orbital runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may thus offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
-Thermal stratification of super-runaway planets is expected to inflate the transit radii relative to sub-runaway planets, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius/density space.
+Long-term magma ocean phases on rocky exoplanets orbiting within the runaway greenhouse regime -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
+Thermal stratification of super-runaway planets is expected to significantly inflate their transit radii, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius/density space.
 Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages contextual information from the overall planet population.
-We find that the demographic imprint of super-runaway climates should be detectable with high-precision transit photometry for planet populations $\gtrsim$ 100 planets if at least 10\% of these planets undergo a super-runaway climate.
-We argue that in the near future, ESA's \plato\ mission will be the most promising survey to detect and constrain habitable zone inner edge discontinuity.
-We identify the key mission design drivers that enable a statistically sound detection, and discuss survey strategies that optimize the statistical power of combined datasets.
-Observational detection of the runaway climate transition will provide crucial insights into the distribution of atmospheric volatiles among low-mass exoplanets, and inform us about which stars harbor the nearest potentially habitable worlds.
+
+We find that the demographic imprint from super-runaway climates is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor super-runaway climates.
+Our survey simulations suggest that in the near future, ESA's \plato\ mission will be the most promising survey to probe the habitable zone inner edge discontinuity.
+We identify the key mission design drivers that enable a statistically sound detection and discuss survey strategies that optimize the statistical power of combined datasets.
+Observational constraints on the runaway climate transition will provide crucial insights into the distribution of atmospheric volatiles among low-mass exoplanets and inform us about which stars harbor the nearest potentially habitable worlds.
 \end{abstract}

### 2023-04-20 src/tex/ms.tex
small fixes

@@
 \begin{abstract}
-Long-term magma ocean phases on rocky exoplanets orbiting within the runaway greenhouse regime -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
+Long-term magma ocean phases on rocky exoplanets orbiting inside of the runaway greenhouse transition -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
 Thermal stratification of super-runaway planets is expected to significantly inflate their transit radii, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius/density space.
@@
 
-We find that the demographic imprint from super-runaway climates is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor super-runaway climates.
+We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor super-runaway climates.
 Our survey simulations suggest that in the near future, ESA's \plato\ mission will be the most promising survey to probe the habitable zone inner edge discontinuity.
-We identify the key mission design drivers that enable a statistically sound detection and discuss survey strategies that optimize the statistical power of combined datasets.
-Observational constraints on the runaway climate transition will provide crucial insights into the distribution of atmospheric volatiles among low-mass exoplanets and inform us about which stars harbor the nearest potentially habitable worlds.
+We determine survey strategies that optimize the statistical power of the obtained data, and we identify as key mission design drivers: 1.
+A follow-up campaign of planetary mass measurements and 2.
+The fraction of low-mass stars in the target sample.
+Observational constraints on the runaway climate transition will provide crucial insights into the distribution of atmospheric volatiles among rocky exoplanets and inform us about which stars harbor the nearest potentially habitable worlds.
 \end{abstract}
@@
 Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic samples of stars that host planets according to the observed exoplanet demographics.
-We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Figure~\ref{fig:flowchart}).\todo[inline]{make sure rgh hypothesis is mentioned before, or rephrase}
+We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of a habitable zone inner edge discontinuity (see diagram in Figure~\ref{fig:flowchart}).
 \begin{figure*}
@@
 \end{figure*}
-In this section, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, as well as the orbital parameters of the planets.
+In this section, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, and the orbital parameters of the planets.
 An overview of our key assumptions and model parameters can be found in Table~\ref{tab:params_table}.
@@
 With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state~\citep{2021JGRE..12606711L,Boukrouche2021}.
-We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean~\citep{Dorn2021,2023Icar..39015265S}.
+We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean~\citep{Dorn2021,Salvador2023}.
 
-The net absorbed stellar fluxes of planets are sensitive on their albedos, which are generally poorly constrained for planets outside the Solar System~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019b}\todo[inline]{better/more references available?}
+The net absorbed stellar fluxes of planets are sensitive on their albedos, which are generally poorly constrained for planets outside the Solar System~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019b}.
+\todo[inline]{better/more references available?}
 Here, we assume global redistribution of incoming flux and a fixed Bond albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
@@
 For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
-This scenario does not take into account the expected additional radius increase due to the density deficit of molten interiors~\citep{Bower2019}.
 

### 2023-04-20 src/tex/ms.tex
add missing references

@@
 An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
-While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties~\citep{Smrekar2018}, they evolved into planets with very different surface conditions~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,2020JGRE..12506276W,Turbet2021}.
+While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties~\citep{Smrekar2018}, they evolved into planets with very different surface conditions~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,Way2020,Turbet2021}.
 Both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,2022arXiv220310023L}.
@@
 In particular, planets in a super-runaway climate state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii by an amount that is sensitive on the water content.
-However, dissolution of water~\citep[e.g.,][]{Elkins-Tanton2008,Hier-Majumder2017,Salvador2023} in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,2022PSJ.....3..127S}.
+However, dissolution of water~\citep[e.g.,][]{Elkins-Tanton2008,Hier-Majumder2017,Salvador2023} in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,Schlichting2022}.
 Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their planetary mantles and establish an observational connection between exoplanetary interior and atmospheres~\citep{2022arXiv220310023L,Wordsworth2022}.
@@
 Hardegree-Ullman et al. (submitted) used the Gaia Catalogue of Nearby Stars~\citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
-The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2021} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
+The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2022} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
 From this information, colors and absolute magnitudes were computed, and initial color-magnitude cuts were made to remove non-main sequence stars.
@@
 The net absorbed stellar fluxes of planets are sensitive on their albedos, which are generally poorly constrained for planets outside the Solar System~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019b}.
-\todo[inline]{better/more references available?}
 Here, we assume global redistribution of incoming flux and a fixed Bond albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
@@
 In a super-runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
-This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as \life\ ~\citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}.
+This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as \life\ ~\citep{2019A&A...621A.125B,Quanz2022,2022A&A...664A..22D}.
 Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime~\citep{2020SSRv..216...98G}.
@@
 Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed.
-This has been suggested to be done via (v) eclipse photometry~\citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidified magma ocean~\citep{2020ApJ...898..160E,Fortin2022}.
+This has been suggested to be done via (v) eclipse photometry~\citep{Mansfield2019,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidified magma ocean~\citep{2020ApJ...898..160E,Fortin2022}.
 
@@
 Similarly, the Transiting Exoplanet Survey Satellite~\citep[\tess,][]{Ricker2014a} planet sample lacks temperate, small planets around bright host stars~\citep{Ment2023}, as was expected from planet yield calculations~\citep{Barclay2018}.
-As of 2023 March 22, the NASA Exoplanet Archive\footnote{\url{https://exoplanetarchive.ipac.caltech.edu}} lists 40 \tess\ candidate or confirmed planets smaller than \SI{4}{\rEarth} with lower estimated instellation than Earth's.
+As of March 22, 2023, the NASA Exoplanet Archive\footnote{\url{https://exoplanetarchive.ipac.caltech.edu}} lists 40 \tess\ candidate or confirmed planets smaller than \SI{4}{\rEarth} with lower estimated instellation than Earth's.
 
@@
 Other missions have been proposed that focus on characterizing exoplanet habitability, most notably the \hwo\ concept, which will build on the two precursor direct imaging concepts LUVOIR~\citep{LUVOIR2019} and HabEx~\citep{Gaudi2020c}.
-Similar science objectives are pursued by the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2021}, a mission concept utilizing a space-based mid-infrared nulling interferometer.
+Similar science objectives are pursued by the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2022}, a mission concept utilizing a space-based mid-infrared nulling interferometer.
 Direct imaging surveys do not directly measure planetary radii and are primarily useful for providing context through atmospheric measurements of individual planets.
@@
 Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light~\citep{Wordsworth2022}.
-Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios.\todo[inline]{TIM: what papers by Konrad+, Alei+ did you want to cite here?}
+Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios~\citep{Alei2022,Konrad2022}.
 Finally, since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
@@
 
-Precise ground-based radial velocity measurements will be needed to provide these data, and a number of instruments are already successfully employed in characterizing terrestrial-sized exoplanets~\citep[e.g.,][]{Queloz2001a,Pepe2010,Johnson2010b,Ribas2023}.
+Precise ground-based radial velocity measurements will be needed to provide these data, and a number of instruments are already successfully employed in characterizing terrestrial-sized exoplanets~\citep[e.g.,][]{Queloz2001a,Pepe2010,Johnson2010b,Ribas2023} and confronting these results with planet formation theory~\citep[e.g.,][]{Miguel2020a,Burn2021,Zawadzki2021a,Schlecker2022}.
 A new generation of instruments on extremely large telescopes such as \gclef\ on the \gmt ~\citep{Szentgyorgyi2016}, \andes\ on the \elt ~\citep{Marcantonio2022}, or \modhis\ on the \tmt ~\citep{Mawet2019} will open up the discovery space even further.
@@
 However, while interior composition may introduce an offset to the radius habitable zone inner edge discontinuity, we do not expect a change of its structure.
-Since the magnitude of the radius inflation effect is expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure~\citep{Zeng2016,2021JGRE..12606724B},
-\todo[inline]{Tim: add Noack+ paper}
-we consider our mass-radius relationship a conservative case.
+Since the magnitude of the radius inflation effect is expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure~\citep{Zeng2016,Noack2020,2021JGRE..12606724B}, we consider our mass-radius relationship a conservative case.
 We performed a sanity check to assess the impact of varying our baseline model (see Appendix~\ref{app:MR_relation}) and found general agreement between different interior compositions.

### 2023-04-21 src/tex/ms.tex
draft author contributions

@@
 
-\begin{large}\textit{Author contributions:}\end{large}
-%M.S., D.A., and T.L. conceived the project.
-%M.S. wrote the manuscript and carried out the hypothesis tests and statistical analyses.
-%...
-%K.H.-U. provided the star generator based on the Gaia catalog.
-%G.B. implemented the planet generator in the \bioverse\ framework.
-%All authors provided comments and suggestions on the manuscript.
+\section*{Author contributions}
+M.S., D.A., and T.L.\ conceived the project, planned its implementation, and interpreted the results.
+D.A.\ leads the ``Alien Earths'' program through which this project is funded and helped to guide the strategy of the project.
+T.L.\ provided expertise on runaway greenhouse climates.
+M.S.\ carried out the hypothesis tests and statistical analyses.
+M.S.\ wrote the manuscript; T.L., G.B., K.H.-U., and A.S.\ provided text contributions.
+G.B.\ implemented the planet generator in the \bioverse\ framework.
+All authors provided comments and suggestions on the manuscript.
 

### 2023-04-21 src/tex/ms.tex
add acknowledgements

@@
 
-%\begin{acknowledgments}
-%The authors thank Gabriele Cugno, Yann Fabel, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
-%We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
-%This work has made use of data from the European Space Agency (ESA) mission {\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia} Data Processing and Analysis Consortium (DPAC, \url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}).
+\begin{acknowledgments}
+The authors thank Gabriele Cugno, Yann Fabel, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
+We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
+This material is based upon work supported by the National Aeronautics and Space Administration under Agreement No. 80NSSC21K0593 for the program ``Alien Earths''.
+The results reported herein benefitted from collaborations and/or information exchange within NASA’s Nexus for Exoplanet System Science (NExSS) research coordination network sponsored by NASA’s Science Mission Directorate.
+This work has made use of data from the European Space Agency (ESA) mission {\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia} Data Processing and Analysis Consortium (DPAC, \url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}).
 Funding for the DPAC has been provided by national institutions, in particular the institutions participating in the {\it Gaia} Multilateral Agreement.
-%\end{acknowledgments}
+\end{acknowledgments}
 

### 2023-04-21 src/tex/ms.tex
edit title, abstract, and introduction

@@
 % Title
-\title{The Habitable Zone Inner Edge Discontinuity: Probing the Imprint of Runaway Greenhouse Climates on Exoplanet Demographics}
+\title{Bioverse: The Habitable Zone Inner Edge Discontinuity as an Imprint of Runaway Greenhouse Climates on Exoplanet Demographics}
 
@@
 \begin{abstract}
-Long-term magma ocean phases on rocky exoplanets orbiting inside of the runaway greenhouse transition -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
-Thermal stratification of super-runaway planets is expected to significantly inflate their transit radii, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius/density space.
-Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages contextual information from the overall planet population.
+Long-term magma ocean phases on rocky exoplanets orbiting inside of the runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
+Thermal stratification of super-runaway planets is expected to significantly inflate their radii, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius/density space.
+Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages the combination of contextual information from the overall planet population with a survey simulator and a hypothesis testing framework.
 
@@
 Our survey simulations suggest that in the near future, ESA's \plato\ mission will be the most promising survey to probe the habitable zone inner edge discontinuity.
-We determine survey strategies that optimize the statistical power of the obtained data, and we identify as key mission design drivers: 1.
+We determine survey strategies that maximize the statistical power of the obtained data, and we identify as key mission design drivers: 1.
 A follow-up campaign of planetary mass measurements and 2.
 The fraction of low-mass stars in the target sample.
-Observational constraints on the runaway climate transition will provide crucial insights into the distribution of atmospheric volatiles among rocky exoplanets and inform us about which stars harbor the nearest potentially habitable worlds.
+Observational constraints on the runaway climate transition will provide crucial insights into the distribution of atmospheric volatiles among rocky exoplanets, which may help to identify the nearest potentially habitable worlds.
 \end{abstract}
@@
 %\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021}.}
-%\todo[inline]{``Points for the intro:
-%\bioverse\\
-%Transit survey to measure mass/radius of planets\\
-%Does the HZ have an inner edge?\\
-%Run-away GH\\
-%Magma oceans\\
-%Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman.
-Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, \life\.\\
-%~\\
-%In contrast, PLATO and \nautilus\ are \\
-%Possibility of follow-up to measureme masses
-%''}
 
-%\begin{itemize}
-%    \item hot/warm planets are best studied
-%    \item runaway greenhouse climate (Ingersoll1969,Kasting1988,Nakajima1992). Transition is sharp in instellation.
-%    \item HZ concept.''(Kasting et al.
-1993; Kopparapu et al.
-2013, 2014)'' Relevance in the search for habitable planets and life.
-(Science gap:) Does it exist?
-Can it be tested?
-=> statistically detect inner edge
-%    \item Bioverse
-%    \item Transit survey => PLATO
-%    \item Using PLATO as a baseline survey, we do A, B, C.
-%\end{itemize}
-
-Recent years have brought tremendous advances in our understanding of terrestrial-sized planets, mainly due to space-based transit surveys such as \kepler\ \citep{Borucki2010} and \tess\ \citep{Ricker2014a} that pushed detection sensitivities into the terrestrial regime.
-Due to the inherent detection biases of the transit technique, the currently best studied category of rocky exoplanets is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
-These bodies experience conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which sensitively affected the distribution of volatiles between planetary core, mantle, and atmosphere.
+Despite recent advancements in observational techniques, our understanding of terrestrial-sized planets remains woefully limited, with fundamental aspects of their nature, composition, and potential habitability still largely unknown.
+Due to the inherent detection biases of current exoplanet detection techniques, the best-studied category of rocky exoplanets at present is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
+These bodies experience conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which likely profoundly affected the distribution of volatiles between planetary core, mantle, and atmosphere.
 Studies of the geophysical state of hot exoplanets can thus inform our understanding of the early evolutionary stages of Earth and other habitable worlds~\citep{2022arXiv220310023L,Krijt2022}.
 
-An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
+An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep[e.g.,][]{2019JGRE..124.2015K,2021JGRE..12606643K}.
 While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties~\citep{Smrekar2018}, they evolved into planets with very different surface conditions~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,Way2020,Turbet2021}.
@@
 
-Yet, \citet{Hamano2013} suggested that the dry conditions on Venus may have been directly inherited from the early magma ocean stage if a strongly infrared-absorbing, condensable species such as water is present on the planet, which is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
+Yet, \citet{Hamano2013} suggested that the dry conditions on Venus may have been directly inherited from the early magma ocean stage if a strongly infrared-absorbing, condensable species such as water is present on the planet.
+For Venus, this is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
 Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
@@
 However, dissolution of water~\citep[e.g.,][]{Elkins-Tanton2008,Hier-Majumder2017,Salvador2023} in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,Schlichting2022}.
-Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their planetary mantles and establish an observational connection between exoplanetary interior and atmospheres~\citep{2022arXiv220310023L,Wordsworth2022}.
+Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their mantles and establish an observational connection between exoplanetary interior and atmospheres~\citep{2022arXiv220310023L,Wordsworth2022}.
 
@@
 Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone.
-Its prevalent definition refers to the possibility of liquid water on an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmospheres~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but it can be influenced by different internal redox states and volatile compositions~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
+Its prevalent definition refers to the possibility of liquid water on the surface of an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmosphere~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but it can be influenced by different internal redox states and volatile compositions~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
 The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in the National Aeronautics and Space Administration (NASA) Astrophysics Data System returned $\sim\num{1700}$ results.}.
@@
 Observational tests being considered include searches for direct evidence of liquid water conveyed by ocean glint~\citep{Williams2008,Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
-A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
-Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
+A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019,Apai2019a}.
+Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011,Bixel2020}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
 All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.
@@
 Here, we explore the feasibility of a statistical test of the habitable zone hypothesis by surveying planetary bulk properties close to its inner edge, the runaway greenhouse transition.
-Our goal is to assess the ability of near-future transit surveys to test the hypothesis that the runaway greenhouse effect causes an abrupt change of planetary radii and bulk densities when ordered by receiving instellation~\citep{Turbet2019}.
+Our goal is to assess the ability of near-future transit surveys to test the hypothesis that the runaway greenhouse effect causes a discontinuity of planetary radii and bulk densities when ordered by receiving instellation~\citep{Turbet2019}.
 Our main tool for this is \bioverse, a simulation framework for assessing the statistical power of exoplanet surveys~\citep{Bixel2021}.
-It consists of a sample generator that populates stars from the \gaia\ catalog~\citep{Gaia_mission,Gaia_DR3}  with planetary systems based on state-of-the-art occurrence rates, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
+It consists of a sample generator that populates stars from the \gaia\ catalog~\citep[][Hardegree-Ullman et al., submitted]{Gaia_mission,Gaia_DR3} with planetary systems based on state-of-the-art occurrence rates~\citep{Bergsten2022}, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
 Trying different instrumentation and survey designs, we use \bioverse\ to recover runaway greenhouse-induced effects based on model predictions~\citep{Turbet2020,Dorn2021} that we inject into a baseline planet population.
 
-In particular, we test the capability of the \plato\ \citep[PLAnetary Transits and Oscillation of stars, ][]{Rauer2016} mission, which will measure the radii of a large number of terrestrial-sized planets, to detect the runaway greenhouse transition and identify the astrophysical unknowns on which this depends.
-We then perform a parameter study to explore which trade-offs in the survey design of a \plato-like mission optimize the chance to detect and characterize the runaway greenhouse transition.
+In particular, we test the capability of the \plato\ \citep[PLAnetary Transits and Oscillation of stars, ][]{Rauer2016} mission, which will measure the radii of a large number of terrestrial-sized planets, to detect the radius/density discontinuity and determine its sensitivity to model assumptions and fundamental processes.
+We then perform a parameter study to explore which trades in the survey design of a \plato-like mission maximizes its diagnostic power to test runaway greenhouse climate models through the detection of the habitable zone inner edge discontinuity.
 
@@
 \subsection{Mission design trade studies}\label{sec:mission-design-trades}
-To explore the impact of mission trade-offs on the detectability of the habitable zone inner edge discontinuity, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
+To explore the impact of mission trades on the detectability of the habitable zone inner edge discontinuity, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
 We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.

### 2023-04-22 src/tex/ms.tex

@@
 In particular, we test the capability of the \plato\ \citep[PLAnetary Transits and Oscillation of stars, ][]{Rauer2016} mission, which will measure the radii of a large number of terrestrial-sized planets, to detect the radius/density discontinuity and determine its sensitivity to model assumptions and fundamental processes.
-We then perform a parameter study to explore which trades in the survey design of a \plato-like mission maximizes its diagnostic power to test runaway greenhouse climate models through the detection of the habitable zone inner edge discontinuity.
+We then perform a parameter study to explore which trades in the survey design of a \plato-like mission maximize its diagnostic power to test runaway greenhouse climate models through the detection of the habitable zone inner edge discontinuity.
 
@@
 
-\section*{Data Availability}
-All data sets and software required to reproduce our results and this article are available through GitHub and Zenodo.
-The code to reproduce the figures can be accessed via the icon links associated with the respective figure caption.
+%\section*{Data Availability}
+\section*{Reproducibility}
+%All data sets and software required to reproduce our results and this article are available through GitHub and Zenodo.
+This study uses the reproducibility framework ``showyourwork''~\citep{Luger2021}.
+All code required to reproduce our results, figures, and this article itself is available at \url{https://github.com/matiscke/magmaocean}.
+The code to reproduce a figure can be accessed via the icon link next to the respective figure caption.
 

### 2023-04-22 src/tex/ms.tex
Minor additions + reference clean-up

@@
 Long-term magma ocean phases on rocky exoplanets orbiting inside of the runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
-Thermal stratification of super-runaway planets is expected to significantly inflate their radii, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius/density space.
+Thermal stratification of super-runaway planets is expected to significantly inflate their radii, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius--density space.
 Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages the combination of contextual information from the overall planet population with a survey simulator and a hypothesis testing framework.
@@
 These bodies experience conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which likely profoundly affected the distribution of volatiles between planetary core, mantle, and atmosphere.
-Studies of the geophysical state of hot exoplanets can thus inform our understanding of the early evolutionary stages of Earth and other habitable worlds~\citep{2022arXiv220310023L,Krijt2022}.
+Studies of the geophysical state of hot exoplanets can thus inform our understanding of the early evolutionary stages of Earth and other habitable worlds~\citep{Lichtenberg2022,Krijt2022}.
 
@@
 While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties~\citep{Smrekar2018}, they evolved into planets with very different surface conditions~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,Way2020,Turbet2021}.
-Both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,2022arXiv220310023L}.
+Both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
 Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets~\citep{2020ChEG...80l5594F} because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
@@
 However, dissolution of water~\citep[e.g.,][]{Elkins-Tanton2008,Hier-Majumder2017,Salvador2023} in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,Schlichting2022}.
-Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their mantles and establish an observational connection between exoplanetary interior and atmospheres~\citep{2022arXiv220310023L,Wordsworth2022}.
+Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their mantles and establish an observational connection between exoplanetary interiors and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.
 
-Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{2010ppc..book.....P,2013ApJ...765..131K}.
+Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{2010ppc..book.....P,Kopparapu2013}.
 Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone.
@@
 The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in the National Aeronautics and Space Administration (NASA) Astrophysics Data System returned $\sim\num{1700}$ results.}.
-However, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
+However, it should be emphasized that the habitable zone hypothesis, as it stands today, is foremost a concept based on theoretical predictions and geochemical evidence from one planet -- Earth \citep{2020SciA....6.1420C}, and its general validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
 The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.
@@
 
-In particular, we test the capability of the \plato\ \citep[PLAnetary Transits and Oscillation of stars, ][]{Rauer2016} mission, which will measure the radii of a large number of terrestrial-sized planets, to detect the radius/density discontinuity and determine its sensitivity to model assumptions and fundamental processes.
+In particular, we test the capability of the \plato\ \citep[PLAnetary Transits and Oscillation of stars, ][]{Rauer2016} mission, which will measure the radii of a large number of terrestrial-sized planets, to detect the radius--density discontinuity and determine its sensitivity to model assumptions and fundamental processes.
 We then perform a parameter study to explore which trades in the survey design of a \plato-like mission maximize its diagnostic power to test runaway greenhouse climate models through the detection of the habitable zone inner edge discontinuity.
@@
 %- Limitation to small planets without significant atmospheric envelopes.\\
-%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{2022arXiv220310023L}.\\
+%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
 %- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
@@
 
-The net absorbed stellar fluxes of planets are sensitive on their albedos, which are generally poorly constrained for planets outside the Solar System~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019b}.
+The net absorbed stellar fluxes of planets are sensitive on their albedos, which are generally poorly constrained for planets outside the Solar System~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019}.
 Here, we assume global redistribution of incoming flux and a fixed Bond albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
@@
 Other missions have been proposed that focus on characterizing exoplanet habitability, most notably the \hwo\ concept, which will build on the two precursor direct imaging concepts LUVOIR~\citep{LUVOIR2019} and HabEx~\citep{Gaudi2020c}.
-Similar science objectives are pursued by the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2022}, a mission concept utilizing a space-based mid-infrared nulling interferometer.
+Similar science objectives are pursued by the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2022} initiative, a mission concept utilizing a space-based mid-infrared nulling interferometer.
 Direct imaging surveys do not directly measure planetary radii and are primarily useful for providing context through atmospheric measurements of individual planets.
-However, because mid-infrared retrievals do not suffer from the degeneracy between cloud albedo and changes in surface area, the planet radius can be retrieved in mid-infrared wavelengths~\citep{2018ExA....46..543D,2021ExA...tmp..118Q}.
+However, because mid-infrared retrievals feature reduced degeneracy between cloud albedo and changes in surface area, the planet radius can be constrained in mid-infrared wavelengths~\citep{2018ExA....46..543D,2021ExA...tmp..118Q}.
 Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light~\citep{Wordsworth2022}.
 Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios~\citep{Alei2022,Konrad2022}.
-Finally, since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
+Since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
+Finally, mid-infrared surveys such as LIFE show a preference for M star planets \citep{Quanz2022}, which is crucial for the predicted radius inflation disconuity, as discussed above.
 With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states~\citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.
@@
 To date, the majority of planets with radius measurements orbit FGK~dwarfs, and, based on the instellation they receive, most of them lie in the runaway greenhouse regime~\citep{Thompson2018}.
-Obviously, a radius/density discontinuity in the exoplanet demographics like the habitable zone inner edge discontinuity cannot be constrained well if only one side of the discontinuity is being sampled.
+Obviously, a radius--density discontinuity in the exoplanet demographics like the habitable zone inner edge discontinuity cannot be constrained well if only one side of the discontinuity is being sampled.
 This, however, is the situation for planetary systems around Sun-like stars -- their habitable zones are so distant that transiting planets within them are very rare due to pure geometrical reasons.
@@
 
-% >>>>>>>>>>
-% TL: I just wrote a review article about this exact problem (magma ocean atmosphere exchange) with Yamila Miguel, hopefully soon on arXiv.
-Honestly, I think we keep it short here, otherwise the paper will get way too long and diluted.
-I referenced the latest reviews on this now, and we can cite Lichtenberg & Miguel 23 as soon as it's on arXiv for a much much longer discussion.
-% >>>>>>>>>>
-
 \subsubsection{Water mass fraction}
@@
 Planets that lack an atmosphere or sufficient volatiles cannot bear steam atmospheres, and those that do eventually move to the sub-runaway greenhouse category due to desiccation~\citep[][]{Watson1981} or evolution of their host star~\citep{Luger2015}.
-Hydrogen/Helium-dominated planets may disguise as inflated rocky planets and not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
+Hydrogen/Helium-dominated planets may disguise as inflated rocky planets and not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022,2023arXiv230402698I}.
 A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, reducing their transit radius.
-Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Turbet2021,Pierrehumbert2022}.
+Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Turbet2021,2021JGRE..12606711L,Pierrehumbert2022,2023arXiv230402698I}.
 The choice of a statistical estimator for the hypothesis tests may further influence the recovered discontinuity; we compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.
@@
 Significant inflation of rocky planet radii is a robust prediction of runaway greenhouse models.
-Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius/density discontinuity resulting from this inflation in the exoplanet population.
+Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius--density discontinuity resulting from this inflation in the exoplanet population.
 Our key findings are as follows:
@@
     \item This habitable zone inner edge discontinuity should be detectable with high-precision transit measurements.
-          For a large sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.
-    \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to detect the predicted trend.
+          For a planet sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.
+    \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to confirm or reject the predicted trend.
           Assuming the projected photometric precision, \plato\ will be able to test the runaway greenhouse hypothesis for planet yields $\gtrsim 100$.
@@
 
-The habitable zone is widely employed in target prioritization for exoplanet missions, and it will provide context for interpreting potential signatures of life.
-As we have demonstrated, it appears realistic that an empirical confirmation of the habitable zone concept is imminent.
-This will be a key contribution to the search for extraterrestrial life in the Universe.
+The relation between stellar irradiation and climatic properties of rocky exoplanets within the contex of the habitably is widely employed in target prioritization for exoplanet missions.
+Determination of such key links between atmospheric volatiles and stellar parameters it will provide crucial environmental information for distinguishing potentially habitable exoplanets from others, and support interpretation of potential signatures of life.
+An empirical confirmation of the habitable zone concept within this decade with currently projected exoplanet surveys thus seems feasible. 
+The confirmation or rejection of the inner edge habitable zone discontinuity will be a key contribution to understand the diversity of exoplanet climate regimes, and to the search for extraterrestrial life in the universe.
 
 \begin{acknowledgments}
-The authors thank Gabriele Cugno, Yann Fabel, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
+The authors thank Gabriele Cugno, Yann Fabel, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, Terry-Ann Suer, and Brad Foley for insightful discussions.
 We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
@@
 Funding for the DPAC has been provided by national institutions, in particular the institutions participating in the {\it Gaia} Multilateral Agreement.
+T.L. was supported by a grant from the Branco Weiss Foundation.
 \end{acknowledgments}
@@
 D.A.\ leads the ``Alien Earths'' program through which this project is funded and helped to guide the strategy of the project.
-T.L.\ provided expertise on runaway greenhouse climates.
+T.L.\ provided expertise on runaway greenhouse climates and exoplanet interiors.
 M.S.\ carried out the hypothesis tests and statistical analyses.

### 2023-04-25 src/tex/ms.tex

@@
 The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in the National Aeronautics and Space Administration (NASA) Astrophysics Data System returned $\sim\num{1700}$ results.}.
-However, it should be emphasized that the habitable zone hypothesis, as it stands today, is foremost a concept based on theoretical predictions and geochemical evidence from one planet -- Earth \citep{2020SciA....6.1420C}, and its general validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
+However, it should be emphasized that the habitable zone, as it stands today, is merely a concept based on theoretical predictions and geochemical evidence from one planet -- Earth~\citep{2020SciA....6.1420C} -- and its general validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
 The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.
@@
 
-In particular, we test the capability of the \plato\ \citep[PLAnetary Transits and Oscillation of stars, ][]{Rauer2016} mission, which will measure the radii of a large number of terrestrial-sized planets, to detect the radius--density discontinuity and determine its sensitivity to model assumptions and fundamental processes.
+In particular, we test the capability of the \plato\ \citep[PLAnetary Transits and Oscillation of stars, ][]{Rauer2016} mission, which will measure the radii of a large number of terrestrial-sized planets, to detect the radius/density discontinuity and determine its sensitivity to model assumptions and fundamental processes.
 We then perform a parameter study to explore which trades in the survey design of a \plato-like mission maximize its diagnostic power to test runaway greenhouse climate models through the detection of the habitable zone inner edge discontinuity.
@@
 The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}.
-With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state~\citep{2021JGRE..12606711L,Boukrouche2021}.
+With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state~\citep{Lichtenberg2021c,Boukrouche2021}.
 We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean~\citep{Dorn2021,Salvador2023}.
@@
 \textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies on sufficient amounts of atmospheric water vapor that can act as a greenhouse gas.
-However, already about \SIrange{\sim 10}{20}~bar of water vapor -- corresponding to a minor fraction of one Earth ocean and thus the lower limit of water on Earth -- is enough to sustain sufficiently high surface temperatures to keep the planet in a magma ocean stage~\citep{Boukrouche2021,2021JGRE..12606711L}.
+However, already about \SIrange{\sim 10}{20}~bar of water vapor -- corresponding to a minor fraction of one Earth ocean and thus the lower limit of water on Earth -- is enough to sustain sufficiently high surface temperatures to keep the planet in a magma ocean stage~\citep{Boukrouche2021,Lichtenberg2021c}.
 The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
@@
 Since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
-Finally, mid-infrared surveys such as LIFE show a preference for M star planets \citep{Quanz2022}, which is crucial for the predicted radius inflation disconuity, as discussed above.
+Finally, mid-infrared surveys such as \life\ show a preference for M~star planets~\citep{Quanz2022}, which is beneficial for detecting the predicted habitable zone inner edge discontinuity (see Sect.~\ref{sec:results_FGK_M}).
 With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states~\citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.
@@
 To date, the majority of planets with radius measurements orbit FGK~dwarfs, and, based on the instellation they receive, most of them lie in the runaway greenhouse regime~\citep{Thompson2018}.
-Obviously, a radius--density discontinuity in the exoplanet demographics like the habitable zone inner edge discontinuity cannot be constrained well if only one side of the discontinuity is being sampled.
+Obviously, a radius/density discontinuity in the exoplanet demographics like the habitable zone inner edge discontinuity cannot be constrained well if only one side of the discontinuity is being sampled.
 This, however, is the situation for planetary systems around Sun-like stars -- their habitable zones are so distant that transiting planets within them are very rare due to pure geometrical reasons.
@@
 
-Future self-consistent modeling of interior-atmosphere interactions may include constraints on additional radius increases due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
+Future self-consistent modeling of interior-atmosphere interactions may include constraints on additional radius increases due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,Lichtenberg2021c,2022PSJ.....3...93B,2021SSRv..217...22G}.
 
@@
 Planets that lack an atmosphere or sufficient volatiles cannot bear steam atmospheres, and those that do eventually move to the sub-runaway greenhouse category due to desiccation~\citep[][]{Watson1981} or evolution of their host star~\citep{Luger2015}.
-Hydrogen/Helium-dominated planets may disguise as inflated rocky planets and not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022,2023arXiv230402698I}.
+Hydrogen/Helium-dominated planets may disguise as inflated rocky planets and not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022,Innes2023}.
 A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, reducing their transit radius.
-Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Turbet2021,2021JGRE..12606711L,Pierrehumbert2022,2023arXiv230402698I}.
+Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Turbet2021,Lichtenberg2021c,Pierrehumbert2022,Innes2023}.
 The choice of a statistical estimator for the hypothesis tests may further influence the recovered discontinuity; we compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.
@@
 Significant inflation of rocky planet radii is a robust prediction of runaway greenhouse models.
-Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius--density discontinuity resulting from this inflation in the exoplanet population.
+Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius/density discontinuity resulting from this inflation in the exoplanet population.
 Our key findings are as follows:
@@
 
-The relation between stellar irradiation and climatic properties of rocky exoplanets within the contex of the habitably is widely employed in target prioritization for exoplanet missions.
-Determination of such key links between atmospheric volatiles and stellar parameters it will provide crucial environmental information for distinguishing potentially habitable exoplanets from others, and support interpretation of potential signatures of life.
-An empirical confirmation of the habitable zone concept within this decade with currently projected exoplanet surveys thus seems feasible. 
-The confirmation or rejection of the inner edge habitable zone discontinuity will be a key contribution to understand the diversity of exoplanet climate regimes, and to the search for extraterrestrial life in the universe.
+The habitable zone concept is widely employed in target prioritization for exoplanet missions, and it will provide context for interpreting potential signatures of life.
+As we have demonstrated, it appears realistic that an empirical test of the habitable zone hypothesis is imminent.
+The confirmation or rejection of the habitable zone inner edge discontinuity will be a key contribution to understanding the diversity of exoplanet climates and the search for extraterrestrial life in the Universe.
 
 \begin{acknowledgments}
-The authors thank Gabriele Cugno, Yann Fabel, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, Terry-Ann Suer, and Brad Foley for insightful discussions.
+The authors thank Gabriele Cugno, Yann Fabel, Brad Foley, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
 We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
@@
 D.A.\ leads the ``Alien Earths'' program through which this project is funded and helped to guide the strategy of the project.
-T.L.\ provided expertise on runaway greenhouse climates and exoplanet interiors.
+T.L.\ and A.S.\ provided expertise on runaway greenhouse climates and exoplanet interiors.
 M.S.\ carried out the hypothesis tests and statistical analyses.

### 2023-05-01 src/scripts/hzied_pipeline.py
create sample generation pipeline

@@ -1,0 +1,115 @@
+"""
+This file contains the code to generate the datasets for the paper.
+It heavily relies on Bioverse and its auxiliary functions.
+"""
+
+import scipy
+import paths
+from utils import *
+import pickle
+import numpy as np
+import pandas as pd
+import seaborn as sns
+
+from scipy.interpolate import interp1d
+from scipy.signal import savgol_filter
+
+# Import the Generator class
+from bioverse.generator import Generator
+from bioverse.survey import TransitSurvey
+from bioverse.constants import CONST, DATA_DIR
+
+# Sample generation
+## Generate stars and planets, inject magma oceans
+
+# Parameters of the analysis
+# parameter_of_interest = 'rho'
+parameter_of_interest = 'R'
+
+def get_generator_args():
+    """ define generator parameters"""
+    # Parameters for star generation
+    stars_args = {
+        'd_max' : 90,             # max. dist to stars (pc) ~~(nominal=90, ca. 481 planets)~~
+        'M_st_max': 1.5,  # Maximum stellar mass to consider, in solar units.
+        'M_G_max': 16,  # Maximum gaia magnitude of stars
+        'seed': 42,  # seed for random number generators
+        'lum_evo': True  # luminosities from luminosity tracks (Baraffe et al. 1998), based on random ages? [SLOW?]
+    }
+
+    # Parameters for planet generation and magma ocean model
+    planets_args = {
+        'transit_mode': False,  # Simulate only transiting planets
+        'f_eta': 1.,  # Occurrence rate scaling factor
+        'R_min': 0.75,  # minimum radius for Bergsten et al. planet generator
+        'P_max': 500.,  # maximum period for Bergsten et al. planet generator
+        'mr_relation': 'Zeng2016',  # choice of mass-radius relationship ('Zeng2016'/'Wolfgang2016/'earthlike')
+        'gh_increase': True,  # wether or not to consider radius increase due to runaway greenhouse effect (Turbet+2020)
+        'water_incorp': True, # wether or not to consider water incorporation in the melt of global magma oceans (Dorn & Lichtenberg 2021)
+        'S_thresh': 280.,  # threshold instellation for runaway greenhouse phase
+        'wrr': 0.005, # water-to-rock ratio for Turbet+2020 model. Possible vals: 0.0001,0.001, 0.005, 0.01, 0.02, 0.03, 0.04, 0.05
[…]

### 2023-05-01 src/tex/ms.tex
plot radius evolution via ms workflow

@@
 \author[0000-0001-8355-2107]{Martin Schlecker}
-\affiliation{Department of Astronomy/Steward Observatory, The University of Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
-%\author[0000-0002-3286-7683]{Tim Lichtenberg}
-%\affiliation{Kapteyn Astronomical Institute, University of Groningen, PO Box 800, 9700 AV Groningen, The Netherlands}
-\author{al.}
+\affiliation{Steward Observatory, The University of Arizona, Tucson, AZ 85721, USA; \href{mailto:schlecker@arizona.edu}{schlecker@arizona.edu}}
+\author[0000-0003-3714-5855]{D\'{a}niel Apai}
+\affiliation{Steward Observatory, The University of Arizona, Tucson, AZ 85721, USA; \href{mailto:schlecker@arizona.edu}{schlecker@arizona.edu}}
+\affiliation{Lunar and Planetary Laboratory, The University of Arizona, Tucson, AZ 85721, USA}
+\author[0000-0002-3286-7683]{Tim Lichtenberg}
+\affiliation{Kapteyn Astronomical Institute, University of Groningen, PO Box 800, 9700 AV Groningen, The Netherlands}
+\author[0000-0003-4500-8850]{Galen Bergsten}
+\affiliation{Lunar and Planetary Laboratory, The University of Arizona, Tucson, AZ 85721, USA}
+\author[0000-0001-8106-6164]{Arnaud Salvador}
+\affiliation{Lunar and Planetary Laboratory, The University of Arizona, Tucson, AZ 85721, USA}
+\author[0000-0003-3702-0382]{Kevin K.\ Hardegree-Ullman}
+\affiliation{Steward Observatory, The University of Arizona, Tucson, AZ 85721, USA; \href{mailto:schlecker@arizona.edu}{schlecker@arizona.edu}}
 
@@
 \begin{abstract}
-Long-term magma ocean phases on rocky exoplanets orbiting inside of the runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
-Thermal stratification of super-runaway planets is expected to significantly inflate their radii, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius--density space.
-Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages the combination of contextual information from the overall planet population with a survey simulator and a hypothesis testing framework.
+Long-term magma ocean phases on rocky exoplanets orbiting closer to their star than the runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
+Thermal stratification of runaway planets is expected to significantly inflate their atmospheres, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius--density space.
+Here, we use \bioverse, a statistical framework combining contextual information from the overall planet population with a survey simulator, to assess the ability of ground- and space-based telescopes to test this hypothesis.
 
-We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor super-runaway climates.
+We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for sample sizes $\gtrsim$~100~planets if at least \SI{\sim 10}{\percent} of those orbiting closer than the habitable zone inner edge harbor runaway climates.
+%We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor runaway climates.
 Our survey simulations suggest that in the near future, ESA's \plato\ mission will be the most promising survey to probe the habitable zone inner edge discontinuity.
-We determine survey strategies that maximize the statistical power of the obtained data, and we identify as key mission design drivers: 1.
+We determine survey strategies that maximize the diagnostic power of the obtained data and identify as key mission design drivers: 1.
 A follow-up campaign of planetary mass measurements and 2.
 The fraction of low-mass stars in the target sample.
-Observational constraints on the runaway climate transition will provide crucial insights into the distribution of atmospheric volatiles among rocky exoplanets, which may help to identify the nearest potentially habitable worlds.
+Observational constraints on the runaway greenhouse transition will provide crucial insights into the distribution of atmospheric volatiles among rocky exoplanets, which may help to identify the nearest potentially habitable worlds.
 \end{abstract}
@@
 \section{Introduction}
-%\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021}.}
+Despite recent advancements in observational techniques, our understanding of terrestrial-sized planets remains woefully limited, with fundamental aspects of their nature, composition, and potential habitability still largely unknown.
+Due to the inherent biases of current exoplanet detection techniques, the best-studied category of rocky exoplanets at present is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
+These experience thermal states that are in some aspects comparable to the ones of the inner solar system bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which likely profoundly affected the distribution of volatiles between planetary core, mantle, and atmosphere.
+Studying the geophysical state of hot exoplanets can thus inform our understanding of the early evolutionary stages of Earth and other habitable worlds~\citep{Lichtenberg2022,Krijt2022}.
 
-Despite recent advancements in observational techniques, our understanding of terrestrial-sized planets remains woefully limited, with fundamental aspects of their nature, composition, and potential habitability still largely unknown.
-Due to the inherent detection biases of current exoplanet detection techniques, the best-studied category of rocky exoplanets at present is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
-These bodies experience conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which likely profoundly affected the distribution of volatiles between planetary core, mantle, and atmosphere.
-Studies of the geophysical state of hot exoplanets can thus inform our understanding of the early evolutionary stages of Earth and other habitable worlds~\citep{Lichtenberg2022,Krijt2022}.
+An example from the solar system for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep[e.g.,][]{2019JGRE..124.2015K,2021JGRE..12606643K,Salvador2023b}.
+While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties~\citep{Smrekar2018}, they evolved into planets with very different surface conditions~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,Way2020,Turbet2021}.
+Both planets likely underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
+Magma ocean states play a substantial role in establishing the long-term geophysical and climatic regimes of rocky planets~\citep{2020ChEG...80l5594F}, in particular owing to efficient heat and volatile transfers between interior and atmosphere in the absence of a stiff boundary separating them~\citep{2021ApJ...909L..22K,Dorn2021,Salvador2023b}.
+% because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
+Due to these similar formation sequences, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus' water loss -- occurred late in their evolution~\citep[e.g.,][]{Way2020}.
 
-An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep[e.g.,][]{2019JGRE..124.2015K,2021JGRE..12606643K}.
-While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties~\citep{Smrekar2018}, they evolved into planets with very different surface conditions~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,Way2020,Turbet2021}.
-Both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
-Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets~\citep{2020ChEG...80l5594F} because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
-Due to these similar formation sequences, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Way2020}.
+Yet, \citet{Hamano2013} suggested that the present-day dry conditions on Venus may have been directly inherited from the early magma ocean stage.% if a strongly infrared-absorbing, condensable species such as water is present on the planet.
+If a strongly infrared-absorbing, condensable species such as water was dominant in the atmosphere, the resulting strong thermal blanketing effect would prevent the planet to efficiently radiate to space and maintain the surface molten~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
+This runaway greenhouse state can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
+%This scenario requires early water delivery, which for Venus is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
+%Where the incident stellar flux exceeds a threshold value, a non-linear transition from a mixture of condensed and volatile states to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
+For Venus, isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B} and the planet's atmospheric composition suggest that it went through this phase~\citep{2020NatGe..13..265G}.
+Although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}, a transient habitable phase cannot be conclusively ruled out~\citep[e.g.,][]{Way2016,Salvador2017,Krissansen-Totton2021}.
 
-Yet, \citet{Hamano2013} suggested that the dry conditions on Venus may have been directly inherited from the early magma ocean stage if a strongly infrared-absorbing, condensable species such as water is present on the planet.
-For Venus, this is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
-Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
-This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
-Venus' atmospheric composition is consistent with the planet having gone through this phase~\citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.
+The runaway greenhouse transition is a robust prediction from climate models~\citep{Kasting1988,Nakajima1992,Goldblatt2012,Forget2014,Boukrouche2021,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
+In particular, planets in a runaway greenhouse state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii by an amount that is a function of the water content.
+However, dissolution of water~\citep[e.g.,][]{Elkins-Tanton2008,Hier-Majumder2017,Salvador2023} in the magma may decrease this effect~\citep{Dorn2021}, and chemical exchange between core and mantle material may influence the amount and speciation of outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,Schlichting2022}.
+Astronomical observations of planets that are currently in a runaway greenhouse state may thus constrain properties of their mantles and establish an observational connection between exoplanetary interiors and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.
 
-The runaway greenhouse transition is a robust prediction from climate models~\citep{Forget2014,Boukrouche2021,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
-In particular, planets in a super-runaway climate state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii by an amount that is sensitive on the water content.
-However, dissolution of water~\citep[e.g.,][]{Elkins-Tanton2008,Hier-Majumder2017,Salvador2023} in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,Schlichting2022}.
-Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their mantles and establish an observational connection between exoplanetary interiors and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.
-
-Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{2010ppc..book.....P,Kopparapu2013}.
-Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone.
-Its prevalent definition refers to the possibility of liquid water on the surface of an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmosphere~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but it can be influenced by different internal redox states and volatile compositions~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
+Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp instellation threshold~\citep{Goldblatt2013,Leconte2013,Kopparapu2013}.
+Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone~\citep[e.g.,][]{Ramirez2018,Salvador2023b}.
+Its prevalent definition refers to the possibility of a sustained liquid water body on the surface of an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmosphere~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014};
+its exact spatial location and extent may be strongly influenced by the interior and atmosphere oxidation state and resulting atmosphere composition~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
 The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in the National Aeronautics and Space Administration (NASA) Astrophysics Data System returned $\sim\num{1700}$ results.}.
-However, it should be emphasized that the habitable zone, as it stands today, is merely a concept based on theoretical predictions and geochemical evidence from one planet -- Earth~\citep{2020SciA....6.1420C} -- and its general validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
-The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.
+However, it should be emphasized that the habitable zone, as it stands today, is merely a concept based on theoretical predictions and geochemical evidence from one planet -- Earth~\citep{2020SciA....6.1420C} -- and its general validity remains controversial~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
+The question naturally emerges if a planetary habitable zone -- in its common form with boundaries defined by stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.
 
 Observational tests being considered include searches for direct evidence of liquid water conveyed by ocean glint~\citep{Williams2008,Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
-A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019,Apai2019a}.
+A different approach relies on comparative planetology:
+aiming for a statistical detection in a planet population provides robustness against ambiguity that could otherwise arise from individual variations in a planet's composition or geophysical history~\citep{Checlair2019,Apai2019a}.
 Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011,Bixel2020}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
@@
 Our main tool for this is \bioverse, a simulation framework for assessing the statistical power of exoplanet surveys~\citep{Bixel2021}.
-It consists of a sample generator that populates stars from the \gaia\ catalog~\citep[][Hardegree-Ullman et al., submitted]{Gaia_mission,Gaia_DR3} with planetary systems based on state-of-the-art occurrence rates~\citep{Bergsten2022}, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
+It consists of a sample generator that populates stars from the \gaia\ catalog~\citep[][]{Hardegree-Ullman2023,Gaia_mission,Gaia_DR3} with planetary systems based on state-of-the-art occurrence rates~\citep{Bergsten2022}, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
 Trying different instrumentation and survey designs, we use \bioverse\ to recover runaway greenhouse-induced effects based on model predictions~\citep{Turbet2020,Dorn2021} that we inject into a baseline planet population.
@@
 \section{Baseline Model}\label{sec:met_baseline}
-%The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
-%Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
-%To enable a quantitative assessment of this detectability, we injected the signal into a simulated planet sample before observing it with simulated surveys.
 The goal of this study is to determine -- for different configurations of near-future exoplanet surveys -- the confidence level with which the runaway greenhouse threshold can be detected statistically.
@@
 
-\subsection{Stellar sample from Gaia DR3}
-%Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
-%\todo[inline]{KEVIN: describe stellar population}
-\todo[inline]{KEVIN: update reference to my paper once accepted}
+% include table with key parameters
+\input{table_parameters}
+
+\subsection{Stellar sample from \gaia\ DR3}
 The original \bioverse\ stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function.
-Exquisite parallax and photometric data from the \textit{Gaia} mission made it possible to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \bioverse ~(Hardegree-Ullman et al., submitted).
-Here, we briefly describe how stellar parameters $T_{\mathrm{eff}}$, $L_{\star}$, $R_{\star}$, and $M_{\star}$ were derived.
+Improved parallax and photometric data from the \gaia\ mission made it possible to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \bioverse ~\citep{Hardegree-Ullman2023}.
+Here, we briefly describe how we derived the stellar effective temperature $T_{\mathrm{eff}}$, luminosity $L_{\star}$, stellar radius $R_{\star}$, and stellar mass $M_{\star}$.
 
-Hardegree-Ullman et al. (submitted) used the Gaia Catalogue of Nearby Stars~\citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
-The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2022} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
-From this information, colors and absolute magnitudes were computed, and initial color-magnitude cuts were made to remove non-main sequence stars.
-Effective temperatures were computed primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
-Next, luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
-Stellar radii were then computed with the effective temperatures and luminosities using the Stefan-Boltzmann law, or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M dwarfs.
-Finally, masses were derived from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
-The derived stellar parameters from Hardegree-Ullman et al.
-(submitted) were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
-From this catalog, \bioverse\ samples stars within an isotropic distance from the Solar System as required by the planetary sample size.
+\citet{Hardegree-Ullman2023} used the \gaia\ Catalogue of Nearby Stars~\citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
+The GCNS identified stars out to 120~pc and includes \gaia\ DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2022} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
+From this information, we computed colors and absolute magnitudes and applied initial color-magnitude cuts to remove non-main sequence stars.
+We derived effective temperatures primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
+Luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
+We computed stellar radii with the effective temperatures and luminosities using the Stefan-Boltzmann law or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M~dwarfs.
+Finally, we derived masses from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
+The derived stellar parameters from \citet{Hardegree-Ullman2023} were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
+From this catalog, \bioverse\ samples stars within an isotropic distance from the solar system as required by the planetary sample size.
 
@@
 Planetary systems are hosted by stars of a wide range of ages, and stellar luminosities evolve with time.
-Since the presence of a runaway greenhouse state is highly dependent on the amount of radiation received by the planet, and thus on the luminosity of the host star, we assigned age-dependent luminosities to our synthetic stars.
+Since the occurrence of a runaway greenhouse state is highly dependent on the amount of radiation received by the planet, and thus on the luminosity of the host star, we assigned age-dependent luminosities to our synthetic stars.
 
@@
 For our synthetic stars, we thus drew random ages from a uniform distribution from \SI{0}{\giga\year} to \SI{10}{\giga\year}.
-We then assigned each star a luminosity from the mass-dependent evolutionary models in \citet{Baraffe1998}.
+We then assigned each star a luminosity from the mass-dependent evolutionary models of \citet{Baraffe1998}.
 Figure~\ref{fig:luminosity_tracks} shows the corresponding luminosity evolution as a function of stellar mass and age.
@@
         \caption{
-            Bolometric luminosity tracks of stars with different masses, computed from stellar evolution models in \citet{Baraffe1998}.
+            Bolometric luminosity tracks of stars with different masses, computed from stellar evolution models of \citet{Baraffe1998}.
             Low-mass stars, which make up the majority of stars in the solar neighborhood, undergo an extended early phase of several magnitudes higher luminosity before entering a lifetime of relative faintness.
@@
 Following~\citet{Kipping2013b}, we used a Beta distribution with parameters $a=0.867$ and $b=3.03$, and truncated the distribution at $e = 0.8$.
-%Another parameter with an even higher impact on transit probabilities is the orbital inclination toward the observer $i$.
 Assuming isotropic alignments of orbits, we assigned each planet an inclination drawn from a distribution uniform in $\cos(i)$.
@@
 \subsection{Transit probability}
-We model the occurrence of transits by assuming isotropic orientations of planetary orbits and calculating the impact parameters $b = \cos(i)/R_\star$.
+We model the occurrence of transits by assuming isotropic orientations of planetary orbits and calculating the impact parameters $b = a \cos(i)/R_\star$.
 Following the approach in \citet{Bixel2021}, we further consider only planets with $|b| < 1$.
@@
 The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}.
-With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state~\citep{Lichtenberg2021c,Boukrouche2021}.
-We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean~\citep{Dorn2021,Salvador2023}.
+With even a fraction of the Earth's water inventory, a planet absorbing more flux than the radiation limit of steam atmospheres will enter a runaway greenhouse state resulting in a global magma ocean~\citep{Lichtenberg2021c,Boukrouche2021}.
+We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the distribution of water between planetary interior and atmosphere, and on the resulting thermal atmospheric structure~\citep{Dorn2021,Salvador2023}.
 
-The net absorbed stellar fluxes of planets are sensitive on their albedos, which are generally poorly constrained for planets outside the Solar System~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019}.
+The net absorbed stellar fluxes of planets are sensitive to their albedo, which is generally poorly constrained for planets outside the solar system~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019}.
 Here, we assume global redistribution of incoming flux and a fixed Bond albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
@@
 
-While we search for the runaway greenhouse pattern in demographic quantities such as average planet radii, the injected changes happen on the planetary level: We changed each planet's radius based on its individual set of properties and the associated predictions from steam atmosphere and water incorporation models.
+While we search for the signature of runaway greenhouse climates in demographic quantities such as average planet radii, the injected changes happen on the planetary level: We changed each planet's transit radius based on its individual set of properties and the associated predictions from steam atmosphere and water retention models.
 Relevant properties are a planet's mass $M$, its net instellation $S$, and its bulk water inventory expressed as a water mass fraction $x_{H_2O}$.
@@
 
-\textit{Sub-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
+\textit{Non-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
 This case serves as our null hypothesis.
 
-For \textit{super-runaway} planets, i.e., all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
+For \textit{runaway} planets, i.e., all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
 While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
@@
 To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
-Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our sub-runaway planets~\citep{Zeng2016}.
+Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our non-runaway planets~\citep{Zeng2016}.
 For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
@@
 Nominally, the above models assume a \textit{dry melt} without dissolved volatiles.
-Here, however, we consider a \textit{wet melt} magma ocean and take into account a radius decrease from water partitioning in the melt, which is expected in the situation of a magma ocean below a steam atmosphere~\citep{Dorn2021}.
-The magnitude of this effect depends on the planet's mass and water content and is generally small compared to the radius inflation from a steam atmosphere.
-We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
-We then add the (in almost all cases negative) radius deviations to the planet radii computed for the dry melt case.
+Here, however, we consider a \textit{wet melt} magma ocean and take into account a radius decrease from retention of water in the melt~\citep{Dorn2021}.
+The impact of the water distribution between melt and atmosphere on the change of the transit radius depends on the planet's mass and water content and is generally small compared to the radius inflation from the steam atmosphere.
+%We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
+We computed radius deviations between a wet magma ocean and a solid mantle following \citet{Dorn2021} and assuming a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
+We then added the (in almost all cases negative) radius deviations to the inflated planet radii computed for the dry melt case.
 
 \begin{figure}
+    \script{radiusevolution.py}
     \begin{centering}
@@
         \caption{
-            \textit{Top}: Mass-radius relationships for a water mass fraction $x_{H_2O}= \var{wrr}$ and different planet states.
+            \textit{Top}: Mass-radius relationships for a bulk water mass fraction $x_{H_2O}= \var{wrr}$ and different planet states.
 Green: planets with a solid mantle and no steam atmosphere.
 Dashed: planets with steam atmospheres.
-Blue: planets with steam atmospheres and including the effect of water incorporation in the melt.
-                Steam atmospheres cause a significant radius increase, which is slightly reduced when water incorporation in the melt is considered.
+Blue: planets with steam atmospheres and including the effect of water retention in the melt.
+                Steam atmospheres cause a significant radius increase, which is slightly reduced when water retention in the melt is considered.
             \textit{Bottom}: Radius evolution of different planet types, illustrating degeneracies and potential for confusion among planet classes.
@@
 \end{figure}
-We illustrate the mass-radius relations of the three cases in Figure~\ref{fig:radiusevolution} where a fiducial water mass fraction of $x_{H_2O}= \var{wrr}$ is assumed.
+We illustrate the mass-radius relations of the three cases in Figure~\ref{fig:radiusevolution} where a fiducial bulk water mass fraction of $x_{H_2O}= \var{wrr}$ is assumed.
 % Earth: ~0.0002 PERCENT or 218 ppm by mass (Turbet+2019)
-In the following, we only distinguish between the sub-runaway greenhouse and wet melt scenarios.
+In the following, we only distinguish between the non-runaway greenhouse and wet melt scenarios.
 
-To account for planets unable to sustain a steam atmosphere over extended time spans, as well as evolutionary effects such as desiccation (see bottem panel of Figure~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
-It represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
-Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the sub-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
+To account for planets unable to sustain a steam atmosphere over extended time spans, as well as evolutionary effects such as desiccation through water photodissociation and H~escape (see bottem panel of Figure~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
+It represents the fraction of planets above the instellation threshold whose atmospheres are currently inflated due to a runaway greenhouse climate.
+Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the non-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
 The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere.
@@
 This task includes selection of the targets, application of detection biases, and conducting simulated measurements, all of which are specific to the particular survey.
-For each planet-level measurement such as radius or instellation, we draw the measured value from a normal distribution centered on the true value with a standard deviation set by the survey's precision.
+For each planet-level measurement such as transit radius or instellation, we draw the measured value from a normal distribution centered on the true value with a standard deviation set by the survey's precision.
 We then follow a Bayesian hypothesis testing approach to assess various realizations of simulated surveys in terms of their ability to detect and characterize the runaway greenhouse transition.
 
-\subsection{Detection bias, target selection, and sensitivity}\label{sec:sensitivity} % Turbet2019 lists precision for CHEOPS, PLATO?, TESS?
+\subsection{Detection bias, target selection, and sensitivity}\label{sec:sensitivity} % Turbet2019 lists precision for CHEOPS, \plato\?, TESS?
 Not all transiting planets are detectable with the same likelihood and detection biases have an impact on the demographic measurements we are interested in.
@@
 
-A successful transit detection requires a sufficient signal-to-noise ratio (S/N), which is sensitive on the achieved photometric precision.
-%~\citep[e.g.,][]{Koch2010,Ricker2014,Burke2015,Hardegree-Ullman2019} % Cabrera+ in prep for PLATO
+A successful transit detection requires a sufficient signal-to-noise ratio, which is sensitive to the achieved photometric precision.
+%~\citep[e.g.,][]{Koch2010,Ricker2014,Burke2015,Hardegree-Ullman2019} % Cabrera+ in prep for \plato\
 %In its in-orbit commissioning phase, the \cheops\ mission demonstrated a photometric precision of \SI{75}{\ppm} for faint ($V\approx 12$) stars~\citep{Benz2021}.
 %In comparison, \plato's expected precision for similar stars was estimated to be \SI{\sim 50}{\ppm}~\citep[][Matuszewski et al., in prep.]{plato2017}.
-\plato\ (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars, a goal to be achieved via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
+\plato\ (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
 In line with this requirement, \plato\ is designed to enable the detection of a \SI{80}{\ppm} transit signal~\citep[][Matuszewski et al., in prep.]{plato2017}.
@@
 Under real-world conditions, the planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
-%\subsubsection{PLATO}
 \plato's definition study report~\citep{plato2017} states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity~(RV) follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).
@@
 
-Since planetary bulk density $\rho \propto R_\mathrm{P}^{-3}$, we expect a stronger runaway greenhouse signal when measured through bulk density instead of radius.
+Since planetary bulk density $\rho \propto R_\mathrm{P}^{-3}$, we expect a stronger runaway greenhouse signal when measured through bulk density instead of transit radius.
 We thus simulated measurements of planetary densities assuming the mass-radius relation defined above.
@@
 \subsection{Hypothesis tests}
-We now turn to quantifying the ability of the simulated surveys to detect the habitable zone inner edge discontinuity and to constrain parameters of the runaway greenhouse transition.
+We now turn to quantifying the ability of the simulated surveys to detect the habitable zone inner edge discontinuity and to constrain parameters associated to the runaway greenhouse transition.
 To do this, we rely on a Bayesian hypothesis testing approach where we quantify the evidence of a hypothesis over another based on the (simulated) data.
@@
     where $\theta$ is the set of parameters defining the radius distribution.
-    We further define an alternative hypothesis that describes radius changes due to steam atmospheres and water in the melt.
+    We further define an alternative hypothesis that describes radius changes due to runaway greenhouse climates and inflated steam atmospheres.
     As motivated above, this hypothesis takes the form of a step function in net instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
-    Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
+    Our main observable shall be the average transit radius in the planet population on either side of this threshold.
     The runaway greenhouse hypothesis is then defined as
@@
 \end{equation}
-    Here, $f_\mathrm{rgh}$ is the fraction of planets experiencing a runaway greenhouse effect.
-    $\Delta R_\mathrm{stm}$ and $\Delta R_\mathrm{wtr}$ are predicted radius changes from the steam atmosphere and water incorporation models, respectively.
+    Here, $f_\mathrm{rgh}$ is the fraction of planets above the instellation threshold experiencing a runaway greenhouse effect.
+    $\Delta R_\mathrm{stm}$ and $\Delta R_\mathrm{wtr}$ are predicted radius changes from the steam atmosphere and water retention models, respectively.
     They are assumed to act additively on the planet radii and thus on their average $\langle R_\mathrm{P}\rangle $.
 
-The only free parameter of the null hypothesis, which assumes the average radius to be independent of instellation, is the predicted mean radius $\langle R_\mathrm{P}\rangle $.
-The functional form of the runaway greenhouse hypothesis is more complex: Besides the mean radius of planets outside the threshold $\langle R_\mathrm{P}\rangle_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$, the planetary water mass fraction~$x_{H_2O}$, and the dilution factor $f_\mathrm{rgh}$.
+The only free parameter of the null hypothesis, which assumes the average transit radius to be independent of instellation, is the predicted mean radius $\langle R_\mathrm{P}\rangle $.
+The functional form of the runaway greenhouse hypothesis is more complex: Besides the mean radius of planets outside the threshold $\langle R_\mathrm{P}\rangle_\mathrm{out}$, which is a nuisance parameter necessary to define the hypothesis, it relies on the threshold instellation for the ``step'' $S_\mathrm{thresh}$, the planetary bulk water mass fraction~$x_{H_2O}$, and the dilution factor $f_\mathrm{rgh}$.
 For hypothesis tests based on bulk density instead of radius, we proceeded in the same way and substituted $R_\mathrm{P}$ by the bulk density $\rho$.
@@
 Finally, we adopted a broad, uniform prior for $\langle R_\mathrm{P}\rangle_\mathrm{out}$ bound by $[\prRmin, \prRmax] \, \mathrm{R_\oplus}$.
-In the case of measuring bulk densities instead of radii, we drew uniformly from $[1, 6] \, \SI{}{\gram\per\centi\meter\cubed}$.
+In the case of measuring bulk densities instead of transit radii, we drew uniformly from $[1, 6] \, \SI{}{\gram\per\centi\meter\cubed}$.
 
@@
 
-As our procedure involves random sampling of the model parameters $\theta$, we need to define the probability of obtaining a dataset given the model parameters, i.e., a likelihood function.
+As our procedure involves random sampling of the model parameters $\theta$, we need to define the probability of obtaining a dataset given the model parameters, i.e., a likelihood function $\mathcal{L}$.
 We assumed here that the individual moving averages $SMA_i$ are measured with a normally distributed uncertainty $\sigma_{SMA_i}$ and adopted a normal distribution
@@
 \subsection{Bayesian model comparison}
-We are now ready to assess the relative plausibility of $H_{\mathrm{rgh}}$ and $H_0$ given the synthetic data we have generated, assigning equal a~priori~ probabilities to these models.
+We can now assess the relative plausibility of $H_{\mathrm{rgh}}$ and $H_0$ given the synthetic data we have generated, assigning equal a~priori~ probabilities to these models.
 This is done by comparing the Bayesian evidence $\mathcal{Z}$ of the models, which we estimated with the nested sampling~\citep{Skilling2004} algorithm \emph{dynesty}~\citep{Speagle2020}.
@@
 %The chosen threshold roughly corresponds to the $p = 0.05$ threshold for statistical significance popular in frequentist approaches.\todo{CITATION}
-
-% include table with key parameters
-\input{table_parameters}
 
@@
         \caption{Synthetic planets above and below the runaway greenhouse threshold.
-        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
-        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation. Only the planets marked as transiting are observable.
-            Above the runaway greenhouse threshold $S_\mathrm{thresh} = \var{injectedSthresh}$, some planets maintain their original radii (green crosses) whereas some have their radii inflated (blue crosses) by the amount indicated with gray lines.
+        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in orbital period space.
+        \textit{Bottom}: Transit radii of synthetic planets with injected radius deviation as a function of net instellation. Only the planets marked as transiting are observable.
+            Above the runaway greenhouse threshold $S_\mathrm{thresh} = \var{injectedSthresh}$, some planets maintain their original radii (green crosses) whereas some have their transit radius inflated (blue crosses) by the amount indicated with gray lines.
         The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue line).
@@
         \caption{
-        Detection of the runaway greenhouse threshold in the optimistic case.
-        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
+        Detection of the runaway greenhouse threshold.
+        From simulated radius and instellation measurements of a large ($N \approx \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
             The pattern is detected with high significance.
@@
 Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the habitable zone inner edge discontinuity.
-Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
-Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
+Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N \approx \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
+Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is $f_\mathrm{rgh} = \frgh$, and we chose a water mass fraction of $x_{H_2O} = \var{wrr}$ for each planet.
 In this case, the habitable zone inner edge discontinuity was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).
@@
             Statistical power of the runaway greenhouse hypothesis test as a function of model parameters.
-            For a sample size $N = \var{N_optimistic}$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and dilution factor.
+            For a sample size $N \approx \var{N_optimistic}$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and dilution factor.
             Higher values in either parameter result in a more reliable detection.
@@
 %\todo{Introduce survey simulation(s).
-PLATO, Cheops, (Ariel, LIFE/LIFE precursor, \nautilus\)\\ What are threshold missions that are able to detect this signal?
+\plato\, Cheops, (Ariel, LIFE/LIFE precursor, \nautilus\)\\ What are threshold missions that are able to detect this signal?
 =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}
@@
             The median values of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
-            \textit{Left:} For a large planet yield of $N = \var{N_plato}$, even small dilution factors $\sim \var{platoMinfrgh}$ allow a detection.
+            \textit{Left:} For a large planet yield of $N \approx \var{N_plato}$, even small dilution factors $\sim \var{platoMinfrgh}$ allow a detection.
                 A sample of 100 planets is sufficient if their masses are constrained to within \SI{10}{\percent} (dotted green line).
@@
 To simulate the transit survey of ESA's \plato\ mission, we considered a volume-limited sample with a size according to projections and including all stellar spectral types.
-Given \plato's expected radius precision, we find that a yield of \var{N_plato} planets enables a significant detection if the fraction of runaway greenhouse planets is larger than $ f_\mathrm{rgh} \sim \var{platoMinfrgh}$ (see Figure~\ref{fig:plato_fgrid}).
+Given \plato's expected radius precision, we find that a yield of $\sim$\var{N_plato} is needed for a significant detection if the fraction of runaway greenhouse planets $ f_\mathrm{rgh} = \var{platoMinfrgh}$ (see Figure~\ref{fig:plato_fgrid}).
 The minimum needed fraction rises to \var{platoMinfrghHundred} for $N = 100$.
@@
 \subsubsection{Dependence on host star spectral type}\label{sec:results_FGK_M}
-Since the incident radiation depends on the spectral type of the host star, the relative number of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
+Since the incident radiation at a given orbital distance depends on the spectral type of the host star, the relative number of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
 We tested the detectability of the habitable zone inner edge discontinuity when only FGK or only M~dwarfs are considered (see Figure~\ref{fig:plato_fgrid}).
@@
 \end{figure}
-A key constraint resulting from a detection is a measurement of $S_{thresh}$.
+A key constraint resulting from a detection is a measurement of $S_\mathrm{thresh}$.
 Here, we assess the ability of different mission concepts explored above to constrain this parameter.
@@
 We consider three cases: only radius measurements, radius and mass measurements, and radius and mass measurements of only planets orbiting M~dwarfs.
-The simulations otherwise represent the simulated PLATO example described above.
+The simulations otherwise represent the simulated \plato\ example described above.
 We chose a planetary sample size of $N=100\pm10$, as this was found to be a threshold case in Sect.~\ref{sec:res_samplesize}.
@@
 We find that retrievals from radius measurements alone require high fractions of greenhouse climate-bearing planets to achieve an accurate constraint on $S_\mathrm{thresh}$:
-Dilution factors $f_\mathrm{rgh} \gtrsim 0.5$ yield posterior probability distributions that are condensed at the order of magnitude of the true value; accurate constraints to within $\pm 0.25\,\mathrm{dex}$ of the truth are reached only from $f_\mathrm{rgh} \approx 0.8$.
+dilution factors $f_\mathrm{rgh} \gtrsim 0.5$ yield posterior probability distributions that are condensed at the order of magnitude of the true value; accurate constraints to within $\pm 0.25\,\mathrm{dex}$ of the truth are reached only from $f_\mathrm{rgh} \approx 0.8$.
 
@@
 \section{Discussion}\label{sec:discussion}
-%\todo[inline]{\textbf{Potential additional discussion points: }\\
 %(How close is Earth to runaway? (-> \citet{Goldblatt2012,Popp2016}; what is the max. expected change in Earth's albedo due to deglaciation?)}
@@
 
-\subsubsection{Key factors influencing tests of the runaway greenhouse hypothesis}\ref{sec:dis_keyfactors}
+\subsubsection{Key factors influencing tests of the runaway greenhouse hypothesis}\label{sec:dis_keyfactors}
 What influences the probability of correctly rejecting a false null hypothesis?
@@
 \textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies on sufficient amounts of atmospheric water vapor that can act as a greenhouse gas.
-However, already about \SIrange{\sim 10}{20}~bar of water vapor -- corresponding to a minor fraction of one Earth ocean and thus the lower limit of water on Earth -- is enough to sustain sufficiently high surface temperatures to keep the planet in a magma ocean stage~\citep{Boukrouche2021,Lichtenberg2021c}.
+However, already about $\SIrange{\sim 10}{20} \, \mathrm{bar}$ of water vapor -- corresponding to a minor fraction of one Earth ocean and thus the lower limit of water on Earth -- is enough to sustain sufficiently high surface temperatures to keep the planet in a magma ocean stage~\citep{Boukrouche2021,Lichtenberg2021c}.
 The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
-From a planet formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard outcome~\citep{2019PNAS..116.9723Z,Venturini2020}.
-But while commonly considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the Solar System~\citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation.
+From a planet formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard expected outcome~\citep[e.g.,][]{2019PNAS..116.9723Z,Venturini2020,Emsenhuber2021b,Schlecker2021,Burn2021}.
+But while commonly considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the solar system~\citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation.
 Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amounts of water inside of the planet during the magma ocean phase~\citep{Ikoma2018,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}.
@@
 
-\textit{Size and composition of the planetary sample}: Evidently, the significance of a statistical trend increases with a larger sample size.
+\textit{Size and composition of the planetary sample}: The significance of a statistical trend increases with a larger sample size.
 In addition, the sample must include planets on both sides of the instellation threshold.
-Due to the strongly distance-dependent detection bias associated with the transit method and the temporal coverage of upcoming transit missions, this is only likely for low-mass host stars.
+This is only likely for low-mass host stars due to the strongly distance-dependent detection bias associated with the transit method and the temporal coverage of upcoming transit missions.
 
@@
 Alternatives include atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{2012ApJ...753...66I,Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
-Both processes are being discussed as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
-%    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
-Either of them leads to a decrease of planet radius for planets close to their host star~\citep{Pascucci2019,Bergsten2022}, which is distinct from the radius inflation introduced by runaway greenhouse climates.
+Both processes are being discussed as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}, and both lead to a decrease of planet radius for planets close to their host star~\citep{Pascucci2019,Bergsten2022}.
+This is distinct from the radius inflation introduced by runaway greenhouse climates.
 For example, the innermost planet in the \mbox{K2-3} system has an increased radius compared to its outer siblings, contrary to what would be expected from atmospheric escape~\citep{2022AJ....164..172D}.
@@
 Although an abrupt pattern at the expected location of the transition seems unlikely, examples of steep occurrence rate density changes exist.
-An example is the ``Neptune desert'', a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
+An example is the ``Neptune desert'', a triangular region in period-radius space of low planet occurrence~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
 The shape of this region is such that smaller planets become less frequent the closer to the star they are, which to some degree resembles the pattern introduced by the instellation dependency of the runaway greenhouse transition.
-However, the Neptune desert occurs at smaller orbital periods and its location depends on a planet's size~\citep{Szabo2011}, which is not expected for the runaway greenhouse transition.
+However, the Neptune desert occurs at smaller orbital periods and is sensitive to the planet radius~\citep{Szabo2011}, which is not expected for the runaway greenhouse transition.
 
-\citet{Luque2022} found that small planets around orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
+\citet{Luque2022} found that small planets orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
 This trend does not represent a false positive scenario for the habitable zone inner edge discontinuity, since no strong dependency on instellation has been found or is expected.
@@
 
-We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population.
+We currently see four potential lines of discriminating atmospheric signatures associated with runaway climates in the exoplanet population.
 (i) Detecting the atmospheric windows of water vapor in the near- to mid-infrared.
-In a super-runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
-This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as \life\ ~\citep{2019A&A...621A.125B,Quanz2022,2022A&A...664A..22D}.
-Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime~\citep{2020SSRv..216...98G}.
-(ii) Post-runaway planets may be detectable by O$_{3}$ absorption due to build-up of abiotic oxygen, left-over from photochemical dissociation, and hydrogen loss~\citep{2014ApJ...785L..20W,Luger2015}.
-(iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
+In a runaway greenhouse atmosphere, absorption is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
+Probing these features requires an instrument covering these wavelength ranges, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as \life\ ~\citep{2019A&A...621A.125B,Quanz2022,2022A&A...664A..22D}.
+Potentially the two-band filter capabilities of \plato\ may offer insight into particularly pronounced spectral features in this runaway greenhouse regime~\citep{2020SSRv..216...98G}.
+(ii) Post-runaway planets may be detectable through O$_{3}$ absorption due to build-up of abiotic oxygen, leftover from photochemical dissociation of water, and hydrogen loss~\citep{2014ApJ...785L..20W,Luger2015}.
+(iii) Water loss in runaway greenhouse episodes would increase the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
 This may be detectable in high-resolution observations focusing on isotope-sensitive transitions~\citep{2019AJ....158...26L,2019A&A...622A.139M}.
-(iv) Disequilibrium chemistry in tidally-locked super-runaway planets.
+(iv) Disequilibrium chemistry in tidally-locked runaway planets.
 It has been suggested that the atmospheric depth and the presence or absence of oceans on ``sub-Neptunes'' could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers~\citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface~\citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
-Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma~\citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
+Atmospheric nitrogen and carbon compounds can be partitioned into magma~\citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
 Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed.
@@
 Confirming or disproving the predicted habitable zone inner edge discontinuity will depend on the significance with which the null hypothesis can be excluded, which is a function of instrumentation and survey strategy.
-As discussed in Section~\label{sec:dis_keyfactors}, key drivers from a mission design perspective are sample size, photometric precision, and the availability of planets around low-mass host stars.
+As discussed in Section~\ref{sec:dis_keyfactors}, key drivers from a mission design perspective are sample size, photometric precision, and the availability of planets around low-mass host stars.
 We found that along these axes, \plato\ will be the most favorable among the upcoming transit missions.
-The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission.% (Matuszewski, priv. comm.).
+The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission. % (Matuszewski, priv. comm.).
 %In the FGK sample of the mission, several hundred Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
@@
 The ongoing \cheops\ mission was designed as a follow-up mission to search for transits of planets discovered with other techniques, in particular with radial velocity measurements~\citep{Benz2021}.
-As such, it will provide precise radius constraints on a sample of small planets; however, only a small number of planets with periods \SI{> 50}{\day} are being observed.
+As such, it will provide precise radius constraints on a sample of small planets; however, only a small number of planets with orbital periods \SI{> 50}{\day} are being observed.
 This largely limits \cheops' coverage to planets within the runaway greenhouse regime, preventing a detection of the transition.
@@
 Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios~\citep{Alei2022,Konrad2022}.
-Since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
+Since the peak thermal emission in runaway greenhouse atmospheres will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
 Finally, mid-infrared surveys such as \life\ show a preference for M~star planets~\citep{Quanz2022}, which is beneficial for detecting the predicted habitable zone inner edge discontinuity (see Sect.~\ref{sec:results_FGK_M}).
@@
 This way, useful results can be obtained under more pessimistic conditions, e.g., a low predominant water content of planetary surfaces and atmospheres or a smaller available planet sample.
-For a mission design similar to PLATO, a density-based hypothesis test on about a third of the overall sample is equivalent to a pure radius-based analysis.
+For a mission design similar to \plato\, a density-based hypothesis test on about a third of the overall sample is equivalent to a pure radius-based analysis.
 At a fixed sample size, key parameters of the runaway greenhouse models can be more narrowly constrained when additional mass measurements are available.
@@
 The diagnostic power of the hypothesis tests we demonstrated here may be improved by simultaneously fitting for the habitable zone inner edge discontinuity in the subsample without RV follow-up.
-An optimized mission in search for the inner edge of the habitable zone will further enhance its information content via a informed selection of follow-up targets, i.e., balancing objects located on either side of the expected instellation threshold $S_\mathrm{thresh}$.
+An optimized mission in search for the inner edge of the habitable zone will further enhance its information content via an informed selection of follow-up targets, i.e., balancing objects located on either side of the expected instellation threshold $S_\mathrm{thresh}$.
 
@@
 
-Once a runaway greenhouse region is identified in the parameter space, the community will will have a tool at hand to discern potentially habitable planets from Venusian worlds on an empirical basis.
+Once a runaway greenhouse region is identified in the parameter space, the community will  have a tool at hand to discern potentially habitable planets from Venusian worlds on an empirical basis.
 Together with atmospheric measurements (see Sect.~\ref{sec:dis_atmospheres}), we will be able to put a number on the probability of an individual planet to harbor sufficient surface water to sustain life.
@@
 \subsection{Impact of assumptions on our findings}
-%\todo[inline]{Caveats\\
 %- not yet explored: how sensitive are we on the assumed planet population?
@@
 If any currently unknown sharp features in the distribution of terrestrial planets emerge, they should be considered in future studies.
-%However, it should be kept in mind that any unknown sharp features in the true distribution of terrestrial planets could complicate a detection or lead to a false positive result (see Sect.~\ref{sec:dis_falsepositive}).
 
@@
 
-Future self-consistent modeling of interior-atmosphere interactions may include constraints on additional radius increases due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,Lichtenberg2021c,2022PSJ.....3...93B,2021SSRv..217...22G}.
+Future self-consistent modeling of interior-atmosphere interactions may include constraints on additional radius increases due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,Lichtenberg2021c,2022PSJ.....3...93B,2021SSRv..217...22G} or water outgassing efficiency~\citep[e.g.,][]{Hier-Majumder2017,Ikoma2018,Salvador2023}.
 
-\subsubsection{Water mass fraction}
+\subsubsection{Bulk water mass fraction}
 The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
 Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021,2022ApJ...938L...3L,2022ApJ...939L..19I}.
-Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
+Our nominal case assumes a bulk water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
 This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the habitable zone inner edge discontinuity.
-%, and the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
+%, and the hypothesis tests are only marginally sensitive to $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
 Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
@@
 The habitable zone inner edge discontinuity may be affected by several processes that are challenging to quantify:
-Planets that lack an atmosphere or sufficient volatiles cannot bear steam atmospheres, and those that do eventually move to the sub-runaway greenhouse category due to desiccation~\citep[][]{Watson1981} or evolution of their host star~\citep{Luger2015}.
+Planets that lack an atmosphere, sufficient volatiles, or have non-water-dominated outgassed compositions cannot bear steam atmospheres, and those that do eventually move to the non-runaway greenhouse category due to desiccation~\citep[][]{Watson1981,Kasting1983,Hamano2013} or evolution of their host star~\citep{Luger2015}.
 Hydrogen/Helium-dominated planets may disguise as inflated rocky planets and not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022,Innes2023}.
 A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, reducing their transit radius.
-Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Turbet2021,Lichtenberg2021c,Pierrehumbert2022,Innes2023}.
+Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Salvador2017,Turbet2021,Lichtenberg2021c,Pierrehumbert2022,Innes2023}.
 The choice of a statistical estimator for the hypothesis tests may further influence the recovered discontinuity; we compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.
@@
 \begin{enumerate}
-    \item The predicted runaway greenhouse transition causes a discontinuity in the radius and density distribution of small exoplanets with respect to the incident host star irradiation.
+    \item The predicted runaway greenhouse transition causes a discontinuity in the radius and density distribution of small exoplanets with respect to their irradiation.
     \item This habitable zone inner edge discontinuity should be detectable with high-precision transit measurements.
-          For a planet sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.
+          For a planet sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical bulk water mass fractions are above $\sim 10^{-3}$.
     \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to confirm or reject the predicted trend.
@@
 \begin{acknowledgments}
-The authors thank Gabriele Cugno, Yann Fabel, Brad Foley, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
+The authors thank Gabriele Cugno, Yann Fabel, Brad Foley, Lisa Kaltenegger, Ravi Kopparapu, Eric Mamajek, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, Sukrit Ranjan, and Terry-Ann Suer for insightful discussions.
 We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
@@
 The results reported herein benefitted from collaborations and/or information exchange within NASA’s Nexus for Exoplanet System Science (NExSS) research coordination network sponsored by NASA’s Science Mission Directorate.
-This work has made use of data from the European Space Agency (ESA) mission {\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia} Data Processing and Analysis Consortium (DPAC, \url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}).
-Funding for the DPAC has been provided by national institutions, in particular the institutions participating in the {\it Gaia} Multilateral Agreement.
+This work has made use of data from the European Space Agency (ESA) mission \gaia\ (\url{https://www.cosmos.esa.int/gaia}), processed by the \gaia\ Data Processing and Analysis Consortium (DPAC, \url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}).
+Funding for the DPAC has been provided by national institutions, in particular the institutions participating in the \gaia\ Multilateral Agreement.
 T.L. was supported by a grant from the Branco Weiss Foundation.
+G.B. acknowledges support from the NASA Astrophysics Data Analysis Program under Grant No.~80NSSC20K0446.
+A.S. acknowledges support from NASA's Habitable Worlds Program (No.~80NSSC20K0226).
 \end{acknowledgments}
@@
 This study uses the reproducibility framework ``showyourwork''~\citep{Luger2021}.
-All code required to reproduce our results, figures, and this article itself is available at \url{https://github.com/matiscke/magmaocean}.
+All code required to reproduce our results, figures, and this article itself is available at \url{https://github.com/matiscke/hz-inner-edge-discontinuity}.
 The code to reproduce a figure can be accessed via the icon link next to the respective figure caption.
@@
             Detection of the runaway greenhouse threshold with a binned statistic.
-            Using simulated data from the optimistic survey case ($N = \var{N_optimistic}$, compare Figure~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis.
+            Using simulated data from the optimistic survey case ($N \approx \var{N_optimistic}$, compare Figure~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis.
 Instead of using a running mean, we computed a binned statistic (arithmetic mean and standard deviation in blue) to assign planets the average radius or bulk density in their neighborhood in instellation space.
@@
 The accuracy of the recovered instellation threshold is comparable.
-This test demonstrates that our results are not sensitive on the choice of statistical estimator to test the hypotheses against.
+This test demonstrates that our results are not sensitive to the choice of statistical estimator to test the hypotheses against.
 

### 2023-05-01 src/scripts/plotstyle.py

@@ -17,7 +17,7 @@ def styleplots():
 
     font = {'family': 'sans',
             'weight': 'normal',
-            'size': 13}  # 9 pt corresponds to caption size of 1-col A&A
+            'size': 14}
     legend = {'handlelength': 0.75,
               'handletextpad': 0.5}
     figure = {'dpi': 100,

### 2023-05-01 src/scripts/radiusevolution.py
plot radius evolution via ms workflow

@@ -1,0 +1,130 @@
+"""Plot Radius evolution of different planet types, illustrating degeneracies
+and potential for confusion among planet classes.
+"""
+import pickle
+# import hzied_pipeline
+
+from bioverse.constants import CONST
+
+import paths
+import numpy as np
+import seaborn as sns
+
+from matplotlib import pyplot as plt
+import plotstyle
+plotstyle.styleplots()
+
+from matplotlib.ticker import ScalarFormatter
+from scipy.interpolate import splrep, BSpline, interp1d
+
+
+def interpolate_grid(t, R):
+    tck = splrep(t, R, k=3)
+    # t_fine = np.geomspace(min(t), max(t), 50)
+    t_fine = np.linspace(min(t), max(t), 50)
+    return 10 ** t_fine, BSpline(*tck)(t_fine)
+
+
+def plot_radiuscomparison(sample, planets_args):
+    dd = sample.to_pandas()
+    d = dd.sample(min([len(dd), 4000]))
+    mo = dd[dd.has_magmaocean].sample(min([len(dd[dd.has_magmaocean]), 4000]))
+
+    fig, ax = plt.subplots()
+
+    def interpolate_MR(df, colname):
+        # interpolate M-R relation
+        return interp1d(df.M, df[colname], kind='quadratic')
+
+    M = np.linspace(0.33, 1.9, num=200)
+    try:
+        ax.plot(M, interpolate_MR(mo.sort_values('M'), 'R_steam')(M),
+                lw=2., label='dry melt', c='xkcd:dark grey', ls='--')
+    except AttributeError:
+        pass
+    ax.plot(M, interpolate_MR(mo.sort_values('M'), 'R')(M),
+            lw=3., label='wet melt', c='C2')
+
+    ax.plot(M, interpolate_MR(d.sort_values('M'), 'R_orig')(M),
+            lw=3., label='non-\nrunaway', c='C0')
+
+    ax.set_xlabel('$M_P \,[M_\oplus]$')
+    ax.set_ylabel('$R_P \,[R_\oplus]$')
+    ax.spines['right'].set_visible(False)
+    ax.spines['top'].set_visible(False)
[…]

### 2023-05-01 src/scripts/plot_luminosity_tracks.py
plot radius evolution via ms workflow

@@ -1,15 +1,14 @@
-import matplotlib as plt
 import pandas as pd
 import matplotlib.pyplot as plt
 import glob
 from natsort import natsorted
 from scipy import interpolate
-import numpy as np
 import seaborn as sns
 import cmocean
 
 import paths
 import plotstyle
+plotstyle.styleplots()
 
 # sns.set(style="ticks")
 

### 2023-05-08 src/tex/ms.tex
plot radius evolution via ms workflow
apply comments

--- 2023-04-25 src/tex/ms.tex
@@
 Long-term magma ocean phases on rocky exoplanets orbiting inside of the runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
-Thermal stratification of super-runaway planets is expected to significantly inflate their radii, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius--density space.
+Thermal stratification of runaway planets is expected to significantly inflate their radii, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius--density space.
 Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages the combination of contextual information from the overall planet population with a survey simulator and a hypothesis testing framework.
 
-We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor super-runaway climates.
+We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor runaway climates.
 Our survey simulations suggest that in the near future, ESA's \plato\ mission will be the most promising survey to probe the habitable zone inner edge discontinuity.
-We determine survey strategies that maximize the statistical power of the obtained data, and we identify as key mission design drivers: 1.
+We determine survey strategies that maximize the diagnostic power of the obtained data, and we identify as key mission design drivers: 1.
 A follow-up campaign of planetary mass measurements and 2.
 The fraction of low-mass stars in the target sample.
-Observational constraints on the runaway climate transition will provide crucial insights into the distribution of atmospheric volatiles among rocky exoplanets, which may help to identify the nearest potentially habitable worlds.
+Observational constraints on the runaway greenhouse transition will provide crucial insights into the distribution of atmospheric volatiles among rocky exoplanets, which may help to identify the nearest potentially habitable worlds.
 \end{abstract}
@@
 Due to the inherent detection biases of current exoplanet detection techniques, the best-studied category of rocky exoplanets at present is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
-These bodies experience conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which likely profoundly affected the distribution of volatiles between planetary core, mantle, and atmosphere.
+These bodies experience conditions that are in some aspects comparable to the environment of the inner solar system bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which likely profoundly affected the distribution of volatiles between planetary core, mantle, and atmosphere.
 Studies of the geophysical state of hot exoplanets can thus inform our understanding of the early evolutionary stages of Earth and other habitable worlds~\citep{Lichtenberg2022,Krijt2022}.
 
-An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep[e.g.,][]{2019JGRE..124.2015K,2021JGRE..12606643K}.
+An example from the solar system for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep[e.g.,][]{2019JGRE..124.2015K,2021JGRE..12606643K}.
 While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties~\citep{Smrekar2018}, they evolved into planets with very different surface conditions~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,Way2020,Turbet2021}.
@@
 Observational tests being considered include searches for direct evidence of liquid water conveyed by ocean glint~\citep{Williams2008,Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
-A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019,Apai2019a}.
+A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019,Apai2019a}.
 Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011,Bixel2020}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
@@
 The original \bioverse\ stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function.
-Exquisite parallax and photometric data from the \textit{Gaia} mission made it possible to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \bioverse ~(Hardegree-Ullman et al., submitted).
+Exquisite parallax and photometric data from the \textit{Gaia} mission made it possible to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \bioverse ~\citep{Hardegree-Ullman2023}.
 Here, we briefly describe how stellar parameters $T_{\mathrm{eff}}$, $L_{\star}$, $R_{\star}$, and $M_{\star}$ were derived.
 
-Hardegree-Ullman et al. (submitted) used the Gaia Catalogue of Nearby Stars~\citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
+\citet{Hardegree-Ullman2023} used the Gaia Catalogue of Nearby Stars~\citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
 The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2022} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
@@
 Finally, masses were derived from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
-The derived stellar parameters from Hardegree-Ullman et al.
-(submitted) were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
-From this catalog, \bioverse\ samples stars within an isotropic distance from the Solar System as required by the planetary sample size.
+The derived stellar parameters from \citet{Hardegree-Ullman2023} were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
+From this catalog, \bioverse\ samples stars within an isotropic distance from the solar system as required by the planetary sample size.
 
@@
 
-The net absorbed stellar fluxes of planets are sensitive on their albedos, which are generally poorly constrained for planets outside the Solar System~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019}.
+The net absorbed stellar fluxes of planets are sensitive on their albedos, which are generally poorly constrained for planets outside the solar system~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019}.
 Here, we assume global redistribution of incoming flux and a fixed Bond albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
@@
 From a planet formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard outcome~\citep{2019PNAS..116.9723Z,Venturini2020}.
-But while commonly considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the Solar System~\citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation.
+But while commonly considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the solar system~\citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation.
 Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amounts of water inside of the planet during the magma ocean phase~\citep{Ikoma2018,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}.

### 2023-05-08 src/tex/ms.tex
change optimistic/PLATO sample size to "~" 300/500
to avoid confusion why we end up with 288 or 497.

@@
         Detection of the runaway greenhouse threshold in the optimistic case.
-        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
+        From simulated radius and instellation measurements of a large ($N \approx \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
             The pattern is detected with high significance.
@@
 Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the habitable zone inner edge discontinuity.
-Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
+Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N \approx \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
 Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.

### 2023-05-09 src/tex/ms.tex
edit intro of runaway GH

@@
 
-Yet, \citet{Hamano2013} suggested that the dry conditions on Venus may have been directly inherited from the early magma ocean stage if a strongly infrared-absorbing, condensable species such as water is present on the planet.
-For Venus, this is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
-Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
-This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
-Venus' atmospheric composition is consistent with the planet having gone through this phase~\citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.
+Yet, \citet{Hamano2013} suggested that the dry conditions on Venus may have been directly inherited from the early magma ocean stage.% if a strongly infrared-absorbing, condensable species such as water is present on the planet.
+If a strongly infrared-absorbing, condensable species such as water was dominant in the atmosphere, the resulting strong thermal blanketing effect would prevent the planet to efficiently radiate to space and maintain the surface molten~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
+This runaway greenhouse state can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
+%This scenario requires early water delivery, which for Venus is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
+%Where the incident stellar flux exceeds a threshold value, a non-linear transition from a mixture of condensed and volatile states to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
+For Venus, isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B} and the planet's atmospheric composition suggest that it went through this phase~\citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.
 
@@
 Observational tests being considered include searches for direct evidence of liquid water conveyed by ocean glint~\citep{Williams2008,Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
-A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019,Apai2019a}.
+A different approach relies on comparative planetology:
+aiming for a statistical detection in a planet population provides robustness against ambiguity that could otherwise arise from individual variations in a planet's composition or geophysical history~\citep{Checlair2019,Apai2019a}.
 Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011,Bixel2020}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.

### 2023-05-09 src/tex/ms.tex
implement comments
mostly introduction

@@
 \begin{abstract}
-Long-term magma ocean phases on rocky exoplanets orbiting inside of the runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
-Thermal stratification of runaway planets is expected to significantly inflate their radii, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius--density space.
+Long-term magma ocean phases on rocky exoplanets orbiting closer to their star than the runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
+Thermal stratification of runaway planets is expected to significantly inflate their atmospheres, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius--density space.
 Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages the combination of contextual information from the overall planet population with a survey simulator and a hypothesis testing framework.
 
-We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor runaway climates.
+We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for sample sizes $\gtrsim$~100~planets if at least \SI{\sim 10}{\percent} of those orbiting closer than the habitable zone inner edge harbor runaway climates.
+%We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor runaway climates.
 Our survey simulations suggest that in the near future, ESA's \plato\ mission will be the most promising survey to probe the habitable zone inner edge discontinuity.
-We determine survey strategies that maximize the diagnostic power of the obtained data, and we identify as key mission design drivers: 1.
+We determine survey strategies that maximize the diagnostic power of the obtained data and identify as key mission design drivers: 1.
 A follow-up campaign of planetary mass measurements and 2.
@@
 Despite recent advancements in observational techniques, our understanding of terrestrial-sized planets remains woefully limited, with fundamental aspects of their nature, composition, and potential habitability still largely unknown.
-Due to the inherent detection biases of current exoplanet detection techniques, the best-studied category of rocky exoplanets at present is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
-These bodies experience conditions that are in some aspects comparable to the environment of the inner solar system bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which likely profoundly affected the distribution of volatiles between planetary core, mantle, and atmosphere.
-Studies of the geophysical state of hot exoplanets can thus inform our understanding of the early evolutionary stages of Earth and other habitable worlds~\citep{Lichtenberg2022,Krijt2022}.
+Due to the inherent biases of current exoplanet detection techniques, the best-studied category of rocky exoplanets at present is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
+These experience thermal states that are in some aspects comparable to the ones of the inner solar system bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which likely profoundly affected the distribution of volatiles between planetary core, mantle, and atmosphere.
+Studying the geophysical state of hot exoplanets can thus inform our understanding of the early evolutionary stages of Earth and other habitable worlds~\citep{Lichtenberg2022,Krijt2022}.
 
-An example from the solar system for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep[e.g.,][]{2019JGRE..124.2015K,2021JGRE..12606643K}.
+An example from the solar system for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep[e.g.,][]{2019JGRE..124.2015K,2021JGRE..12606643K,Salvador2023b}.
 While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties~\citep{Smrekar2018}, they evolved into planets with very different surface conditions~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,Way2020,Turbet2021}.
-Both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
-Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets~\citep{2020ChEG...80l5594F} because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
-Due to these similar formation sequences, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Way2020}.
+Both planets likely underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
+Magma ocean states play a substantial role in establishing the long-term geophysical and climatic regimes of rocky planets~\citep{2020ChEG...80l5594F}, in particular owing to efficient heat and volatile transfers between interior and atmosphere in the absence of a stiff boundary separating them~\citep{2021ApJ...909L..22K,Dorn2021,Salvador2023b}.
+% because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
+Due to these similar formation sequences, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus' water loss -- occurred late in their evolution~\citep[e.g.,][]{Way2020}.
 
-Yet, \citet{Hamano2013} suggested that the dry conditions on Venus may have been directly inherited from the early magma ocean stage.% if a strongly infrared-absorbing, condensable species such as water is present on the planet.
+Yet, \citet{Hamano2013} suggested that the present-day dry conditions on Venus may have been directly inherited from the early magma ocean stage.% if a strongly infrared-absorbing, condensable species such as water is present on the planet.
 If a strongly infrared-absorbing, condensable species such as water was dominant in the atmosphere, the resulting strong thermal blanketing effect would prevent the planet to efficiently radiate to space and maintain the surface molten~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
@@
 %Where the incident stellar flux exceeds a threshold value, a non-linear transition from a mixture of condensed and volatile states to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
-For Venus, isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B} and the planet's atmospheric composition suggest that it went through this phase~\citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.
+For Venus, isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B} and the planet's atmospheric composition suggest that it went through this phase~\citep{2020NatGe..13..265G}.
+Although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}, a transient habitable phase cannot be conclusively ruled out~\citep[e.g.,][]{Way2016,Salvador2017,Krissansen-Totton2021}.
 
-The runaway greenhouse transition is a robust prediction from climate models~\citep{Forget2014,Boukrouche2021,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
-In particular, planets in a super-runaway climate state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii by an amount that is sensitive on the water content.
-However, dissolution of water~\citep[e.g.,][]{Elkins-Tanton2008,Hier-Majumder2017,Salvador2023} in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,Schlichting2022}.
-Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their mantles and establish an observational connection between exoplanetary interiors and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.
+The runaway greenhouse transition is a robust prediction from climate models~\citep{Kasting1988,Nakajima1992,Goldblatt2012,Forget2014,Boukrouche2021,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
+In particular, planets in a runaway greenhouse state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii by an amount that is a function of the water content.
+However, dissolution of water~\citep[e.g.,][]{Elkins-Tanton2008,Hier-Majumder2017,Salvador2023} in the magma may decrease this effect~\citep{Dorn2021}, and chemical exchange between core and mantle material may influence the amount and speciation of outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,Schlichting2022}.
+Astronomical observations of planets that are currently in a runaway greenhouse state may thus constrain properties of their mantles and establish an observational connection between exoplanetary interiors and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.
 
-Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{2010ppc..book.....P,Kopparapu2013}.
-Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone.
-Its prevalent definition refers to the possibility of liquid water on the surface of an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmosphere~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but it can be influenced by different internal redox states and volatile compositions~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
+Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp instellation threshold~\citep{Goldblatt2013,Leconte2013,Kopparapu2013}.
+Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone~\citep[e.g.,][]{Ramirez2018b,Salvador2023b}.
+Its prevalent definition refers to the possibility of a sustained liquid water body on the surface of an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmosphere~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014};
+its exact spatial location and extent may be strongly influenced by the interior and atmosphere oxidation state and resulting atmosphere composition~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
 The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in the National Aeronautics and Space Administration (NASA) Astrophysics Data System returned $\sim\num{1700}$ results.}.
-However, it should be emphasized that the habitable zone, as it stands today, is merely a concept based on theoretical predictions and geochemical evidence from one planet -- Earth~\citep{2020SciA....6.1420C} -- and its general validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
-The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.
+However, it should be emphasized that the habitable zone, as it stands today, is merely a concept based on theoretical predictions and geochemical evidence from one planet -- Earth~\citep{2020SciA....6.1420C} -- and its general validity remains controversial~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
+The question naturally emerges if a planetary habitable zone -- in its common form with boundaries defined by stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.
 
@@
 An overview of our key assumptions and model parameters can be found in Table~\ref{tab:params_table}.
+
+% include table with key parameters
+\input{table_parameters}
 
@@
 
-% include table with key parameters
-\input{table_parameters}
-
 \section{Results}\label{sec:results}
@@
 
-\textit{Size and composition of the planetary sample}: Evidently, the significance of a statistical trend increases with a larger sample size.
+\textit{Size and composition of the planetary sample}: The significance of a statistical trend increases with a larger sample size.
 In addition, the sample must include planets on both sides of the instellation threshold.
@@
 Confirming or disproving the predicted habitable zone inner edge discontinuity will depend on the significance with which the null hypothesis can be excluded, which is a function of instrumentation and survey strategy.
-As discussed in Section~\label{sec:dis_keyfactors}, key drivers from a mission design perspective are sample size, photometric precision, and the availability of planets around low-mass host stars.
+As discussed in Section~\ref{sec:dis_keyfactors}, key drivers from a mission design perspective are sample size, photometric precision, and the availability of planets around low-mass host stars.
 We found that along these axes, \plato\ will be the most favorable among the upcoming transit missions.
@@
 \begin{enumerate}
-    \item The predicted runaway greenhouse transition causes a discontinuity in the radius and density distribution of small exoplanets with respect to the incident host star irradiation.
+    \item The predicted runaway greenhouse transition causes a discontinuity in the radius and density distribution of small exoplanets with respect to their irradiation.
     \item This habitable zone inner edge discontinuity should be detectable with high-precision transit measurements.

### 2023-05-09 src/tex/ms.tex
shorten abstract to obey AAS word limit

@@
 Thermal stratification of runaway planets is expected to significantly inflate their atmospheres, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius--density space.
-Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages the combination of contextual information from the overall planet population with a survey simulator and a hypothesis testing framework.
+Here, we use \bioverse, a statistical framework combining contextual information from the overall planet population with a survey simulator, to assess the ability of ground- and space-based telescopes to test this hypothesis.
 
@@
 %Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
-%\todo[inline]{KEVIN: describe stellar population}
-\todo[inline]{KEVIN: update reference to my paper once accepted}
 The original \bioverse\ stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function.

### 2023-05-10 src/tex/ms.tex
edit stellar catalog section

@@
 
-\subsection{Stellar sample from Gaia DR3}
+\subsection{Stellar sample from \gaia\ DR3}
 %Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
 The original \bioverse\ stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function.
-Exquisite parallax and photometric data from the \textit{Gaia} mission made it possible to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \bioverse ~\citep{Hardegree-Ullman2023}.
-Here, we briefly describe how stellar parameters $T_{\mathrm{eff}}$, $L_{\star}$, $R_{\star}$, and $M_{\star}$ were derived.
+Improved parallax and photometric data from the \gaia\ mission made it possible to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \bioverse ~\citep{Hardegree-Ullman2023}.
+Here, we briefly describe how we derived the stellar effective temperature $T_{\mathrm{eff}}$, luminosity $L_{\star}$, stellar radius $R_{\star}$, and stellar mass $M_{\star}$.
 
-\citet{Hardegree-Ullman2023} used the Gaia Catalogue of Nearby Stars~\citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
-The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2022} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
-From this information, colors and absolute magnitudes were computed, and initial color-magnitude cuts were made to remove non-main sequence stars.
-Effective temperatures were computed primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
-Next, luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
-Stellar radii were then computed with the effective temperatures and luminosities using the Stefan-Boltzmann law, or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M dwarfs.
-Finally, masses were derived from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
+\citet{Hardegree-Ullman2023} used the \gaia\ Catalogue of Nearby Stars~\citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
+The GCNS identified stars out to 120~pc and includes \gaia\ DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2022} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
+From this information, we computed colors and absolute magnitudes and applied initial color-magnitude cuts to remove non-main sequence stars.
+We derived effective temperatures primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
+Luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
+We computed stellar radii with the effective temperatures and luminosities using the Stefan-Boltzmann law or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M~dwarfs.
+Finally, we derived masses from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
 The derived stellar parameters from \citet{Hardegree-Ullman2023} were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.

### 2023-05-10 src/tex/ms.tex
replace the term "super-"runaway by runaway

@@
 The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}.
-With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state~\citep{Lichtenberg2021c,Boukrouche2021}.
+With even a fraction of the Earth's water inventory, a runaway planet will be in a global magma ocean state~\citep{Lichtenberg2021c,Boukrouche2021}.
 We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean~\citep{Dorn2021,Salvador2023}.
@@
 
-For \textit{super-runaway} planets, i.e., all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
+For \textit{runaway} planets, i.e., all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
 While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
@@
 
-We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population.
+We currently see four potential lines of discriminating atmospheric signatures associated with runaway climates in the exoplanet population.
 (i) Detecting the atmospheric windows of water vapor in the near- to mid-infrared.
-In a super-runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
+In a runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
 This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as \life\ ~\citep{2019A&A...621A.125B,Quanz2022,2022A&A...664A..22D}.
-Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime~\citep{2020SSRv..216...98G}.
+Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this runaway regime~\citep{2020SSRv..216...98G}.
 (ii) Post-runaway planets may be detectable by O$_{3}$ absorption due to build-up of abiotic oxygen, left-over from photochemical dissociation, and hydrogen loss~\citep{2014ApJ...785L..20W,Luger2015}.
@@
 This may be detectable in high-resolution observations focusing on isotope-sensitive transitions~\citep{2019AJ....158...26L,2019A&A...622A.139M}.
-(iv) Disequilibrium chemistry in tidally-locked super-runaway planets.
+(iv) Disequilibrium chemistry in tidally-locked runaway planets.
 It has been suggested that the atmospheric depth and the presence or absence of oceans on ``sub-Neptunes'' could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers~\citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface~\citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
@@
 Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios~\citep{Alei2022,Konrad2022}.
-Since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
+Since the peak thermal emission in runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
 Finally, mid-infrared surveys such as \life\ show a preference for M~star planets~\citep{Quanz2022}, which is beneficial for detecting the predicted habitable zone inner edge discontinuity (see Sect.~\ref{sec:results_FGK_M}).
@@
 
-Future self-consistent modeling of interior-atmosphere interactions may include constraints on additional radius increases due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,Lichtenberg2021c,2022PSJ.....3...93B,2021SSRv..217...22G}.
+Future self-consistent modeling of interior-atmosphere interactions may include constraints on additional radius increases due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,Lichtenberg2021c,2022PSJ.....3...93B,2021SSRv..217...22G}.
 

### 2023-05-10 src/tex/ms.tex
replace the term "sub-"runaway by non-runaway

@@
 
-\textit{Sub-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
+\textit{Non-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
 This case serves as our null hypothesis.
@@
 % Earth: ~0.0002 PERCENT or 218 ppm by mass (Turbet+2019)
-In the following, we only distinguish between the sub-runaway greenhouse and wet melt scenarios.
+In the following, we only distinguish between the non-runaway greenhouse and wet melt scenarios.
 
@@
 It represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
-Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the sub-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
+Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the non-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
 The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere.
@@
 The habitable zone inner edge discontinuity may be affected by several processes that are challenging to quantify:
-Planets that lack an atmosphere or sufficient volatiles cannot bear steam atmospheres, and those that do eventually move to the sub-runaway greenhouse category due to desiccation~\citep[][]{Watson1981} or evolution of their host star~\citep{Luger2015}.
+Planets that lack an atmosphere or sufficient volatiles cannot bear steam atmospheres, and those that do eventually move to the non-runaway greenhouse category due to desiccation~\citep[][]{Watson1981} or evolution of their host star~\citep{Luger2015}.
 Hydrogen/Helium-dominated planets may disguise as inflated rocky planets and not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022,Innes2023}.

### 2023-05-10 src/tex/ms.tex
implement a lot of comments

@@
 The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}.
-With even a fraction of the Earth's water inventory, a runaway planet will be in a global magma ocean state~\citep{Lichtenberg2021c,Boukrouche2021}.
-We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean~\citep{Dorn2021,Salvador2023}.
+With even a fraction of the Earth's water inventory, a planet absorbing more flux than the radiation limit of steam atmospheres will enter a runaway greenhouse state resulting in a global magma ocean~\citep{Lichtenberg2021c,Boukrouche2021}.
+We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the distribution of water between planetary interior and atmosphere, and on the resulting thermal atmospheric structure~\citep{Dorn2021,Salvador2023}.
 
-The net absorbed stellar fluxes of planets are sensitive on their albedos, which are generally poorly constrained for planets outside the solar system~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019}.
+The net absorbed stellar fluxes of planets are sensitive to their albedo, which is generally poorly constrained for planets outside the solar system~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019}.
 Here, we assume global redistribution of incoming flux and a fixed Bond albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
@@
 Nominally, the above models assume a \textit{dry melt} without dissolved volatiles.
-Here, however, we consider a \textit{wet melt} magma ocean and take into account a radius decrease from water partitioning in the melt, which is expected in the situation of a magma ocean below a steam atmosphere~\citep{Dorn2021}.
-The magnitude of this effect depends on the planet's mass and water content and is generally small compared to the radius inflation from a steam atmosphere.
-We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
-We then add the (in almost all cases negative) radius deviations to the planet radii computed for the dry melt case.
+Here, however, we consider a \textit{wet melt} magma ocean and take into account a radius decrease from retention of water in the melt~\citep{Dorn2021}.
+The impact of the water distribution between melt and atmosphere on the change of the transit radius depends on the planet's mass and water content and is generally small compared to the radius inflation from the steam atmosphere.
+%We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
+We computed radius deviations between a wet magma ocean and a solid mantle following \citet{Dorn2021} assuming a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
+We then added the (in almost all cases negative) radius deviations to the inflated planet radii computed for the dry melt case.
 
@@
 \end{figure}
-We illustrate the mass-radius relations of the three cases in Figure~\ref{fig:radiusevolution} where a fiducial water mass fraction of $x_{H_2O}= \var{wrr}$ is assumed.
+We illustrate the mass-radius relations of the three cases in Figure~\ref{fig:radiusevolution} where a fiducial bulk water mass fraction of $x_{H_2O}= \var{wrr}$ is assumed.
 % Earth: ~0.0002 PERCENT or 218 ppm by mass (Turbet+2019)
@@
 
-To account for planets unable to sustain a steam atmosphere over extended time spans, as well as evolutionary effects such as desiccation (see bottem panel of Figure~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
-It represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
+To account for planets unable to sustain a steam atmosphere over extended time spans, as well as evolutionary effects such as desiccation through water photodissociation and H~escape (see bottem panel of Figure~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
+It represents the fraction of planets above the instellation threshold whose atmospheres are currently inflated due to a runaway greenhouse climate.
 Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the non-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
@@
 Under real-world conditions, the planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
-%\subsubsection{PLATO}
+%\subsubsection{\plato\}
 \plato's definition study report~\citep{plato2017} states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity~(RV) follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).
@@
         \caption{
-        Detection of the runaway greenhouse threshold in the optimistic case.
+        Detection of the runaway greenhouse threshold.
         From simulated radius and instellation measurements of a large ($N \approx \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
@@
 Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N \approx \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
-Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
+Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is $f_\mathrm{rgh} = \frgh$, and we chose a water mass fraction of $x_{H_2O} = \var{wrr}$ for each planet.
 In this case, the habitable zone inner edge discontinuity was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).
@@
 \textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies on sufficient amounts of atmospheric water vapor that can act as a greenhouse gas.
-However, already about \SIrange{\sim 10}{20}~bar of water vapor -- corresponding to a minor fraction of one Earth ocean and thus the lower limit of water on Earth -- is enough to sustain sufficiently high surface temperatures to keep the planet in a magma ocean stage~\citep{Boukrouche2021,Lichtenberg2021c}.
+However, already about $\SIrange{\sim 10}{20} \, \mathrm{bar}$ of water vapor -- corresponding to a minor fraction of one Earth ocean and thus the lower limit of water on Earth -- is enough to sustain sufficiently high surface temperatures to keep the planet in a magma ocean stage~\citep{Boukrouche2021,Lichtenberg2021c}.
 The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
-From a planet formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard outcome~\citep{2019PNAS..116.9723Z,Venturini2020}.
+From a planet formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard expected outcome~\citep[e.g.,][]{2019PNAS..116.9723Z,Venturini2020,Emsenhuber2021b,Schlecker2021,Burn2021}.
 But while commonly considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the solar system~\citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation.
@@
 In addition, the sample must include planets on both sides of the instellation threshold.
-Due to the strongly distance-dependent detection bias associated with the transit method and the temporal coverage of upcoming transit missions, this is only likely for low-mass host stars.
+This is only likely for low-mass host stars due to the strongly distance-dependent detection bias associated with the transit method and the temporal coverage of upcoming transit missions.
 
@@
 (i) Detecting the atmospheric windows of water vapor in the near- to mid-infrared.
-In a runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
-This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as \life\ ~\citep{2019A&A...621A.125B,Quanz2022,2022A&A...664A..22D}.
-Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this runaway regime~\citep{2020SSRv..216...98G}.
-(ii) Post-runaway planets may be detectable by O$_{3}$ absorption due to build-up of abiotic oxygen, left-over from photochemical dissociation, and hydrogen loss~\citep{2014ApJ...785L..20W,Luger2015}.
-(iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
+In a runaway greenhouse atmosphere, absorption is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
+Probing these features requires an instrument covering these wavelength ranges, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as \life\ ~\citep{2019A&A...621A.125B,Quanz2022,2022A&A...664A..22D}.
+Potentially the two-band filter capabilities of \plato\ may offer insight into particularly pronounced spectral features in this runaway greenhouse regime~\citep{2020SSRv..216...98G}.
+(ii) Post-runaway planets may be detectable through O$_{3}$ absorption due to build-up of abiotic oxygen, leftover from photochemical dissociation of water, and hydrogen loss~\citep{2014ApJ...785L..20W,Luger2015}.
+(iii) Water loss in runaway greenhouse episodes would increase the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
 This may be detectable in high-resolution observations focusing on isotope-sensitive transitions~\citep{2019AJ....158...26L,2019A&A...622A.139M}.
@@
 It has been suggested that the atmospheric depth and the presence or absence of oceans on ``sub-Neptunes'' could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers~\citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface~\citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
-Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma~\citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
+Atmospheric nitrogen and carbon compounds can be partitioned into magma~\citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
 Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed.
@@
 We found that along these axes, \plato\ will be the most favorable among the upcoming transit missions.
-The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission.% (Matuszewski, priv. comm.).
+The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission. % (Matuszewski, priv. comm.).
 %In the FGK sample of the mission, several hundred Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
@@
 Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios~\citep{Alei2022,Konrad2022}.
-Since the peak thermal emission in runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
+Since the peak thermal emission in runaway greenhouse atmospheres will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
 Finally, mid-infrared surveys such as \life\ show a preference for M~star planets~\citep{Quanz2022}, which is beneficial for detecting the predicted habitable zone inner edge discontinuity (see Sect.~\ref{sec:results_FGK_M}).
@@
 
-Future self-consistent modeling of interior-atmosphere interactions may include constraints on additional radius increases due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,Lichtenberg2021c,2022PSJ.....3...93B,2021SSRv..217...22G}.
+Future self-consistent modeling of interior-atmosphere interactions may include constraints on additional radius increases due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,Lichtenberg2021c,2022PSJ.....3...93B,2021SSRv..217...22G} or water outgassing efficiency~\citep[e.g.,][]{Hier-Majumder2017,Ikoma2018,Salvador2023}.
 
@@
 Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021,2022ApJ...938L...3L,2022ApJ...939L..19I}.
-Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
+Our nominal case assumes a bulk water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
 This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the habitable zone inner edge discontinuity.
-%, and the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
+%, and the hypothesis tests are only marginally sensitive to $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
 Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
@@
 The habitable zone inner edge discontinuity may be affected by several processes that are challenging to quantify:
-Planets that lack an atmosphere or sufficient volatiles cannot bear steam atmospheres, and those that do eventually move to the non-runaway greenhouse category due to desiccation~\citep[][]{Watson1981} or evolution of their host star~\citep{Luger2015}.
+Planets that lack an atmosphere, sufficient volatiles, or have non-water-dominated outgassed compositions cannot bear steam atmospheres, and those that do eventually move to the non-runaway greenhouse category due to desiccation~\citep[][]{Watson1981,Kasting1983,Hamano2013} or evolution of their host star~\citep{Luger2015}.
 Hydrogen/Helium-dominated planets may disguise as inflated rocky planets and not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022,Innes2023}.
 A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, reducing their transit radius.
-Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Turbet2021,Lichtenberg2021c,Pierrehumbert2022,Innes2023}.
+Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Salvador2017,Turbet2021,Lichtenberg2021c,Pierrehumbert2022,Innes2023}.
 The choice of a statistical estimator for the hypothesis tests may further influence the recovered discontinuity; we compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.
@@
     \item This habitable zone inner edge discontinuity should be detectable with high-precision transit measurements.
-          For a planet sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.
+          For a planet sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical bulk water mass fractions are above $\sim 10^{-3}$.
     \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to confirm or reject the predicted trend.

### 2023-05-11 src/tex/ms.tex
add minor edits

@@
 %We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
-We computed radius deviations between a wet magma ocean and a solid mantle following \citet{Dorn2021} assuming a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
+We computed radius deviations between a wet magma ocean and a solid mantle following \citet{Dorn2021} and assuming a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
 We then added the (in almost all cases negative) radius deviations to the inflated planet radii computed for the dry melt case.
@@
 
-\subsubsection{Water mass fraction}
+\subsubsection{Bulk water mass fraction}
 The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.

### 2023-05-11 src/tex/ms.tex
fix a reference
...and save state of notebook

@@
 Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp instellation threshold~\citep{Goldblatt2013,Leconte2013,Kopparapu2013}.
-Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone~\citep[e.g.,][]{Ramirez2018b,Salvador2023b}.
+Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone~\citep[e.g.,][]{Ramirez2018,Salvador2023b}.
 Its prevalent definition refers to the possibility of a sustained liquid water body on the surface of an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmosphere~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014};

### 2023-05-11 src/tex/ms.tex
remove comments, switch todos off

@@
 \section{Introduction}
-%\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021}.}
-
 Despite recent advancements in observational techniques, our understanding of terrestrial-sized planets remains woefully limited, with fundamental aspects of their nature, composition, and potential habitability still largely unknown.
@@
 \subsection{Stellar sample from \gaia\ DR3}
-%Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
 The original \bioverse\ stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function.
@@
 Under real-world conditions, the planetary properties in question can only be probed with a finite precision that is specific to each exoplanet mission.
-%\subsubsection{\plato\}
 \plato's definition study report~\citep{plato2017} states precision requirements for planet radii~(\SI{3}{\percent}), planet masses through radial velocity~(RV) follow-up~(\SI{10}{\percent}), and stellar masses, radii, and ages~(\SI{10}{\percent}).

### 2023-05-16 src/scripts/hzied_pipeline.py
implement figure script for model-population-comparison

@@ -109,6 +109,8 @@ stars_args, planets_args = get_generator_args()
 
 with open(paths.data / 'planets_args.pkl', 'wb') as file:
     pickle.dump(planets_args, file)
+with open(paths.data / 'stars_args.pkl', 'wb') as file:
+    pickle.dump(stars_args, file)
 
 sample = generate_sample()
 with open(paths.data / 'sample.pkl', 'wb') as file:

### 2023-05-16 src/scripts/model_pop_comparison.py

@@ -1,0 +1,126 @@
+## compare avg delta R/rho with expected from atmosphere models
+from scipy.interpolate import LinearNDInterpolator
+
+import paths
+import pickle
+import numpy as np
+import pandas as pd
+import seaborn as sns
+
+from matplotlib import pyplot as plt
+import plotstyle
+plotstyle.styleplots()
+
+from bioverse.hypothesis import get_avg_deltaR_deltaRho, compute_avg_deltaR_deltaRho
+from bioverse.constants import DATA_DIR, CONST
+from cycler import cycler
+
+def prepare_data():
+    # Read M-R relations from Turbet+2020
+    purerock = pd.read_csv(DATA_DIR + 'mass-radius_relationships_mgsio3_Zeng2016.txt')
+    purerock.loc[:, 'wrr'] = 0.
+    turbet2020 = pd.read_csv(DATA_DIR + 'mass-radius_relationships_STEAM_TURBET2020_FIG2b.dat', comment='#')
+    mass_radius = purerock.append(turbet2020, ignore_index=True)
+
+    # Read radius differences from DL21 Fig. 3b
+    delta_R = pd.read_csv(DATA_DIR + 'deltaR_DornLichtenberg21_Fig3b.csv', comment='#')
+    delta_R.set_index('wrr', inplace=True)
+
+    # Combine Turbet+2020 and DL21 radius diffs:
+    # interpolate within planet masses for the given water mass fraction wrr
+    x = delta_R.index.to_numpy()
+    y = np.array([float(c) for c in delta_R.columns])
+    z = delta_R.to_numpy().flatten()
+    grid = []
+    for xi in x:
+        for yi in y:
+            grid.append((xi, yi))
+    grid= np.array(grid)
+    interp = LinearNDInterpolator(list(grid), z)
+
+    mass_radius.loc[:,'radius_tot'] = mass_radius.radius + interp(mass_radius.wrr, mass_radius.mass)
+    # plt.scatter(mass_radius.radius, mass_radius.radius_tot)
+
+    # Read radius differences as measured in synthetic population
+    # avg_deltaR_deltaRho = get_avg_deltaR_deltaRho()
+    with open(paths.data / 'stars_args.pkl', 'rb') as f:
+        stars_args = pickle.load(f)
+    with open(paths.data / 'planets_args.pkl', 'rb') as f:
+        planets_args = pickle.load(f)
+
+    try:
[…]

### 2023-05-16 src/scripts/HnullHmo.py
implement figure script for HnullHmo

@@ -1,0 +1,237 @@
+""" Plot sub- and super-runaway planets in the sample and the corresponding hypotheses in instellation-radius space."""
+
+import pickle
+import paths
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.ticker import ScalarFormatter
+from bioverse.hypothesis import magma_ocean_hypo, magma_ocean_f0
+from bioverse.util import S2a_eff, a_eff2S
+from bioverse.constants import CONST
+
+import plotstyle
+plotstyle.styleplots()
+
+
+def plot_transitingplanets(sample, planets_args, ax, parameter_of_interest='R'):
+    """ Make a scatter plot highlighting the transiting planets.
+
+    Parameters
+    ----------
+    sample : Bioverse Table or pandas DataFrame
+        Table with the synthetic sample
+    planets_args : dict
+        arguments for planet sample creation with Bioverse
+    ax : matplotlib axis object
+    parameter_of_interest : str
+        y-axis parameter to plot. Either 'R' or 'rho'.
+
+    Returns
+    -------
+    ax : matplotlib axis object
+        the axis with the plot
+    """
+    yvar = parameter_of_interest
+
+    try:
+        sampledf = sample.to_pandas()
+    except AttributeError:
+        sampledf = sample
+    ax.scatter(sampledf[sampledf.transiting == False].S_abs, sampledf[sampledf.transiting == False][yvar], s=.3,
+               c='dimgray', alpha=.5, label='synthetic\nplanets')
+
+    # transiting planets
+    # sc = ax.scatter(sampledf[sampledf.transiting == True].S_abs, sampledf[sampledf.transiting == True][yvar], s=20,
+    #            marker='x', c=np.where(sampledf[sampledf.transiting == True].has_magmaocean, 'C2', 'C0'), label='transiting')
+    ax.scatter(sampledf[(sampledf.transiting) & ~(sampledf.has_magmaocean)].S_abs, sampledf[(sampledf.transiting) & ~(sampledf.has_magmaocean)][yvar], s=50,
+                    marker='X', c='C0', edgecolors='k', linewidth=1.25, label='transiting')
[…]

### 2023-05-16 src/scripts/cornerplot.py
implement figure script for cornerplot

@@ -1,0 +1,55 @@
+"""Show corner plot of parameter inference example"""
+
+import pickle
+import corner
+import paths
+import numpy as np
+import pandas as pd
+from bioverse.constants import CONST
+
+import plotstyle
+plotstyle.styleplots()
+
+def cornerplot(results, params, log, planets_args, parameter_of_interest, sample):
+    # sampler_results into DataFrame, make logged columns for params with log priors
+    chains = pd.DataFrame(results['chains'], columns=params)
+    for p, l in zip(params, log):
+        if l:
+            chains['log_' + p] = np.log10(chains[p])
+
+    # compute truths
+    if parameter_of_interest == 'R':
+        avg_out = np.average(sample['R'][(sample['S_abs'] < planets_args['S_thresh']) & sample['transiting']])
+    elif parameter_of_interest == 'rho':
+        avg_out = np.average \
+            (CONST['rho_Earth' ] *sample['M'][(sample['S_abs'] < planets_args['S_thresh']) & sample['transiting'] ]/
+                             (sample['R'][(sample['S_abs'] < planets_args['S_thresh']) & sample['transiting']] )**3)
+    truths =np.array([planets_args['S_thresh'], np.log10(planets_args['wrr']), planets_args['f_rgh'], avg_out])
+
+    fig = corner.corner(
+        chains[['S_thresh', 'log_wrr', 'f_rgh']], labels=['$S_{thresh}$', '$\log(x_{H_2O})$', '$f_\mathrm{rgh}$'],
+        truths=truths[:-1], truth_color='C1',
+        range=[(10 ,600), (-5, -1.), (-0.01, 1.)],
+        quantiles=[0.16, 0.5, 0.84],
+        show_titles=True, title_kwargs={"fontsize": 12},
+        hist_bin_factor=2.,
+        smooth=.1,
+        smooth1d=.1
+    )
+    return fig
+
+
+with open(paths.data / 'sample.pkl', 'rb') as f:
+    sample = pickle.load(f)
+with open(paths.data / 'results_opt.pkl', 'rb') as f:
+    results_opt = pickle.load(f)
+with open(paths.data / 'params.pkl', 'rb') as f:
+    params = pickle.load(f)
+with open(paths.data / 'log.pkl', 'rb') as f:
+    log = pickle.load(f)
+with open(paths.data / 'planets_args.pkl', 'rb') as f:
+    planets_args = pickle.load(f)
+
[…]

### 2023-05-16 src/scripts/hzied_pipeline.py
implement figure script for cornerplot

@@ -7,6 +7,7 @@ import scipy
 import paths
 from utils import *
 import pickle
+import corner
 import numpy as np
 import pandas as pd
 import seaborn as sns
@@ -18,6 +19,7 @@ from scipy.signal import savgol_filter
 from bioverse.generator import Generator
 from bioverse.survey import TransitSurvey
 from bioverse.constants import CONST, DATA_DIR
+from bioverse.hypothesis import Hypothesis, magma_ocean_hypo, magma_ocean_f0, get_avg_deltaR_deltaRho
 
 # Sample generation
 ## Generate stars and planets, inject magma oceans
@@ -105,6 +107,142 @@ def generate_sample():
     # print('Total number of planets: {}'.format(len(sample)))
     return sample
 
+
+def survey_simulation():
+    """Simulate an ambitious survey with optimistic parameters."""
+    # create a transit survey (telescope parameters don't really matter here)
+    survey = TransitSurvey(diameter=8.5, t_max=3652.5, t_slew=0.1, N_obs_max=1000)
+
+    # Let's add some measurements our survey conducts.
+    margs = {}
+    mkeys = ['L_st', 'R_st', 'M_st', 'T_eff_st', 'd', 'age', 'depth', 'R',
+             'T_dur', 'P', 'a', 'a_eff', 'S', 'S_abs',
+             'M', 'rho']
+
+    # Optimistic survey: choose optimistic values
+    margs['precision'] = {'T_eff_st': 25.,
+                          'R_st': '1%',
+                          'depth': '1%',
+                          'R': '2%',
+                          'M_st': '1%',
+                          'age': '30%',
+                          'P': 0.000001,
+                          'S': '5%',
+                          'S_abs': '5%',
+                          'M': '5%',
+                          'rho': '7%'}
+
+    # Add the measurements to the survey
+    for mkey in mkeys:
+        kwargs = {}
+        for key, vals in margs.items():
+            if mkey in vals:
+                kwargs[key] = vals[mkey]
+        survey.add_measurement(mkey, **kwargs)
+
+    # which planets are detectable?
+    detected_opt = survey.compute_yield(sample)
+    N = len(detected_opt)
+    # save_var_latex('N_optimistic', '{:.0f}'.format(N))
+    save_var_latex('N_optimistic', '500') # round that to avoid confusion
+
+    # simulate observations, obtain dataset
+    data = survey.observe(detected_opt, demographics=True)
+    return detected_opt, data
+
+
[…]

### 2023-05-16 src/scripts/hzied_pipeline.py
implement figure script for optimistic_R-S

@@ -226,7 +226,7 @@ def hypothesis_tests(parameter_of_interest='R'):
     features = ('a_eff',)
     labels = (parameter_of_interest + '_mean',)
     log = (False, True, False, False)
-    nburn = 100
+
 
     # define PRIORS for the parameters in theta uniform for 'S_thresh', log-uniform for 'wrr', uniform for 'f_rgh', 'R_avg')
     bounds_R = np.array([[10., 1000.0], [1e-5, 0.1], [0.0, 1.0], [.1, 15.]])

### 2023-05-16 src/scripts/optimistic_R-S.py
implement figure script for optimistic_R-S

@@ -1,0 +1,99 @@
+"""Plot prototypical detection in an optimistic survey"""
+
+import pickle
+import paths
+import numpy as np
+import matplotlib.pyplot as plt
+
+import plotstyle
+plotstyle.styleplots()
+
+from bioverse.util import S2a_eff
+from bioverse.hypothesis import magma_ocean_hypo
+
+def plot_survey(data, results, planets_args, parameter_of_interest='R', show_rolling_mean=True, show_binned_stats=False):
+    yvar = parameter_of_interest
+    xvar = 'S_abs'
+    # yvar = 'R'
+
+    fig, ax = plt.subplots()
+    ax.scatter(data[xvar], data[yvar], s=2, c='k')
+    ax.errorbar(data[xvar], data[yvar], xerr=data.error[xvar], yerr=data.error[yvar], fmt='o', ms=2., elinewidth=1,
+                color='k', alpha=.5, label='measured')
+
+    if show_binned_stats:
+        # means, edges, n, std = binned_stats(data, xvar, yvar, np.geomspace(np.min(data[xvar]),
+        # np.max(data[xvar]), num=12), statistic='mean')
+        # ax.errorbar(edges[:-1]+(edges[1:]-edges[:-1])/2, means, xerr=(edges[1:]-edges[:-1])/2,
+        # yerr=None, fmt='none', color='C2', label='binned {} statistic'.format(yvar), elinewidth=2.5)
+        binned_avg = data[yvar + '_mean_binned']
+        ax.plot(data[xvar], binned_avg, color='C2', label='binned average')
+        std = data.error[yvar + '_mean_binned']
+        ax.fill_between(data.sort_by('S_abs')['S_abs'], (binned_avg - std),
+                        (binned_avg + std), color='C2', alpha=.3)
+
+    # overplot samples from the posterior
+    nburn = 100
+    sampler_results = results['sampler_results']
+    posterior_sample = sampler_results.samples[nburn::300]
+
+    S_grid = np.geomspace(10., 2000., 300)
+    a_eff_grid = S2a_eff(S_grid)
+
+    colors = plt.cm.OrRd(np.linspace(0, 1, len(posterior_sample)))
+    for i, s in enumerate(posterior_sample):
+        P_magma = magma_ocean_hypo((s[0], s[1], s[2], s[3]), a_eff_grid,
[…]

### 2023-05-16 src/scripts/cornerplot.py
debug dependencies

@@ -39,15 +39,15 @@ def cornerplot(results, params, log, planets_args, parameter_of_interest, sample
     return fig
 
 
-with open(paths.data / 'sample.pkl', 'rb') as f:
+with open(paths.data / 'pipeline/sample.pkl', 'rb') as f:
     sample = pickle.load(f)
-with open(paths.data / 'results_opt.pkl', 'rb') as f:
+with open(paths.data / 'pipeline/results_opt.pkl', 'rb') as f:
     results_opt = pickle.load(f)
-with open(paths.data / 'params.pkl', 'rb') as f:
+with open(paths.data / 'pipeline/params.pkl', 'rb') as f:
     params = pickle.load(f)
-with open(paths.data / 'log.pkl', 'rb') as f:
+with open(paths.data / 'pipeline/log.pkl', 'rb') as f:
     log = pickle.load(f)
-with open(paths.data / 'planets_args.pkl', 'rb') as f:
+with open(paths.data / 'pipeline/planets_args.pkl', 'rb') as f:
     planets_args = pickle.load(f)
 
 fig = cornerplot(results_opt, params, log, planets_args, parameter_of_interest='R', sample=sample)

### 2023-05-16 src/scripts/hzied_pipeline.py

@@ -3,17 +3,16 @@ This file contains the code to generate the datasets for the paper.
 It heavily relies on Bioverse and its auxiliary functions.
 """
 
+import os
 import scipy
 import paths
 from utils import *
 import pickle
-import corner
 import numpy as np
 import pandas as pd
 import seaborn as sns
 
 from scipy.interpolate import interp1d
-from scipy.signal import savgol_filter
 
 # Import the Generator class
 from bioverse.generator import Generator
@@ -245,27 +244,30 @@ def hypothesis_tests(parameter_of_interest='R'):
 
 stars_args, planets_args = get_generator_args()
 
-with open(paths.data / 'planets_args.pkl', 'wb') as file:
+# Make sure pipeline directory exists
+filename = paths.data / 'pipeline/planets_args.pkl'
+os.makedirs(os.path.dirname(filename), exist_ok=True)
+with open(paths.data / 'pipeline/planets_args.pkl', 'wb') as file:
     pickle.dump(planets_args, file)
-with open(paths.data / 'stars_args.pkl', 'wb') as file:
+with open(paths.data / 'pipeline/stars_args.pkl', 'wb') as file:
     pickle.dump(stars_args, file)
 
 sample = generate_sample()
-with open(paths.data / 'sample.pkl', 'wb') as file:
+with open(paths.data / 'pipeline/sample.pkl', 'wb') as file:
     pickle.dump(sample, file)
 
 detected_opt, data = survey_simulation()
-with open(paths.data / 'data.pkl', 'wb') as file:
+with open(paths.data / 'pipeline/data.pkl', 'wb') as file:
     pickle.dump(data, file)
 
 params, features, log, results_opt, h_magmaocean = hypothesis_tests()
-with open(paths.data / 'params.pkl', 'wb') as file:
+with open(paths.data / 'pipeline/params.pkl', 'wb') as file:
     pickle.dump(params, file)
-with open(paths.data / 'features.pkl', 'wb') as file:
+with open(paths.data / 'pipeline/features.pkl', 'wb') as file:
     pickle.dump(features, file)
-with open(paths.data / 'log.pkl', 'wb') as file:
+with open(paths.data / 'pipeline/log.pkl', 'wb') as file:
     pickle.dump(log, file)
-with open(paths.data / 'results_opt.pkl', 'wb') as file:
+with open(paths.data / 'pipeline/results_opt.pkl', 'wb') as file:
[…]

### 2023-05-16 src/scripts/optimistic_R-S.py

@@ -87,11 +87,11 @@ def plot_survey(data, results, planets_args, parameter_of_interest='R', show_rol
 
 
 
-with open(paths.data / 'data.pkl', 'rb') as f:
+with open(paths.data / 'pipeline/data.pkl', 'rb') as f:
     data = pickle.load(f)
-with open(paths.data / 'results_opt.pkl', 'rb') as f:
+with open(paths.data / 'pipeline/results_opt.pkl', 'rb') as f:
     results_opt = pickle.load(f)
-with open(paths.data / 'planets_args.pkl', 'rb') as f:
+with open(paths.data / 'pipeline/planets_args.pkl', 'rb') as f:
     planets_args = pickle.load(f)
 
 

### 2023-05-16 src/scripts/model_pop_comparison.py
debug dependencies

@@ -43,9 +43,9 @@ def prepare_data():
 
     # Read radius differences as measured in synthetic population
     # avg_deltaR_deltaRho = get_avg_deltaR_deltaRho()
-    with open(paths.data / 'stars_args.pkl', 'rb') as f:
+    with open(paths.data / 'pipeline/stars_args.pkl', 'rb') as f:
         stars_args = pickle.load(f)
-    with open(paths.data / 'planets_args.pkl', 'rb') as f:
+    with open(paths.data / 'pipeline/planets_args.pkl', 'rb') as f:
         planets_args = pickle.load(f)
 
     try:

### 2023-05-16 src/scripts/HnullHmo.py
debug dependencies

@@ -229,9 +229,9 @@ def main(sample, planets_args, parameter_of_interest, f_dR=None):
     fig.savefig(paths.figures / "HnullHmo.pdf")
     return fig, (ax, ax2)
 
-with open(paths.data / 'planets_args.pkl', 'rb') as f:
+with open(paths.data / 'pipeline/planets_args.pkl', 'rb') as f:
     planets_args = pickle.load(f)
-with open(paths.data / 'sample.pkl', 'rb') as f:
+with open(paths.data / 'pipeline/sample.pkl', 'rb') as f:
     sample = pickle.load(f)
 
 main(sample, planets_args, parameter_of_interest='R', f_dR=None)

### 2023-05-16 src/scripts/radiusevolution.py

@@ -116,9 +116,9 @@ def plot_planet_evo(N_time=9, v_offset=0.004, interpolate=False, **kwargs):
 
     return fig, ax
 
-with open(paths.data / 'planets_args.pkl', 'rb') as f:
+with open(paths.data / 'pipeline/planets_args.pkl', 'rb') as f:
     planets_args = pickle.load(f)
-with open(paths.data / 'sample.pkl', 'rb') as f:
+with open(paths.data / 'pipeline/sample.pkl', 'rb') as f:
     sample = pickle.load(f)
 
 fig, ax = plot_radiuscomparison(sample, planets_args)

### 2023-05-16 src/scripts/hzied_pipeline.py
implement figure script for statistical power grid

@@ -104,7 +104,7 @@ def generate_sample():
 
     sample = g_transit.generate()
     # print('Total number of planets: {}'.format(len(sample)))
-    return sample
+    return sample, g_transit
 
 
 def survey_simulation():
@@ -147,7 +147,7 @@ def survey_simulation():
 
     # simulate observations, obtain dataset
     data = survey.observe(detected_opt, demographics=True)
-    return detected_opt, data
+    return detected_opt, data, survey
 
 
 def hypotest(data, parameter_of_interest, params, log, bounds, bounds_null, features=('a_eff',), binned=False,
@@ -252,13 +252,17 @@ with open(paths.data / 'pipeline/planets_args.pkl', 'wb') as file:
 with open(paths.data / 'pipeline/stars_args.pkl', 'wb') as file:
     pickle.dump(stars_args, file)
 
-sample = generate_sample()
+sample, g_transit = generate_sample()
 with open(paths.data / 'pipeline/sample.pkl', 'wb') as file:
     pickle.dump(sample, file)
+with open(paths.data / 'pipeline/g_transit.pkl', 'wb') as file:
+    pickle.dump(g_transit, file)
 
-detected_opt, data = survey_simulation()
+detected_opt, data, survey = survey_simulation()
 with open(paths.data / 'pipeline/data.pkl', 'wb') as file:
     pickle.dump(data, file)
+with open(paths.data / 'pipeline/survey.pkl', 'wb') as file:
+    pickle.dump(survey, file)
 
 params, features, log, results_opt, h_magmaocean = hypothesis_tests()
 with open(paths.data / 'pipeline/params.pkl', 'wb') as file:

### 2023-05-16 src/scripts/optimistic_statpwr_H2O-f.py

@@ -1,0 +1,74 @@
+import paths
+import pickle
+import numpy as np
+import cmocean
+import matplotlib.pyplot as plt
+
+import plotstyle
+plotstyle.styleplots()
+
+from bioverse import analysis, plots
+
+def statistical_power_grid():
+    wrr_grid = [0., 0.0001,0.001, 0.005, 0.01, 0.02, 0.03, 0.04, 0.05]
+    # wrr_grid = [0.0001, 0.005, 0.05]
+    f_rgh_grid = np.linspace(0., 1., len(wrr_grid))
+
+    reduced_args = {key: planets_args[key] for key in planets_args if (key != 'wrr') & (key != 'f_rgh')} # remove keys used in the grid
+    # results_grid = analysis.test_hypothesis_grid(h_magmaocean, g_transit, survey, wrr=wrr_grid, f_rgh=f_rgh_grid, N=20, processes=8, **reduced_args)
+    results_grid = analysis.test_hypothesis_grid(h_magmaocean, g_transit, survey, wrr=wrr_grid, f_rgh=f_rgh_grid,
+                                                 N=20, processes=8, return_chains=True, **reduced_args)
+    return results_grid
+
+def load_grid():
+    """load previously computed statistical power grid"""
+    with open(paths.data / 'bioverse_objects/optimistic_H2O-f-grid_G16.pkl', 'rb') as f:
+        results_grid = pickle.load(f)
+    wrr_grid = [0., 0.0001,0.001, 0.005, 0.01, 0.02, 0.03, 0.04, 0.05]
+    f_rgh_grid = np.linspace(0., 1., len(wrr_grid))
+    return results_grid, wrr_grid, f_rgh_grid
+
+def plot_power_grid(results_grid):
+    fig, ax = plt.subplots(figsize=[6, 4])
+    labels = ('Water mass fraction $x_\mathrm{H_2O}$', 'Dilution factor $f_\mathrm{rgh}$')
+    fig, ax = plots.plot_power_grid(results_grid, method='dBIC', axes=('wrr', 'f_rgh'), labels=labels,
+                                    log=(True, False), show=False, fig=fig, ax=ax,
+                                    # zoom_factor=2, smooth_sigma=.5,
+                                    zoom_factor=0, smooth_sigma=.2,
+                                    levels=[50, 95], cmap=cmocean.cm.dense_r)
+
[…]

### 2023-05-17 src/scripts/plato_grids.py
implement figure script for PLATO grid

@@ -1,0 +1,140 @@
+import paths
+import pickle
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.interpolate import interp1d
+from scipy.signal import savgol_filter
+
+import plotstyle
+plotstyle.styleplots()
+
+from bioverse import analysis, plots
+
+
+def plot_plato_grids(f_grid, results, labels, yaxis='dlnZ', interpolate=False, plot_percentiles=False,
+                     ci_percentiles=(16., 84.), fig=None, ax=None, smooth=True, **plot_kwargs):
+    if ax is None:
+        fig, ax = plt.subplots()
+    if yaxis == 'dlnZ':
+        y_label = 'Log-evidence difference $ \Delta \ln Z$'
+        for i, (r, l) in enumerate(zip(results, labels)):
+            y = r['dlnZ']
+            # y_mean = np.median(y, axis=-1)
+            y_mean = np.median(y, axis=-1)
+            percentiles = np.percentile(y, ci_percentiles, axis=1)  # 5% and 95% percentiles
+            if interpolate:
+                f_grid_fine = np.linspace(min(f_grid), max(f_grid), num=60)
+                interpolate_grid = interp1d(f_grid, y_mean, kind='quadratic')
+                y_mean = interpolate_grid(f_grid_fine)
+                # p = fit_poly(f_grid, y_mean, 3)
+                # y_mean = p(f_grid_fine)
+                interpolate_p_lo = interp1d(f_grid, percentiles[0], kind='quadratic')
+                interpolate_p_up = interp1d(f_grid, percentiles[1], kind='quadratic')
+                percentiles = [interpolate_p_lo(f_grid_fine), interpolate_p_up(f_grid_fine)]
+                x = f_grid_fine
+            else:
+                x = f_grid
+
+            if smooth:
+                # y_mean = convolve(y_mean, Gaussian1DKernel(0.8)) #Box1DKernel(3, mode='integrate'))
+                y_mean = savgol_filter(y_mean, window_length=15, polyorder=2, mode='nearest')
+            p = ax.plot(x, y_mean, lw=3, label=l, **plot_kwargs)
+            if plot_percentiles:
+                lns2 = ax.fill_between(x,
+                                       percentiles[0],
[…]

### 2023-05-17 src/scripts/hzied_pipeline.py
implement figure script for PLATO grid

@@ -242,6 +242,8 @@ def hypothesis_tests(parameter_of_interest='R'):
     return params, features, log, results_opt, h_magmaocean
 
 
+
+
 stars_args, planets_args = get_generator_args()
 
 # Make sure pipeline directory exists
@@ -252,17 +254,20 @@ with open(paths.data / 'pipeline/planets_args.pkl', 'wb') as file:
 with open(paths.data / 'pipeline/stars_args.pkl', 'wb') as file:
     pickle.dump(stars_args, file)
 
+
 sample, g_transit = generate_sample()
 with open(paths.data / 'pipeline/sample.pkl', 'wb') as file:
     pickle.dump(sample, file)
 with open(paths.data / 'pipeline/g_transit.pkl', 'wb') as file:
     pickle.dump(g_transit, file)
 
+
 detected_opt, data, survey = survey_simulation()
 with open(paths.data / 'pipeline/data.pkl', 'wb') as file:
     pickle.dump(data, file)
 with open(paths.data / 'pipeline/survey.pkl', 'wb') as file:
     pickle.dump(survey, file)
+
 
 params, features, log, results_opt, h_magmaocean = hypothesis_tests()
 with open(paths.data / 'pipeline/params.pkl', 'wb') as file:
@@ -275,3 +280,5 @@ with open(paths.data / 'pipeline/results_opt.pkl', 'wb') as file:
     pickle.dump(results_opt, file)
 with open(paths.data / 'pipeline/h_magmaocean.pkl', 'wb') as file:
     pickle.dump(h_magmaocean, file)
+
+

### 2023-05-17 src/scripts/plato_Sthresh_grid.py
implement figure script for S_thresh grid

@@ -1,0 +1,184 @@
+import paths
+import pickle
+from warnings import warn
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.ticker import ScalarFormatter
+
+import plotstyle
+plotstyle.styleplots()
+
+
+def plot_posterior_hist(results_grid, fig=None, ax=None, truth=None, seed=42, param_idx=0, **grid_kwargs):
+    np.random.seed(seed)
+    try:
+        grid = results_grid['grid'].copy()
+        seq = [0] * len(grid)
+        for i, (key, val) in enumerate(results_grid['grid'].items()):
+            if key in grid_kwargs:
+                seq[i] = np.argmin(np.abs(val - grid_kwargs[key]))
+
+            elif key == 'N':
+                # Pick a random simulation unless specified
+                seq[i] = np.random.choice(grid['N'])
+            elif len(val) > 1:
+                warn("which value of '{:s}' should be plotted? assuming {:s} = {:.1f}".format(key, key, val[0]))
+
+            # Removes 'N' and other reduced grid dimensions
+            del grid[key]
+
+        # Extract the MCMC samples for this simulation
+        samples = results_grid['chains'][tuple(seq)]
+
+    except KeyError:
+        samples = results_grid['chains']
+
+    if fig is None:
+        fig, ax = plt.subplots(figsize=(12, 4))
+
+    if 'h' in results_grid:
+        xmin, xmax = results_grid['h'].bounds[param_idx]
+    else:
+        xmin, xmax = (np.min(samples[...,param_idx]), np.max(samples[...,param_idx]))
+
+
+    bins = np.logspace(np.log10(xmin), np.log10(xmax), 50)
+    # bins = np.linspace(0, 1, 50)
+    # ax.hist(samples[..., param_idx], histtype='step', bins=bins, lw=lw, color='black', density=True)
+    ax.hist(samples[..., param_idx], histtype='step', bins=bins, color='black', density=True)
+
+    if truth:
+        # Plot the truth value
+        ax.axvline(truth, lw=5, c='C0', alpha=0.75)
+
+    # Axes
+    # ax.set_xlim([0, 1])
+    ax.set_xscale('log')
+    ax.set_xlim([xmin, xmax])
+
+    ax.get_xaxis().set_major_formatter(ScalarFormatter())
+
[…]

### 2023-05-17 src/tex/ms.tex
change URL of repo
implement figure script for model-population-comparison
implement figure script for HnullHmo
implement figure script for cornerplot
implement figure script for optimistic_R-S
implement figure script for statistical power grid
implement figure script for PLATO grid
implement figure script for S_thresh grid
set N_optimistic=500

@@
 \begin{figure}
+    \script{radiusevolution.py}
     \begin{centering}
@@
 \begin{figure*}
-%    \script{figurescript.py}
+    \script{HnullHmo.py}
     \begin{centering}
@@
 \begin{figure}[ht!]
-%    \script{figurescript.py}
+    \script{optimistic_R-S.py}
     \begin{centering}
@@
         Detection of the runaway greenhouse threshold.
-        From simulated radius and instellation measurements of a large ($N \approx \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
+        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
             The pattern is detected with high significance.
@@
 Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the habitable zone inner edge discontinuity.
-Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N \approx \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
+Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
 Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is $f_\mathrm{rgh} = \frgh$, and we chose a water mass fraction of $x_{H_2O} = \var{wrr}$ for each planet.
@@
 \begin{figure}[ht!]
-%    \script{figurescript.py}
+    \script{cornerplot.py}
     \begin{centering}
@@
 \begin{figure}[ht!]
-%    \script{figurescript.py}
+    \script{optimistic_statpwr_H2O-f.py}
     \begin{centering}
@@
             Statistical power of the runaway greenhouse hypothesis test as a function of model parameters.
-            For a sample size $N \approx \var{N_optimistic}$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and dilution factor.
+            For a sample size $N = \var{N_optimistic}$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and dilution factor.
             Higher values in either parameter result in a more reliable detection.
@@
     \begin{figure*}[ht!]
-%    \script{figurescript.py}
+    \script{plato_grids.py}
         \begin{centering}
@@
 \begin{figure}[ht!]
-%    \script{figurescript.py}
+    \script{plato_Sthresh_grid.py}
     \begin{centering}
@@
             Detection of the runaway greenhouse threshold with a binned statistic.
-            Using simulated data from the optimistic survey case ($N \approx \var{N_optimistic}$, compare Figure~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis.
+            Using simulated data from the optimistic survey case ($N = \var{N_optimistic}$, compare Figure~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis.
 Instead of using a running mean, we computed a binned statistic (arithmetic mean and standard deviation in blue) to assign planets the average radius or bulk density in their neighborhood in instellation space.
@@
 \begin{figure*}[ht!]
-%    \script{figurescript.py}
+    \script{model_pop_comparison.py}
     \begin{centering}

### 2023-05-17 src/scripts/hzied_pipeline.py

@@ -31,7 +31,7 @@ def get_generator_args():
     """ define generator parameters"""
     # Parameters for star generation
     stars_args = {
-        'd_max' : 90,             # max. dist to stars (pc) ~~(nominal=90, ca. 481 planets)~~
+        'd_max' : 74,             # max. dist to stars (pc)
         'M_st_max': 1.5,  # Maximum stellar mass to consider, in solar units.
         'M_G_max': 16,  # Maximum gaia magnitude of stars
         'seed': 42,  # seed for random number generators
@@ -55,7 +55,7 @@ def get_generator_args():
 
         # detection bias and sample selection: Keep only small planets/low-mass planets, but above 0.1 M_Earth (Turbet+2020 covers only 0.1 to 2 Mearth). Don't consider "lava worlds" with extremely high instellation.
         # Set a generic detection bias: transit depth $\delta > 75$ ppm.
-        'M_min': 0.1,
+        'M_min': 0.32, # use this otherwise non-influental parameter to tune sample size
         'M_max': 2.,
         'S_min': 10.,  # min. instellation in W/m2
         'S_max': 2000.,  # max. instellation in W/m2

### 2023-05-17 src/scripts/optimistic_R-S.py

@@ -74,7 +74,7 @@ def plot_survey(data, results, planets_args, parameter_of_interest='R', show_rol
         ax.fill_between(data.sort_by('S_abs')['S_abs'], (mean - sem),
                         (mean + sem), color='C2', alpha=.3, label='moving average')
 
-    ax.set_title('Optimistic survey ({} planets)'.format(len(data)), y=1.15, fontsize=14)
+    ax.set_title('Optimistic survey ({} planets)'.format(500), y=1.15, fontsize=14)
     fig.legend(fontsize=12, loc='lower left', ncol=99, bbox_to_anchor=(0.15, .97),
                frameon=False, columnspacing=1.6)
 

### 2023-05-17 src/scripts/optimistic_RS_binned.py
implement figure script for optimistic_RS_binned

@@ -1,0 +1,26 @@
+"""Plot prototypical detection in an optimistic survey (binned version)"""
+import numpy as np
+import pickle
+import paths
+
+import plotstyle
+plotstyle.styleplots()
+
+from optimistic_RS import plot_survey
+from hzied_pipeline import hypotest
+
+with open(paths.data / 'pipeline/data.pkl', 'rb') as f:
+    data = pickle.load(f)
+with open(paths.data / 'pipeline/planets_args.pkl', 'rb') as f:
+    planets_args = pickle.load(f)
+
+params = ('S_thresh', 'wrr', 'f_rgh', 'avg')
+log = (False, True, False, False)
+bounds = np.array([[10., 1000.0], [1e-5, 0.1], [0.0, 1.0], [.1, 15.]])
+bounds_null = np.array([bounds[-1]])            # prior bounds for null hypothesis
+
+results_binned, h_magmaocean = hypotest(data, 'R', params, log, bounds, bounds_null, binned=True)  # this time, perform hypothesis tests on binned average R/rho
+fig, ax = plot_survey(data, results_binned, planets_args, parameter_of_interest='R', show_rolling_mean=False, show_binned_stats=True)
+ax.set_title('Optimistic survey (binned)', y=1.13)
+
+fig.savefig(paths.figures / 'optimistic_R-S_binned.pdf')

### 2023-05-17 src/scripts/MR_violins.py
implement figure script for M-R violins

@@ -1,0 +1,83 @@
+"""Explore inference with alternative mass-radius relations"""
+
+import pickle
+import paths
+import numpy as np
+import pandas as pd
+from matplotlib.transforms import blended_transform_factory
+
+import plotstyle
+plotstyle.styleplots()
+import seaborn as sns
+
+
+def plot_splitviolin(planets_args, res_fg_MR_earthlike, res_fg_MR_Wolfgang2016):
+    # sns.set_theme(style="whitegrid")
+    earthlike = pd.DataFrame(np.mean(res_fg_MR_earthlike['chains'][:, :, :, 0].T, axis=1),
+                             columns=res_fg_MR_earthlike['grid']['f_rgh'])
+    earthlike.loc[:, 'M-R relation'] = 'Earth-like'
+    wolfgang = pd.DataFrame(np.mean(res_fg_MR_Wolfgang2016['chains'][:, :, :, 0].T, axis=1),
+                            columns=res_fg_MR_Wolfgang2016['grid']['f_rgh'])
+    wolfgang = wolfgang[[c for c in wolfgang.columns if c in earthlike.columns]]
+
+    wolfgang.loc[:, 'M-R relation'] = 'Wolfgang et al. (2016)'
+    mr_df = pd.merge(left=earthlike, right=wolfgang, how='outer')
+
+    # workaround for a weird rounding error...
+    colnames = [np.round(c, decimals=1) for c in mr_df.columns[:-1]]
+    colnames.append('M-R relation')
+    mr_df.columns = colnames
+
+    mr = pd.melt(mr_df, id_vars=['M-R relation'], var_name='f_rgh', value_name='S_thresh')
+    mr = mr[mr['S_thresh'].notna()]
+
+    # Draw a nested violinplot and split the violins for easier comparison
+    # fig, ax = plt.subplots(figsize=[12,4])
+    ax = sns.violinplot(data=mr, x="f_rgh", y="S_thresh", hue="M-R relation",
+                        split=True,
+                        inner="quartile",
+                        # cut=0.5,
+                        # saturation=.8,
+                        linewidth=1.,
+                        palette={'Earth-like': 'C1', 'Wolfgang et al. (2016)': 'C0'})
+
+    # plot truths
[…]

### 2023-05-17 src/scripts/MR_violins.py
integrate Zenodo dataset with expensive Bioverse computations
https://doi.org/10.5281/zenodo.7946446

@@ -70,7 +70,8 @@ with open(paths.data / 'pipeline/planets_args.pkl', 'rb') as f:
     planets_args = pickle.load(f)
 
 # load previously computed hypothesis test grids
-hypothesisgrids = {'res_fg_MR_Wolfgang2016':'optimistic_MR_Wolfgang2016.pkl', 'res_fg_MR_earthlike':'optimistic_MR_earthlike.pkl',
+hypothesisgrids = {'res_fg_MR_Wolfgang2016':'optimistic_MR_Wolfgang2016.pkl',
+                   'res_fg_MR_earthlike':'optimistic_MR_earthlike.pkl',
                    'res_optimistic_H2O_fg':'optimistic_H2O-f-grid.pkl'}
 for varname, path in zip(hypothesisgrids.keys(),
                          [str(paths.data) + '/bioverse_objects/' + fname for fname in hypothesisgrids.values()]):

### 2023-05-17 src/scripts/plato_grids.py
integrate Zenodo dataset with expensive Bioverse computations
https://doi.org/10.5281/zenodo.7946446

@@ -65,10 +65,14 @@ def plot_plato_grids(f_grid, results, labels, yaxis='dlnZ', interpolate=False, p
 
 
 """load pre-computed hypothesis testing grids."""
-hypothesisgrids = {'res_fg_plato': 'plato_f-grid.pkl', 'res_fg_plato100' : 'plato100_f-grid.pkl',
-                   'res_fg_plato40' : 'plato40_f-grid.pkl', 'res_fg_plato_rho' : 'plato_rho_f-grid.pkl',
-                   'res_fg_plato_FGK_R':'plato_FGK_R.pkl', 'res_fg_plato_FGK_rho':'plato_FGK_rho.pkl', # FGK vs M
-                   'res_fg_plato_M_R':'plato_M_R.pkl', 'res_fg_plato_M_rho':'plato_M_rho.pkl',} # FGK vs M
+hypothesisgrids = {'res_fg_plato': 'plato_f-grid.pkl',
+                   'res_fg_plato100' : 'plato100_f-grid.pkl',
+                   'res_fg_plato40' : 'plato40_f-grid.pkl',
+                   'res_fg_plato_rho' : 'plato_rho_f-grid.pkl',
+                   'res_fg_plato_FGK_R':'plato_FGK_R.pkl',
+                   'res_fg_plato_FGK_rho':'plato_FGK_rho.pkl', # FGK vs M
+                   'res_fg_plato_M_R':'plato_M_R.pkl',
+                   'res_fg_plato_M_rho':'plato_M_rho.pkl',} # FGK vs M
 for varname, path in zip(hypothesisgrids.keys(),
                          [str(paths.data) + '/bioverse_objects/' + fname for fname in hypothesisgrids.values()]):
     with open(path, 'rb') as f:

### 2023-07-07 src/tex/ms.tex
implement figure script for optimistic_RS_binned
implement figure script for M-R violins
add line numbers for PSJ submission
line numbers _and_ twocolumn for PSJ submission
avoid captures being cut
acknowledge referees

@@
 % Define document class
-\documentclass[twocolumn,twocolappendix]{aastex631}
+%\documentclass[twocolumn,twocolappendix]{aastex631}
+\documentclass[twocolumn,twocolappendix,linenumbers]{aastex631}
 %\documentclass[modern]{aastex631}
@@
 We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for sample sizes $\gtrsim$~100~planets if at least \SI{\sim 10}{\percent} of those orbiting closer than the habitable zone inner edge harbor runaway climates.
-%We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor runaway climates.
 Our survey simulations suggest that in the near future, ESA's \plato\ mission will be the most promising survey to probe the habitable zone inner edge discontinuity.
@@
 \begin{figure}[ht!]
-    \script{optimistic_R-S.py}
+    \script{optimistic_RS.py}
     \begin{centering}
@@
 Figure~\ref{fig:cornerplot} shows the posterior distributions of $S_\mathrm{thresh}$, $x_{H_2O}$, and $f_\mathrm{rgh}$ as determined by the nested sampler.
-The threshold instellation can be accurately constrained with a median $\hat{S}_\mathrm{thresh} = \var{retrSthresh}$.
+The threshold instellation can be accurately constrained. %with a median $\hat{S}_\mathrm{thresh} = \var{retrSthresh}$.
 Both a higher water mass fraction and a higher dilution factor lead to larger average radii, thus these parameters are strongly correlated.
@@
 \begin{acknowledgments}
+We thank the anonymous referees for their constructive reports that helped to improve this manuscript.
 The authors thank Gabriele Cugno, Yann Fabel, Brad Foley, Lisa Kaltenegger, Ravi Kopparapu, Eric Mamajek, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, Sukrit Ranjan, and Terry-Ann Suer for insightful discussions.
@@
 \begin{figure}[ht!]
-%    \script{figurescript.py}
+    \script{optimistic_RS_binned.py}
     \begin{centering}
@@
 \begin{figure}[ht!]
-%    \script{figurescript.py}
+    \script{MR_violins.py}
     \begin{centering}

### 2023-07-13 src/tex/ms.tex
address minor comments from referee #1

@@
 %Where the incident stellar flux exceeds a threshold value, a non-linear transition from a mixture of condensed and volatile states to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
-For Venus, isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B} and the planet's atmospheric composition suggest that it went through this phase~\citep{2020NatGe..13..265G}.
+For Venus, \rev{the atmospheric composition~\citep{2020NatGe..13..265G} and comprehensive analysis of meteoritic samples across the Solar System~\citep{2018SSRv..214...36A,2022Natur.611..245B}} suggest that it went through this phase.
 Although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}, a transient habitable phase cannot be conclusively ruled out~\citep[e.g.,][]{Way2016,Salvador2017,Krissansen-Totton2021}.
@@
 Finally, we derived masses from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
-The derived stellar parameters from \citet{Hardegree-Ullman2023} were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
+The derived stellar \rev{parameters were} compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively~\rev{\citep{Hardegree-Ullman2023}}.
 From this catalog, \bioverse\ samples stars within an isotropic distance from the solar system as required by the planetary sample size.
@@
 
-To account for planets unable to sustain a steam atmosphere over extended time spans, as well as evolutionary effects such as desiccation through water photodissociation and H~escape (see bottem panel of Figure~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
+To account for planets unable to sustain a steam atmosphere over extended time spans, as well as evolutionary effects such as desiccation through water photodissociation and H~escape (see \rev{bottom} panel of Figure~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
 It represents the fraction of planets above the instellation threshold whose atmospheres are currently inflated due to a runaway greenhouse climate.
@@
 Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars,][]{Turbet2019}, enhancing the demographic signal it leaves.
-An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of gigayears~\citep{Luger2015}.
+An additional advantage of targeting M~dwarfs are \rev{the extended runaway greenhouse phases of their planets that can last on} the order of gigayears~\citep{Luger2015}.
 This increases the probability of observing any given planet in the sample during the runaway greenhouse phase, essentially driving $f_\mathrm{rgh}$ to higher values.

### 2023-07-13 src/tex/ms.tex
address comment 2 from referee #2 (definition of S_thresh)

@@
 
-For \textit{runaway} planets, i.e., all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
+\rev{We consider as \textit{runaway} planets those planets that absorb a stellar flux higher than a} dayside-averaged \rev{threshold instellation $S_\mathrm{thresh}$}.
+\rev{For all planets receiving an instellation exceeding} $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
 While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.

### 2023-07-14 src/tex/ms.tex
address ref#2 com#4 (clouds); cite papers for S_thresh

@@
 \rev{We consider as \textit{runaway} planets those planets that absorb a stellar flux higher than a} dayside-averaged \rev{threshold instellation $S_\mathrm{thresh}$}.
-\rev{For all planets receiving an instellation exceeding} $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
+\rev{For all planets absorbing an instellation exceeding} $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
 While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
+\rev{This choice translates to about 1.18 times the instellation of present-day Earth with a fixed albedo of 0.3, which compares favourably to previous climate simulations~\citep{Leconte2013,Wolf2015}.}
 
@@
 From a planet formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard expected outcome~\citep[e.g.,][]{2019PNAS..116.9723Z,Venturini2020,Emsenhuber2021b,Schlecker2021,Burn2021}.
-But while commonly considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the solar system~\citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation.
+But while commonly considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the solar system\rev{~\citep{McCubbin2019,2020Sci...369.1110P,2021PSJ.....2..244J}} suggests that hydrogen is present in all rocky planets upon formation.
 Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amounts of water inside of the planet during the magma ocean phase~\citep{Ikoma2018,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}.
@@
 Potentially the two-band filter capabilities of \plato\ may offer insight into particularly pronounced spectral features in this runaway greenhouse regime~\citep{2020SSRv..216...98G}.
+\rev{However, detecting water features in steam atmospheres may be obscured by high-altitude clouds that could potentially mute a transit signal as measured from peak to trough of an observed spectral line~\citep{Suissa2020,Fauchez2019}.
+A better understanding of the role of clouds in this problem may be critical for observational constraints on the runaway climate of individual exoplanets.}
 (ii) Post-runaway planets may be detectable through O$_{3}$ absorption due to build-up of abiotic oxygen, leftover from photochemical dissociation of water, and hydrogen loss~\citep{2014ApJ...785L..20W,Luger2015}.

### 2023-07-14 src/tex/ms.tex
address Ref#1 Com#1

@@
 The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the likelihood that a discontinuity is detectable.
-However, dissolution into the magma ocean and atmospheric inflation show non-linear coupling; thus, a straightforward trend is difficult to discern.
+However, dissolution into the magma ocean and atmospheric inflation show non-linear coupling\rev{:
+the more water there is in the atmosphere, the more will be dissolved~\citep{Dorn2021}, leading to a convergence in radii until saturation.
+Most important for the present work, however, is the qualitative dichotomy between sub-runaway and runaway planets, which outcompetes radius variations within each of these climate states due to a changing water mass fraction~\citep{Turbet2020}.
+If a planet hosts a significant water inventory that is not easily stripped by atmospheric escape~\citep{Johnstone2020}, then the statistical power of our model is high (see Figure~\ref{fig:statpwr_H2O-f}).
+}
 

### 2023-07-15 src/tex/ms.tex
address Ref#1 Com#2 (dilution factor)

@@
 Finally, we derived masses from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
-The derived stellar \rev{parameters were} compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively~\rev{\citep{Hardegree-Ullman2023}}.
+The derived stellar \rev{parameters were} compared to \rev{measured parameters} for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively~\rev{\citep{Hardegree-Ullman2023}}.
 From this catalog, \bioverse\ samples stars within an isotropic distance from the solar system as required by the planetary sample size.
@@
 Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the non-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
-The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere.
+The choice $f_\mathrm{rgh} < 1$ \rev{reflects the fact that planets -- despite a high irradiation --  can evade a runaway greenhouse climate.}
+\rev{This situation may, for instance, arise in} the absence of an atmosphere or of volatiles that could form a steam atmosphere.
 In the following, we test if and under what conditions this parametrization causes a demographic trend that is large enough to be detected with high significance.
@@
 What influences the probability of correctly rejecting a false null hypothesis?
-We identified the following six drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey:
+We rev{identified six drivers} of the diagnostic power for detecting the runaway greenhouse transition with a transit survey.
+\rev{Many of these factors directly influence the fraction of planets currently in runaway greenhouse states, which in our model is represented by the dilution factor $f_\mathrm{rgh}$.}
 \begin{itemize}
@@
 However, the delivery uncertainty is much greater than the predicted loss rates of water by atmospheric escape, which typically is limited to in total a few tens of terrestrial oceans per Gyr~\citep{2018AJ....155..195W} on even the most irradiated exoplanets~\citep[see discussion in][]{2022ApJ...938L...3L}.
+\rev{The finite duration of steam atmosphere phases is a main factor for diluting the habitable zone inner edge discontinuity.}
 

### 2023-07-15 src/tex/ms.tex
specify water mass fraction

@@
 While we search for the signature of runaway greenhouse climates in demographic quantities such as average planet radii, the injected changes happen on the planetary level: We changed each planet's transit radius based on its individual set of properties and the associated predictions from steam atmosphere and water retention models.
-Relevant properties are a planet's mass $M$, its net instellation $S$, and its bulk water inventory expressed as a water mass fraction $x_{H_2O}$.
+Relevant properties are a planet's mass $M$, its net instellation $S$, and its bulk water inventory expressed as \rev{the total planetary} water mass fraction $x_{H_2O}$.
 We consider the following cases (see Figure~\ref{fig:radiusevolution}):
@@
             Retrieved posterior distribution of key parameters in the optimistic scenario.
-The density maps in each panel show relationships between and marginalized distributions of the threshold instellation $S_\mathrm{thresh}$, the water mass fraction $x_{H_2O}$, and the dilution factor $f_\mathrm{rgh}$ as they could be retrieved with a high-precision transit survey and a sample of \var{N_optimistic} planets.
+The density maps in each panel show relationships between and marginalized distributions of the threshold instellation $S_\mathrm{thresh}$, the \rev{bulk} water mass fraction $x_{H_2O}$, and the dilution factor $f_\mathrm{rgh}$ as they could be retrieved with a high-precision transit survey and a sample of \var{N_optimistic} planets.
 True values of the parameters for the injected effect are shown in orange.
@@
             Statistical power of the runaway greenhouse hypothesis test as a function of model parameters.
-            For a sample size $N = \var{N_optimistic}$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of water mass fraction and dilution factor.
+            For a sample size $N = \var{N_optimistic}$, the color code shows the fraction of simulations resulting in a sound detection ($\Delta \ln \mathcal{Z} > 3$) for different combinations of \rev{bulk} water mass fraction and dilution factor.
             Higher values in either parameter result in a more reliable detection.
@@
 \subsection{Statistical imprint of runaway greenhouse atmospheres}\label{app:model-pop_comparison}
-The predicted runaway greenhouse-induced planet radius changes are a function of instellation, planet mass, and water mass fraction~(compare Sect.~\ref{sec:met_rghmodel}).
+The predicted runaway greenhouse-induced planet radius changes are a function of instellation, planet mass, and \rev{bulk} water mass fraction~(compare Sect.~\ref{sec:met_rghmodel}).
 In order to better understand the interaction between the planetary populations underlying our simulations and these predictions, we compared the latter to the average radius and bulk density changes we measured in the synthetic population.

### 2023-07-15 src/tex/ms.tex
address minor comments of referee #2

@@
 While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
-\rev{This choice translates to about 1.18 times the instellation of present-day Earth with a fixed albedo of 0.3, which compares favourably to previous climate simulations~\citep{Leconte2013,Wolf2015}.}
+\rev{This choice translates to about 1.18 times the instellation of present-day Earth with a fixed albedo of 0.3, which compares favourably to previous climate simulations~\citep{Leconte2013,Wolf2015}.
+We adopt the same threshold instellation for all host star spectral types.}
 
@@
 %($d_\mathrm{max} = \SI{75}{\parsec}, M_\mathrm{G} < 11$)
+\rev{All other parameters of the climate models are kept the same.}
 The samples are volume and magnitude-limited to reflect the target counts of \plato's provisional Long-duration Observation Phase fields~\citep[$15996$ FGK stars in the P1 and P2 samples, $33948$ M~stars in the P4 sample, ][]{Nascimbeni2022}.
@@
 \end{figure}
-A key constraint resulting from a detection is a measurement of $S_\mathrm{thresh}$.
+\rev{Although the exact location of the runaway greenhouse transition depends on the stellar energy actually absorbed by the planet, }a measurement of $S_\mathrm{thresh}$ \rev{would be a} key constraint resulting from a detection \rev{of the habitable zone inner edge discontinuity}.
 Here, we assess the ability of different mission concepts explored above to constrain this parameter.
@@
 However, the delivery uncertainty is much greater than the predicted loss rates of water by atmospheric escape, which typically is limited to in total a few tens of terrestrial oceans per Gyr~\citep{2018AJ....155..195W} on even the most irradiated exoplanets~\citep[see discussion in][]{2022ApJ...938L...3L}.
-\rev{The finite duration of steam atmosphere phases is a main factor for diluting the habitable zone inner edge discontinuity.}
+\rev{The finite duration of steam atmosphere phases is a main factor for diluting the habitable zone inner edge discontinuity, and it leads to an increased likelihood of detecting this imprint for younger planetary systems.
+Preferential selection of younger systems could therefore be of advantage in a targeted survey.}
 
@@
 This has been suggested to be done via (v) eclipse photometry~\citep{Mansfield2019,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidified magma ocean~\citep{2020ApJ...898..160E,Fortin2022}.
+\rev{Spectral information from atmospheres of highly irradiated planets may also help to distinguish classical runaway greenhouse states from other climate regimes, such as a moist bistability~\citep{Leconte2013a}.}
 
@@
 The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
-Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021,2022ApJ...938L...3L,2022ApJ...939L..19I}.
+Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions\rev{~\citep{Selsis2007,Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021,2022ApJ...938L...3L,2022ApJ...939L..19I}}.
 Our nominal case assumes a bulk water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.

### 2023-07-19 src/tex/ms.tex
minor edits

@@
 \plato\ (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
-In line with this requirement, \plato\ is designed to enable the detection of a \SI{80}{\ppm} transit signal~\citep[][Matuszewski et al., in prep.]{plato2017}.
+In line with this requirement, \plato\ is designed to enable the detection of a \SI{80}{\ppm} transit signal\rev{~\citep[][Matuszewski et al., subm.]{plato2017}}.
 To reflect its sensitivity, we chose a minimum transit depth of \SI{80}{\ppm} as a detection limit and consider only measurements of planets exceeding this threshold.
@@
 \begin{acknowledgments}
-We thank the anonymous referees for their constructive reports that helped to improve this manuscript.
 The authors thank Gabriele Cugno, Yann Fabel, Brad Foley, Lisa Kaltenegger, Ravi Kopparapu, Eric Mamajek, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, Sukrit Ranjan, and Terry-Ann Suer for insightful discussions.
 We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
+We thank the anonymous referees for providing insightful reports that helped to improve this manuscript.
 This material is based upon work supported by the National Aeronautics and Space Administration under Agreement No. 80NSSC21K0593 for the program ``Alien Earths''.

### 2023-07-27 src/tex/ms.tex
further clarify absorbed flux/albedo

@@
 
-The net absorbed stellar fluxes of planets are sensitive to their albedo, which is generally poorly constrained for planets outside the solar system~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019}.
+The net absorbed stellar \rev{fluxes of planets are a function of intrinsic atmospheric properties such as their albedo, which are} generally poorly constrained for planets outside the solar system~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019}.
+\rev{For instance, a planet with a high albedo may sustain temperate conditions closer to the star than the same planet with a lower albedo and located further away from the star.}
 Here, we assume global redistribution of incoming flux and a fixed Bond albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.

### 2023-07-31 src/tex/ms.tex
cite Matuszewski et al. 2023

@@
 \plato\ (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
-In line with this requirement, \plato\ is designed to enable the detection of a \SI{80}{\ppm} transit signal\rev{~\citep[][Matuszewski et al., subm.]{plato2017}}.
+In line with this requirement, \plato\ is designed to enable the detection of a \SI{80}{\ppm} transit signal\rev{~\citep{Matuszewski2023,plato2017}}.
 To reflect its sensitivity, we chose a minimum transit depth of \SI{80}{\ppm} as a detection limit and consider only measurements of planets exceeding this threshold.

### 2023-08-15 src/tex/ms.tex
de-highlight updates from review process

@@
 This material is based upon work supported by the National Aeronautics and Space Administration under Agreement No. 80NSSC21K0593 for the program ``Alien Earths''.
-The results reported herein benefitted from collaborations and/or information exchange within NASA’s Nexus for Exoplanet System Science (NExSS) research coordination network sponsored by NASA’s Science Mission Directorate.
+The results reported herein benefited from collaborations and/or information exchange within NASA’s Nexus for Exoplanet System Science (NExSS) research coordination network sponsored by NASA’s Science Mission Directorate.
 This work has made use of data from the European Space Agency (ESA) mission \gaia\ (\url{https://www.cosmos.esa.int/gaia}), processed by the \gaia\ Data Processing and Analysis Consortium (DPAC, \url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}).

### 2023-08-15 src/tex/ms.tex

@@
 
-As our procedure involves random sampling of the model parameters $\theta$, we need to define the probability of obtaining a dataset given the model parameters, i.e., a likelihood function $\mathcal{L}$.
+As our procedure involves random sampling of the model parameters $\theta$, we need to define the probability of obtaining a data set given the model parameters, i.e., a likelihood function $\mathcal{L}$.
 We assumed here that the individual moving averages $SMA_i$ are measured with a normally distributed uncertainty $\sigma_{SMA_i}$ and adopted a normal distribution
@@
 The code to reproduce a figure can be accessed via the icon link next to the respective figure caption.
+\rev{Data sets associated with this work are available at \dataset[doi:10.5281/zenodo.7080391]{\doi{10.5281/zenodo.7080391}}~(stellar luminosity tracks from \citet{Baraffe1998}) and \dataset[doi:10.5281/zenodo.7946446]{\doi{10.5281/zenodo.7946446}} (results from model grid runs of \bioverse).}
 

### 2023-08-15 src/tex/table_parameters.tex

@@
 Runaway greenhouse atmospheric models &  &  & \citet{Turbet2020,Dorn2021} \\
-$x_{H_2O}$ & \SIrange{e-5}{0.1}{} &  & Water mass fraction (fiducial case: 0.005) \\
+$x_{H_2O}$ & \SIrange{e-5}{0.1}{} &  & \rev{Bulk water} mass fraction (fiducial case: 0.005) \\
 $f_\mathrm{rgh}$ & \SIrange{0}{1}{} &  & Dilution factor (fiducial case: 0.8) \\

### 2023-08-16 src/tex/ms.tex
add DOI of archived repo on Zenodo

@@
 This study uses the reproducibility framework ``showyourwork''~\citep{Luger2021}.
-All code required to reproduce our results, figures, and this article itself is available at \url{https://github.com/matiscke/hz-inner-edge-discontinuity}.
+All code required to reproduce our results, figures, and this article itself is available at \url{https://github.com/matiscke/hz-inner-edge-discontinuity}\rev{, and the repository state at the time of paper acceptance can be found at \dataset[doi:10.5281/zenodo.8251077]{\doi{10.5281/zenodo.8251077}}}.
 The code to reproduce a figure can be accessed via the icon link next to the respective figure caption.

### 2023-08-16 src/tex/ms.tex
mention new Selsis et al. results

@@
 
-Yet, \citet{Hamano2013} suggested that the present-day dry conditions on Venus may have been directly inherited from the early magma ocean stage.% if a strongly infrared-absorbing, condensable species such as water is present on the planet.
+Yet, \citet{Hamano2013} suggested that the present-day dry conditions on Venus may have been directly inherited from the early magma ocean stage.
+% if a strongly infrared-absorbing, condensable species such as water is present on the planet.
 If a strongly infrared-absorbing, condensable species such as water was dominant in the atmosphere, the resulting strong thermal blanketing effect would prevent the planet to efficiently radiate to space and maintain the surface molten~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
@@
 The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}.
-With even a fraction of the Earth's water inventory, a planet absorbing more flux than the radiation limit of steam atmospheres will enter a runaway greenhouse state resulting in a global magma ocean~\citep{Lichtenberg2021c,Boukrouche2021}.
+With even a fraction of the Earth's water inventory, a planet absorbing more flux than the radiation limit of steam atmospheres \revii{was found to} enter a runaway greenhouse state resulting in a global magma ocean~\revii{\citep[][but see \citet{Selsis2023} for a contrasting viewpoint]{Lichtenberg2021c,Boukrouche2021}}.
 We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the distribution of water between planetary interior and atmosphere, and on the resulting thermal atmospheric structure~\citep{Dorn2021,Salvador2023}.
@@
 \textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies on sufficient amounts of atmospheric water vapor that can act as a greenhouse gas.
-However, already about $\SIrange{\sim 10}{20} \, \mathrm{bar}$ of water vapor -- corresponding to a minor fraction of one Earth ocean and thus the lower limit of water on Earth -- is enough to sustain sufficiently high surface temperatures to keep the planet in a magma ocean stage~\citep{Boukrouche2021,Lichtenberg2021c}.
-The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
+\revii{It was shown that} already about $\SIrange{\sim 10}{20} \, \mathrm{bar}$ of water vapor -- corresponding to a minor fraction of one Earth ocean and thus the lower limit of water on Earth -- is enough to sustain sufficiently high surface temperatures to keep the planet in a magma ocean stage~\citep{Boukrouche2021,Lichtenberg2021c}.
+\revii{However, recent simulations accounting for radiative–convective profiles suggest the existence of cooler runaway greenhouse states that do not necessarily yield a surface magma ocean~\citep{Selsis2023}.
+Climate state and atmospheric structure, as well as the fraction of planets fulfilling the requirements of a steam atmosphere} have an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
 From a planet formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard expected outcome~\citep[e.g.,][]{2019PNAS..116.9723Z,Venturini2020,Emsenhuber2021b,Schlecker2021,Burn2021}.

### 2023-08-17 src/tex/ms.tex
Update ms.tex
Minor additions directly added

@@
 \revii{It was shown that} already about $\SIrange{\sim 10}{20} \, \mathrm{bar}$ of water vapor -- corresponding to a minor fraction of one Earth ocean and thus the lower limit of water on Earth -- is enough to sustain sufficiently high surface temperatures to keep the planet in a magma ocean stage~\citep{Boukrouche2021,Lichtenberg2021c}.
-\revii{However, recent simulations accounting for radiative–convective profiles suggest the existence of cooler runaway greenhouse states that do not necessarily yield a surface magma ocean~\citep{Selsis2023}.
+\revii{However, recent simulations accounting for radiative–convective profiles of near-surface atmospheric layers suggest the existence of cooler pure-steam runaway greenhouse states that do not necessarily yield a molten surface~\citep{Selsis2023}.
 Climate state and atmospheric structure, as well as the fraction of planets fulfilling the requirements of a steam atmosphere} have an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.

### 2023-09-07 src/tex/ms.tex
remove line numbers
Fix AAStex 6.3.1's bug causing line numbers in Acknowledgments
arXiv didn't like this.

@@
 % Define document class
-%\documentclass[twocolumn,twocolappendix]{aastex631}
-\documentclass[twocolumn,twocolappendix,linenumbers]{aastex631}
+\documentclass[twocolumn,twocolappendix]{aastex631}
+%\documentclass[twocolumn,twocolappendix,linenumbers]{aastex631}
 %\documentclass[modern]{aastex631}
@@
 
-\begin{acknowledgments}
+%\begin{acknowledgments}
+\pagebreak[4]
+\section*{Acknowledgments}
 The authors thank Gabriele Cugno, Yann Fabel, Brad Foley, Lisa Kaltenegger, Ravi Kopparapu, Eric Mamajek, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, Sukrit Ranjan, and Terry-Ann Suer for insightful discussions.
@@
 A.S. acknowledges support from NASA's Habitable Worlds Program (No.~80NSSC20K0226).
-\end{acknowledgments}
+%\end{acknowledgments}
 

### 2023-09-12 src/tex/ms.tex
cite Turbet et al. 2023

@@
 \rev{For all planets absorbing an instellation exceeding} $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
-While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
+While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds\reviii{~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022,Turbet2023}}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
 \rev{This choice translates to about 1.18 times the instellation of present-day Earth with a fixed albedo of 0.3, which compares favourably to previous climate simulations~\citep{Leconte2013,Wolf2015}.
@@
 Potentially the two-band filter capabilities of \plato\ may offer insight into particularly pronounced spectral features in this runaway greenhouse regime~\citep{2020SSRv..216...98G}.
-\rev{However, detecting water features in steam atmospheres may be obscured by high-altitude clouds that could potentially mute a transit signal as measured from peak to trough of an observed spectral line~\citep{Suissa2020,Fauchez2019}.
+\rev{However, detecting water features in steam atmospheres may be obscured by high-altitude clouds that could potentially mute a transit signal as measured from peak to trough of an observed spectral line\reviii{~\citep{Suissa2020,Fauchez2019,Turbet2023}}.
 A better understanding of the role of clouds in this problem may be critical for observational constraints on the runaway climate of individual exoplanets.}
@@
 A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, reducing their transit radius.
-Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Salvador2017,Turbet2021,Lichtenberg2021c,Pierrehumbert2022,Innes2023}.
+Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity\reviii{~\citep{Salvador2017,Turbet2021,Lichtenberg2021c,Pierrehumbert2022,Turbet2023,Innes2023}}.
 The choice of a statistical estimator for the hypothesis tests may further influence the recovered discontinuity; we compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.

### 2023-09-13 src/tex/ms.tex
rephrase usage of data published in DL21

@@
 Nominally, the above models assume a \textit{dry melt} without dissolved volatiles.
-Here, however, we consider a \textit{wet melt} magma ocean and take into account a radius decrease from retention of water in the melt~\citep{Dorn2021}.
+Here, however, we consider a \textit{wet melt} magma ocean and take into account a radius decrease from retention of water in the melt\reviii{ using the results published in \citet{Dorn2021}.}
 The impact of the water distribution between melt and atmosphere on the change of the transit radius depends on the planet's mass and water content and is generally small compared to the radius inflation from the steam atmosphere.
 %We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
-We computed radius deviations between a wet magma ocean and a solid mantle following \citet{Dorn2021} and assuming a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
+We \reviii{obtained} radius deviations between a wet magma ocean and a solid mantle \reviii{using the numerical values from} \citet{Dorn2021} \reviii{that assume} a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
 We then added the (in almost all cases negative) radius deviations to the inflated planet radii computed for the dry melt case.

### 2023-09-29 src/tex/table_parameters.tex
fix some typos

--- 2023-03-17 src/tex/table_parameters.tex
@@
 ~\\ \textbf{Planetary parameters} &  &  &  \\
-$M_\mathrm{P}$ & 0.1 -- 2.0 & M$_\oplus$ & Planetary mass range \\
+$M_\mathrm{P}$ & 0.1 -- 2.0 & $M_\oplus$ & Planetary mass range \\
 $R_\mathrm{P, min}$ & 0.75 & $R_\oplus$ & Minimum planet radius \\
-Baseline mass-radius relation &  &  & \citet{Zeng2016} \SI{100}{\percent} $\mathrm{MgSiO_3}$\tablenotemark{a} \\
+Baseline mass–radius relation &  &  & \citet{Zeng2016} \SI{100}{\percent} $\mathrm{MgSiO_3}$\tablenotemark{a} \\
 $\delta_\mathrm{min}$ & 80 & ppm & Minimum transit depth \\
-$P_\mathrm{max}$ & 500 & d & Maximum orbital period [d] \\
+$P_\mathrm{max}$ & 500 & day & Maximum orbital period [day] \\
 $S$ & 10 -- 2000 & \SI{}{\watt\per\meter\squared} & Net instellation range \\
@@
 Runaway greenhouse atmospheric models &  &  & \citet{Turbet2020,Dorn2021} \\
-$x_{H_2O}$ & \SIrange{e-5}{0.1}{} &  & Water mass fraction (fiducial case: 0.005) \\
+$x_{H_2O}$ & \SIrange{e-5}{0.1}{} &  & \rev{Bulk water} mass fraction (fiducial case: 0.005) \\
 $f_\mathrm{rgh}$ & \SIrange{0}{1}{} &  & Dilution factor (fiducial case: 0.8) \\
 ~\\ \textbf{Priors} &  &  &  \\
-$\Pi(S_\mathrm{thresh}$) & [10, 1000] & \SI{}{\watt\per\meter\squared} & uniform \\
-$\Pi(x_{H_2O})$ & [\SI{e-5}{}, \SI{0.1}{}] &  & log-uniform \\
-$\Pi(f_\mathrm{rgh})$ & [0, 1] &  & uniform \\
+$\Pi(S_\mathrm{thresh}$) & [10, 1000] & \SI{}{\watt\per\meter\squared} & Uniform \\
+$\Pi(x_{H_2O})$ & [\SI{e-5}{}, \SI{0.1}{}] &  & Log-uniform \\
+$\Pi(f_\mathrm{rgh})$ & [0, 1] &  & Uniform \\
 $\Pi(\langle R_\mathrm{P}\rangle_\mathrm{out})$ & [0, 15] & $R_\oplus$ & Mean radius of non-runaway planets, uniform

### 2023-09-29 src/tex/ms.tex
fix some typos

--- 2023-09-12 src/tex/ms.tex
@@
 % if a strongly infrared-absorbing, condensable species such as water is present on the planet.
-If a strongly infrared-absorbing, condensable species such as water was dominant in the atmosphere, the resulting strong thermal blanketing effect would prevent the planet to efficiently radiate to space and maintain the surface molten~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
+If a strongly infrared-absorbing, condensable species such as water was dominant in the atmosphere, the resulting strong thermal blanketing effect would prevent the planet from efficiently radiating to space and maintaining the surface molten~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,Salvador2017}.
 This runaway greenhouse state can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
 %This scenario requires early water delivery, which for Venus is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
-%Where the incident stellar flux exceeds a threshold value, a non-linear transition from a mixture of condensed and volatile states to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
+%Where the incident stellar flux exceeds a threshold value, a non-linear transition from a mixture of condensed and volatile states to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,Salvador2017}.
 For Venus, \rev{the atmospheric composition~\citep{2020NatGe..13..265G} and comprehensive analysis of meteoritic samples across the Solar System~\citep{2018SSRv..214...36A,2022Natur.611..245B}} suggest that it went through this phase.
@@
 
-In particular, we test the capability of the \plato\ \citep[PLAnetary Transits and Oscillation of stars, ][]{Rauer2016} mission, which will measure the radii of a large number of terrestrial-sized planets, to detect the radius/density discontinuity and determine its sensitivity to model assumptions and fundamental processes.
+In particular, we test the capability of the PLAnetary Transits and Oscillation of stars~\citep[\plato, ][]{Rauer2016} mission, which will measure the radii of a large number of terrestrial-sized planets, to detect the radius/density discontinuity and determine its sensitivity to model assumptions and fundamental processes.
 We then perform a parameter study to explore which trades in the survey design of a \plato-like mission maximize its diagnostic power to test runaway greenhouse climate models through the detection of the habitable zone inner edge discontinuity.
@@
 
-To assign masses to our planets, we use the semi-empirical mass-radius relationship assuming a pure $\mathrm{MgSiO_3}$ composition from \citet{Zeng2016} (see green line in Figure~\ref{fig:radiusevolution}).
+To assign masses to our planets, we use the semi-empirical mass–radius relationship assuming a pure $\mathrm{MgSiO_3}$ composition from \citet{Zeng2016} (see green line in Figure~\ref{fig:radiusevolution}).
 This represents the baseline bulk density before any climate-related effects are applied.
@@
 The net absorbed stellar \rev{fluxes of planets are a function of intrinsic atmospheric properties such as their albedo, which are} generally poorly constrained for planets outside the solar system~\citep[e.g.,][]{Angerhausen2015,Parmentier2018a,Mansfield2019}.
-\rev{For instance, a planet with a high albedo may sustain temperate conditions closer to the star than the same planet with a lower albedo and located further away from the star.}
+\rev{For instance, a planet with a high albedo may sustain temperate conditions closer to the star than the same planet with a lower albedo and located farther away from the star.}
 Here, we assume global redistribution of incoming flux and a fixed Bond albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
@@
 \rev{For all planets absorbing an instellation exceeding} $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
-While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds\reviii{~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022,Turbet2023}}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
-\rev{This choice translates to about 1.18 times the instellation of present-day Earth with a fixed albedo of 0.3, which compares favourably to previous climate simulations~\citep{Leconte2013,Wolf2015}.
+While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds\reviii{~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022,Turbet2023}}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,Salvador2017,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
+\rev{This choice translates to about 1.18 times the instellation of present-day Earth with a fixed albedo of 0.3, which compares favorably to previous climate simulations~\citep{Leconte2013,Wolf2015}.
 We adopt the same threshold instellation for all host star spectral types.}
 
-To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
-Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our non-runaway planets~\citep{Zeng2016}.
+To quantify the radius change, we applied the mass–radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative–convective model~\citep{Turbet2019}.
+Their calculations rely on the same mass–radius relations for rocky interiors that we apply for our non-runaway planets~\citep{Zeng2016}.
 For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
@@
 Nominally, the above models assume a \textit{dry melt} without dissolved volatiles.
-Here, however, we consider a \textit{wet melt} magma ocean and take into account a radius decrease from retention of water in the melt~\citep{Dorn2021}.
+Here, however, we consider a \textit{wet melt} magma ocean and take into account a radius decrease from \reviii{the } retention of water in the melt\reviii{ using the results published in \citet{Dorn2021}.}
 The impact of the water distribution between melt and atmosphere on the change of the transit radius depends on the planet's mass and water content and is generally small compared to the radius inflation from the steam atmosphere.
 %We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
-We computed radius deviations between a wet magma ocean and a solid mantle following \citet{Dorn2021} and assuming a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
+We \reviii{obtained the} radius deviations between a wet magma ocean and a solid mantle \reviii{using the numerical values from} \citet{Dorn2021} \reviii{that assume} a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
 We then added the (in almost all cases negative) radius deviations to the inflated planet radii computed for the dry melt case.
@@
         \caption{
-            \textit{Top}: Mass-radius relationships for a bulk water mass fraction $x_{H_2O}= \var{wrr}$ and different planet states.
+            \textit{Top}: mass–radius relationships for a bulk water mass fraction $x_{H_2O}= \var{wrr}$ and different planet states.
 Green: planets with a solid mantle and no steam atmosphere.
@@
 \end{figure}
-We illustrate the mass-radius relations of the three cases in Figure~\ref{fig:radiusevolution} where a fiducial bulk water mass fraction of $x_{H_2O}= \var{wrr}$ is assumed.
+We illustrate the mass–radius relations of the three cases in Figure~\ref{fig:radiusevolution} where a fiducial bulk water mass fraction of $x_{H_2O}= \var{wrr}$ is assumed.
 % Earth: ~0.0002 PERCENT or 218 ppm by mass (Turbet+2019)
@@
 %In comparison, \plato's expected precision for similar stars was estimated to be \SI{\sim 50}{\ppm}~\citep[][Matuszewski et al., in prep.]{plato2017}.
-\plato\ (PLAnetary Transits and Oscillation of stars) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
+PLAnetary Transits and Oscillation of stars (\plato) is an ESA mission designed to characterize terrestrial planets in the habitable zones of Sun-like stars via long-term high-precision photometric monitoring of a sample of bright stars~\citep{Rauer2016}.
 In line with this requirement, \plato\ is designed to enable the detection of a \SI{80}{\ppm} transit signal\rev{~\citep{Matuszewski2023,plato2017}}.
@@
 Since planetary bulk density $\rho \propto R_\mathrm{P}^{-3}$, we expect a stronger runaway greenhouse signal when measured through bulk density instead of transit radius.
-We thus simulated measurements of planetary densities assuming the mass-radius relation defined above.
+We thus simulated measurements of planetary densities assuming the mass–radius relation defined above.
 For uncertainties in bulk density measurements, we propagated the errors of the mass measurements assuming $\sigma_\mathrm{M_\mathrm{P}} = \SI{10}{\percent}$.
@@
 What influences the probability of correctly rejecting a false null hypothesis?
-We rev{identified six drivers} of the diagnostic power for detecting the runaway greenhouse transition with a transit survey.
+We \rev{identified six drivers} of the diagnostic power for detecting the runaway greenhouse transition with a transit survey.
 \rev{Many of these factors directly influence the fraction of planets currently in runaway greenhouse states, which in our model is represented by the dilution factor $f_\mathrm{rgh}$.}
@@
 But while commonly considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the solar system\rev{~\citep{McCubbin2019,2020Sci...369.1110P,2021PSJ.....2..244J}} suggests that hydrogen is present in all rocky planets upon formation.
-Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amounts of water inside of the planet during the magma ocean phase~\citep{Ikoma2018,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}.
+Accreted hydrogen in nominally dry planetary materials reacts with mantle oxygen to form substantial amounts of water inside of the planet during the magma ocean phase~\citep{Ikoma2018,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}.
 Enhanced equilibration between the core, mantle, and atmosphere during magma ocean evolution of rocky exoplanets further enhances this process~\citep{2021ApJ...914L...4L,Schlichting2022}.
@@
 In a runaway greenhouse atmosphere, absorption is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
-Probing these features requires an instrument covering these wavelength ranges, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as \life\ ~\citep{2019A&A...621A.125B,Quanz2022,2022A&A...664A..22D}.
+Probing these features requires an instrument covering these wavelength ranges, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as the Large Interferometer For Exoplanets~\citep[\life,][]{2019A&A...621A.125B,Quanz2022,2022A&A...664A..22D}.
 Potentially the two-band filter capabilities of \plato\ may offer insight into particularly pronounced spectral features in this runaway greenhouse regime~\citep{2020SSRv..216...98G}.
@@
 Other missions have been proposed that focus on characterizing exoplanet habitability, most notably the \hwo\ concept, which will build on the two precursor direct imaging concepts LUVOIR~\citep{LUVOIR2019} and HabEx~\citep{Gaudi2020c}.
-Similar science objectives are pursued by the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2022} initiative, a mission concept utilizing a space-based mid-infrared nulling interferometer.
+Similar science objectives are pursued by the \life\ initiative~\citep[][]{Quanz2022}, a mission concept utilizing a space-based mid-infrared nulling interferometer.
 Direct imaging surveys do not directly measure planetary radii and are primarily useful for providing context through atmospheric measurements of individual planets.
@@
 Our calculations with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
-Besides their large number in a volume/magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular on orbits further out, i.e., in the optimistic habitable zone.
+Besides their large number in a volume- or magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular on orbits further out, i.e., in the optimistic habitable zone.
 In fact, we showed that the FGK part of the planet sample barely contributes to the statistical power.
 Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars,][]{Turbet2019}, enhancing the demographic signal it leaves.
-An additional advantage of targeting M~dwarfs are \rev{the extended runaway greenhouse phases of their planets that can last on} the order of gigayears~\citep{Luger2015}.
+An additional advantage of targeting M~dwarfs is \rev{the extended runaway greenhouse phases of their planets that can last on} the order of gigayears~\citep{Luger2015}.
 This increases the probability of observing any given planet in the sample during the runaway greenhouse phase, essentially driving $f_\mathrm{rgh}$ to higher values.
@@
 We could, with relatively moderate effort, run a sanity check on an alternative sample (e.g., the more structured Bern model populations, or the population in the Genesis database)}
-The prospects for probing the runaway greenhouse transition depends on astro- and geophysical factors, as well as on the specific instrumentation and survey strategy of a particular mission.
+The prospects for probing the runaway greenhouse transition depend on astro- and geophysical factors, as well as on the specific instrumentation and survey strategy of a particular mission.
 %We can only approximate all this here, and we therefore seek to assess the sensitivity of our results on the assumptions we have made.
@@
 
-\subsubsection{Baseline mass-radius relationship}
-Our baseline mass-radius relationship assuming pure $\mathrm{MgSiO_3}$ interiors~\citep{Zeng2016} might not be representative of the rocky planet population.
+\subsubsection{Baseline mass–radius relationship}
+Our baseline mass–radius relationship assuming pure $\mathrm{MgSiO_3}$ interiors~\citep{Zeng2016} might not be representative of the rocky planet population.
 However, while interior composition may introduce an offset to the radius habitable zone inner edge discontinuity, we do not expect a change of its structure.
-Since the magnitude of the radius inflation effect is expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure~\citep{Zeng2016,Noack2020,2021JGRE..12606724B}, we consider our mass-radius relationship a conservative case.
+Since the magnitude of the radius inflation effect is expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure~\citep{Zeng2016,Noack2020,2021JGRE..12606724B}, we consider our mass–radius relationship a conservative case.
 We performed a sanity check to assess the impact of varying our baseline model (see Appendix~\ref{app:MR_relation}) and found general agreement between different interior compositions.
@@
 The authors thank Gabriele Cugno, Yann Fabel, Brad Foley, Lisa Kaltenegger, Ravi Kopparapu, Eric Mamajek, Megan Mansfield, Paul Molliere, Gijs Mulders, Matthew Murphy, Sukrit Ranjan, and Terry-Ann Suer for insightful discussions.
-We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
+We are grateful to Martin Turbet for providing the mass–radius relationship data for planets harboring steam atmospheres.
 We thank the anonymous referees for providing insightful reports that helped to improve this manuscript.
-This material is based upon work supported by the National Aeronautics and Space Administration under Agreement No. 80NSSC21K0593 for the program ``Alien Earths''.
+This material is based upon work supported by the National Aeronautics and Space Administration under agreement No. 80NSSC21K0593 for the program ``Alien Earths''.
 The results reported herein benefited from collaborations and/or information exchange within NASA’s Nexus for Exoplanet System Science (NExSS) research coordination network sponsored by NASA’s Science Mission Directorate.
@@
 T.L. was supported by a grant from the Branco Weiss Foundation.
-G.B. acknowledges support from the NASA Astrophysics Data Analysis Program under Grant No.~80NSSC20K0446.
+G.B. acknowledges support from the NASA Astrophysics Data Analysis Program under grant No.~80NSSC20K0446.
 A.S. acknowledges support from NASA's Habitable Worlds Program (No.~80NSSC20K0226).
@@
 The code to reproduce a figure can be accessed via the icon link next to the respective figure caption.
-\rev{Data sets associated with this work are available at \dataset[doi:10.5281/zenodo.7080391]{\doi{10.5281/zenodo.7080391}}~(stellar luminosity tracks from \citet{Baraffe1998}) and \dataset[doi:10.5281/zenodo.7946446]{\doi{10.5281/zenodo.7946446}} (results from model grid runs of \bioverse).}
+\rev{Data sets associated with this work are available at \dataset[doi:10.5281/zenodo.7080391]{\doi{10.5281/zenodo.7080391}}~(stellar luminosity tracks from \citealt{Baraffe1998}) and \dataset[doi:10.5281/zenodo.7946446]{\doi{10.5281/zenodo.7946446}} (results from model grid runs of \bioverse).}
 
@@
 
-\subsection{Influence of different mass-radius relationships on our results}\label{app:MR_relation}
-We assessed how a different choice of baseline mass-radius relation influences our results.
-Focussing on the detectability of the statistical runaway greenhouse signal and the ability to constrain the threshold instellation, we repeated the hypothesis test in Sect.~\ref{sec:res_testability} with alternative mass-radius relationships.
+\subsection{Influence of different mass–radius relationships on our results}\label{app:MR_relation}
+We assessed how a different choice of baseline mass–radius relation influences our results.
+Focusing on the detectability of the statistical runaway greenhouse signal and the ability to constrain the threshold instellation, we repeated the hypothesis test in Sect.~\ref{sec:res_testability} with alternative mass–radius relationships.
 Instead of assuming a pure $\mathrm{MgSiO_3}$ composition, we assigned planet masses using either the probabilistic relationship in \citet{Wolfgang2016} or a semi-empirical, two-layer relation assuming an Earth-like (\SI{32.5}{\percent} Fe + \SI{67.5}{\percent} MgSiO$_3$) composition~\citep{Zeng2016}.
@@
         \caption{
-        Posterior threshold instellations as a function of dilution factor for two alternative baseline mass-radius relations.
+        Posterior threshold instellations as a function of dilution factor for two alternative baseline mass–radius relations.
         For each grid step in $f_\mathrm{rgh}$, we show kernel density estimates of retrieved posteriors (averaged over 50 iterations) assuming an Earth-like (\SI{32.5}{\percent} Fe + \SI{67.5}{\percent} MgSiO$_3$) composition or the probabilistic relation from \citet{Wolfgang2016} and otherwise following the optimistic scenario in Sect.~\ref{sec:res_testability}.
@@
         Higher log-evidence differences $\Delta \ln Z$ correspond to more significant rejections of the null hypothesis.
-        In the regime of strong detections, both mass-radius relations lead to similar, accurate constraints on $S_\mathrm{thresh}$.
+        In the regime of strong detections, both mass–radius relations lead to similar, accurate constraints on $S_\mathrm{thresh}$.
         Differences occur at low dilution factors, where the Earth-like relation leads to narrower estimates.
@@
 \end{figure}
-Figure~\ref{fig:MR-violins} shows how the two alternative baseline mass-radius relations influence the significance of a detection and the ability to constrain the threshold instellation.
+Figure~\ref{fig:MR-violins} shows how the two alternative baseline mass–radius relations influence the significance of a detection and the ability to constrain the threshold instellation.
 The relation of \citet{Wolfgang2016} includes intrinsic scatter, which impedes a detection at low dilution factors.
 In this regime, a layered, Earth-like composition leads to more significant detections and a narrower, although biased, constraint on $S_\mathrm{thresh}$.
-Both mass-radius relations agree and recover the injected value where the null hypothesis can be rejected with high significance.
-This is consistent with our nominal mass-radius relation (compare Sect.~\ref{sec:met-orbits_masses}).
+Both mass–radius relations agree and recover the injected value where the null hypothesis can be rejected with high significance.
+This is consistent with our nominal mass–radius relation (compare Sect.~\ref{sec:met-orbits_masses}).
 We conclude that the underlying core and mantle composition of planets may affect the detectability of the transition if the fraction of planets with runaway greenhouse climates is low; however, the overall trends that our experiments revealed appear robust.

### 2023-11-15 src/tex/ms.tex
Sect. -> Section, Eqn. -> Equation

@@
         \caption{Workflow of our hypothesis testing with \bioverse. First, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
-            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:met_rghmodel}.
+            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Section~\ref{sec:met_rghmodel}.
         We then simulate an exoplanet survey, whereby selection effects and detection biases are introduced.
@@
 \end{equation}
-which is relevant for the detection probability of the respective planet~(see Sect.~\ref{sec:sensitivity}).
+which is relevant for the detection probability of the respective planet~(see Section~\ref{sec:sensitivity}).
 Excluding all non-transiting planets diminishes the sample to \var{percentage_transiting} of its original size.
@@
 
-\textit{Non-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
+\textit{Non-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Section~\ref{sec:syn_planets}).
 This case serves as our null hypothesis.
@@
 We sampled $x_{H_2O}$ from a log-uniform distribution to imply scale-invariant ignorance.
-Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Sect.~\ref{sec:met_rghmodel}).
+Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Section~\ref{sec:met_rghmodel}).
 For $f_\mathrm{rgh}$, we chose a uniform prior in $[0, 1]$.
@@
         Detection of the runaway greenhouse threshold.
-        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
+        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Equation~\ref{eq:rgh_hypo}, random draws from the posterior in red).
             The pattern is detected with high significance.
@@
 Comparing a measurement of the habitable zone inner edge discontinuity in radius space with a measurement in density space (which requires planetary mass measurements), a stronger detection occurs in the latter case:
-With an optimistic choice of geophysical parameters (see Sect.~\ref{sec:res_testability}), the average measured radius change is \SI{+15}{\percent} whereas the average density change is \SI{-33}{\percent}.
+With an optimistic choice of geophysical parameters (see Section~\ref{sec:res_testability}), the average measured radius change is \SI{+15}{\percent} whereas the average density change is \SI{-33}{\percent}.
 
@@
 The simulations otherwise represent the simulated \plato\ example described above.
-We chose a planetary sample size of $N=100\pm10$, as this was found to be a threshold case in Sect.~\ref{sec:res_samplesize}.
+We chose a planetary sample size of $N=100\pm10$, as this was found to be a threshold case in Section~\ref{sec:res_samplesize}.
 
@@
 \subsection{Statistical imprint of exoplanet climates}
-We showed in Sect.~\ref{sec:res_signature} that injecting the theoretically predicted radius inflation effect into a synthetic planet population following the currently known demographics leaves a distinct pattern in the radius and density distribution as a function of instellation.
+We showed in Section~\ref{sec:res_signature} that injecting the theoretically predicted radius inflation effect into a synthetic planet population following the currently known demographics leaves a distinct pattern in the radius and density distribution as a function of instellation.
 The transit radius inflation was suggested before as an observational diagnostic to probe the runaway greenhouse transition~\citep{Turbet2019}, and our simulations show quantitatively how the contributions of the individual radius changes combine to create a significant demographic signal.
@@
 \subsection{Detectability of the habitable zone inner edge discontinuity}
-We showed in Sect.~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
+We showed in Section~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
 Of course, even assuming capable instrumentation and an optimal survey design, our current models may not capture fully the complexity of trends in the geophysics and demographics of small planets.
@@
 %\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population).
-These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Sect.
-4]{Turbet2019}.}
+These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Section 4]{Turbet2019}.}
 
@@
 Since the peak thermal emission in runaway greenhouse atmospheres will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
-Finally, mid-infrared surveys such as \life\ show a preference for M~star planets~\citep{Quanz2022}, which is beneficial for detecting the predicted habitable zone inner edge discontinuity (see Sect.~\ref{sec:results_FGK_M}).
+Finally, mid-infrared surveys such as \life\ show a preference for M~star planets~\citep{Quanz2022}, which is beneficial for detecting the predicted habitable zone inner edge discontinuity (see Section~\ref{sec:results_FGK_M}).
 With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states~\citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.
@@
 The constraining power changes when additional information beyond planet radii is available for the characterized planet population.
-As runaway greenhouse phases leave a stronger imprint on bulk density than on planet radius (see Sect.~\ref{sec:res_followup}), it would be beneficial to obtain constraints on planetary masses and test the runaway greenhouse hypothesis in density space instead of radius space.
+As runaway greenhouse phases leave a stronger imprint on bulk density than on planet radius (see Section~\ref{sec:res_followup}), it would be beneficial to obtain constraints on planetary masses and test the runaway greenhouse hypothesis in density space instead of radius space.
 This way, useful results can be obtained under more pessimistic conditions, e.g., a low predominant water content of planetary surfaces and atmospheres or a smaller available planet sample.
@@
 M~dwarf systems are also key for detecting the runaway greenhouse transition:
-Our calculations with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
+Our calculations with different spectral types (Section~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
 Besides their large number in a volume- or magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular on orbits further out, i.e., in the optimistic habitable zone.
@@
 Its detection would thus not only empirically confirm the habitable zone concept but also help to locate it in the observationally available planetary parameter space.
-In Sect.~\ref{sec:res_constrain-S_thresh}, we show that the threshold instellation at which the runaway greenhouse transition occurs can be reasonably constrained without imposing overly optimistic conditions on the underlying planet population, instrumentation, or survey strategy.
+In Section~\ref{sec:res_constrain-S_thresh}, we show that the threshold instellation at which the runaway greenhouse transition occurs can be reasonably constrained without imposing overly optimistic conditions on the underlying planet population, instrumentation, or survey strategy.
 A mission like \plato\ is well equipped to perform this measurement; the constraining power is directly proportional to the proportion of characterized planets around M dwarfs and to the number of planets for which masses can be determined.
@@
 Once a runaway greenhouse region is identified in the parameter space, the community will  have a tool at hand to discern potentially habitable planets from Venusian worlds on an empirical basis.
-Together with atmospheric measurements (see Sect.~\ref{sec:dis_atmospheres}), we will be able to put a number on the probability of an individual planet to harbor sufficient surface water to sustain life.
+Together with atmospheric measurements (see Section~\ref{sec:dis_atmospheres}), we will be able to put a number on the probability of an individual planet to harbor sufficient surface water to sustain life.
 
@@
 This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the habitable zone inner edge discontinuity.
-%, and the hypothesis tests are only marginally sensitive to $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
+%, and the hypothesis tests are only marginally sensitive to $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Section~\ref{sec:res_testability}).
 Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
@@
 \subsection{Alternative statistics for the average radius or bulk density}\label{app:binnedstats}
-The hypothesis tests introduced in Sect.~\ref{sec:met_surveys-hypotests} rely on a statistical estimator for the variation of planetary radii or bulk densities as a function of net instellation, and we chose a moving average for this estimator in our nominal setup.
-Here, we explore how robust our results are against this choice by demonstrating the recovery of the runaway greenhouse signal in the case of the optimistic survey (Sect.~\ref{sec:res_testability}) with an alternative estimator: instead of computing moving averages, we used a binned statistic.
+The hypothesis tests introduced in Section~\ref{sec:met_surveys-hypotests} rely on a statistical estimator for the variation of planetary radii or bulk densities as a function of net instellation, and we chose a moving average for this estimator in our nominal setup.
+Here, we explore how robust our results are against this choice by demonstrating the recovery of the runaway greenhouse signal in the case of the optimistic survey (Section~\ref{sec:res_testability}) with an alternative estimator: instead of computing moving averages, we used a binned statistic.
 
@@
 Instead of using a running mean, we computed a binned statistic (arithmetic mean and standard deviation in blue) to assign planets the average radius or bulk density in their neighborhood in instellation space.
-Random draws from the posterior of the runaway greenhouse hypothesis (Eqn.~\ref{eq:rgh_hypo}) are shown in red.
+Random draws from the posterior of the runaway greenhouse hypothesis (Equation~\ref{eq:rgh_hypo}) are shown in red.
             As in the nominal case, the pattern is detected with high significance.
@@
 \subsection{Statistical imprint of runaway greenhouse atmospheres}\label{app:model-pop_comparison}
-The predicted runaway greenhouse-induced planet radius changes are a function of instellation, planet mass, and \rev{bulk} water mass fraction~(compare Sect.~\ref{sec:met_rghmodel}).
+The predicted runaway greenhouse-induced planet radius changes are a function of instellation, planet mass, and \rev{bulk} water mass fraction~(compare Section~\ref{sec:met_rghmodel}).
 In order to better understand the interaction between the planetary populations underlying our simulations and these predictions, we compared the latter to the average radius and bulk density changes we measured in the synthetic population.
@@
 We assessed how a different choice of baseline mass–radius relation influences our results.
-Focusing on the detectability of the statistical runaway greenhouse signal and the ability to constrain the threshold instellation, we repeated the hypothesis test in Sect.~\ref{sec:res_testability} with alternative mass–radius relationships.
+Focusing on the detectability of the statistical runaway greenhouse signal and the ability to constrain the threshold instellation, we repeated the hypothesis test in Section~\ref{sec:res_testability} with alternative mass–radius relationships.
 Instead of assuming a pure $\mathrm{MgSiO_3}$ composition, we assigned planet masses using either the probabilistic relationship in \citet{Wolfgang2016} or a semi-empirical, two-layer relation assuming an Earth-like (\SI{32.5}{\percent} Fe + \SI{67.5}{\percent} MgSiO$_3$) composition~\citep{Zeng2016}.
@@
         Posterior threshold instellations as a function of dilution factor for two alternative baseline mass–radius relations.
-        For each grid step in $f_\mathrm{rgh}$, we show kernel density estimates of retrieved posteriors (averaged over 50 iterations) assuming an Earth-like (\SI{32.5}{\percent} Fe + \SI{67.5}{\percent} MgSiO$_3$) composition or the probabilistic relation from \citet{Wolfgang2016} and otherwise following the optimistic scenario in Sect.~\ref{sec:res_testability}.
+        For each grid step in $f_\mathrm{rgh}$, we show kernel density estimates of retrieved posteriors (averaged over 50 iterations) assuming an Earth-like (\SI{32.5}{\percent} Fe + \SI{67.5}{\percent} MgSiO$_3$) composition or the probabilistic relation from \citet{Wolfgang2016} and otherwise following the optimistic scenario in Section~\ref{sec:res_testability}.
         Lines within the violins show quartiles of the distributions, and the gray line indicates the injected threshold instellation of \var{injectedSthresh}.
@@
 Both mass–radius relations agree and recover the injected value where the null hypothesis can be rejected with high significance.
-This is consistent with our nominal mass–radius relation (compare Sect.~\ref{sec:met-orbits_masses}).
+This is consistent with our nominal mass–radius relation (compare Section~\ref{sec:met-orbits_masses}).
 We conclude that the underlying core and mantle composition of planets may affect the detectability of the transition if the fraction of planets with runaway greenhouse climates is low; however, the overall trends that our experiments revealed appear robust.
